### 2024-05-28 proposal/Proposal.tex
prop

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[super]{nth}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}

\title{Master's Thesis Proposal}
\author{Marcel Gietzmann-Sanders}
\date{}
\setcounter{tocdepth}{1}
\begin{document}
\maketitle
\tableofcontents

\section{Introduction}

\subsection{Breaking Down Modeling}

One of the functions of science is to build models of the world around us. Specifically to look for functions of the form:

$$v=F(\theta)$$

where $\theta$ is some vector of information, $v$ is a particular outcome given that information, and $F$ is the model in question. This we might term a \textit{deterministic model} because for any specific $\theta$ there is one single outcome $v$.

However, life is rarely so kind to us. From quantum physics to fisheries science the world is replete with examples where the same information does not always lead to the same conclusion. Sometimes that is because we do not possess all of the relevant information but sometimes things are just truly random. As such, the model above will in some sense be lying to us in its deterministic certainty - while we will always predict $v$, $v$ is not always what we will receive.

Instead suppose that for a specific $\theta$ there exist a set of possible outcomes $V(\theta)=\lbrace v \rbrace$. We can define a \textit{stochastic model} as:

$$P(v|\theta)=F(v, \theta)$$

where now the model is predicting the probability of $v$ given $\theta$ instead of just predicting $v$ itself.

Now what is especially interesting about models of this form is the fact that in some sense for any combination of $V$ and $\theta$ they always exist. Whereas to get a deterministic model we need to know precisely what kind of information ($\theta$) we need to make a deterministic prediction, we can always make a stochastic one, even in the presence of no information. The model can always give us a correct answer, it's just a question of how useful that answer is. If you base it off of better data it will become more useful.

Unfortunately, unless we ourselves have built the part of the world we're studying, $F$ is not known to us. Instead our models are really just proposals $\hat{F}$ on what $F$ could be. This presents us with an immediate problem, how do we evaluate any specific $\hat{F}$ if we don't know $F$ itself? Well suppose we've captured a whole series of pairs of $\theta_i$ and $v_i$ where $i$ allows us to index the pair in question. Given we know $\hat{F}$ we can directly ask what the likelihood of the data we have collected is, given $\hat{F}$:

$$\mathcal{L}=\prod_i \hat{F}(v_i, \theta_i)$$

or equivalently (and more easily computed)

$$\ln \mathcal{L} = \sum_i \ln \hat{F}(v_i, \theta_i)$$

$\hat{F}$'s with higher $\ln \mathcal{L}$ (log likelihood) are better fits to the given data and as we have more and more comprehensive data those data better and better represent $F$ (see Appendix 1).

Therefore finding the "true" $F$ can be summarized with two steps:

\begin{enumerate}
\item Collect large amounts of comprehensive data.
\item Find the $\hat{F}$ that maximizes the likelihood of that data.
\end{enumerate}

So is this all there is to modeling? Absolutely not. Everything we've said so far is dependent upon a chosen information set $\theta$. For each choice there is a $F$ and obviously the usefulness of these is a balance between how accessible $\theta$ is when the model needs to be used and how predictive $F$ is (i.e. how much variance is represented by it's predicted $P(v|\theta)$. Determining the best $\theta$ is really what makes any kind of modeling an art form.

\subsection{Searching for Maximal Functions - Machine Learning}

While the first step outlined is obviously highly dependent on what we are trying to model, once the data has been collected

\section{Appendices}
\subsection{Proof of Log Likelihood Maximization}

\begin{theorem}
Suppose we have a set of possible outcomes $V=\lbrace v_k \rbrace$ with probabilities $P_k$. Such that:

$$\sum_k P_k = P$$

Next suppose we concoct a new series of probabilities $U_k = P_k \alpha_k$ s.t.

$$\sum_k U_k = P-\epsilon$$

where $\epsilon \geq 0$. There does not exist a set of $\alpha_k$ s.t.

$$\prod_k \left(\frac{P_k\alpha_k}{P_k}\right)^{P_kN}>1$$

\end{theorem}

\begin{proof}

We proceed by induction. \newline

First note that from the equation above we get:

$$\sum_k P_k \ln \alpha_k > 0$$

and subtracting $\sum_k U_k$ we arrive at:

$$\sum_k P_k \left( \ln \alpha_k - \alpha_k \right) > -P+\epsilon$$

Now suppose for $k$ outcomes we know no such $\alpha_k$ exist as to satisfy the above. Suppose we incorporate a new $k+1$ term s.t:

$$P_{k+1} + \sum_k {P_k} = P_{k+1} + P$$

and:

$$P_{k+1}\alpha_{k+1} + \sum_k {P_k}\alpha_k = P_{k+1} + P - \epsilon = P_{k+1}\alpha_{k+1} + \left(P - P_{k+1}(\alpha_{k+1}-1) \right) - \epsilon$$

Now we need:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) + \sum_k P_k \left( \ln \alpha_k - \alpha_k \right) > -P_{k+1}-P+\epsilon$$

However given our inductive assumption we know that at best,

$$\sum_k P_k \left( \ln \alpha_k - \alpha_k \right)=-P$$

therefore at best:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P > -P - P_{k+1} + \epsilon$$

Furthermore the easiest case for us is if $\epsilon=0$. If it cannot be satisfied than this certainly doesn't work for $\epsilon>0$. So let's consider:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P > -P - P_{k+1}$$

or equivalently:

$$\ln \alpha_{k+1}> \alpha_{k+1}-1$$

Now if $\alpha_{k+1}=1$ we have:

$$\ln 1 = 1-1$$

Further consider the derivatives of each side:

$$\partial_{\alpha}\ln \alpha_{k+1}=\frac{1}{\alpha_{k+1}}$$

$$\partial_{\alpha}(\alpha_{k+1}-1) = 1$$

If $\alpha_{k+1}> 1$ then the log component rises more slowly than the constant component. I.e. our left side will be larger than the right. Likewise if $\alpha_{k+1}<1$ the log component will shrink faster that the constant component which means it will also be less than the constant component. Therefore given our assumptions:

$$\ln \alpha_{k+1} \not\ge \alpha_{k+1}-1$$

and so:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P \not\ge -P - P_{k+1} + \epsilon$$

So much for the $k+1$th case. What about $k=1$. This is trivial because any $\alpha_1$ that satisfies:

$$\sum_k U_k = P-\epsilon$$

must be less than or equal to $1$ and therefore our product of quotients:

$$\prod_k \left(\frac{P_k\alpha_k}{P_k}\right)^{P_kN}>1$$

must be less than or equal to one.

\end{proof}

With this proof in hand we can now show how in the limit as the number of samples taken $N$ goes to infinity our likelihood is only maximized if $\hat{F} \rightarrow F$. \newline

For some given information $\theta$ and outcomes $V=\lbrace v_k \rbrace$ we have, in gathering our data, observed $v_k$ $N_k$ times. Therefore our overall likelihood is:

$$\mathcal{L} = \prod_k \hat{F}(v_k, \theta)^{N_k}$$

Now suppose that we represent:

$$\hat{F}(v_k, \theta)=F(v_k, \theta)\alpha_k$$

That is the ratio of our likelihoods between $\hat{F}$ and $F$ is given by:

$$\prod_k \left(\frac{F(v_k, \theta)\alpha_k}{F(v_k, \theta)}\right)^{N_k}$$

but we also know that:

$$\sum_k F(v_k, \theta) = \sum_k F(v_k, \theta)\alpha_k = 1$$

and that $\lim_{N\rightarrow \inf}N_k = F(v_k, \theta)N$ so that we have:

$$\prod_k \left(\frac{F(v_k, \theta)\alpha_k}{F(v_k, \theta)}\right)^{F(v_k, \theta)N}$$

which is now in the exact same form as our theorem above. And we now know that this ratio is maximized when $\alpha_k\equiv 1$. So as $N\rightarrow \inf$ a higher $\mathcal{L}$ means we are closer to approximating $F$.

\end{document}

### 2024-05-28 proposal/Proposal.tex
first objective

@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{objective}{Objective}
 
@@
 \date{}
-\setcounter{tocdepth}{1}
+\setcounter{tocdepth}{2}
 \begin{document}
@@
 
-\section{Introduction}
+\section{Objectives}
 
-\subsection{Breaking Down Modeling}
+\begin{objective}
+To provide tooling that allows machine learning methods to fit processes of the form
+
+$$P(v|\theta)=F(v, \theta)$$
+
+using maximizing a likelihood function as the objective:
+
+$$\mathcal{L}=\prod_i \hat{F}(v_i, \theta_i)$$
+
+\end{objective}
+
+\subsection{Rationale for Objective 1}
 
@@
 
-So is this all there is to modeling?
-Absolutely not.
-Everything we've said so far is dependent upon a chosen information set $\theta$.
-For each choice there is a $F$ and obviously the usefulness of these is a balance between how accessible $\theta$ is when the model needs to be used and how predictive $F$ is (i.e.
-how much variance is represented by it's predicted $P(v|\theta)$.
-Determining the best $\theta$ is really what makes any kind of modeling an art form.
+All of this, however, has been conditional on the form of $\theta$ having been chosen.
+Obviously, the information we choose to collect and build our model with has an impact on how good the model is from a purely predictive perspective.
+So how can we compare the predictive power of different $\theta$?
+Well as we find $\theta$ that are more predictive the corresponding $F$'s will have greater and greater likelihoods $\mathcal{L}$, given the data.
+But this presents us with a chicken and the egg style problem - to know the value of $\theta$ we must know $F$, but to know $F$ we must have fit $\hat{F}$ which requires a reasonable amount of data to have been collected.
+I.e.
+I cannot know in advance if the data I am collecting is going to be useful or not.
+Instead anyone doing modeling finds themselves in the cycle illustrated by Fig.
+\ref{fig:model_cycle}.
 
-\subsection{Searching for Maximal Functions - Machine Learning}
+\begin{figure}[h!] 
+  \includegraphics[width=\linewidth]{model_cycle.png}
+  \caption{Modeling Cycle}
+  \label{fig:model_cycle}
+\end{figure}
 
-While the first step outlined is obviously highly dependent on what we are trying to model, once the data has been collected
+Each step in this cycle is highly non-trivial. However special interest should be paid to step 3 (finding an $\hat{F}$ that best approximates $F$).
+
+The standard method here has been to propose a family of possible $F$ which are defined by a set of parameters $\mu$.
+Then using some choice of likelihood maximization method the set of $\mu$ are searched for that maximize the likelihood of the data $\mathcal{L}$.
+While this method has borne considerable fruit it still requires a great deal of effort on the side of the scientist to envision various hypotheses on what families of functions would make sense, coding those up, fitting them, and then evaluating them post-fit.
+Furthermore if the true form of $F$ is not contained in those set of hypotheses tested then $F$ is never found.
+Lots of effort without any kind of guarantee that $F$ will be discovered.
+
+The field of Machine Learning (ML) on the other hand has been specifically occupied with finding ways to maximize objectives without having to assume much, if anything, about the functional form of the prediction function.
+Therefore it provides an opportunity to lessen the toil and uncertainty in step 3 and allow scientists to focus more of the efforts on the other steps (and thereby accelerate the whole cycle).
+
+However, ML has traditionally not been occupied with loss functions like our likelihood $\mathcal{L}$. Instead the field has been more concerned with models of the form we first introduced:
+
+$$v=F(\theta)$$
+
+and therefore have been more focused on such things as accuracy and precision (as defined in the field).
+
+There is therefore work to be done in modifying standard ML methods to follow the form:
+
+$$P(v|\theta)=F(v, \theta)$$
+
+and optimize for $\mathcal{L}$.
+But if it can be done that would turn step 3 into an automated process and thereby significantly improve the likelihood that $F$ is actually well approximated while reducing the labor involved the whole cycle thereby accelerating it.
+
+Indeed, at that point, modeling becomes almost completely about gathering, hypothesizing on, and cleaning the data itself.
 

### 2024-05-28 proposal/Proposal.tex
expanding on objective 1

@@
 \begin{objective}
-To provide tooling that allows machine learning methods to fit processes of the form
+To provide tooling that allows for using machine learning methods to fit models of the form
 
-$$P(v|\theta)=F(v, \theta)$$
+$$\psi_k = G(\eta_k)$$
 
-using maximizing a likelihood function as the objective:
+that maximize the following objective:
 
-$$\mathcal{L}=\prod_i \hat{F}(v_i, \theta_i)$$
+$$\mathcal{L}=\prod_i P'(v_i | \eta_i)$$
+
+where:
+
+$$P'(v_i|\eta_i) = \frac{\psi_i}{\sum_k \psi_k}$$
+
+These models will be known as odds models as they predict the "odds for" each outcome $v_k$ given the information contained in $\eta_k$.
 
@@
 \subsection{Rationale for Objective 1}
+
+\subsubsection{Value of Machine Learning in Model Building}
 
@@
 
-The field of Machine Learning (ML) on the other hand has been specifically occupied with finding ways to maximize objectives without having to assume much, if anything, about the functional form of the prediction function.
+The field of Machine Learning (ML) (specifically probabilistic machine learning) on the other hand has been specifically occupied with finding ways to maximize objectives without having to assume much, if anything, about the functional form of the prediction function.
 Therefore it provides an opportunity to lessen the toil and uncertainty in step 3 and allow scientists to focus more of the efforts on the other steps (and thereby accelerate the whole cycle).
 
-However, ML has traditionally not been occupied with loss functions like our likelihood $\mathcal{L}$. Instead the field has been more concerned with models of the form we first introduced:
+\subsubsection{Issues with Probabilistic Machine Learning when Applied to Behavior}
 
-$$v=F(\theta)$$
+Probabilistic machine classification of the kind we were pointing to in the last section has two pitfalls when it comes to modeling behavior.
 
-and therefore have been more focused on such things as accuracy and precision (as defined in the field).
+The first comes from the fact that such models predict on the same set of outcomes each time, regardless of $\theta$.
+For example, if we were trying to predict which drink a person will buy at their local cafe we would not only have to include everything currently on the menu but also everything that could be on the menu across all time points we are interested in predicting.
+If suddenly a new menu item appears that we have not trained on, our probabilistic classifier will be at a loss.
 
-There is therefore work to be done in modifying standard ML methods to follow the form:
+The second issue comes from the "curse of dimensionality".
+Basically this is an issue where as the dimensionality of $\theta$ increases, the amount of data required to find real relationships in that data increases exponentially.
+This is a particular problem for behavioral modeling because we in all likelihood need to include data on each of the possible decisions before us.
+So if I can choose between 5 different drinks I need to provide features on each of those five drinks.
+This leads to very high dimensional spaces very quickly.
+\newline
 
-$$P(v|\theta)=F(v, \theta)$$
+We can get around both of these problems by introducing what we will term an "odds model".
+Specifically if we imagine $\theta$ now as a variable length vector composed of vectors $\eta_k$ for each of our possible outcomes $v_k$ then the odds model is given by the function:
 
-and optimize for $\mathcal{L}$.
-But if it can be done that would turn step 3 into an automated process and thereby significantly improve the likelihood that $F$ is actually well approximated while reducing the labor involved the whole cycle thereby accelerating it.
+$$\psi_k=G(\eta_k)$$
 
-Indeed, at that point, modeling becomes almost completely about gathering, hypothesizing on, and cleaning the data itself.
+where we now say that:
+
+$$P'(v_k|\theta) = \frac{\psi_k}{\sum_j \psi_j}$$
+
+We can see why we chose the name "odds model" as the $\psi_k$ are simply the "odds for" outcome $v_k$. \newline
+
+Now note it is not necessarily true that:
+
+$$P'(v_k|\eta_k) = P(v_k|\theta)$$
+
+because it is possible that information in the other $\eta_j$ informs the probability of $v_k$. However it should be possible to include such information in the $\eta_k$ if necessary.
+
+By making this sacrifice, however, we have achieved two things:
+
+\begin{enumerate}
+\item So long as we can provide a $\eta_k$ for a newly seen outcome we can predict on it. Therefore we need not have a fixed set of possible outcomes (or a fixed number of such outcomes per $\theta$).
+\item Compared to the corresponding probabilistic machine learning classifier, if our total number of outcomes were $|V|$ we've now reduced the dimensionality of our feature space by $|V|$. 
+\end{enumerate}
+
+And this means that unless we have to pass large amounts of information between the $\eta_k$ the amount of data we need to collect to fit our model has decreased tremendously.
+And in cases where data is hard to come by (like animal behavior), any reduction in data requirements is of the utmost importance.
+
+So finally to model behavior we'd like to fit an odds model of the form:
+
+$$\psi_k = G(\eta_k)$$
+
+that maximizes:
+
+$$\mathcal{L}=\prod_i P'(v_i | \eta_i)$$
+
+where
+
+$$P'(v_i|\eta_i) = \frac{\psi_i}{\sum_k \psi_k}$$\newline
+
+To the author's knowledge there is presently no tooling to fit and diagnose such models.
 

### 2024-05-29 proposal/Proposal.tex
objective 2

@@
 \tableofcontents
+\newpage
 
@@
 \end{objective}
+
+\begin{objective}
+To provide tooling that allows for running spatio-temporal simulations of management strategies using movement models as defined above. Such simulations should allow for:
+
+\begin{enumerate}
+\item Flexible initial assumptions that capture heterogeneity in the stock.
+\item The ability to impose spatio-temporal mortality (both fishing and natural).
+\item The ability to capture heterogeneity in fishing vulnerability.
+\end{enumerate}
+
+Likewise these simulations need to be quick enough to allow for capturing uncertainty in initial conditions through bootstrapping.\newline
+
+These simulations will be known as diffusion models as they model the probabilistic behavior of cohorts as opposed to individual fish or schools. 
+\end{objective}
+
+\newpage
+
+\section{Demonstrations}
+
+\newpage
+
+\section{Rationales}
 
@@
 
+\subsection{Rationale for Objective 2}
+
+From studies that started with Beverton and Holt's work on recruitment in fisheries we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to a pattern as illustrated in Fig.
+\ref{fig:recruitment}.
+
+\begin{figure}[h!] 
+  \includegraphics[width=\linewidth]{recruitment.png}
+  \caption{Recruitment}
+  \label{fig:recruitment}
+\end{figure}
+
+What we see is a clear pattern in recruitment verses spawners for low spawners and the eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
+Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out.
+\newline
+
+The implication for fisheries management is pretty clear - sustainable fishing is in large part driven by just maintaining the spawning stock biomass (SSB) above that knee in the curve.
+Indeed this is well noted as the basis of why fisheries \textit{can} exist in the first place.
+If taking spawners always led to decreases in recruits, we'd quickly fish a stock out of existence.
+It is only through the compensation we get on the asymptote that sustainable fisheries are possible.
+\newline
+
+However, two very simple observations of fish biology will show how actually managing SSB can be made difficult.
+
+The first, and simplest, is age structure.
+Obviously we're talking here about maintaining a specific level of spawning stock biomass.
+But today's juveniles are tomorrows SSB and as fish grow they typically become more fecund.
+This means that in order to maintain our SSB over time we have to be careful about how we target each age group.
+This gets even more complicated for fish like Pacific salmon or many species of groupers where life history is intimately entwined with spawning.
+The salmon obviously only breed once and several species of groupers are protogynous hermaphrodites meaning that if you target a specific age group you are effectively targeting a specific sex.
+Clearly understanding and managing our age class selectivity matters when it comes to maintaining a healthy SSB for years to come.
+\newline
+
+The other observation is that many species of fish actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries).
+Pacific salmon are once again a good example in that the home to specific streams and only breed with other salmon in those streams.
+Bluefin tuna are also known to form specific breeding aggregations, as are many tropical fishes.
+Recent research too also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior.
+
+What does this mean?
+It means that if we trust our beverton-holt style model that instead of having a single SSB to manage we really have several SSB's - one for each sub stock.
+Obviously, if we set one global piece of management like an Allowable Biological Catch (ABC) that makes sense for a single SSB but then we have biased fishing mortality where one sub stock is getting fished out far more than the others we may drop below the SSB for that one stock without actually exceeding our expected ABC for the whole fishery.
+And this can certainly happen as different sub stocks will often exhibit quite different spatial behaviors in movement.
+\newline
+
+All of this to say that understanding heterogeneity in vulnerability by things like age class and sub stock is of the utmost importance for managing a fisheries and truly ensuring an appropriate amount of SSB remains year over year.
+
+It is also worth noting that this knowledge around changes in selectivity in space and time is not only useful for considerations around SSB but also for other management issues such as bycatch avoidance and growth overfishing to name just two.
+
+What kind of tooling do we need in order to address these issues and test out ideas for "selectivity aware" fisheries management? Clearly we need some kind of reliable simulation. \newline
+
+Suppose we built models of spatio-temporal behavior (specifically movement) as expressed in Objective 1.
+These models we know can be fed information that captures both environmental variables but also factors we'd like to categorize by in order to understand heterogeneity in selectivity.
+Such variables could include features like the age and genetic cohorts that we just described.
+Then we could take some baseline assumptions about abundance in space at a given time (say providing recruits at spawning grounds or taking a spatio-temporal assessment as a starting point) and then using our behavioral models to simulate time-steps forward.
+As we do these time steps we could account for natural mortality as well to complete the "environment" portion of our simulation.
+
+It is important to note that while this may seem like a Lagrangian approach to simulation it actually falls more into the Eulerian camp as, in order to reduce our computational costs, we'd simply have cohorts as a function of space.
+Remember that our model is only taking the categories that separate one group from the next thereby allowing us to model cohorts instead of single individuals.
+Furthermore instead of modeling the specific singular behaviors our behavior model as outlined in Objective 1 allows us to understand the "many world" probabilistic outcomes of behavior.
+We are not following a single fish but instead are watching as that cohort diffuses across its decision space.
+Therefore we'll term this simulation a diffusion model.
+
+Into our diffusion model we can now incorporate whatever management policy we like as a spatio-temporal specification of fishing mortality - which can now be modulated by our understanding of, say, gear selectivity as a function of space and time.
+So if we want to implement a marine protected area we can take a baseline fishing pattern and exclude specific areas in space.
+If we want to see what would happen if we redistribute the fleet we can as well.
+If we just want to understand the selectivity biases already present we can simulate fleet patterns as they occur today.
+
+The point is that with such a simulation we are able to test different hypotheses of management on data driven understanding of fish behavior.
+
+\newpage
+
 \section{Appendices}

### 2024-05-30 proposal/Proposal.tex
demonstrations and timeline

@@
 \newtheorem{objective}{Objective}
+\newtheorem{model}{Model}
 
@@
 
+Our demonstrations of the above objectives will be split into two chapters.
+
+\subsection{Chapter 1}
+
+Chapter 1 will be concerned with demonstrating the tooling for creating odds models for spatio-temporal movement behavior. It will pursue two specific modeling examples:
+
+\begin{model}
+A depth occupancy model for Chinook salmon in the Gulf of Alaska (GOA). \newline
+
+\textbf{Target:} Depth range. \newline
+
+\textbf{Features:} Environmental data derived either from environmental simulations or from remote sensing such as Google Earth, cohort data about the fish themselves, and temporal features such as month, season, or time of day.
+\newline
+
+\textbf{Data:} PSAT tagging data from the GOA. \newline
+
+\textbf{Purpose:} To study the heterogeneity in depth occupancy as a function of cohort, environment, and time.  
+\end{model}
+
+\begin{model}
+A movement model for Chinook salmon in the Gulf of Alaska (GOA). \newline
+
+\textbf{Target:} Spatial grid cell selection as movement. Any state feature updates needed to capture internal states of a cohort. \newline
+
+\textbf{Features:} Environmental data derived either from environmental simulations or from remote sensing such as Google Earth, cohort data about the fish themselves, temporal features such as month, season, or time of day, and internal state information about the cohort.
+\newline
+
+\textbf{Data:} PAST tagging data from the GOA. \newline
+
+\textbf{Purpose:} To study the heterogeneity in spatial movement/preference as a function of cohort, environment, and time.\newline
+
+\end{model}
+
+In building these models our primary purpose is to demonstrate the tooling exists to prudently build odds models.
+We can frame this goal in terms of a "standard" machine learning pipeline and work-flow.
+
+\begin{enumerate}
+\item Formatting Data for Odds Modeling 
+\item Class Balancing
+\item Splitting Data into Train, Validation, Test
+\item Training and Hyperparameter Tuning Using Standard Model Types
+\item Diagnoses of Fit and Comparison of Different Model Types
+\item Identifying Sample Groups that are More or Less Fit
+\end{enumerate}
+
+We will want to explicitly demonstrate each of these processes, provide guidance on using the tools for them, and use them to build the models listed above. \newline
+
+The outcome will be a package in Python for building, diagnosing, and running odds models.
+
+\subsection{Chapter 2}
+
+Chapter 2 will be concerned with taking these models for Chinook in the Gulf of Alaska and building a spatio-temporal movement simulation that can be used to investigate a series of management questions.
+It is worth noting that we will use the simulation to ask questions that may come up in general management settings even if they may not be directly applicable to the issues currently faced by Chinook management in the GOA.
+This is just to demonstrate general applicability.
+\newline
+
+The overview of the components of this simulation are outlined in Fig.
+\ref{fig:simulation}.
+Boxes with rounded edges would be inputs specified by the user.
+The hexagons are functions provided by the simulation and the trapezoids are data generated during the main loop.
+
+\begin{figure}[h!] 
+  \includegraphics[width=\linewidth]{simulation.png}
+  \caption{Simulation Outline}
+  \label{fig:simulation}
+\end{figure}
+
+With this simulation we will attempt to demonstrate the following:
+
+\begin{enumerate}
+\item \textbf{Connectivity:} by looking at the vector flows of fish throughout the space we will be able to identify which areas are highly connected and which are more or less separated by environment, cohort, or other factors of fish behavior.
+\item \textbf{Vulnerability:} by especially looking at changes in depth occupancy and preference for habitat we will attempt to quantify the degree of vulnerability in different parts of the space. 
+\item \textbf{Cumulative Mortality:} we would like to demonstrate that we can model the effect of fishing the same cohorts through space as opposed to distributing the fishing mortality across more of the stock.
+\item \textbf{Protected Areas/Times:} we will attempt to demonstrate that we can simulate the effects of MPA or other kinds of closures in space and time. 
+\item \textbf{Cohort Bias:} we will show (models willing) that we can understand and model differences in the above by cohort.
+\item \textbf{Uncertainty:} we will show that such a simulation can be run at scale, bootstrapping can be applied, and therefore uncertainty in simulation outcomes quantified. 
+\end{enumerate}
+
+The outcome will be a package in Python for deploying, running, and analyzing diffusion models.
+
+\newpage
+
+\section{Timeline}
+
+\subsection{Class Work}
+
+\begin{enumerate}
+\item \textbf{Fall 2023:} (University of Florida) FAS 6355c Fisheries Management, FAS 6337c Population Dynamics
+\item \textbf{Spring 2024:} FISH F621 Estimation of Fish Abundance, FISH F427 Ichthyology
+\item \textbf{Fall 2024:} STAT F641 Bayesian Statistics, FISH F692 Seminar
+\item \textbf{Spring 2025:} FISH F631 Data Analysis in Community Ecology, FISH 699 Thesis 
+\item \textbf{Fall 2025:} STAT F651 Statistical Theory I, FISH 699 Thesis
+\item \textbf{Spring 2026:} FISH F626  Behavioral Ecology of Fishes, STAT F605 Spatial Statistics 
+\end{enumerate}
+
+\subsection{Thesis}
+
+\begin{enumerate}
+\item \textbf{Spring 2024:} Depth and Movement Models v0, Thesis Exploration
+\item \textbf{Fall 2024:} Graduate Study Plan, Self Evaluation, Thesis Proposal, Depth and Movement Models v1
+\item \textbf{Spring 2025:} Thesis Proposal Defense, Simulation v1
+\item \textbf{Fall 2025:} Self Evaluation, Write Chapters, Publish Package(s)
+\item \textbf{Spring 2026:} Thesis Defense
+\end{enumerate}
+
 \newpage
@@
 
-\subsection{Rationale for Objective 1}
+\subsection{Objective 1}
 
@@
 
-\subsection{Rationale for Objective 2}
+\subsection{Objective 2}
 
@@
 
+\subsection{Chinook Salmon PSAT Data from GOA}
+
+\newpage
+
+\section{Related Work}
+
 \newpage
@@
 \section{Appendices}
-\subsection{Proof of Log Likelihood Maximization}
+
+\subsection{Value of Log Likelihood Maximization}
 

### 2024-05-30 proposal/Proposal.tex
notes on PSAT and Chinook

@@
 
-\section{Rationales}
+\section{Rationale}
 
@@
 
-\subsection{Chinook Salmon PSAT Data from GOA}
+\subsection{PSAT Data}
+
+Pop-up Satellite Archival Tags (PSATs) are a great way of collecting behavior data. They are defined by three key properties:
+
+\begin{enumerate}
+\item \textbf{Archival:} Transmitting data is only possible if animals come to the surface. Therefore these tags archive collected data for later transmission. 
+\item \textbf{Pop Up:} These tags are programmed to release from a fish at a specific time or under specific conditions to facilitate retrieval. 
+\item \textbf{Satellite:} On pop-up the tags then broadcast data (usually summarized) to Argos satellites which is then relayed to a database for retrieval.
+Such tags usually also broadcast some positioning information which allows for direct retrieval and access to the full archived dataset.
+\end{enumerate}
+
+The normal lifecycle then is to catch an animal, attach the tag, allow the animal to go about its life all while collecting data, have the tag pop-up, transfer data to satellite or retrieve the tag, and then clean and analyze the data.
+\newline
+
+For our purposes these tags critically capture:
+
+\begin{enumerate}
+\item Temperature
+\item Depth
+\item Light Levels
+\end{enumerate}
+
+The latter two are then used along with tag release location to predict animal movement (latitude and longitude) using a proprietary statistical model.
+
+We will look at movements on a day by day basis and depths at the transmitted resolution (2-5 minutes).
+For those tags retrieved, a much higher resolution is available (1-3 seconds) however most fish had data only from transmission and not from retrieval.
+\newline
+
+Clearly then this kind of data suites our modeling and simulation purposes well as we get behavior outcomes directly.
+
+\subsection{Chinook in the GOA}
+
+At this point in time the main interest here is that Chinook present an interesting model species for testing out the efficacy of these models and simulations.
+Chinook are highly motile and undergo clear migration behaviors.
+Therefore they should provide a good challenge to attempts at behavioral modeling and simulation.
+It is the author's hope that patterns useful to Chinook management issues may arise from this research but it is not a necessity as the intention here is to develop tools that justify additional data collection for species where spatial management is of interest.
 

### 2024-05-30 proposal/Proposal.tex
breaking up the lit review work
better title page
navigating the document

@@
 
-\title{Master's Thesis Proposal}
-\author{Marcel Gietzmann-Sanders}
-\date{}
+\makeatletter
+\renewcommand{\maketitle}{
+\begin{center}
+
+\pagestyle{empty}
+\phantom{.}  %necessary to add space on top before the title
+\vspace{3cm}
+
+{\Huge \bf \@title\par}
+\vspace{2.5cm}
+
+{\LARGE Marcel Gietzmann-Sanders}\\[1cm]
+
+{\Large\@date}
+
+\vspace{2.5cm}
+{\Large Dr. Andrew Seitz}\hspace{2cm}{\Large Dr. Curry Cunningham}\\[2cm]{\Large Michael Courtney, M.S.}\\[2cm]
+College of Fisheries and Ocean Sciences\\
+University of Alaska Fairbanks
+
+\end{center}
+}\makeatother
+
+\title{Master's Research Proposal}
+
+\date{2024}
 \setcounter{tocdepth}{2}
@@
 \maketitle
+\newpage
+
+\section{Navigating This Document}
+
+This document is organized in order to allow you, the reader, to read as little or as much as you need/want. \newline
+
+It begins with the \textbf{objectives} so that you know immediately exactly what this proposal is hoping to achieve and whether this is of interest to you.  \newline
+
+Then it dives into how we will go about demonstrating success along those objectives, this will inform you on the specific \textbf{outcomes} to be expected. \newline
+
+Finally there is a \textbf{timeline} provided to give a sense of when these outcomes can be expected. \newline
+
+If these are compelling enough there is no need to read any further. \newline
+
+However if you are interested in the rationales behind these objectives or diving into related work has already been done by others, the \textbf{Rationales} and \textbf{Related Work} sections are for you.
+
+\newpage
+
 \tableofcontents
@@
 
+\subsection{Chinook Ecology}
+
+\subsection{Chinook Management}
+
+\subsection{Studies Using PSAT Data}
+
+\subsection{Spatial Modeling}
+
+\subsection{Spatial Management}
+
 \newpage

### 2024-05-30 proposal/Proposal.tex
typo

@@
 
-However if you are interested in the rationales behind these objectives or diving into related work has already been done by others, the \textbf{Rationales} and \textbf{Related Work} sections are for you.
+However if you are interested in the rationales behind these objectives or diving into related work has already been done by others, the \textbf{Rationale} and \textbf{Related Work} sections are for you.
 

### 2024-05-30 proposal/Proposal.tex
ultimate objective

@@
 \section{Objectives}
+
+The ultimate objective of this research is to provide a new pathway by which collections of high resolution spatio-temporal movement data that are representative of a species can be used to make a direct impact on fisheries management.
+It is hoped that this will provide an incentive for creating/enlarging such datasets.
 

### 2024-06-01 proposal/Proposal.tex
management

@@
 
-\title{Master's Research Proposal}
+\title{Master's Research Proposal (IN PROGRESS)}
 
@@
 
+\subsection{Spatial Management}
+
+Fisheries management is as diverse a subject as the species and cultures to which it is applied.
+It simultaneously has to deal with creatures from Bluefin Tuna who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters.
+There are species that school and species that consume those schools.
+Some live entirely at sea, some in inland rivers and lakes, and some, like Salmon and eels spend time in both.
+There are species whose lives last the span of a year to others that can live longer than any of us human beings.
+
+And to capture this vast array of diversity are similarly diverse strategies and gears.
+From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial bands, and everything in between the variety in strategies to catch fish are numerous.
+
+And then there are the stakeholders of these fisheries and their multitude of objectives.
+What kinds of fish sell the best?
+Is selling fish even the point or is subsistence a big concern?
+Is this a primary source of livelihood or are fishers also fishing for other things, or occupied with entirely different forms of work during other parts of the year?
+Where do conservation groups fit in?
+What about the natural predators and prey lodged nearby in the food web?
+Is the stock in need of rebuilding?
+What requirements of management are there?
+
+Fisheries management is a field in which the phrase "the devil is in the details" is incredibly relevant. \newline
+
+As with any such endeavor that can come in so many forms and captures so many different people, interests, and potential problems and opportunities, the best creative solutions come when everyone is working together to build and understand those solutions.
+And of utmost importance here is the inclusion of those who actually go out and do the fishing.
+Fisheries management that creates a sense of inclusion and involvement with the fishing community is the best kind of management because of a number of things:
+
+\begin{enumerate}
+\item Fishers often have incredible amounts of insight into the species they are targeting 
+\item Fishers have an interest in the long term sustainability of their work and therefore often have ideas on how to achieve this
+\item There are plenty of examples of fishers helping to self enforce policies if they agree with said policy
+\item Fishers are key when it comes to collecting the kinds of data required to do management
+\item Fishers are one of the primary stakeholders impacted and therefore must have a say in what is going to be done
+\end{enumerate}
+
+But in general creating cooperation improves the overall solution whereas exclusion simply creates conflict that works against any kind of management.
+Therefore the first priority on our list when it comes to doing good management (spatial or otherwise) has to be creating engagement.
+\newline
+
+To this end simulations and models of the kind that we are proposing should be helpful.
+
+\subsubsection{Driving Engagement}
+
+First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world.
+By having simulations that can simply be shown and laid down on a map that places the spatio-temporal modeling in familiar context one should be able to make objective or confirmation extremely easy.
+Rather than having to attempt to condense down loads of parameters, summary statistics, and the like in a model, one can simply show a map and folks can point at things that line up or are in disagreement with their intuition and experience.
+In this way mapped out views of the simulation work to be a useful device for consensus building among multiple groups of stakeholders.
+Not everyone may be familiar with the nuances of a specific kind of model but everyone in the problem can point to a map and understand what's going on.
+\newline
+
+With the simulation providing an easier to understand jumping off point to create feedback around a model, the models themselves provide a reasonably straightforward method for taking that feedback and working it into the model itself.
+This is because these models rely only on the information input as opposed to requiring modelers to determine the relationship between the information and the outcomes.
+
+For example suppose that we have a more standard model that has been hand crafted by a scientist familiar with the species in question.
+They may have included specific reactions to, say, oceanic currents, temperatures, and primary productivity.
+But then if someone looking at the simulation points out that they think the species' behavior is also being driven by day/night patterns the scientist then has to go back to the drawing board, figure out how to add this into their overall model, code it up, tune parameters, and then return - all without any real guarantee that whoever provided the feedback is right.
+And this would be for one piece of feedback.
+If there are several this only gets more difficult.
+
+Instead by using machine learning as proposed here, such feedback would amount to perhaps playing around with a few day light features, pushing those to the model, retraining using the same automation as for any of the other feature combinations, and then just reporting if that information helped or not.
+Feedback can be incorporated much more easily allowing for direct contribution to the model on the part of those who may have no modeling experience but have lots of practical experience that could suggest forms of information that would improve how targeted the model is.
+\newline
+
+In summary the combination of clear, relevant visuals from the simulation, and ease of feature incorporation model side allow for a nice iterative loop to drive engagement and input from all stakeholders into the underlying spatial understanding of the fishery.
+
+\subsubsection{Spatial Management Today}
+
+While we've already noted that the creative solutions in fisheries management are as diverse as the people, species, and objectives present, there are some specific kinds of strategies that are used across the world to do spatial management.
+By and large they are comprised of performing spatial closures, however those closures differ with regards to who they effect and for how long.
+\newline
+
+\begin{enumerate}
+\item \textbf{Marine Protected Areas (MPAs)} - MPAs are usually permanent closures intended to reduce the fishing pressure in specific areas in the hope that those areas can become "sources" of larvae and stock to help build up the surrounding areas or to help scientists understand the "unfished" condition.
+Sometimes these areas are total no-go zones for all fishing activity, sometimes specific "low pressure" activities are allowed.
+In general it is clear that the efficacy of MPAs is largely dependent on the spatio-temporal behavior and spatio-temporal aspects of the species' life cycle as well as the same aspects of behavior and lifestyle of the species' predators and prey.
+Species that move about dramatically may not be helped by MPAs and species that are "too" sessile may not actually do much to contribute to stocks outside of the MPA itself.
+Therefore it is clear that a good understanding of movement is extremely important when both thinking about the overall efficacy of an MPA and also where to place them.
+\item \textbf{Territorial Use Rights for Fisheries (TURFs)} - TURFs delineate a fishery into spatial regions and then give exclusive access rights to specific communities of fishers to each of the defined regions.
+The intention with exclusivity is to align the incentives of the fishers with that of sustainability and is a way to address "tragedy of the commons" style issues that arise with completely open access.
+However, as with MPAs, it is clear that understanding the spatio-temporal behavior of the species in question is of the utmost importance here because for species that exhibit considerable movement, small regions would mean that while spatial rights exist those rights don't really have anything to do with the underlying fish.
+\item \textbf{Temporary Closures} - These kinds of closures are designed to deal with specific issues directly related to the target species (or related bycatch's) ecology.
+For example seasonal closures in specific areas might exist to protect spawning aggregations or other kinds of aggregations that create a particularly dangerous kind of vulnerability.
+Or these closures might protect against vulnerability by another species to fishing activities such as preventing hits on whales or bycatch of an endangered species.
+Here again, clear spatio-temporal understanding is necessary for defining and determining the efficacy of these kinds of measures.
+\item \textbf{Real Time Management} - This is a particularly interesting kind of management as instead of trying to predict where closures are needed, in this strategy metrics from the fishers themselves help dictate where closures are needed temporarily.
+For example if bycatch avoidance is the goal then catch logs (that are usually reported with some frequency in these cases) are used to determine areas where bycatch abundance has spiked and then the fishery can call for a temporary closure last days to months in order to avoid that aggregation.
+In general this strategy is comprised of measuring some overall catch statistic and then responding to that statistic in real time as needs be.
+Here models of the kind we're proposing are arguably less relevant as prediction is not \textit{required} for this kind of strategy to work.
+However they could certainly help in both defining the extent of the closures in both space and time and potentially helping fishers avoid such closures in the first place by helping drive understanding of where things like high bycatch may occur.
+\end{enumerate}
+
+Unsurprisingly then, for any kind of spatial management tool in use today, spatio-temporal understanding of the stock is key to both definition and efficacy.
+Therefore it seems likely that bringing high resolution spatio-temporal data to the bear on the issue can only help make such techniques more nuanced, precise, and thereby effective.
+
+\subsubsection{"Weather" Forecasting}
+
+Finally it is worth noting that the potential in combining this modeling with fisheries dependent information is quite interesting on its own.
+Any forecasting model is always going to have issues predicting farther and farther into the future but fisheries dependent data could provide a means for making corrections to keep the model in line.
+This would amount to something like weather forecast where for the next few days the forecast is accurate enough to precise predict specific kinds of movement helpful to fishers themselves.
+
+For example it could help them predict where to go to avoid high levels of bycatch, help optimize their use of fuel and labor when searching for fish, or, depending on the degree of subtlety in the model might even be able to help them target catches that are more or less valuable on the market itself.
+
+With such direct incentives to fishers this might provide a jumping off point to bring them into helping collect the kinds and quantities of data that would allow for increased precision in these models.
+With such "weather forecasting" any such increases would directly benefit them in addition to providing benefits to management.
+
 \subsection{Chinook Ecology}
@@
 
-\subsection{Studies Using PSAT Data}
+\subsection{Other Applications of PSAT Data}
 
@@
 
-\subsection{Spatial Management}
+\subsection{Spatial Assessment}
 

### 2024-06-04 proposal/Proposal.tex
spatial assessment and movement modeling

@@
 
+\newpage
+
+\subsection{Spatial Assessments}
+
+\subsubsection{Assessments in Short}
+
+Fisheries assessments are one way in which unknown parameters important to management are estimated.
+By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so are arise from the state model.
+Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit and the observation model is encoded as a set of distributions dependent on those parameters thus allowing for a likelihood to be computed against known data for any given set of parameters.
+This, in turn, allows for maximum likelihood estimation (MLE) routines to then fit the likely parameters of the assessment.
+\newline
+
+So, as an example, you might have your state model be comprised of age cohorts that have a specific growth pattern (fit externally) and then you leave as parameters natural mortality and recruitment.
+Then you might model vulnerability to a couple different kinds of gear, provide an expected distribution for catches given your underlying abundance by cohort, and then "simulate" a few years.
+Matching this back to real catch data would then allow you to estimate your natural mortality and recruitment by year.
+\newline
+
+In short the pattern is (1) design a sim that has your unknown statistics as parameters, (2) specify how data is drawn from the sim, (3) fit your parameters using MLE.
+Then those desired statistics are extracted from the assessment (usually with some notion of uncertainty attached to them) and used to inform management.
+
+\subsubsection{Features of Spatial Assessments}
+
+As has been noted by several authors most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit.
+However from studies of fish behavior and the fact that fish don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases.
+
+Therefore in recent years there has been work in creating new assessments that are able to capture spatial heterogeneity and movement.
+There are effectively two different kinds of heterogeneity that such models have to capture - that in recruitment and that in post recruitment movement and connectivity (or lack thereof).
+As such these models are usually structured as having multiple sub regions within the overall assessment region and then two deviations from the standard assessment follow:
+
+\begin{enumerate}
+\item First sub-region specific mortality, growth, recruitment, or other sub-population specific parameters may be estimated.
+The degree to which these are shared between or constrained by other regions is often related to the data available and how flexible the model is allowed to be.
+\item Movement among regions at each time step is also modeled, usually as a matrix $X$ such that the elements $X_{i,j}$ signify movement rates from sub-region $i$ to sub-region $j$. 
+\end{enumerate}
+
+Different assessment paradigms will have varying degree of resolution in the sub-regions, will estimate certain parameters freely to not at all, leave $X$ free or constrain it on other kinds of assumptions (indeed some assessments don't model movement at all).
+
+Ultimately what is bounding the freedom in parametrizing these assessments is the availability of data to actually fit all of these parameters - a subject to which we turn next.
+
+\subsubsection{Integrated Assessments}
+
+A primary issue with spatial assessments is their hunger for data.
+This makes sense given the combination of the "curse of dimensionality" and the fact that adding per region parameters and across region movements introduces not just new parameters but many multiples of parameters as compared to a non-spatial assessment model.
+
+Therefore a huge step forward in being able to run these kinds of assessments was the development of "integrated assessments" - assessments that can take advantage of pretty much any kind of data collected from the fishery.
+The process here falls neatly into our earlier description of how an assessment works - for some new kind of data (say acoustic tag data) one must add a new component to our observation model that describes the distribution that would generate our acoustic tag data.
+With this in hand we've connected a new data type to our underlying parameters and can therefore use that data to help fit our parameters.
+
+There are of course issues that arise in terms of how to weight one data type against another when doing MLE but by and large integrated assessments are a boon because they increase the data available to fit assessments and have led to several tag integrated assessments that include electronic tag data in addition to more standard mark and recapture tag data.
+
+\subsubsection{Validating Spatial Assessments}
+
+With the recency of their development and overall complexity it has been noted that diagnosing goodness of fit and relevance of spatial assessments is an art that is still in development (although several papers exist that given guidance on best practices in choosing and using spatial assessment paradigms).
+
+Every assessment paradigm is of course a simplification of reality but what's important to know is whether those simplifications are obscuring important structure of the stock and sub-stocks.
+I.e.
+are the simplifications introducing bias or errors that would lead to contradictory results from sound management based on the advice of the assessment?
+
+Given that ultimately we do \textit{not} know the underlying structure (models such as these are built specifically to elucidate such meaningful structure) we find ourselves with a "Catch-22".
+We need spatial components to understand spatial structure, but need to know the spatial structure in order to understand what simplifications are allowable.
+Note that it is not enough to simply design the most complicated spatial model possible given the data will likely not exist to fit that model with any degree of confidence.
+
+The way out of this hole, as many authors have noted, is to use simulations.
+Specifically one can use simulations to do hypothesis testing of the form - if my stock had spatial structure of the form X would my specific assessment capture that at all or would it introduce biases and/or errors as a result?
+This is a very powerful technique that has been responsible for illustrating just how important spatial assessments are to management in the first place.\newline
+
+Here then is another potential application of the kind of model proposed in this document.
+As we mentioned earlier these assessments are trying to model at least three things - recruitment, local population dynamics parameters such as growth or natural mortality, and then movement.
+By having a movement simulation based on real observations of the species in question we would be able to help build a high quality simulation of at least the last of those three.
+Rather than having to assume one form of a movement or another we'd instead be able to see whether the assessment paradigm in question can capture the relevant aspects of structure in the \textit{real} movement of our stocks and sub-stocks.
+And this in turn would help in selecting the assessments best suited to the job (as well as illustrate what uncertainties exist if a lack of specific kinds of data requires us to use an over-simplified assessment paradigm).
+
+\subsubsection{Comments on Interpretation}
+
+Something that any student of statistical or Machine Learning learns early on is that models can be deceiving in interpretation due to correlations in different kinds of data or parameters.
+For example a common critic of using "feature importance" on random forests is that models like random forests can extract the value of a set of correlated features out of a single feature.
+So, for example if studying vulnerability to a specific kind of disease, a low feature importance for something like exercise may not mean anything given exercise patterns may be heavily correlated with things like, say, diet.
+As models get more and more complicated and include more features the likelihood that situations like these arise becomes greater and greater.
+
+For such things as fisheries assessments this is particularly concerning as the specific ranges of parameters derived from the models are used directly to set such things as allowable biological catch.
+Yet given we are often estimated recruitment, abundance, mortality, and in the case of spatial assessments things like movement and spatial heterogeneity all in the same model the only guarantee we have is that this choice of parameters fits the current data, not that they are parameters that are indicative of reality itself.
+
+In order to get a better guarantee that the parameters of the model are indicative of reality rather than just parameters that cause fit in the overall model one would need to estimate the parameters independently so cross feature "compensation" cannot occur and obscure the interpretation of the results.
+
+It has been noted that one of the most difficult aspects to fit is the movement matrix $X$ described above.
+So here too is another application of the model proposed in this document.
+Rather than having to depend on fitting $X$ through the assessment MLE routine such a fit could be taken directly from the behavioral model proposed herein thereby reducing the data requirements for the assessment itself and guaranteeing that $X$ is not being used to improve fit of the model in compensation for other important life history and abundance patterns.
+
+\subsubsection{Other Applications of a Movement Model}
+
+There are many decisions that have to be made when designing a spatial assessment and some of these would directly benefit from the kind of view our proposed model would provide:
+
+\begin{enumerate}
+\item \textbf{Choosing sub-regions} - Selecting the scale of sub-regions, the boundaries of those regions, and whether those regions should move over time requires intimate knowledge of fish movement.
+\item \textbf{Parameters to make regional} - Not all population dynamic parameters need be regional and obviously being able to share information across regions reduce the data requirements for the assessments.
+By knowing what kinds of covariates actually drive movement, data driven decision about what to keep regional versus not could be made.
+\item \textbf{Temporal resolution} - What time step to choose or whether only specific seasons need to modeled (and what those season are) would be another place in which data driven decisions would be aided by models of the kind proposed herein.
+\end{enumerate}
+
+\newpage
+
+\subsection{Modeling Movement}
+
+Modeling movement has traditionally divided into two categories - the Eulerian and Lagrangian approaches.
+In brief the former deals with movement as a series of flows throughout a space (that may or may not be discretized) while the latter treats each animal independently.
+We will provide a brief introduction to both approaches.
+
+\subsubsection{Eulerian Approach}
+
+The typical strategy behind the Eulerian approach is to subdivide the region in question into discrete subregions and then model movements between the boundaries of these regions as the result of advection and diffusion where advection is a directed movement and diffusion is not.
+
+These advection and diffusion terms are represented by partial differential equations where each may be parametrized by life history or environmental factors depending on the model in question.
+Data is included in the model as boundary conditions, for example where we were fish released and where were they recovered.
+Finally time too is discretized in order to allow finite difference methods to be used to solve the PDE's in question.
+
+The result of solution is a description of the advection and diffusion terms parametrized by whatever was included in the original model.
+With these in hand simulations can then be run to represent the overall movement modeled.
+
+The advantage of the Eulerian approach is that it is specifically attempting to catch population level movement patterns and therefore does not need as thorough parametrization as the langragian approaches.
+It is also therefore less data intense in its requirements and generally stock assessments are only interested in population or sub-population level movement anyways.
+
+\subsubsection{Lagrangian Approach}
+
+In contrast the Eulerian approach the Lagrangian approach attempts to model at the individual level. Therefore you'll see the terms Lagrangian and Individual Based Model (IBM) used interchangeably.
+
+The approaches used to build these kinds of models are extremely variable (a point we'll get back to momentarily) but follow the general paradigm of:
+
+\begin{enumerate}
+\item Collect some form of movement data (could be very high resolution or quite low resolution in some cases).
+\item Collect environmental data such as temperatures, measures of abundance, measures of productivity, flow/current data, etc.
+\item Craft a model that relates the specific queues from the environment into specific movements.
+\item Tune any parameters of those models through some kind of fitting paradigm.
+\end{enumerate}
+
+All of these steps are presented so vaguely because there really are no general paradigms for how these are done.
+For example, in the last step - tuning the parameters - strategies range from MLE style routines, to "by-eye" tuning, to weighted averages across multiple different allowable parameters.
+\newline
+
+Such approaches however have been shown to be able to model very fine scale movement dynamics and have been especially useful in applications in rivers where structures like dams are being designed and the ecological impacts are wanting to be understood.
+These kinds of fine scale dynamics are not possible with Eulerian approaches.
+
+IBM's can also capture much finer detail about \textit{why} specific movements occur because they are effectively modeling the individual itself.
+Whereas such detail ends up getting washed out by Eulerian approaches the focus on the individual means very specific details of behavior have to be sorted out and understood.
+A feature which, depending on the situation, can be a pro or a con.
+
+Finally it has been noted that one can derive Eulerian models from Lagrangian ones make the latter perhaps the more general strategy. \newline
+
+\subsubsection{Struggles in Building IBM's}
+
+There are a series of issues that have arisen in building IBM's of which we will note a few key issues that relate to our proposal:
+
+\begin{enumerate}
+\item \textbf{High Precision as a Requirement} - IBM's typically predict what an animal \textit{will} do as opposed to the likelihoods of what it \textit{could} do.
+This creates the requirement for capturing \textit{all} the meaningful drivers of variability in the behavior of the animal.
+To see why this is consider the example of a creature that can either choose to stay put or move to another location.
+If the more likely of the two actions is to stay put a deterministic model \textit{must} predict that the animal will \textit{always} stay put.
+But this is obviously non-sensical.
+However because the model is deterministic in nature the modeler cannot depend on stochasticity to save the day - instead they must find whatever is driving the difference between staying put and moving on.
+And this can cascade into greater and greater requirements for modeling the "interal state" of an animal.
+\item \textbf{Obscurity} - given the inherent complexity of IBM's that arises from the point above it gets increasingly difficult to make IBM's clear to new users or potential developers.
+There can often be many dozens of different parameters, functions, and "internal states" to keep track of without even considering all of the external modeling that has to go in place to provide for all of this.
+And this makes for not just complicated descriptions but equally if not more complicated code.
+\item \textbf{Hand Made} - almost all IBM's that the author has come across have been hand-coded in the sense that every functional relationship is specifically crafted and outlined by the modeler.
+This is obviously a very laborious process but it also means that no two IBM's are the same and that lack of generalization or abstraction makes exchanging lessons learned near difficult to impossible.
+\item \textbf{No Generalization of Fitting or Diagnostics} - because each IBM is effectively unique the methods for fitting, the data used, and the diagnostics developed are all unique as well.
+This once again increases the workload of the modeler, makes exchanging lessons learned and best practices extremely difficult, and also makes starting a new model quite prohibitive.
+\end{enumerate}
+
+\subsubsection{Comparison to Proposed Odds Model}
+
+Our odds modeling approach looks to directly address these issues by:
+
+\begin{enumerate}
+\item \textbf{Modeling Likelihoods} - by modeling likelihoods instead of having a deterministic model we no longer \textit{have} to model each kind of relevant variance, instead the variance can be encoded through stochasticity.
+While we would obviously like to know what creates the variance this allows us to build our model iteratively and to whatever complexity is allowable by the data.
+It also means we can build models that can be used in contexts where specific kinds of environmental or "internal state" data may be missing.
+\item \textbf{Information as the Model} - by using machine learning (ML) we are drawing the attention away from the specifics of the model itself to what information was used to train it and what metrics were used to select the best fit.
+This makes the model far clearer and creates a generalized paradigm for describing any such model.
+\item \textbf{Machine Learning} - as described in our rationale for creating these models, ML allows us to remove much of the labor in finding the approximate function that relates our inputs and outputs and instead allows us to place our focus on creating quality information to input to the model.
+\item \textbf{MLE in All Cases} - odds modeling allows us to use the same measure of fitness every single time - likelihood.
+And because all of the models share the same format the form of diagnostics and fitting can also be shared.
+\end{enumerate}
+
+\newpage
+
 \subsection{Chinook Ecology}
@@
 \subsection{Other Applications of PSAT Data}
-
-\subsection{Spatial Modeling}
-
-\subsection{Spatial Assessment}
 

### 2024-06-05 proposal/Proposal.tex
starter chinook eco

@@
 
-\subsection{Chinook Ecology}
+\subsection{Other Applications of PSAT Data}
 
-\subsection{Chinook Management}
+\newpage
 
-\subsection{Other Applications of PSAT Data}
+\subsection{Chinook}
+
+Chinook salmon (\textit{Oncorhynchus tshawytscha}) is the largest species of Pacific salmon.
+Like other Pacific salmon the species is both anadromous and semelparous leading to a very interesting and distinct life history.
+
+Juvenile Chinook are born in rivers where they remain for some time before heading to coastal waters.
+When they make this migration down river is quite variable and dependent on the region in question with juveniles entering anywhere from their first to second years of life.
+This leads to a specific system of determining age that includes both years in freshwater and years at sea.
+For example if a particular fish spends one winter in freshwater and so far 2 at sea it would be a 1.2 age fish.
+If it spent zero winters in freshwater and 3 at sea it would be a 0.3 age fish.
+Clearly both of these examples are the same overall age.
+
+The immature fish then spend time (up to five years) growing in the marine environment before maturing and then migrating back to the river where they were born to breed and subsequently die.
+The timing of such homing behavior is clearly temperature dependent with species further north having to avoid ice build up during the cooler months.
+The mechanism behind this homing ability is still a question under active research.
+\newline
+
+Given this homing behavior Chinook populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies.
+Indeed using tagging data it has been demonstrated that both initial and later marine distribution is highly dependent on the river of origin for the fish in question.
+
+Beyond this large scale analysis of fisheries dependent data has also demonstrated some of the likely migration patterns the species exhibits during its rearing phase at sea and electronic tagging studies have shown how quickly and with what accuracy natal homing can occur.
+
+All in all then, Chinook represent a highly mobile species of fish.\newline
+
+Broadly distributed through out the Northern pacific, chinook are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side.
+
+Marine residency in the North Eastern pacific is primarily coastal from California up to Western Alaska and then spans into much of the Eastern Bering Sea. \newline
+
+Chinook are well known piscivores with analysis of stomach contents finding primarily fish and squid as their main sources of prey.
+It has been noted that this diet is similar to that of Coho salmon but that it is unlikely that the two species experience much cross-species competition for these resources.
+
+Chinook themselves are prey for a few species such as Salmon sharks and whales and dolphins but it is still not clear how much mortality in later life stages is due to predation as opposed to other sources.
+Size selective predation on earlier life stages is a well known driver of natural mortality.
+
+Other drivers of natural mortality include a variety of known diseases, starvation during low productivity months, and potential competition for food with species like whales.\newline
+
+Chinook have been the subject of hatchery programs since the 1950s with the past decade seeing hatchery releases of 254million individuals on average.
+The vast majority of these individuals (~200million) have come from hatchery programs in the United States with the remaining significant portion from programs in Canada.
+
+Fish are released at weights of around 6-9 grams and a large number of these have a micro-wire with an identifiable code etched into them inserted into their snouts.
+These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook sub-populations.
+The etchings are such that each cohort from each river can be identified allowing for long term studies of fish movement.
+
+The initial purpose of these hatcheries programs was to supplement losses in juveniles due to targeting adult fish in river systems however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact.
+For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments.
+\newline
+
+There are valuable tribal, commercial, and recreational fisheries of Chinook salmon through the Pacific.
+
+Given the highly mobile nature of Chinook independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Fishery Management Council.
+Federal management councils (NOAA for the United States) then collaborate with the Pacific Fishery Management Council to set harvest quotas and other management objectives.
+The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals.
+
+These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook form so many reasonably separated genetic populations.
+Therefore management is further complicated by the fact that some sub populations are considered overfished while others are not.
+Indeed several sub populations are actually listed as endangered under the Endangered species act.
+
+This latter fact as serious implications for non-Chinook fisheries such as the incredibly lucrative Walleye Pollock fishery which, if enough endangered Chinook are caught, can experience a moratorium.
+Therefore there is a great deal of interest in limiting bycatch of Chinook in these fisheries.
+
+To manage this diversity the United States uses a variety of tools including but not limited to - differentiated management zones, minimum size limits, bag limits for recreational fisheries, allowed seasons and quotas, and fishing gear restrictions.
+\newline
+
+Finally, Chinook salmon are especially prized for their large size which has led to serious questions about size selective mortality from fishing as both recreational and commercial fisheries have incentives for targeting larger fish.
+While it is not yet known whether fisheries are responsible there is clear evidence that fish size in Chinook populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook's primary predators.
+\newline
+
+All in all then given their ecology and the corresponding complexities of management, Chinook represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
 

### 2024-06-06 proposal/Proposal.tex
ready for bibliography
bib for ecology

--- 2024-06-04 proposal/Proposal.tex
@@
 
-\subsection{Chinook Ecology}
+\subsection{Chinook}
 
-\subsection{Chinook Management}
+Chinook salmon (\textit{Oncorhynchus tshawytscha}) are the largest species of Pacific salmon \cite{oceanchinook}.
+Like other Pacific salmon the species is both anadromous and semelparous leading to a very interesting and distinct life history.
 
-\subsection{Other Applications of PSAT Data}
+Juvenile Chinook are born in the rivers where their parents spawned.
+Sometime in their first or second year of life they transition to the marine environment where they will spend the rest of their life before maturation \cite{oceanchinook}.
+This leads to a specific system of determining age that includes both years in freshwater and years at sea.
+For example if a particular fish spends one winter in freshwater and so far 2 at sea it would be a 1.2 age fish.
+If it spent zero winters in freshwater and 3 at sea it would be a 0.3 age fish.
+Clearly both of these examples are the same overall age.
+
+The immature fish then spend time (up to five years) growing in the marine environment before maturing and then migrating back to the river where they were born to breed and subsequently die \cite{oceanchinook}.
+The timing of such homing behavior is clearly season dependent with species further north having to avoid ice build up during the cooler months \cite{oceanchinook}.
+The mechanism behind this homing ability is still a question under active research \cite{bracis2012}.
+\newline
+
+Given this homing behavior, Chinook populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies.
+Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook is highly dependent on the region of origin for the fish in question \cite{shelton2019} \cite{tucker2019}.
+
+Beyond this, large scale analysis of fisheries dependent data has also demonstrated some of the likely migration patterns the species exhibits during its rearing phase at sea \cite{langan2024}.
+
+Studies like these point to Chinook being a highly mobile, spatially heterogeneous species.\newline
+
+Broadly distributed through out the Northern pacific, chinook are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side \cite{oceanchinook}.
+
+Marine residency in the North Eastern pacific is primarily coastal from California up to Western Alaska and then spans into much of the Eastern Bering Sea \cite{oceanchinook} \cite{langan2024}.
+\newline
+
+Chinook are well known piscivores with several studies of stomach contests finding primarily fish and squid as their main sources of prey.
+It has been noted that this diet is similar to that of Coho salmon but that it is unlikely that the two species experience much cross-species competition for these resources \cite{oceanchinook}.
+
+Chinook themselves are prey for a few species such as Salmon sharks, whales, and dolphins but it is still not unclear how much mortality in later life stages is due to predation as opposed to other sources.
+Size selective predation on earlier life stages is a well known driver of natural mortality \cite{oceanchinook}.
+
+Other drivers of natural mortality include a variety of known diseases, starvation during periods of low ocean productivity, and potential competition for food with species like whales \cite{oceanchinook}.\newline
+
+Chinook have been the subject of hatchery programs since the 1950s with the past decade seeing hatchery releases of 254million individuals on average.
+The vast majority of these individuals (~200million) have come from hatchery programs in the United States with the remaining significant portion from programs in Canada \cite{oceanchinook}.
+
+Fish are released at weights of around 6-9 grams and a large number of these have a micro-wire with an identifiable code etched into them inserted into their snouts.
+These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook sub-populations \cite{shelton2019}.
+The etchings provide a unique label for each cohort from each hatchery.
+
+The initial purpose of these hatcheries programs was to supplement losses in juveniles due to targeting of adult fish in river systems, however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact.
+For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments \cite{oceanchinook}.
+\newline
+
+There are valuable tribal, commercial, and recreational fisheries of Chinook salmon throughout the Pacific \cite{salmonplan}.
+
+Given the highly mobile nature of Chinook independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Fishery Management Council.
+Federal management councils (NOAA for the United States) then collaborate with the Pacific Fishery Management Council to set harvest quotas and other management objectives.
+The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals \cite{NOAA22}.
+
+These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook form so many different genetic sub-populations \cite{salmonplan} - a fact that further complicates management.
+Indeed several sub populations are actually listed as endangered under the Endangered species act \cite{NOAA24}.
+
+This combined with other threatened stocks \cite{NOAA24} means that bycatch avoidance is a serious issue for Chinook management and for other overlapping fisheries such as the Walleye Pollock fishery \cite{ianelli2015}.
+
+To manage this diversity the United States uses a variety of tools including but not limited to - differentiated management zones, minimum size limits, bag limits for recreational fisheries, allowed seasons and quotas, and fishing gear restrictions \cite{salmonplan}.
+\newline
+
+Finally, Chinook salmon are especially prized for their large size which has led to serious questions about size selective mortality from fishing as both recreational and commercial fisheries have incentives for targeting larger fish.
+While it is not yet known whether fisheries are responsible there is clear evidence that fish size in Chinook populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook's primary predators \cite{kendall2011}\cite{oceanchinook}.
+\newline
+
+All in all then given their ecology and the corresponding complexities of management, Chinook represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
 
@@
 
+\newpage
+
+\section{Bibliography}
+
+\bibliographystyle{apalike}
+\bibliography{reference}
 \end{document}

### 2024-06-06 proposal/Proposal.tex
cite for related work

@@
 
-Fisheries management is as diverse a subject as the species and cultures to which it is applied.
-It simultaneously has to deal with creatures from Bluefin Tuna who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters.
+Fisheries management is as diverse a subject as the species and cultures to which it is applied \cite{waltersmartell} \cite{king}.
+It simultaneously has to deal with creatures from Bluefin Tuna who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \cite{prince2010}.
 There are species that school and species that consume those schools.
 Some live entirely at sea, some in inland rivers and lakes, and some, like Salmon and eels spend time in both.
-There are species whose lives last the span of a year to others that can live longer than any of us human beings.
+There are species whose lives last the span of a year to others that can live longer than any us human beings.
 
 And to capture this vast array of diversity are similarly diverse strategies and gears.
-From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial bands, and everything in between the variety in strategies to catch fish are numerous.
+From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial bands, and everything in between the variety in strategies that exist to catch fish are numerous \cite{waltersmartell} \cite{king}.
 
@@
 What requirements of management are there?
+\cite{waltersmartell} \cite{king}
 
@@
 
-As with any such endeavor that can come in so many forms and captures so many different people, interests, and potential problems and opportunities, the best creative solutions come when everyone is working together to build and understand those solutions.
+As with any such endeavor that can come in so many forms and captures so many different people, interests, and potential problems and opportunities, the best creative solutions come when everyone is working together to build and understand those solutions \cite{prince2010} \cite{waltersmartell}.
 And of utmost importance here is the inclusion of those who actually go out and do the fishing.
@@
 
-First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world.
+First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world \cite{prince2010} \cite{lorenzen2010}.
 By having simulations that can simply be shown and laid down on a map that places the spatio-temporal modeling in familiar context one should be able to make objective or confirmation extremely easy.
@@
 While we've already noted that the creative solutions in fisheries management are as diverse as the people, species, and objectives present, there are some specific kinds of strategies that are used across the world to do spatial management.
-By and large they are comprised of performing spatial closures, however those closures differ with regards to who they effect and for how long.
+By and large they are comprised of performing spatial closures, however those closures differ with regards to who they effect and for how long \cite{selig2016} \cite{little2014}.
 \newline
@@
 
-Unsurprisingly then, for any kind of spatial management tool in use today, spatio-temporal understanding of the stock is key to both definition and efficacy.
+Unsurprisingly then, for any kind of spatial management tool in use today, spatio-temporal understanding of the stock is key to both definition and efficacy \cite{selig2016} \cite{little2014}.
 Therefore it seems likely that bringing high resolution spatio-temporal data to the bear on the issue can only help make such techniques more nuanced, precise, and thereby effective.
@@
 Fisheries assessments are one way in which unknown parameters important to management are estimated.
-By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so are arise from the state model.
-Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit and the observation model is encoded as a set of distributions dependent on those parameters thus allowing for a likelihood to be computed against known data for any given set of parameters.
+By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so on arise from the state model \cite{king}.
+Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit and the observation model is encoded as a set of distributions dependent on those parameters thus allowing for a likelihood to be computed against known data for any given set of parameters \cite{sippel2014}.
 This, in turn, allows for maximum likelihood estimation (MLE) routines to then fit the likely parameters of the assessment.
@@
 So, as an example, you might have your state model be comprised of age cohorts that have a specific growth pattern (fit externally) and then you leave as parameters natural mortality and recruitment.
-Then you might model vulnerability to a couple different kinds of gear, provide an expected distribution for catches given your underlying abundance by cohort, and then "simulate" a few years.
-Matching this back to real catch data would then allow you to estimate your natural mortality and recruitment by year.
+Vulnerability might then be modeled to capture a couple different kinds of gear, provide an expected distribution for catches given your underlying abundance by cohort, and then "simulate" a few years.
+Matching this back to real catch data would then allow you to estimate your natural mortality and recruitment by year \cite{king} \cite{ASAP}.
 \newline
@@
 
-As has been noted by several authors most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit.
-However from studies of fish behavior and the fact that fish don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases.
+As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit \cite{sippel2014} \cite{punt2019} \cite{goethel2023}.
+However from studies of fish behavior and the fact that fish don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \cite{lorenzen2010}.
 
-Therefore in recent years there has been work in creating new assessments that are able to capture spatial heterogeneity and movement.
+Therefore in recent years there has been work in creating new assessments that are able to capture spatial heterogeneity and movement \cite{punt2019} \cite{cadrin2020}.
 There are effectively two different kinds of heterogeneity that such models have to capture - that in recruitment and that in post recruitment movement and connectivity (or lack thereof).
@@
 
-Different assessment paradigms will have varying degree of resolution in the sub-regions, will estimate certain parameters freely to not at all, leave $X$ free or constrain it on other kinds of assumptions (indeed some assessments don't model movement at all).
+Different assessment paradigms will have varying degree of resolution in the sub-regions, will estimate certain parameters freely to not at all, leave $X$ free or constrain it on other kinds of assumptions (indeed some assessments don't model movement at all) \cite{punt2019} \cite{goethel2023}.
 
-Ultimately what is bounding the freedom in parametrizing these assessments is the availability of data to actually fit all of these parameters - a subject to which we turn next.
+Ultimately what is bounding the freedom in parametrizing these assessments is the availability of data to actually fit all of these parameters \cite{goethel2023} - a subject to which we turn next.
 
@@
 
-Therefore a huge step forward in being able to run these kinds of assessments was the development of "integrated assessments" - assessments that can take advantage of pretty much any kind of data collected from the fishery.
+Therefore a huge step forward in being able to run these kinds of assessments was the development of "integrated assessments" - assessments that can take advantage of pretty much any kind of data collected from the fishery \cite{sippel2014}.
 The process here falls neatly into our earlier description of how an assessment works - for some new kind of data (say acoustic tag data) one must add a new component to our observation model that describes the distribution that would generate our acoustic tag data.
@@
 
-There are of course issues that arise in terms of how to weight one data type against another when doing MLE but by and large integrated assessments are a boon because they increase the data available to fit assessments and have led to several tag integrated assessments that include electronic tag data in addition to more standard mark and recapture tag data.
+There are of course issues that arise in terms of how to weight one data type against another when doing MLE \cite{sippel2014} \cite{punt2019} but by and large integrated assessments are a boon because they increase the data available to fit assessments and have led to several tag integrated assessments that include electronic tag data in addition to more standard mark and recapture tag data.
 
@@
 
-With the recency of their development and overall complexity it has been noted that diagnosing goodness of fit and relevance of spatial assessments is an art that is still in development (although several papers exist that given guidance on best practices in choosing and using spatial assessment paradigms).
+With the recency of their development and overall complexity it has been noted that diagnosing goodness of fit and relevance of spatial assessments is an art that is still in development (although several papers exist that given guidance on best practices in choosing and using spatial assessment paradigms \cite{goethel2023} \cite{punt2019}).
 
@@
 I.e.
-are the simplifications introducing bias or errors that would lead to contradictory results from sound management based on the advice of the assessment?
+are the simplifications introducing bias or errors that would lead to contradictory results from sound management based on the advice of the assessment \cite{cadrin2020}?
 
@@
 
-The way out of this hole, as many authors have noted, is to use simulations.
+The way out of this hole, as many authors have noted, is to use simulations \cite{cadrin2020} \cite{goethel2023}.
 Specifically one can use simulations to do hypothesis testing of the form - if my stock had spatial structure of the form X would my specific assessment capture that at all or would it introduce biases and/or errors as a result?
@@
 Something that any student of statistical or Machine Learning learns early on is that models can be deceiving in interpretation due to correlations in different kinds of data or parameters.
-For example a common critic of using "feature importance" on random forests is that models like random forests can extract the value of a set of correlated features out of a single feature.
+For example a common critique of using "feature importance" on random forests is that models like random forests can extract the value of a set of correlated features out of a single feature.
 So, for example if studying vulnerability to a specific kind of disease, a low feature importance for something like exercise may not mean anything given exercise patterns may be heavily correlated with things like, say, diet.
@@
 
-For such things as fisheries assessments this is particularly concerning as the specific ranges of parameters derived from the models are used directly to set such things as allowable biological catch.
+For such things as fisheries assessments this is particularly concerning as the specific ranges of parameters derived from the models are used directly to set such things as allowable biological catch \cite{king}.
 Yet given we are often estimated recruitment, abundance, mortality, and in the case of spatial assessments things like movement and spatial heterogeneity all in the same model the only guarantee we have is that this choice of parameters fits the current data, not that they are parameters that are indicative of reality itself.
@@
 
-It has been noted that one of the most difficult aspects to fit is the movement matrix $X$ described above.
+It has been noted that one of the most difficult aspects to fit is the movement matrix $X$ described above \cite{punt2019}.
 So here too is another application of the model proposed in this document.
@@
 
-There are many decisions that have to be made when designing a spatial assessment and some of these would directly benefit from the kind of view our proposed model would provide:
+There are many decisions that have to be made when designing a spatial assessment \cite{cadrin2020} \cite{goethel2023} and some of these would directly benefit from the kind of view our proposed model would provide:
 
@@
 Modeling movement has traditionally divided into two categories - the Eulerian and Lagrangian approaches.
-In brief the former deals with movement as a series of flows throughout a space (that may or may not be discretized) while the latter treats each animal independently.
+In brief the former deals with movement as a series of flows throughout a space (that may or may not be discretized) while the latter treats each animal independently \cite{waltersmartell}.
 We will provide a brief introduction to both approaches.
@@
 
-The typical strategy behind the Eulerian approach is to subdivide the region in question into discrete subregions and then model movements between the boundaries of these regions as the result of advection and diffusion where advection is a directed movement and diffusion is not.
+The typical strategy behind the Eulerian approach is to subdivide the region in question into discrete subregions and then model movements between the boundaries of these regions as the result of advection and diffusion where advection is a directed movement and diffusion is not \cite{sibert1999} \cite{lehodey2008}.
 
-These advection and diffusion terms are represented by partial differential equations where each may be parametrized by life history or environmental factors depending on the model in question.
+These advection and diffusion terms are represented by partial differential equations where each may be parametrized by life history or environmental factors depending on the model in question \cite{lehodey2008}.
 Data is included in the model as boundary conditions, for example where we were fish released and where were they recovered.
-Finally time too is discretized in order to allow finite difference methods to be used to solve the PDE's in question.
+Finally time too is discretized in order to allow finite difference methods to be used to solve the PDE's in question \cite{sibert1999}.
 
@@
 The advantage of the Eulerian approach is that it is specifically attempting to catch population level movement patterns and therefore does not need as thorough parametrization as the langragian approaches.
-It is also therefore less data intense in its requirements and generally stock assessments are only interested in population or sub-population level movement anyways.
+It is also therefore less data intense in its requirements and generally stock assessments are only interested in population or sub-population level movement anyways \cite{lehodey2008}.
 
@@
 
-The approaches used to build these kinds of models are extremely variable (a point we'll get back to momentarily) but follow the general paradigm of:
+The approaches used to build these kinds of models are extremely variable \cite{grimm} (a point we'll get back to momentarily) but follow the general paradigm of:
 
@@
 All of these steps are presented so vaguely because there really are no general paradigms for how these are done.
-For example, in the last step - tuning the parameters - strategies range from MLE style routines, to "by-eye" tuning, to weighted averages across multiple different allowable parameters.
+For example, in the last step - tuning the parameters - strategies range from integration of deterministic models against boundary conditions \cite{phillipsa2018}, to "by-eye" tuning \cite{goodwin2006}, to weighted averages across multiple different allowable parameters \cite{burke2016}.
 \newline
 
-Such approaches however have been shown to be able to model very fine scale movement dynamics and have been especially useful in applications in rivers where structures like dams are being designed and the ecological impacts are wanting to be understood.
+Such approaches however have been shown to be able to model very fine scale movement dynamics and have been especially useful in applications in rivers where structures like dams are being designed and the ecological impacts are wanting to be understood \cite{goodwin2006} \cite{grimm}.
 These kinds of fine scale dynamics are not possible with Eulerian approaches.
@@
 
-Finally it has been noted that one can derive Eulerian models from Lagrangian ones make the latter perhaps the more general strategy. \newline
+Finally it has been noted that one can derive Eulerian models from Lagrangian ones make the latter perhaps more general \cite{phillipsa2018}. \newline
 
@@
 
-There are a series of issues that have arisen in building IBM's of which we will note a few key issues that relate to our proposal:
+There are a series of issues \cite{grimm} that have arisen in building IBM's of which we will note a few key issues that relate to our proposal:
 

### 2024-06-06 proposal/Proposal.tex

@@
 
-\title{Master's Research Proposal (IN PROGRESS)}
+\title{Master's Research Proposal (DRAFT)}
 
@@
 
-However if you are interested in the rationales behind these objectives or diving into related work has already been done by others, the \textbf{Rationale} and \textbf{Related Work} sections are for you.
+However if you are interested in the rationales behind these objectives or diving into related work has already been done by others, the \textbf{Rationale} and \textbf{Related Context} sections are for you.
 
@@
 
-Probabilistic machine classification of the kind we were pointing to in the last section has two pitfalls when it comes to modeling behavior.
+Probabilistic machine classification \cite{durr} of the kind we were pointing to in the last section has two pitfalls when it comes to modeling behavior.
 
@@
 From studies that started with Beverton and Holt's work on recruitment in fisheries we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to a pattern as illustrated in Fig.
-\ref{fig:recruitment}.
+\ref{fig:recruitment} \cite{waltersmartell}.
 
@@
 What we see is a clear pattern in recruitment verses spawners for low spawners and the eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
-Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out.
+Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \cite{waltersmartell}.
 \newline
 
-The implication for fisheries management is pretty clear - sustainable fishing is in large part driven by just maintaining the spawning stock biomass (SSB) above that knee in the curve.
+The implication for fisheries management is pretty clear - sustainable fishing is in large part driven by just maintaining the spawning stock biomass (SSB) above that knee in the curve \cite{waltersmartell} \cite{king}.
 Indeed this is well noted as the basis of why fisheries \textit{can} exist in the first place.
@@
 Obviously we're talking here about maintaining a specific level of spawning stock biomass.
-But today's juveniles are tomorrows SSB and as fish grow they typically become more fecund.
+But today's juveniles are tomorrow's SSB and as fish grow they typically become more fecund.
 This means that in order to maintain our SSB over time we have to be careful about how we target each age group.
 This gets even more complicated for fish like Pacific salmon or many species of groupers where life history is intimately entwined with spawning.
-The salmon obviously only breed once and several species of groupers are protogynous hermaphrodites meaning that if you target a specific age group you are effectively targeting a specific sex.
+The salmon obviously only breed once \cite{NOAA24} and some species of groupers are protogynous hermaphrodites \cite{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex.
 Clearly understanding and managing our age class selectivity matters when it comes to maintaining a healthy SSB for years to come.
@@
 
-The other observation is that many species of fish actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries).
-Pacific salmon are once again a good example in that the home to specific streams and only breed with other salmon in those streams.
-Bluefin tuna are also known to form specific breeding aggregations, as are many tropical fishes.
-Recent research too also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior.
+The other observation is that many species of fish actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \cite{lorenzen2010} \cite{salmonplan} \cite{prince2010}.
+Pacific salmon are once again a good example in that the home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \cite{erisman2017}.
+Research also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior \cite{armstrong2001}.
 
 What does this mean?
-It means that if we trust our beverton-holt style model that instead of having a single SSB to manage we really have several SSB's - one for each sub stock.
+It means that if we trust our beverton-holt style model, instead of having a single SSB to manage we really have several SSB's - one for each sub stock.
 Obviously, if we set one global piece of management like an Allowable Biological Catch (ABC) that makes sense for a single SSB but then we have biased fishing mortality where one sub stock is getting fished out far more than the others we may drop below the SSB for that one stock without actually exceeding our expected ABC for the whole fishery.
-And this can certainly happen as different sub stocks will often exhibit quite different spatial behaviors in movement.
+And this can certainly happen as different sub stocks will often exhibit quite different spatial behaviors in movement \cite{tucker2019} \cite{punt2019} \cite{cadrin2020}.
 \newline
@@
 
-It is also worth noting that this knowledge around changes in selectivity in space and time is not only useful for considerations around SSB but also for other management issues such as bycatch avoidance and growth overfishing to name just two.
+It is also worth noting that this knowledge around changes in selectivity in space and time is not only useful for considerations around SSB but also for other management issues such as bycatch avoidance and growth overfishing.
 
@@
 
-Pop-up Satellite Archival Tags (PSATs) are a great way of collecting behavior data. They are defined by three key properties:
+Pop-up Satellite Archival Tags (PSATs) are a great way of collecting behavior data \cite{PSAT}. They are defined by three key properties:
 
@@
 We will look at movements on a day by day basis and depths at the transmitted resolution (2-5 minutes).
-For those tags retrieved, a much higher resolution is available (1-3 seconds) however most fish had data only from transmission and not from retrieval.
+For those tags retrieved, a much higher resolution is available (1-3 seconds) however most data is only guaranteed from transmission and not from retrieval \cite{PSAT}.
 \newline
@@
 At this point in time the main interest here is that Chinook present an interesting model species for testing out the efficacy of these models and simulations.
-Chinook are highly motile and undergo clear migration behaviors.
+Chinook are highly motile and undergo clear migration behaviors \cite{tucker2019} \cite{langan2024}.
 Therefore they should provide a good challenge to attempts at behavioral modeling and simulation.
@@
 
-\section{Related Work}
+\section{Related Context}
 
@@
 
-\subsection{Chinook}
+\subsection{Chinook Ecology and Management}
 

### 2024-06-07 proposal/Proposal.tex

@@
 
-It begins with the \textbf{objectives} so that you know immediately exactly what this proposal is hoping to achieve and whether this is of interest to you.  \newline
+It begins with the \textbf{objectives} so that you immediately know exactly what this proposal is hoping to achieve and whether this is of interest to you.  \newline
 
-Then it dives into how we will go about demonstrating success along those objectives, this will inform you on the specific \textbf{outcomes} to be expected. \newline
+Then it dives into how we will go about demonstrating success in those objectives. This will inform you on the specific \textbf{outcomes} to be expected. \newline
 
@@
 
-If these are compelling enough there is no need to read any further. \newline
+If these sections are compelling enough there is no need to read any further. \newline
 
-However if you are interested in the rationales behind these objectives or diving into related work has already been done by others, the \textbf{Rationale} and \textbf{Related Context} sections are for you.
+However if you are interested in the rationales behind these objectives or diving into the related context, the \textbf{Rationale} and \textbf{Related Context} sections are for you.
 
@@
 
-\textbf{Data:} PAST tagging data from the GOA. \newline
+\textbf{Data:} PSAT tagging data from the GOA. \newline
 
@@
 
-One of the functions of science is to build models of the world around us. Specifically to look for functions of the form:
+One of the functions of science is to build models of the world around us, specifically to look for functions of the form:
 
@@
 However, life is rarely so kind to us.
-From quantum physics to fisheries science the world is replete with examples where the same information does not always lead to the same conclusion.
+From quantum physics to fisheries science, the world is replete with examples where the same information does not always lead to the same conclusion.
 Sometimes that is because we do not possess all of the relevant information but sometimes things are just truly random.
@@
 I.e.
-I cannot know in advance if the data I am collecting is going to be useful or not.
+we cannot know in advance if the data we are collecting is going to be useful or not.
 Instead anyone doing modeling finds themselves in the cycle illustrated by Fig.
@@
 
-So finally to model behavior we'd like to fit an odds model of the form:
+So finally, to model behavior we'd like to fit an odds model of the form:
 
@@
 
-To the author's knowledge there is presently no tooling to fit and diagnose such models.
-
 \subsection{Objective 2}
 
-From studies that started with Beverton and Holt's work on recruitment in fisheries we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to a pattern as illustrated in Fig.
+From studies that started with Beverton and Holt's work on recruitment in fisheries, we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to a pattern as illustrated in Fig.
 \ref{fig:recruitment} \cite{waltersmartell}.
@@
 
-What we see is a clear pattern in recruitment verses spawners for low spawners and the eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
+What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
 Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \cite{waltersmartell}.
@@
 The other observation is that many species of fish actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \cite{lorenzen2010} \cite{salmonplan} \cite{prince2010}.
-Pacific salmon are once again a good example in that the home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \cite{erisman2017}.
+Pacific salmon are once again a good example in that they home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \cite{erisman2017}.
 Research also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior \cite{armstrong2001}.
@@
 What does this mean?
-It means that if we trust our beverton-holt style model, instead of having a single SSB to manage we really have several SSB's - one for each sub stock.
-Obviously, if we set one global piece of management like an Allowable Biological Catch (ABC) that makes sense for a single SSB but then we have biased fishing mortality where one sub stock is getting fished out far more than the others we may drop below the SSB for that one stock without actually exceeding our expected ABC for the whole fishery.
+It means that if we trust our Beverton-Holt style model, instead of having a single SSB to manage we really have several SSB's - one for each sub stock.
+Obviously, if we set one global piece of management like an Allowable Biological Catch (ABC) that makes sense for a single SSB but then have biased fishing mortality where one sub stock is getting fished out far more than the others we may drop below the SSB for that one stock without actually exceeding our expected ABC for the whole fishery.
 And this can certainly happen as different sub stocks will often exhibit quite different spatial behaviors in movement \cite{tucker2019} \cite{punt2019} \cite{cadrin2020}.
@@
 
-All of this to say that understanding heterogeneity in vulnerability by things like age class and sub stock is of the utmost importance for managing a fisheries and truly ensuring an appropriate amount of SSB remains year over year.
+All of this to say that understanding heterogeneity in vulnerability by things like age class and sub stock is of the utmost importance for managing a fishery and truly ensuring an appropriate amount of SSB remains year over year.
 
@@
 Such variables could include features like the age and genetic cohorts that we just described.
-Then we could take some baseline assumptions about abundance in space at a given time (say providing recruits at spawning grounds or taking a spatio-temporal assessment as a starting point) and then using our behavioral models to simulate time-steps forward.
-As we do these time steps we could account for natural mortality as well to complete the "environment" portion of our simulation.
+Then we could take some baseline assumptions about abundance in space at a given time (say providing recruits at spawning grounds or taking a spatio-temporal assessment as a starting point) and use our behavioral models to simulate time-steps forward.
+As we do, we could account for natural mortality as well to complete the "environment" portion of our simulation.
 
@@
 
-The latter two are then used along with tag release location to predict animal movement (latitude and longitude) using a proprietary statistical model.
+The latter two are then used along with tag release location to predict animal movement (latitude and longitude) using a proprietary statistical model \cite{PSAT}.
 
@@
 And of utmost importance here is the inclusion of those who actually go out and do the fishing.
-Fisheries management that creates a sense of inclusion and involvement with the fishing community is the best kind of management because of a number of things:
+Fisheries management that creates a sense of inclusion and involvement with the fishing community is the best kind of management due to a number of things:
 
@@
 
-But in general creating cooperation improves the overall solution whereas exclusion simply creates conflict that works against any kind of management.
+In general, creating cooperation improves the overall solution whereas exclusion simply creates conflict that works against any kind of management.
 Therefore the first priority on our list when it comes to doing good management (spatial or otherwise) has to be creating engagement.
@@
 First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world \cite{prince2010} \cite{lorenzen2010}.
-By having simulations that can simply be shown and laid down on a map that places the spatio-temporal modeling in familiar context one should be able to make objective or confirmation extremely easy.
+By having simulations that can simply be shown and laid down on a map that places the spatio-temporal modeling in familiar context, one should be able to make objections or confirmation much easier.
 Rather than having to attempt to condense down loads of parameters, summary statistics, and the like in a model, one can simply show a map and folks can point at things that line up or are in disagreement with their intuition and experience.
@@
 
-For example suppose that we have a more standard model that has been hand crafted by a scientist familiar with the species in question.
+For example, suppose that we have a more standard model that has been hand crafted by a scientist familiar with the species in question.
 They may have included specific reactions to, say, oceanic currents, temperatures, and primary productivity.
@@
 And this would be for one piece of feedback.
-If there are several this only gets more difficult.
+If there is a lot of feedback this only gets more difficult.
 
-Instead by using machine learning as proposed here, such feedback would amount to perhaps playing around with a few day light features, pushing those to the model, retraining using the same automation as for any of the other feature combinations, and then just reporting if that information helped or not.
-Feedback can be incorporated much more easily allowing for direct contribution to the model on the part of those who may have no modeling experience but have lots of practical experience that could suggest forms of information that would improve how targeted the model is.
+Instead, by using machine learning as proposed here, such feedback would amount to playing around with a few day light features, pushing those to the model, retraining using the same automation as for any of the other feature combinations, and then just reporting if that information helped or not.
+Feedback can be incorporated much more easily, allowing for direct contribution to the model on the part of those who may have no modeling experience but have lots of practical experience that could suggest forms of information that would improve how targeted the model is.
 \newline
@@
 Any forecasting model is always going to have issues predicting farther and farther into the future but fisheries dependent data could provide a means for making corrections to keep the model in line.
-This would amount to something like weather forecast where for the next few days the forecast is accurate enough to precise predict specific kinds of movement helpful to fishers themselves.
+This would amount to something like weather forecast where for the next few days the forecast is accurate enough to precisely predict specific kinds of movement helpful to fishers themselves.
 
-For example it could help them predict where to go to avoid high levels of bycatch, help optimize their use of fuel and labor when searching for fish, or, depending on the degree of subtlety in the model might even be able to help them target catches that are more or less valuable on the market itself.
+For example it could help them predict where to go to avoid high levels of bycatch, help optimize their use of fuel and labor when searching for fish, or, depending on the degree of subtlety in the model, might even be able to help them target catches that are more or less valuable on the market itself.
 
 With such direct incentives to fishers this might provide a jumping off point to bring them into helping collect the kinds and quantities of data that would allow for increased precision in these models.
-With such "weather forecasting" any such increases would directly benefit them in addition to providing benefits to management.
+With such "weather forecasting" any such increases would directly benefit them in addition to providing overall benefits to management.
 
@@
 By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so on arise from the state model \cite{king}.
-Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit and the observation model is encoded as a set of distributions dependent on those parameters thus allowing for a likelihood to be computed against known data for any given set of parameters \cite{sippel2014}.
+Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit.
+The observation model on the other hand is encoded as a set of distributions dependent on those parameters (and including some parameters of its own) thus allowing for a likelihood to be computed against known data for any given set of parameters \cite{sippel2014}.
 This, in turn, allows for maximum likelihood estimation (MLE) routines to then fit the likely parameters of the assessment.
@@
 As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit \cite{sippel2014} \cite{punt2019} \cite{goethel2023}.
-However from studies of fish behavior and the fact that fish don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \cite{lorenzen2010}.
+However from studies of fish behavior and the fact that fish obviously don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \cite{lorenzen2010}.
 
@@
 There are effectively two different kinds of heterogeneity that such models have to capture - that in recruitment and that in post recruitment movement and connectivity (or lack thereof).
-As such these models are usually structured as having multiple sub regions within the overall assessment region and then two deviations from the standard assessment follow:
+As such these models are usually structured as having multiple sub regions within the overall assessment region and then model the following:
 
 \begin{enumerate}
-\item First sub-region specific mortality, growth, recruitment, or other sub-population specific parameters may be estimated.
+\item Sub-region specific mortality, growth, recruitment, or other sub-population specific parameters may be estimated.
 The degree to which these are shared between or constrained by other regions is often related to the data available and how flexible the model is allowed to be.
@@
 
-With the recency of their development and overall complexity it has been noted that diagnosing goodness of fit and relevance of spatial assessments is an art that is still in development (although several papers exist that given guidance on best practices in choosing and using spatial assessment paradigms \cite{goethel2023} \cite{punt2019}).
+With the recency of their development and overall complexity it has been noted that diagnosing goodness of fit and relevance of spatial assessments is an art that is still in development (although several papers exist that give guidance on best practices in choosing and using spatial assessment paradigms \cite{goethel2023} \cite{punt2019}).
 
@@
 For such things as fisheries assessments this is particularly concerning as the specific ranges of parameters derived from the models are used directly to set such things as allowable biological catch \cite{king}.
-Yet given we are often estimated recruitment, abundance, mortality, and in the case of spatial assessments things like movement and spatial heterogeneity all in the same model the only guarantee we have is that this choice of parameters fits the current data, not that they are parameters that are indicative of reality itself.
+Yet given we are often estimating recruitment, abundance, mortality, and in the case of spatial assessments things like movement and spatial heterogeneity all in the same model the only guarantee we have is that this choice of parameters fits the current data, not that they are parameters that are indicative of reality itself.
 
@@
 \item \textbf{Choosing sub-regions} - Selecting the scale of sub-regions, the boundaries of those regions, and whether those regions should move over time requires intimate knowledge of fish movement.
-\item \textbf{Parameters to make regional} - Not all population dynamic parameters need be regional and obviously being able to share information across regions reduce the data requirements for the assessments.
+\item \textbf{Parameters to make regional} - Not all population dynamic parameters need be regional and obviously being able to share information across regions reduces the data requirements for the assessments.
 By knowing what kinds of covariates actually drive movement, data driven decision about what to keep regional versus not could be made.
@@
 Modeling movement has traditionally divided into two categories - the Eulerian and Lagrangian approaches.
-In brief the former deals with movement as a series of flows throughout a space (that may or may not be discretized) while the latter treats each animal independently \cite{waltersmartell}.
+In brief, the former deals with movement as a series of flows throughout a space (that may or may not be discretized) while the latter treats each animal independently \cite{waltersmartell}.
 We will provide a brief introduction to both approaches.
@@
 The result of solution is a description of the advection and diffusion terms parametrized by whatever was included in the original model.
-With these in hand simulations can then be run to represent the overall movement modeled.
+With these in hand, simulations can then be run to represent the overall movement modeled.
 
-The advantage of the Eulerian approach is that it is specifically attempting to catch population level movement patterns and therefore does not need as thorough parametrization as the langragian approaches.
+The advantage of the Eulerian approach is that it is specifically attempting to catch population level movement patterns and therefore does not need as thorough parametrization as the Langragian approaches.
 It is also therefore less data intense in its requirements and generally stock assessments are only interested in population or sub-population level movement anyways \cite{lehodey2008}.
@@
 
-In contrast the Eulerian approach the Lagrangian approach attempts to model at the individual level. Therefore you'll see the terms Lagrangian and Individual Based Model (IBM) used interchangeably.
+In contrast to the Eulerian approach the Lagrangian approach attempts to model at the individual level. Therefore you'll see the terms Lagrangian and Individual Based Model (IBM) used interchangeably.
 
@@
 
-Finally it has been noted that one can derive Eulerian models from Lagrangian ones make the latter perhaps more general \cite{phillipsa2018}. \newline
+Finally it has been noted that one can derive Eulerian models from Lagrangian ones making the latter perhaps more general \cite{phillipsa2018}. \newline
 
@@
 
-There are a series of issues \cite{grimm} that have arisen in building IBM's of which we will note a few key issues that relate to our proposal:
+There are a series of key issues \cite{grimm} that have arisen in building IBM's of which we will note a few that relate to our proposal:
 
@@
 However because the model is deterministic in nature the modeler cannot depend on stochasticity to save the day - instead they must find whatever is driving the difference between staying put and moving on.
-And this can cascade into greater and greater requirements for modeling the "interal state" of an animal.
+And this can cascade into greater and greater requirements for modeling the "internal state" of an animal.
 \item \textbf{Obscurity} - given the inherent complexity of IBM's that arises from the point above it gets increasingly difficult to make IBM's clear to new users or potential developers.
@@
 And this makes for not just complicated descriptions but equally if not more complicated code.
-\item \textbf{Hand Made} - almost all IBM's that the author has come across have been hand-coded in the sense that every functional relationship is specifically crafted and outlined by the modeler.
+\item \textbf{Hand Made} - almost all IBM's have been hand-coded in the sense that every functional relationship is specifically crafted and outlined by the modeler.
 This is obviously a very laborious process but it also means that no two IBM's are the same and that lack of generalization or abstraction makes exchanging lessons learned near difficult to impossible.
 \item \textbf{No Generalization of Fitting or Diagnostics} - because each IBM is effectively unique the methods for fitting, the data used, and the diagnostics developed are all unique as well.
-This once again increases the workload of the modeler, makes exchanging lessons learned and best practices extremely difficult, and also makes starting a new model quite prohibitive.
+This once again increases the workload of the modeler, makes exchanging lessons learned and best practices extremely difficult, and also makes starting a new model from scratch quite prohibitive.
 \end{enumerate}
@@
 \item \textbf{Modeling Likelihoods} - by modeling likelihoods instead of having a deterministic model we no longer \textit{have} to model each kind of relevant variance, instead the variance can be encoded through stochasticity.
-While we would obviously like to know what creates the variance this allows us to build our model iteratively and to whatever complexity is allowable by the data.
+While we would obviously like to know what creates the variance, this allows us to build our model iteratively and to whatever complexity is allowable by the data.
 It also means we can build models that can be used in contexts where specific kinds of environmental or "internal state" data may be missing.
@@
 \item \textbf{MLE in All Cases} - odds modeling allows us to use the same measure of fitness every single time - likelihood.
-And because all of the models share the same format the form of diagnostics and fitting can also be shared.
+And because all of the models share the same format, the form of diagnostics and fitting can also be shared.
 \end{enumerate}
@@
 Sometime in their first or second year of life they transition to the marine environment where they will spend the rest of their life before maturation \cite{oceanchinook}.
-This leads to a specific system of determining age that includes both years in freshwater and years at sea.
+This leads to a specific system of enumerating age that includes both years in freshwater and years at sea.
 For example if a particular fish spends one winter in freshwater and so far 2 at sea it would be a 1.2 age fish.
@@
 
-Broadly distributed through out the Northern pacific, chinook are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side \cite{oceanchinook}.
+Broadly distributed through out the Northern pacific, Chinook are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side \cite{oceanchinook}.
 
@@
 
-Chinook are well known piscivores with several studies of stomach contests finding primarily fish and squid as their main sources of prey.
+Chinook are well known piscivores with several studies of stomach contents finding primarily fish and squid as their main sources of prey.
 It has been noted that this diet is similar to that of Coho salmon but that it is unlikely that the two species experience much cross-species competition for these resources \cite{oceanchinook}.
 
-Chinook themselves are prey for a few species such as Salmon sharks, whales, and dolphins but it is still not unclear how much mortality in later life stages is due to predation as opposed to other sources.
+Chinook themselves are prey for animal groups such as Salmon sharks, whales, and dolphins but it is still unclear how much mortality in later life stages is due to predation as opposed to other sources.
 Size selective predation on earlier life stages is a well known driver of natural mortality \cite{oceanchinook}.
@@
 Chinook have been the subject of hatchery programs since the 1950s with the past decade seeing hatchery releases of 254million individuals on average.
-The vast majority of these individuals (~200million) have come from hatchery programs in the United States with the remaining significant portion from programs in Canada \cite{oceanchinook}.
+The vast majority of these individuals (approx.
+200million) have come from hatchery programs in the United States with the remaining significant portion from programs in Canada \cite{oceanchinook}.
 
@@
 These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook form so many different genetic sub-populations \cite{salmonplan} - a fact that further complicates management.
-Indeed several sub populations are actually listed as endangered under the Endangered species act \cite{NOAA24}.
+Indeed several sub populations are actually listed as endangered under the Endangered Species Act \cite{NOAA24}.
 
@@
 Finally, Chinook salmon are especially prized for their large size which has led to serious questions about size selective mortality from fishing as both recreational and commercial fisheries have incentives for targeting larger fish.
-While it is not yet known whether fisheries are responsible there is clear evidence that fish size in Chinook populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook's primary predators \cite{kendall2011}\cite{oceanchinook}.
+While it is not yet known whether fisheries are responsible, there is clear evidence that fish size in Chinook populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook's primary predators \cite{kendall2011}\cite{oceanchinook}.
 \newline
 
-All in all then given their ecology and the corresponding complexities of management, Chinook represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
+All in all then, given their ecology and the corresponding complexities of management, Chinook represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
 

### 2024-06-09 report/Report.tex
fitting an odds model

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[super]{nth}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{objective}{Objective}
\newtheorem{procedure}{Procedure}
\newtheorem{model}{Model}

\makeatletter
\renewcommand{\maketitle}{
\begin{center}

\pagestyle{empty}
\phantom{.}  %necessary to add space on top before the title
\vspace{3cm}

{\Huge \bf \@title\par}
\vspace{2.5cm}

{\LARGE Marcel Gietzmann-Sanders}\\[1cm]

{\Large\@date}

\vspace{2.5cm}
{\Large Dr. Andrew Seitz}\hspace{2cm}{\Large Dr. Curry Cunningham}\\[2cm]{\Large Michael Courtney, M.S.}\\[2cm]
College of Fisheries and Ocean Sciences\\
University of Alaska Fairbanks

\end{center}
}\makeatother

\title{Spatio-Temporal Modeling and Simulation}

\date{2024}
\setcounter{tocdepth}{2}
\begin{document}
\maketitle
\newpage
\tableofcontents
\newpage

\section{Odds Modeling}

\begin{objective}
To provide tooling that allows for using machine learning methods to fit models of the form

$$\psi_k = G(\eta_k)$$

that maximize the following objective:

$$\mathcal{L}=\prod_i P'(v_i | \eta_i)$$

where:

$$P'(v_i|\eta_i) = \frac{\psi_i}{\sum_k \psi_k}$$

These models will be known as odds models as they predict the "odds for" each outcome $v_k$ given the information contained in $\eta_k$.

\end{objective}

\subsection{Fitting an Odds Model}

All standard Machine Learning (ML) pipelines assume that you have at least two things - targets and features. In our case we certainly have the latter but our target $\psi_k$ is both unknown to us and also unmeasureable. So how are we to fit ML models if we have no target? In short, through iteration. Let's see how this can be done.\newline

First some notation to help us. Our data is composed of a series of decisions $D_j=\lbrace v_{jk} \rbrace$ where $j$ indicates each of the specific decisions and $k$ the options within each decision. For each iteration we will build a model $\hat{G_i}(\eta_{jk})$ using the pairs $\psi_{jk(i-1)},\eta_{jk}$. We will designate the outcome of that model $\phi_{jki}$:

$$\phi_{jki} = \hat{G_i}(\eta_{jk})$$

Then our proposed probabilities are:

$$P_i'(v_{jk}|\eta_{jk})=\frac{\phi_{jki}}{\sum_p \phi_{jpi}}$$

Then if the $k$ selected per decision $D_j$ is given by $s_j$ we want to maximize:

$$\mathcal{L}=\prod_j P_i'(v_{js_j} | \eta_{js_j})\rightarrow \sum_j \ln{\left[ P_i'(v_{js_j} | \eta_{js_j}) \right]}$$

Using this information we will then propose an update $u_{jki}$ s.t. $\psi_{jki}=\phi_{jki} + u_{jki}$ and repeat our iteration loop.

With that notation cleared up we can begin our iteration procedure. \newline

First, let's assume we already have a guess for $\psi_{jk(i-1)}$. We can therefore train our model off of the $\psi_{jk0},\eta_{jk}$ pairs in standard ML fashion. Specifically we will fit a model that optimizes Mean Squared Error (the most common objective across ML software packages):

$$\min{\left[ \sum_{jk}\left(\hat{G_i}(\eta_{jk}) - \psi_{jk(i-1)}\right)^2 \right]}$$

We now need to choose a set of updates $u_{jki}$. To get these we will turn to our overall objective function:

$$\sum_j \ln{\left[ P_i'(v_{js_j} | \eta_{js_j}) \right]}= \sum_j \ln{\left[ \frac{\phi_{js_ji}}{\sum_p \phi_{jpi}} \right]}$$

Let's look at the gradient of this w.r.t the $\phi_{jki}$. There are two cases:

$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{\sum_p \phi_{jpi}-\phi_{js_ji}}{\left( \sum_p \phi_{jpi} \right)^2}=\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}$$

$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{-\phi_{js_ji}}{\left( \sum_p \phi_{jpi} \right)^2} = \frac{-1 }{\sum_p \phi_{jpi}}$$

Next for point of illustration let's suppose that there are a set of $\psi_{jki}$ which we'll designate as $Z$ which share the same features $\eta$, i.e. our model has to give a single $\phi_{jki}$ for all such options. Our derivative then for that collection $Z$ is given by:

$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}-\sum_{\phi_{j\not{s_j}i}\in Z}\frac{1 }{\sum_p \phi_{jpi}}$$

Given classic optimization tactics we know that our function will be maximized when these sums are $0$ (and technically we'd also want to know that the second derivative was negative but we'll assume that's the case given how our iterations will work). \newline

With this in mind let's now propose that our updates are given by:

$$u_{jki} = \alpha_i \partial_{\phi_{jki}} \ln{\mathcal{L}} $$

where $\alpha_i$ is a constant we'll call our "learning rate". Note that by using this update we will increase our $\psi_{jki}$ guess where it corresponds to a taken option ($s_j$) and will decrease it where it corresponds to an option not taken ($\not{s_j}$). This will therefore push us towards maximizing $\ln{\mathcal{L}}$ as opposed to minimizing it. \newline

We are left with a final question - will our iteration sequence end when we've found the maximizing guesses of $\psi_{jki}$? To answer this we turn back to the term we are maximizing when fitting the $\hat{G_i}$:

$$\min{\left[ \sum_{jk}\left(\hat{G_i}(\eta_{jk}) - \psi_{jk(i-1)}\right)^2 \right]}$$

Our new fit will look like:

$$\sum_{jk}\left(\hat{G}_{i+1}(\eta_{jk}) - (\hat{G}_i(\eta_{jk}) + u_{jki})\right)^2=\sum_{jk}\left( \delta \hat{G}_{jk} - u_{jki}\right)^2$$

Given our $Z$ once again we can take the derivative w.r.t $\delta \hat{G}_{jk}$ where $\phi_{jki} \in Z$.

$$\partial_Z \left[ \sum_{jk}\left( \delta \hat{G}_{jk} - u_{jki}\right)^2 \right] = \sum_{\phi_{jki} \in Z} 2\left( \delta \hat{G}_{jk} - u_{jki}\right) = \sum_{\phi_{jki} \in Z} \delta \hat{G}_{jk} - \sum_{\phi_{jki} \in Z} u_{jki}$$

But now remember that if we've maximized w.r.t $Z$ that:

$$\sum_{\phi_{jki} \in Z} u_{jki} = 0$$

which means that in order for our partial derivative above to be zero (and therefore our error term be at a minimum) that $\delta \hat{G}_{jk}=0$. And this means our iteration will have stopped!

\newpage

\begin{procedure}{Fitting an Odds Model}
\begin{enumerate}
\item Collect decisions $D_j$ and corresponding features $\eta_{jk}$, options $v_{jk}$, and selection $s_j$. 
\item Make an initial guess $\psi_{jk0} = 1$.
\item Fit $\hat{G}_i$ on the pairs of $\psi_{jk(i-1)},\eta_{jk}$ using MSE to produce the $\phi_{jki}$.
\item Generate the $u_{jki} = \alpha_i \partial_{\phi_{jki}} \ln{\mathcal{L}}$ and produce a new set of $\psi_{jki}$.
\item Repeat 3 and 4 until, varying $\alpha_i$ until convergence.

$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}$$

$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=\frac{-1 }{\sum_p \phi_{jpi}}$$ 
\end{enumerate}

\end{procedure}

\end{document}

### 2024-06-10 mirrorverse/odds_model/tests/test_odds_model.py
new odds model

@@ -1,0 +1,144 @@
+"""
+Odds Model Tests
+"""
+
+# pylint: disable=missing-function-docstring, invalid-name, missing-class-docstring
+
+import unittest
+
+import pandas as pd
+import numpy as np
+from sklearn.ensemble import RandomForestRegressor
+
+from mirrorverse.odds_model.odds_model import (
+    get_central_likelihood,
+    get_probability,
+    get_proposed_utility,
+    OddsModel,
+)
+
+
+def test_get_central_likelihood():
+    X = pd.DataFrame(
+        {
+            "_selected": [False, True, True, False],
+            "probability": [0.1, 0.2, 0.3, 0.4],
+        }
+    )
+    print(X)
+    likelihood = get_central_likelihood(X)
+    assert likelihood == np.exp((np.log(0.2) + np.log(0.3)) / 2)
+
+
+def test_get_probability():
+    X = pd.DataFrame({"_decision": [0, 1, 1, 2], "utility": [1, 2, 3, 4]})
+    get_probability(X)
+    assert (X["probability"] == [1, 2 / 5, 3 / 5, 1]).all()
+    assert (X["sum_utility"] == [1, 5, 5, 4]).all()
+
+
+def test_get_proposed_utility():
+    X = pd.DataFrame(
+        {
+            "_decision": [0, 1, 1, 2],
+            "_selected": [True, False, True, True],
+            "utility": [1, 2, 3, 4],
+            "sum_utility": [1, 5, 5, 4],
+            "probability": [1, 2 / 5, 3 / 5, 1],
+        }
+    )
+    learning_rate = 0.9
+    get_proposed_utility(X, learning_rate)
+    assert (
+        abs(X["partial"] - [0, -1 / 5, 1 / 5 * 5 / 3 * (1 - 3 / 5), 0]).max() < 10**-10
+    )
+    factor = 2 / (1 / 5) * learning_rate
+    assert (X["step"] == X["partial"] * factor).all()
+    assert (X["proposed"] == X["utility"] + X["step"]).all()
+
+
+class TestOddsModel(unittest.TestCase):
+
+    def setUp(self):
+        self.X_train = pd.concat(
+            [
+                pd.DataFrame(  # 75% of the time 0 is selected
+                    {
+                        "_decision": list(range(75)) + list(range(75)),
+                        "_selected": [True] * 75 + [False] * 75,
+                        "feature": [0] * 75 + [1] * 75,
+                    }
+                ),
[…]

### 2024-06-10 mirrorverse/odds_model/tests/test_odds_model.py
woops

@@ -25,7 +25,6 @@ def test_get_central_likelihood():
             "probability": [0.1, 0.2, 0.3, 0.4],
         }
     )
-    print(X)
     likelihood = get_central_likelihood(X)
     assert likelihood == np.exp((np.log(0.2) + np.log(0.3)) / 2)
 

### 2024-06-10 report/Report.tex

@@
 
-$$\partial_Z \left[ \sum_{jk}\left( \delta \hat{G}_{jk} - u_{jki}\right)^2 \right] = \sum_{\phi_{jki} \in Z} 2\left( \delta \hat{G}_{jk} - u_{jki}\right) = \sum_{\phi_{jki} \in Z} \delta \hat{G}_{jk} - \sum_{\phi_{jki} \in Z} u_{jki}$$
+$$\partial_Z \left[ \sum_{jk}\left( \delta \hat{G} - u_{jki}\right)^2 \right] = \sum_{\phi_{jki} \in Z} 2\left( \delta \hat{G} - u_{jki}\right) = 2\sum_{\phi_{jki} \in Z} \delta \hat{G} - 2\sum_{\phi_{jki} \in Z} u_{jki}$$
 
@@
 
-which means that in order for our partial derivative above to be zero (and therefore our error term be at a minimum) that $\delta \hat{G}_{jk}=0$. And this means our iteration will have stopped!
+which means that in order for our partial derivative above to be zero (and therefore our error term be at a minimum) that $\delta \hat{G}=0$. And this means our iteration will have stopped!
 

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
started on dvc pipeline for depth model

@@ -1,0 +1,28 @@
+"""
+API for the Chinook Depth Model
+"""
+
+import click
+
+from mirrorverse.models.chinook_depth.data import (
+    load_depth_data,
+    add_depth_classes,
+)
+
+
+@click.command()
+@click.option("--function", required=True, type=str)
+@click.option("--output_file", type=str)
+@click.option("--input_file", type=str)
+@click.option("--depth_classes", type=str)
+def main(**kwargs):
+    functions = {
+        "load_depth_data": (load_depth_data, ["output_file"]),
+        "add_depth_classes": (
+            add_depth_classes,
+            ["input_file", "depth_classes", "output_file"],
+        ),
+    }
+    _callable, required_args = functions[kwargs["function"]]
+    args = [kwargs[arg] for arg in required_args]
+    _callable(*args)

### 2024-06-11 mirrorverse/models/chinook_depth/data.py

@@ -1,0 +1,80 @@
+"""
+Loading and Prepping Data for the Chinook Depth Model
+"""
+
+import pandas as pd
+import numpy as np
+from scipy.stats import norm
+
+from mirrorverse.warehouse.utils import get_engine
+
+
+def load_depth_data(output_file):
+    """
+    Inputs:
+    - output_file: str, path to save the output file
+
+    Loads the depth data from the warehouse and saves it to a csv file.
+    """
+    sql = """
+    select 
+        tag_key,
+        date_key,
+        depth,
+        epoch
+    from 
+        tag_depths
+    """
+    depth = pd.read_sql(sql, get_engine())
+    depth = depth[~np.isnan(depth["depth"])]
+    depth.to_csv(output_file, index=False)
+
+
+def select_a_class(depth, depth_classes):
+    """
+    Inputs:
+    - depth: float, the depth of the fish as recorded
+    - depth_classes: np.array, the depth classes to choose from
+
+    Outputs:
+    - int, the selected depth class
+
+    Selects a depth class based on the depth of the fish.
+
+    It turns out that PSAT summary data bins the depth into
+    intervals so the actual depth is not known. However
+    given the recorded depth we can estimate the depth classes
+    it could belong to and the likelihoods of each.
+    """
+    sd = (
+        depth * 0.08 / 1.96
+    )  # ~two standard deviations gives our 95% confidence interval
+    if sd == 0:
+        division = np.zeros(len(depth_classes))
+        division[0] = 1
+    else:
+        # we're going to assume the depth classes are sorted
+        z = (depth_classes - depth) / sd
+        division = norm.cdf(z)
+        division[1:] = division[1:] - division[:-1]
+    # if there aren't quite enough depth classes the
+    # probabilities may not sum to 1, so we'll normalize
+    division = division / division.sum()
+    return np.random.choice(depth_classes, p=division)
+
+
+def add_depth_classes(input_file, depth_classes, output_file):
+    """
+    Inputs:
+    - input_file: str, path to the input file
+    - depth_classes: str, string of depth classes as an array
+    - output_file: str, path to save the output file
+
+    Adds a depth class to the depth data and saves it to a csv file.
+    """
+    depth_data = pd.read_csv(input_file)
+    depth_classes = np.array(eval(depth_classes))
+    depth_data["depth_class"] = depth_data["depth"].apply(
+        lambda x: select_a_class(x, depth_classes)
+    )
+    depth_data.to_csv(output_file, index=False)

### 2024-06-11 setup.py

@@ -42,6 +42,7 @@ setup(
             "mirrorverse_build_blob=mirrorverse.blob.tree.blob_root:build_blob",
             "blob_train=mirrorverse.blob.train:main",
             "blob_states=mirrorverse.blob.states:main",
+            "mirrorverse_chinook_depth=mirrorverse.models.chinook_depth.api:main",
         ]
     },
 )

### 2024-06-11 mirrorverse/models/chinook_depth/warehouse.py
warehouse data

@@ -78,3 +78,45 @@ def add_depth_classes(input_file, depth_classes, output_file):
         lambda x: select_a_class(x, depth_classes)
     )
     depth_data.to_csv(output_file, index=False)
+
+
+def load_context_data(output_file):
+    """
+    Inputs:
+    - output_file: str, path to save the output file
+
+    Loads the context data from the warehouse and saves it to a csv file.
+    """
+    sql = """
+    select 
+        tt.*,
+        h.home_region,
+        e.elevation
+    from 
+        tag_tracks tt 
+        left join home_regions h
+            on tt.tag_key = h.tag_key
+        left join elevation e 
+            on tt.h3_level_4_key = e.h3_level_4_key
+    """
+    context = pd.read_sql(sql, get_engine())
+    context.to_csv(output_file, index=False)
+
+
+def join_in_context_data(input_files, output_file):
+    """
+    Inputs:
+    - input_files: list of str, paths to the input files
+    - output_file: str, path to save the output file
+
+    Joins the context data to the depth data and saves it to a csv file.
+    """
+    input_files = input_files.split(",")
+    depth_data = pd.read_csv(input_files[0])
+    context_data = pd.read_csv(input_files[1])
+    data = depth_data.merge(
+        context_data[
+            ["tag_key", "date_key", "longitude", "latitude", "home_region", "elevation"]
+        ]
+    )
+    data.to_csv(output_file, index=False)

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
not sure what happened...

@@ -7,6 +7,8 @@ import click
 from mirrorverse.models.chinook_depth.data import (
     load_depth_data,
     add_depth_classes,
+    load_context_data,
+    join_in_context_data,
 )
 
 
@@ -21,6 +23,14 @@ def main(**kwargs):
         "add_depth_classes": (
             add_depth_classes,
             ["input_file", "depth_classes", "output_file"],
+        ),
+        "load_context_data": (
+            load_context_data,
+            ["output_file"],
+        ),
+        "join_in_context_data": (
+            join_in_context_data,
+            ["input_files", "output_file"],
         ),
     }
     _callable, required_args = functions[kwargs["function"]]

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
not sure what happened...

@@ -4,7 +4,7 @@ API for the Chinook Depth Model
 
 import click
 
-from mirrorverse.models.chinook_depth.data import (
+from mirrorverse.models.chinook_depth.warehouse import (
     load_depth_data,
     add_depth_classes,
     load_context_data,

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
time features

@@ -10,6 +10,7 @@ from mirrorverse.models.chinook_depth.warehouse import (
     load_context_data,
     join_in_context_data,
 )
+from mirrorverse.models.chinook_depth.time import add_time_features
 
 
 @click.command()
@@ -31,6 +32,10 @@ def main(**kwargs):
         "join_in_context_data": (
             join_in_context_data,
             ["input_files", "output_file"],
+        ),
+        "add_time_features": (
+            add_time_features,
+            ["input_file", "output_file"],
         ),
     }
     _callable, required_args = functions[kwargs["function"]]

### 2024-06-11 mirrorverse/models/chinook_depth/time.py

@@ -1,0 +1,64 @@
+"""
+Time Features for the Chinook Depth Model
+"""
+
+import pandas as pd
+import numpy as np
+from suntimes import SunTimes
+
+
+def get_sunrise(lat, lon, date):
+    """
+    Inputs:
+    - lat: float, latitude
+    - lon: float, longitude
+    - date: str, date
+
+    Outputs:
+    - int, hour of sunrise
+    """
+    return SunTimes(longitude=lon, latitude=lat, altitude=0).risewhere(date, "UTC").hour
+
+
+def get_sunset(lat, lon, date):
+    """
+    Inputs:
+    - lat: float, latitude
+    - lon: float, longitude
+    - date: str, date
+
+    Outputs:
+    - int, hour of sunset
+    """
+    return SunTimes(longitude=lon, latitude=lat, altitude=0).setwhere(date, "UTC").hour
+
+
+def add_time_features(input_file, output_file):
+    """
+    Inputs:
+    - input_file: str, path to the input file
+    - output_file: str, path to save the output file
+
+    Adds time features to the data and saves it to a csv file.
+    """
+    data = pd.read_csv(input_file)
+    data["datetime"] = pd.to_datetime(data["epoch"], utc=True, unit="s")
+    data["date"] = data["datetime"].dt.date
+    data = data[np.abs(data["longitude"]) <= 180]
+    data["sunrise"] = data.apply(
+        lambda r: get_sunrise(r["latitude"], r["longitude"], r["date"]), axis=1
+    )
+    data["sunset"] = data.apply(
+        lambda r: get_sunset(r["latitude"], r["longitude"], r["date"]), axis=1
+    )
+
+    data["hour"] = data["datetime"].dt.hour
+
+    data.loc[data["sunrise"] > data["sunset"], "daytime"] = (
+        data["hour"] < data["sunset"]
+    ) | (data["hour"] > data["sunrise"])
+    data.loc[data["sunrise"] < data["sunset"], "daytime"] = (
+        data["hour"] > data["sunrise"]
+    ) & (data["hour"] < data["sunset"])
+
+    data.to_csv(output_file, index=False)

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
train working

@@ -11,13 +11,21 @@ from mirrorverse.models.chinook_depth.warehouse import (
     join_in_context_data,
 )
 from mirrorverse.models.chinook_depth.time import add_time_features
+from mirrorverse.models.chinook_depth.choices import fill_out_choices
+from mirrorverse.models.chinook_depth.model import split_data, train_model
 
 
 @click.command()
 @click.option("--function", required=True, type=str)
 @click.option("--output_file", type=str)
+@click.option("--output_files", type=str)
 @click.option("--input_file", type=str)
+@click.option("--input_files", type=str)
 @click.option("--depth_classes", type=str)
+@click.option("--train_fraction", type=float)
+@click.option("--features", type=str)
+@click.option("--learning_rate", type=float)
+@click.option("--iterations", type=int)
 def main(**kwargs):
     functions = {
         "load_depth_data": (load_depth_data, ["output_file"]),
@@ -36,6 +44,18 @@ def main(**kwargs):
         "add_time_features": (
             add_time_features,
             ["input_file", "output_file"],
+        ),
+        "fill_out_choices": (
+            fill_out_choices,
+            ["input_file", "output_file", "depth_classes"],
+        ),
+        "split_data": (
+            split_data,
+            ["input_file", "output_file", "train_fraction"],
+        ),
+        "train_model": (
+            train_model,
+            ["input_files", "output_files", "features", "learning_rate", "iterations"],
         ),
     }
     _callable, required_args = functions[kwargs["function"]]

### 2024-06-11 mirrorverse/models/chinook_depth/choices.py

@@ -1,0 +1,30 @@
+"""
+Creates Full Array of Choices
+"""
+
+import pandas as pd
+import numpy as np
+
+
+def fill_out_choices(input_file, output_file, depth_classes):
+    depth_classes = np.array(eval(depth_classes))
+    choices = pd.read_csv(input_file)
+    choices = (
+        choices.reset_index(drop=True)
+        .reset_index()
+        .rename(
+            {
+                "index": "_decision",
+                "depth_class": "selected_class",
+                "tag_key": "_identifier",
+            },
+            axis=1,
+        )
+    )
+    all_choices = choices[["_decision"]].merge(
+        pd.DataFrame({"depth_class": depth_classes}), how="cross"
+    )
+    choices = choices.merge(all_choices, how="outer", on="_decision")
+    choices["_selected"] = choices["depth_class"] == choices["selected_class"]
+    del choices["selected_class"]
+    choices.to_csv(output_file, index=False)

### 2024-06-11 mirrorverse/models/chinook_depth/model.py

@@ -1,0 +1,56 @@
+"""
+Building the Model Itself
+"""
+
+import json
+import pickle
+
+import pandas as pd
+import numpy as np
+from sklearn.ensemble import RandomForestRegressor
+
+from mirrorverse.odds_model.odds_model import OddsModel
+
+
+def split_data(input_file, output_file, train_fraction):
+    """
+    Inputs:
+    - input_file: str, path to the input file
+    - output_file: str, path to save the output file
+    - train_fraction: float, fraction of data to use for training
+
+    Splits the data into training and testing sets and saves them to csv files.
+    """
+    data = pd.read_csv(input_file)
+    ids = data["_identifier"].unique()
+    train_ids = np.random.choice(ids, int(train_fraction * len(ids)), replace=False)
+    test_ids = np.array(list(set(ids) - set(train_ids)))
+
+    print(f"Training set size: {len(train_ids)}")
+    print(f"Testing set size: {len(test_ids)}")
+
+    train_data = data[data["_identifier"].isin(train_ids)]
+    test_data = data[data["_identifier"].isin(test_ids)]
+    train_data.to_csv(f"train_{output_file}", index=False)
+    test_data.to_csv(f"test_{output_file}", index=False)
+
+
+def train_model(input_files, output_files, features, learning_rate, iterations):
+    input_files = input_files.split(",")
+    train_data = pd.read_csv(input_files[0])
+    test_data = pd.read_csv(input_files[1])
+    with open(input_files[2], "r") as f:
+        model_params = json.load(f)
+
+    features = eval(features)
+    assert "_decision" not in features
+    assert "_selected" not in features
+    features += ["_decision", "_selected"]
+
+    model = OddsModel(RandomForestRegressor(**model_params))
+    model.fit(train_data[features], test_data[features], learning_rate, iterations)
+
+    output_files = output_files.split(",")
+    model.diagnostics.to_csv(output_files[0], index=False)
+    with open(output_files[1], "wb") as f:
+        pickle.dump(model, f)

### 2024-06-11 mirrorverse/models/chinook_depth/warehouse.py

@@ -114,6 +114,8 @@ def join_in_context_data(input_files, output_file):
     input_files = input_files.split(",")
     depth_data = pd.read_csv(input_files[0])
     context_data = pd.read_csv(input_files[1])
+    depth_data["tag_key"] = depth_data["tag_key"].astype(str)
+    context_data["tag_key"] = context_data["tag_key"].astype(str)
     data = depth_data.merge(
         context_data[
             ["tag_key", "date_key", "longitude", "latitude", "home_region", "elevation"]

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
linting

@@ -2,6 +2,8 @@
 API for the Chinook Depth Model
 """
 
+# pylint: disable=eval-used
+
 import click
 
 from mirrorverse.models.chinook_depth.warehouse import (
@@ -27,6 +29,9 @@ from mirrorverse.models.chinook_depth.model import split_data, train_model
 @click.option("--learning_rate", type=float)
 @click.option("--iterations", type=int)
 def main(**kwargs):
+    """
+    Just a Router
+    """
     functions = {
         "load_depth_data": (load_depth_data, ["output_file"]),
         "add_depth_classes": (

### 2024-06-11 mirrorverse/models/chinook_depth/choices.py

@@ -2,11 +2,21 @@
 Creates Full Array of Choices
 """
 
+# pylint: disable=eval-used
+
 import pandas as pd
 import numpy as np
 
 
 def fill_out_choices(input_file, output_file, depth_classes):
+    """
+    Inputs:
+    - input_file: str, path to the input file
+    - output_file: str, path to save the output file
+    - depth_classes: str, string of depth classes as an array
+
+    Fills out the choices with all possible depth classes and saves it to a csv file.
+    """
     depth_classes = np.array(eval(depth_classes))
     choices = pd.read_csv(input_file)
     choices = (

### 2024-06-11 mirrorverse/models/chinook_depth/model.py

@@ -2,6 +2,8 @@
 Building the Model Itself
 """
 
+# pylint: disable=eval-used
+
 import json
 import pickle
 
@@ -36,6 +38,16 @@ def split_data(input_file, output_file, train_fraction):
 
 
 def train_model(input_files, output_files, features, learning_rate, iterations):
+    """
+    Inputs:
+    - input_files: str, paths to the input files
+    - output_files: str, paths to save the output files
+    - features: str, list of feature names
+    - learning_rate: float, learning rate for the model
+    - iterations: int, number of iterations to run
+
+    Trains the model and saves the diagnostics and model to csv and pickle files.
+    """
     input_files = input_files.split(",")
     train_data = pd.read_csv(input_files[0])
     test_data = pd.read_csv(input_files[1])

### 2024-06-11 mirrorverse/models/chinook_depth/warehouse.py

@@ -2,6 +2,8 @@
 Loading and Prepping Data for the Chinook Depth Model
 """
 
+# pylint: disable=eval-used
+
 import pandas as pd
 import numpy as np
 from scipy.stats import norm

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
downsampling and grid search

@@ -13,8 +13,11 @@ from mirrorverse.models.chinook_depth.warehouse import (
     join_in_context_data,
 )
 from mirrorverse.models.chinook_depth.time import add_time_features
-from mirrorverse.models.chinook_depth.choices import fill_out_choices
-from mirrorverse.models.chinook_depth.model import split_data, train_model
+from mirrorverse.models.chinook_depth.choices import fill_out_choices, downsample_data
+from mirrorverse.models.chinook_depth.model import (
+    split_data,
+    train_model,
+)
 
 
 @click.command()
@@ -28,6 +31,7 @@ from mirrorverse.models.chinook_depth.model import split_data, train_model
 @click.option("--features", type=str)
 @click.option("--learning_rate", type=float)
 @click.option("--iterations", type=int)
+@click.option("--size_per_individual", type=int)
 def main(**kwargs):
     """
     Just a Router
@@ -61,6 +65,10 @@ def main(**kwargs):
         "train_model": (
             train_model,
             ["input_files", "output_files", "features", "learning_rate", "iterations"],
+        ),
+        "downsample_data": (
+            downsample_data,
+            ["input_file", "output_file", "size_per_individual"],
         ),
     }
     _callable, required_args = functions[kwargs["function"]]

### 2024-06-11 mirrorverse/models/chinook_depth/choices.py

@@ -8,6 +8,28 @@ import pandas as pd
 import numpy as np
 
 
+def downsample_data(input_file, output_file, size_per_individual):
+    """
+    Inputs:
+    - input_file: str, path to the input file
+    - output_file: str, path to save the output file
+    - size_per_individual: int, number of samples to take per individual
+
+    Downsamples the data and saves it to a csv file.
+    """
+    data = pd.read_csv(input_file)
+    print("Current Size: ", len(data))
+    num_ids = len(data["tag_key"].unique())
+    print("Expected Size: ", num_ids * size_per_individual)
+    for i, _identifier in enumerate(data["tag_key"].unique()):
+        df = data[data["tag_key"] == _identifier]
+        df = df.sample(size_per_individual, replace=True)
+        if i == 0:
+            df.to_csv(output_file, index=False)
+        else:
+            df.to_csv(output_file, mode="a", header=False, index=False)
+
+
 def fill_out_choices(input_file, output_file, depth_classes):
     """
     Inputs:

### 2024-06-11 mirrorverse/models/chinook_depth/api.py
full pipeline

@@ -17,6 +17,7 @@ from mirrorverse.models.chinook_depth.choices import fill_out_choices, downsampl
 from mirrorverse.models.chinook_depth.model import (
     split_data,
     train_model,
+    do_search,
 )
 
 
@@ -32,6 +33,7 @@ from mirrorverse.models.chinook_depth.model import (
 @click.option("--learning_rate", type=float)
 @click.option("--iterations", type=int)
 @click.option("--size_per_individual", type=int)
+@click.option("--num_param_sets", type=int)
 def main(**kwargs):
     """
     Just a Router
@@ -69,6 +71,10 @@ def main(**kwargs):
         "downsample_data": (
             downsample_data,
             ["input_file", "output_file", "size_per_individual"],
+        ),
+        "do_search": (
+            do_search,
+            ["input_files", "output_files", "features", "iterations", "num_param_sets"],
         ),
     }
     _callable, required_args = functions[kwargs["function"]]

### 2024-06-11 mirrorverse/models/chinook_depth/model.py

@@ -12,6 +12,10 @@ import numpy as np
 from sklearn.ensemble import RandomForestRegressor
 
 from mirrorverse.odds_model.odds_model import OddsModel
+from mirrorverse.odds_model.search import randomized_odds_model_search
+
+
+pd.options.mode.chained_assignment = None
 
 
 def split_data(input_file, output_file, train_fraction):
@@ -66,3 +70,61 @@ def train_model(input_files, output_files, features, learning_rate, iterations):
     model.diagnostics.to_csv(output_files[0], index=False)
     with open(output_files[1], "wb") as f:
         pickle.dump(model, f)
+
+
+def clean(param_set):
+    """
+    Inputs:
+    - param_set: dict, parameter set
+
+    Outputs:
+    - dict, cleaned parameter set
+    """
+    _type_map = {
+        np.bool_: bool,
+        np.int64: int,
+    }
+    return {
+        k: _type_map[type(v)](v) if type(v) in _type_map else v
+        for k, v in param_set.items()
+    }
+
+
+def do_search(input_files, output_files, features, iterations, num_param_sets):
+    """
+    Inputs:
+    - input_files: str, paths to the input files
+    - output_files: str, paths to save the output files
+    - features: str, list of feature names
+    - iterations: int, number of iterations to run per fit
+    - num_param_sets: int, number of parameter sets to generate
+
+    Searches for the best model parameters and saves
+    the diagnostics and parameters to csv and json files.
+    """
+    input_files = input_files.split(",")
+    train_data = pd.read_csv(input_files[0])
+    test_data = pd.read_csv(input_files[1])
+    with open(input_files[2], "r") as f:
+        param_grid = json.load(f)
+
+    features = eval(features)
+    assert "_decision" not in features
+    assert "_selected" not in features
+    features += ["_decision", "_selected"]
+
+    param_sets, diagnostics = randomized_odds_model_search(
+        train_data[features],
+        test_data[features],
+        iterations,
+        RandomForestRegressor,
+        param_grid,
+        num_param_sets,
+        100,
+    )
+
+    output_files = output_files.split(",")
+    diagnostics.to_csv(output_files[0], index=False)
+    param_sets = [clean(param_set) for param_set in param_sets]
+    with open(output_files[1], "w") as f:
+        json.dump(param_sets, f, indent=4, sort_keys=True)

### 2024-06-11 mirrorverse/models/chinook_depth/time.py

@@ -61,4 +61,8 @@ def add_time_features(input_file, output_file):
         data["hour"] > data["sunrise"]
     ) & (data["hour"] < data["sunset"])
 
+    data["month"] = data["datetime"].dt.month
+    data["interval"] = 24 - data["sunrise"] + data["sunset"]
+    data["daytime"] = data["daytime"].astype(float)
+
     data.to_csv(output_file, index=False)

### 2024-06-11 setup.py
missing dependency

@@ -23,6 +23,7 @@ setup(
         "eralchemy2==1.3.8",
         "netCDF4==1.6.5",
         "earthengine-api==0.1.390",
+        "suntimes==1.1.2",
     ],
     entry_points={
         "console_scripts": [

### 2024-06-12 mirrorverse/models/chinook_depth/api.py
train model and infer

@@ -66,7 +66,7 @@ def main(**kwargs):
         ),
         "train_model": (
             train_model,
-            ["input_files", "output_files", "features", "learning_rate", "iterations"],
+            ["input_files", "output_files", "features", "iterations"],
         ),
         "downsample_data": (
             downsample_data,

### 2024-06-12 mirrorverse/models/chinook_depth/model.py

@@ -41,37 +41,6 @@ def split_data(input_file, output_file, train_fraction):
     test_data.to_csv(f"test_{output_file}", index=False)
 
 
-def train_model(input_files, output_files, features, learning_rate, iterations):
-    """
-    Inputs:
-    - input_files: str, paths to the input files
-    - output_files: str, paths to save the output files
-    - features: str, list of feature names
-    - learning_rate: float, learning rate for the model
-    - iterations: int, number of iterations to run
-
-    Trains the model and saves the diagnostics and model to csv and pickle files.
-    """
-    input_files = input_files.split(",")
-    train_data = pd.read_csv(input_files[0])
-    test_data = pd.read_csv(input_files[1])
-    with open(input_files[2], "r") as f:
-        model_params = json.load(f)
-
-    features = eval(features)
-    assert "_decision" not in features
-    assert "_selected" not in features
-    features += ["_decision", "_selected"]
-
-    model = OddsModel(RandomForestRegressor(**model_params))
-    model.fit(train_data[features], test_data[features], learning_rate, iterations)
-
-    output_files = output_files.split(",")
-    model.diagnostics.to_csv(output_files[0], index=False)
-    with open(output_files[1], "wb") as f:
-        pickle.dump(model, f)
-
-
 def clean(param_set):
     """
     Inputs:
@@ -128,3 +97,41 @@ def do_search(input_files, output_files, features, iterations, num_param_sets):
     param_sets = [clean(param_set) for param_set in param_sets]
     with open(output_files[1], "w") as f:
         json.dump(param_sets, f, indent=4, sort_keys=True)
+
+
+def train_model(input_files, output_files, features, iterations):
+    """
+    Inputs:
+    - input_files: str, paths to the input files
+    - output_files: str, paths to save the output files
+    - features: str, list of feature names
+    - iterations: int, number of iterations to run
+
+    Trains the model and saves the diagnostics and model to csv and pickle files.
+    """
+    input_files = input_files.split(",")
+    train_data = pd.read_csv(input_files[0])
+    test_data = pd.read_csv(input_files[1])
+    with open(input_files[2], "r") as f:
+        param_sets = json.load(f)
+    diagnostics = pd.read_csv(input_files[3])
+
+    model_params = param_sets[diagnostics[diagnostics["best"]]["_param_set"].values[0]]
+    learning_rate = model_params.pop("learning_rate")
+
+    features = eval(features)
+    assert "_decision" not in features
[…]

### 2024-06-12 mirrorverse/grouper.py
bugfixes

@@ -10,8 +10,9 @@ from multiprocessing import Pool
 import click
 import pandas as pd
 import numpy as np
-from sklearn.metrics import explained_variance_score
 from tqdm import tqdm
+
+from mirrorverse.odds_model.odds_model import get_central_likelihood
 
 
 # pylint: disable=missing-class-docstring, missing-function-docstring
@@ -71,7 +72,7 @@ def crossover(ind1, ind2, crossover_gene_rate):
             ind1[i], ind2[i] = ind2[i], ind1[i]
 
 
-def get_explained_variances(groups, identifiers, data):
+def get_central_likelihoods(groups, identifiers, data):
     """
     Inputs:
     - groups (list): the groups
@@ -80,25 +81,23 @@ def get_explained_variances(groups, identifiers, data):
         should have columns "_identifier", "selected",
         and "probability"
 
-    Returns the explained variances of the groups and the
+    Returns the central likelihoods of the groups and the
     counts of the groups.
 
     dict, dict
     """
     groups = pd.DataFrame({"_identifier": identifiers, "group": groups})
     merged_data = data.merge(groups, on="_identifier")
-    variances = {}
+    likelihoods = {}
     counts = {}
     for group in merged_data["group"].unique():
         filtered_data = merged_data[merged_data["group"] == group]
-        variances[group] = explained_variance_score(
-            filtered_data["selected"], filtered_data["probability"]
-        )
+        likelihoods[group] = get_central_likelihood(filtered_data)
         counts[group] = filtered_data["_identifier"].nunique()
-    return variances, counts
+    return likelihoods, counts
 
 
-def evaluate(individual, identifiers, data, explained_variance):
+def evaluate(individual, identifiers, data, central_likelihood):
     """
     Inputs:
     - individual (list): the individual to evaluate
@@ -106,16 +105,16 @@ def evaluate(individual, identifiers, data, explained_variance):
     - data (pd.DataFrame): the data to evaluate over
         should have columns "_identifier", "selected",
         and "probability"
-    - explained_variance (float): the explained variance
+    - central_likelihood (float): the central likelihood
         of the whole dataset
 
-    Returns the average difference between the explained
+    Returns the average difference between the central
     variance of the groups and the whole dataset.
     """
-    variances, counts = get_explained_variances(individual, identifiers, data)
+    likelihoods, counts = get_central_likelihoods(individual, identifiers, data)
     differences, counts_array = [], []
[…]

### 2024-06-12 mirrorverse/models/chinook_depth/api.py

@@ -19,6 +19,7 @@ from mirrorverse.models.chinook_depth.model import (
     train_model,
     do_search,
 )
+from mirrorverse.models.chinook_depth.grouping import merge_back
 
 
 @click.command()
@@ -75,6 +76,10 @@ def main(**kwargs):
         "do_search": (
             do_search,
             ["input_files", "output_files", "features", "iterations", "num_param_sets"],
+        ),
+        "merge_back": (
+            merge_back,
+            ["input_files", "output_file"],
         ),
     }
     _callable, required_args = functions[kwargs["function"]]

### 2024-06-12 mirrorverse/models/chinook_depth/grouping.py

@@ -1,0 +1,22 @@
+"""
+Grouping of the Chinook Depth model.
+"""
+
+import pandas as pd
+
+
+def merge_back(input_files, output_file):
+    """
+    Inputs:
+    - input_files (str): comma-separated list of input files
+    - output_file (str): output file
+
+    Merges the predictions with the choices.
+    """
+    input_files = input_files.split(",")
+    preds = pd.read_csv(input_files[0])[
+        ["_decision", "depth_class", "probability", "utility"]
+    ]
+    choices = pd.read_csv(input_files[1])
+    result = choices.merge(preds, on=["_decision", "depth_class"], how="inner")
+    result.to_csv(output_file, index=False)

### 2024-06-12 mirrorverse/models/chinook_depth/warehouse.py

@@ -29,6 +29,7 @@ def load_depth_data(output_file):
     """
     depth = pd.read_sql(sql, get_engine())
     depth = depth[~np.isnan(depth["depth"])]
+    depth["tag_key"] = "tk" + depth["tag_key"].astype(str)
     depth.to_csv(output_file, index=False)
 
 
@@ -102,6 +103,7 @@ def load_context_data(output_file):
             on tt.h3_level_4_key = e.h3_level_4_key
     """
     context = pd.read_sql(sql, get_engine())
+    context["tag_key"] = "tk" + context["tag_key"].astype(str)
     context.to_csv(output_file, index=False)
 
 

### 2024-06-12 mirrorverse/tests/test_grouper.py

@@ -12,7 +12,7 @@ from mirrorverse.grouper import (
     create_individual,
     mutate,
     crossover,
-    get_explained_variances,
+    get_central_likelihoods,
     evaluate,
 )
 
@@ -65,18 +65,18 @@ def test_crossover():
     assert ind2.fitness is None
 
 
-def test_get_explained_variances():
+def test_get_central_likelihoods():
     groups = [1, 1, 2, 2, 3]
     identifiers = [1, 2, 3, 3, 4]
     data = {
         "_identifier": [1, 2, 3, 3, 4],
-        "selected": [0, 1, 0, 1, 0],
-        "probability": [0, 1, 0.0, 0.5, 0],
+        "_selected": [False, True, False, True, True],
+        "probability": [0, 1, 0.0, 0.5, 0.25],
     }
     data = pd.DataFrame(data)
-    explained_variances, counts = get_explained_variances(groups, identifiers, data)
+    likelihoods, counts = get_central_likelihoods(groups, identifiers, data)
 
-    assert explained_variances == {1: 1.0, 2: 0.75, 3: 1.0}
+    assert likelihoods == {1: 1.0, 2: 0.5, 3: 0.25}
     assert counts == {1: 2, 2: 1, 3: 1}
 
 
@@ -85,10 +85,10 @@ def test_evaulate():
     identifiers = [1, 2, 3, 3, 4]
     data = {
         "_identifier": [1, 2, 3, 3, 4],
-        "selected": [0, 1, 0, 1, 0],
-        "probability": [0, 1, 0.0, 0.5, 0],
+        "_selected": [False, True, False, True, True],
+        "probability": [0, 1, 0.0, 0.5, 0.25],
     }
     data = pd.DataFrame(data)
     result = evaluate(groups, identifiers, data, 0.5)
-    expected_result = ((1 - 0.5) * 2 + (0.75 - 0.5) * 1 + (1 - 0.5) * 1) / 4
+    expected_result = ((1 - 0.5) * 2 + (0.5 - 0.5) * 1 + (0.5 - 0.25) * 1) / 4
     assert result == expected_result

### 2024-06-12 report/Report.tex
log odds boi

@@
 
-\section{Odds Modeling}
+\section{Log Odds Modeling}
 
@@
 
-$$P'(v_i|\eta_i) = \frac{\psi_i}{\sum_k \psi_k}$$
+$$P'(v_i|\eta_i) = \frac{e^{\psi_i}}{\sum_k e^{\psi_k}}$$
 
-These models will be known as odds models as they predict the "odds for" each outcome $v_k$ given the information contained in $\eta_k$.
+These models will be known as log odds models as they predict the "log odds for" each outcome $v_k$ given the information contained in $\eta_k$.
 
@@
 
-Then our proposed probabilities are:
+Our probability is therefore:
 
-$$P_i'(v_{jk}|\eta_{jk})=\frac{\phi_{jki}}{\sum_p \phi_{jpi}}$$
+$$P_i'(v_{jk}|\eta_{jk})=\frac{e^{\phi_{jki}}}{\sum_p e^{\phi_{jpi}}}$$
 
-Then if the $k$ selected per decision $D_j$ is given by $s_j$ we want to maximize:
+Now if the $k$ selected per decision $D_j$ is given by $s_j$ we want to maximize:
 
@@
 
-$$\sum_j \ln{\left[ P_i'(v_{js_j} | \eta_{js_j}) \right]}= \sum_j \ln{\left[ \frac{\phi_{js_ji}}{\sum_p \phi_{jpi}} \right]}$$
+$$\sum_j \ln{\left[ P_i'(v_{js_j} | \eta_{js_j}) \right]}= \sum_j \ln{\left[ \frac{e^{\phi_{js_ji}}}{\sum_p e^{\phi_{jpi}}} \right]}$$
 
@@
 
-$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{\sum_p \phi_{jpi}-\phi_{js_ji}}{\left( \sum_p \phi_{jpi} \right)^2}=\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}$$
+$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}
+=
+\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{\sum_p e^{\phi_{jpi}}-e^{\phi_{js_ji}}}{\left( \sum_p e^{\phi_{jpi}} \right)^2} e^{\phi_{js_ji}}
+=
+\frac{1}{\sum_p e^{\phi_{jpi}}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})} e^{\phi_{js_ji}}$$
 
-$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{-\phi_{js_ji}}{\left( \sum_p \phi_{jpi} \right)^2} = \frac{-1 }{\sum_p \phi_{jpi}}$$
+$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}
+=
+\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{-e^{\phi_{js_ji}}}{\left( \sum_p e^{\phi_{jpi}} \right)^2} e^{\phi_{j\not{s_j}i}}
+= 
+\frac{-1 }{\sum_p e^{\phi_{jpi}}}e^{\phi_{j\not{s_j}i}}$$
 
@@
 
-$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}-\sum_{\phi_{j\not{s_j}i}\in Z}\frac{1 }{\sum_p \phi_{jpi}}$$
+$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}e^{\phi_{js_ji}}-\sum_{\phi_{j\not{s_j}i}\in Z}\frac{1 }{\sum_p \phi_{jpi}}e^{\phi_{j\not{s_j}i}}$$
 
@@
 
-\begin{procedure}{Fitting an Odds Model}
+\begin{procedure}{Fitting a Log Odds Model}
 \begin{enumerate}
 \item Collect decisions $D_j$ and corresponding features $\eta_{jk}$, options $v_{jk}$, and selection $s_j$. 
-\item Make an initial guess $\psi_{jk0} = 1$.
+\item Make an initial guess $\psi_{jk0} = 0$.
 \item Fit $\hat{G}_i$ on the pairs of $\psi_{jk(i-1)},\eta_{jk}$ using MSE to produce the $\phi_{jki}$.
@@
 
-$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}$$
+$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=\frac{1}{\sum_p e^{\phi_{jpi}}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}e^{\phi_{js_ji}}$$
 
-$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=\frac{-1 }{\sum_p \phi_{jpi}}$$ 
+$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=\frac{-1 }{\sum_p e^{\phi_{jpi}}}e^{\phi_{j\not{s_j}i}}$$ 
 \end{enumerate}

### 2024-06-12 mirrorverse/grouper.py

@@ -12,7 +12,7 @@ import pandas as pd
 import numpy as np
 from tqdm import tqdm
 
-from mirrorverse.odds_model.odds_model import get_central_likelihood
+from mirrorverse.log_odds_model.log_odds_model import get_central_likelihood
 
 
 # pylint: disable=missing-class-docstring, missing-function-docstring

### 2024-06-12 mirrorverse/log_odds_model/log_odds_model.py

@@ -24,55 +24,52 @@ def get_central_likelihood(dataframe):
 def get_probability(dataframe):
     """
     Inputs:
-    - dataframe (pd.DataFrame): a dataframe with columns "utility"
+    - dataframe (pd.DataFrame): a dataframe with columns "log_odds"
         and "_decision"
 
     Adds/modifies the following columns to the dataframe:
-    - "sum_utility": the sum of the "utility" column for each decision
+    - "odds": the odds for of selecting each decision
+    - "sum_odds": the sum of the "odds" column for each decision
     - "probability": the probability of selecting each decision
     """
-    dataframe["sum_utility"] = dataframe.groupby("_decision")["utility"].transform(
-        "sum"
-    )
-    dataframe["probability"] = dataframe["utility"] / dataframe["sum_utility"]
+    dataframe["odds"] = np.exp(dataframe["log_odds"])
+    dataframe["sum_odds"] = dataframe.groupby("_decision")["odds"].transform("sum")
+    dataframe["probability"] = dataframe["odds"] / dataframe["sum_odds"]
 
 
-def get_proposed_utility(dataframe, learning_rate):
+def get_proposed_log_odds(dataframe, learning_rate):
     """
     Inputs:
-    - dataframe (pd.DataFrame): a dataframe with columns "utility",
-        "_selected", "sum_utility", "probability",
-    - learning_rate (float): a float between 0 and 1 that determines
-        how much the proposed utility values should be adjusted
+    - dataframe (pd.DataFrame): a dataframe with columns "log_odds",
+        "odds", "_selected", "sum_odds", "probability",
+    - learning_rate (float): a float between that determines
+        how much the proposed log odds values should be adjusted
 
     Adds/modifies the following columns to the dataframe:
-    - "partial": the partial derivative of the probability with respect to the utility
-    - "step": the maximum step that can be taken without causing the probability to be negative
-    - "proposed": the proposed utility for each decision
+    - "partial": the partial derivative of the probability with respect to the log odds
+    - "step": the step that will be taken in the direction of the gradient
+    - "proposed": the proposed log odds for each decision
     """
-    assert 0 < learning_rate < 1, "learning_rate must be between 0 and 1"
 
     # calculate the partial derivative of the probability with respect to the utility
     dataframe["partial"] = (
         1
-        / dataframe["sum_utility"]
+        / dataframe["sum_odds"]
         * (
[…]

### 2024-06-12 mirrorverse/log_odds_model/search.py

@@ -7,7 +7,7 @@ Hyperparameter search for the odds model.
 import numpy as np
 import pandas as pd
 
-from mirrorverse.odds_model.odds_model import OddsModel
+from mirrorverse.log_odds_model.log_odds_model import LogOddsModel
 
 
 def build_randomized_param_sets(param_grids, M, max_attempts):
@@ -65,7 +65,7 @@ def randomized_odds_model_search(
     for i, param_set in enumerate(param_sets):
         print(f"Running parameter set {i + 1}/{M}")
         to_pass = {k: v for k, v in param_set.items() if k != "learning_rate"}
-        model = OddsModel(model_class(**to_pass))
+        model = LogOddsModel(model_class(**to_pass))
         model.fit(train_data, test_data, param_set["learning_rate"], iterations)
         diagnostics = model.diagnostics
         for param, val in param_set.items():

### 2024-06-12 mirrorverse/log_odds_model/tests/test_search.py

@@ -9,7 +9,7 @@ import unittest
 import pandas as pd
 from sklearn.ensemble import RandomForestRegressor
 
-from mirrorverse.odds_model.search import (
+from mirrorverse.log_odds_model.search import (
     build_randomized_param_sets,
     randomized_odds_model_search,
 )

### 2024-06-12 mirrorverse/models/chinook_depth/model.py

@@ -11,8 +11,8 @@ import pandas as pd
 import numpy as np
 from sklearn.ensemble import RandomForestRegressor
 
-from mirrorverse.odds_model.odds_model import OddsModel
-from mirrorverse.odds_model.search import randomized_odds_model_search
+from mirrorverse.log_odds_model.log_odds_model import LogOddsModel
+from mirrorverse.log_odds_model.search import randomized_odds_model_search
 
 
 pd.options.mode.chained_assignment = None
@@ -124,7 +124,7 @@ def train_model(input_files, output_files, features, iterations):
     assert "_selected" not in features
     features += ["_decision", "_selected"]
 
-    model = OddsModel(RandomForestRegressor(**model_params))
+    model = LogOddsModel(RandomForestRegressor(**model_params))
     model.fit(train_data[features], test_data[features], learning_rate, iterations)
 
     output_data = test_data[features].copy()

### 2024-06-15 report/Report.tex
wip

@@
 
-\subsection{Fitting an Odds Model}
+\subsection{Fitting a Log Odds Model}
 
@@
 =
-\frac{1}{\sum_p e^{\phi_{jpi}}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})} e^{\phi_{js_ji}}$$
+1-P_i'(v_{js_j} | \eta_{js_j})$$
 
@@
 = 
-\frac{-1 }{\sum_p e^{\phi_{jpi}}}e^{\phi_{j\not{s_j}i}}$$
+-P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
 
@@
 
-$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\frac{1}{\sum_p \phi_{jpi}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}e^{\phi_{js_ji}}-\sum_{\phi_{j\not{s_j}i}\in Z}\frac{1 }{\sum_p \phi_{jpi}}e^{\phi_{j\not{s_j}i}}$$
+$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\left( 1-P_i'(v_{js_j} | \eta_{js_j}) \right)-\sum_{\phi_{j\not{s_j}i}\in Z}P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
 
@@
 
-$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=\frac{1}{\sum_p e^{\phi_{jpi}}}\frac{\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)}{P_i'(v_{js_j} | \eta_{js_j})}e^{\phi_{js_ji}}$$
+$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=1-P_i'(v_{js_j} | \eta_{js_j})$$
 
-$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=\frac{-1 }{\sum_p e^{\phi_{jpi}}}e^{\phi_{j\not{s_j}i}}$$ 
+$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=-P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$ 
 \end{enumerate}

### 2024-06-15 mirrorverse/log_odds_model/log_odds_model.py

@@ -52,18 +52,29 @@ def get_proposed_log_odds(dataframe, learning_rate):
     """
 
     # calculate the partial derivative of the probability with respect to the utility
-    dataframe["partial"] = (
-        1
-        / dataframe["sum_odds"]
-        * (
-            dataframe["_selected"]
-            * ((1 - dataframe["probability"]) / dataframe["probability"])
-            - (1 - dataframe["_selected"])
-        )
-        * dataframe["odds"]
+    # dataframe["partial"] = (
+    #    1
+    #    / dataframe["sum_odds"]
+    #    * (
+    #        dataframe["_selected"]
+    #        * ((1 - dataframe["probability"]) / dataframe["probability"])
+    #        - (1 - dataframe["_selected"])
+    #    )
+    #    * dataframe["odds"]
+    # )
+
+    dataframe["partial"] = dataframe["_selected"] * (1 - dataframe["probability"]) - (
+        (1 - dataframe["_selected"]) * dataframe["probability"]
     )
 
-    dataframe["step"] = dataframe["partial"] * learning_rate
+    dataframe["curvature"] = dataframe["probability"] * (1 - dataframe["probability"])
+    dataframe["total_curvature"] = dataframe.groupby("log_odds")["curvature"].transform(
+        "mean"
+    )
+
+    dataframe["step"] = (
+        dataframe["partial"] / dataframe["total_curvature"] * learning_rate
+    )
 
     # propose new utility values
     dataframe["proposed"] = dataframe["log_odds"] + dataframe["step"]
@@ -105,8 +116,8 @@ class LogOddsModel:
         diagnostics = []
 
         # setup an initial guess
-        X_train["log_odds"] = 0
-        X_test["log_odds"] = 0
+        X_train["log_odds"] = 3
+        X_test["log_odds"] = 3
 
         get_probability(X_train)
         get_probability(X_test)
@@ -121,11 +132,59 @@ class LogOddsModel:
             }
         )
 
+        CHANGE = None
+
         for i in tqdm(range(1, iterations + 1)):
-            self.model.fit(X_train[features], X_train["proposed"])
+            print(get_central_likelihood(X_train))
+            update = X_train[["_selected", "_decision", "proposed"]].rename(
+                {"proposed": "log_odds"}, axis=1
+            )
+            get_probability(update)
+            print(get_central_likelihood(update))
+
+            # TODO: remove me
+            print(X_train.groupby("log_odds")["curvature"].max().describe())
+
[…]

### 2024-06-15 mirrorverse/models/chinook_depth/warehouse.py

@@ -125,4 +125,10 @@ def join_in_context_data(input_files, output_file):
             ["tag_key", "date_key", "longitude", "latitude", "home_region", "elevation"]
         ]
     )
+
+    data['home_region'] = data['home_region'].fillna('UNKNOWN')
+    for option in ["SEAK", "UNKNOWN", "WA/OR", "BC"]:
+        data[option.lower()] = (data["home_region"] == option).astype(float)
+    data[['home_region', 'seak', 'unknown', 'wa/or', 'bc']]
+
     data.to_csv(output_file, index=False)

### 2024-06-15 book/Perspectives on Stewardship.tex
title page

--- 2024-06-15 report/Report.tex
@@
-\documentclass[11pt]{article}
+\documentclass[11pt,a5paper]{book}
 \usepackage[utf8]{inputenc}
@@
 \usepackage[super]{nth}
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{objective}{Objective}
-\newtheorem{procedure}{Procedure}
-\newtheorem{model}{Model}
 
-\makeatletter
-\renewcommand{\maketitle}{
-\begin{center}
-
-\pagestyle{empty}
-\phantom{.}  %necessary to add space on top before the title
-\vspace{3cm}
-
-{\Huge \bf \@title\par}
-\vspace{2.5cm}
-
-{\LARGE Marcel Gietzmann-Sanders}\\[1cm]
-
-{\Large\@date}
-
-\vspace{2.5cm}
-{\Large Dr. Andrew Seitz}\hspace{2cm}{\Large Dr. Curry Cunningham}\\[2cm]{\Large Michael Courtney, M.S.}\\[2cm]
-College of Fisheries and Ocean Sciences\\
-University of Alaska Fairbanks
-
-\end{center}
-}\makeatother
-
-\title{Spatio-Temporal Modeling and Simulation}
-
-\date{2024}
-\setcounter{tocdepth}{2}
+\title{Perspectives on Stewardship}
+\author{Marcel Gietzmann-Sanders}
+\date{}
+\setcounter{tocdepth}{1}
 \begin{document}
 \maketitle
-\newpage
 \tableofcontents
@@
 
-\section{Log Odds Modeling}
-
-\begin{objective}
-To provide tooling that allows for using machine learning methods to fit models of the form
-
-$$\psi_k = G(\eta_k)$$
-
-that maximize the following objective:
-
-$$\mathcal{L}=\prod_i P'(v_i | \eta_i)$$
-
-where:
-
-$$P'(v_i|\eta_i) = \frac{e^{\psi_i}}{\sum_k e^{\psi_k}}$$
-
-These models will be known as log odds models as they predict the "log odds for" each outcome $v_k$ given the information contained in $\eta_k$.
-
-\end{objective}
-
-\subsection{Fitting a Log Odds Model}
-
-All standard Machine Learning (ML) pipelines assume that you have at least two things - targets and features.
-In our case we certainly have the latter but our target $\psi_k$ is both unknown to us and also unmeasureable.
-So how are we to fit ML models if we have no target?
-In short, through iteration.
-Let's see how this can be done.\newline
-
-First some notation to help us.
-Our data is composed of a series of decisions $D_j=\lbrace v_{jk} \rbrace$ where $j$ indicates each of the specific decisions and $k$ the options within each decision.
-For each iteration we will build a model $\hat{G_i}(\eta_{jk})$ using the pairs $\psi_{jk(i-1)},\eta_{jk}$.
-We will designate the outcome of that model $\phi_{jki}$:
-
-$$\phi_{jki} = \hat{G_i}(\eta_{jk})$$
-
-Our probability is therefore:
-
-$$P_i'(v_{jk}|\eta_{jk})=\frac{e^{\phi_{jki}}}{\sum_p e^{\phi_{jpi}}}$$
-
-Now if the $k$ selected per decision $D_j$ is given by $s_j$ we want to maximize:
-
-$$\mathcal{L}=\prod_j P_i'(v_{js_j} | \eta_{js_j})\rightarrow \sum_j \ln{\left[ P_i'(v_{js_j} | \eta_{js_j}) \right]}$$
-
-Using this information we will then propose an update $u_{jki}$ s.t. $\psi_{jki}=\phi_{jki} + u_{jki}$ and repeat our iteration loop.
-
-With that notation cleared up we can begin our iteration procedure. \newline
-
-First, let's assume we already have a guess for $\psi_{jk(i-1)}$.
-We can therefore train our model off of the $\psi_{jk0},\eta_{jk}$ pairs in standard ML fashion.
-Specifically we will fit a model that optimizes Mean Squared Error (the most common objective across ML software packages):
-
-$$\min{\left[ \sum_{jk}\left(\hat{G_i}(\eta_{jk}) - \psi_{jk(i-1)}\right)^2 \right]}$$
-
-We now need to choose a set of updates $u_{jki}$. To get these we will turn to our overall objective function:
-
-$$\sum_j \ln{\left[ P_i'(v_{js_j} | \eta_{js_j}) \right]}= \sum_j \ln{\left[ \frac{e^{\phi_{js_ji}}}{\sum_p e^{\phi_{jpi}}} \right]}$$
-
-Let's look at the gradient of this w.r.t the $\phi_{jki}$. There are two cases:
-
-$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}
-=
-\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{\sum_p e^{\phi_{jpi}}-e^{\phi_{js_ji}}}{\left( \sum_p e^{\phi_{jpi}} \right)^2} e^{\phi_{js_ji}}
-=
-1-P_i'(v_{js_j} | \eta_{js_j})$$
-
-$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}
-=
-\frac{1}{P_i'(v_{js_j} | \eta_{js_j})}\frac{-e^{\phi_{js_ji}}}{\left( \sum_p e^{\phi_{jpi}} \right)^2} e^{\phi_{j\not{s_j}i}}
-= 
--P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
-
-Next for point of illustration let's suppose that there are a set of $\psi_{jki}$ which we'll designate as $Z$ which share the same features $\eta$, i.e.
-our model has to give a single $\phi_{jki}$ for all such options.
-Our derivative then for that collection $Z$ is given by:
-
-$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\left( 1-P_i'(v_{js_j} | \eta_{js_j}) \right)-\sum_{\phi_{j\not{s_j}i}\in Z}P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
-
-Given classic optimization tactics we know that our function will be maximized when these sums are $0$ (and technically we'd also want to know that the second derivative was negative but we'll assume that's the case given how our iterations will work).
-\newline
-
-With this in mind let's now propose that our updates are given by:
-
-$$u_{jki} = \alpha_i \partial_{\phi_{jki}} \ln{\mathcal{L}} $$
-
-where $\alpha_i$ is a constant we'll call our "learning rate".
-Note that by using this update we will increase our $\psi_{jki}$ guess where it corresponds to a taken option ($s_j$) and will decrease it where it corresponds to an option not taken ($\not{s_j}$).
-This will therefore push us towards maximizing $\ln{\mathcal{L}}$ as opposed to minimizing it.
-\newline
-
-We are left with a final question - will our iteration sequence end when we've found the maximizing guesses of $\psi_{jki}$?
-To answer this we turn back to the term we are maximizing when fitting the $\hat{G_i}$:
-
-$$\min{\left[ \sum_{jk}\left(\hat{G_i}(\eta_{jk}) - \psi_{jk(i-1)}\right)^2 \right]}$$
-
-Our new fit will look like:
-
-$$\sum_{jk}\left(\hat{G}_{i+1}(\eta_{jk}) - (\hat{G}_i(\eta_{jk}) + u_{jki})\right)^2=\sum_{jk}\left( \delta \hat{G}_{jk} - u_{jki}\right)^2$$
-
-Given our $Z$ once again we can take the derivative w.r.t $\delta \hat{G}_{jk}$ where $\phi_{jki} \in Z$.
-
-$$\partial_Z \left[ \sum_{jk}\left( \delta \hat{G} - u_{jki}\right)^2 \right] = \sum_{\phi_{jki} \in Z} 2\left( \delta \hat{G} - u_{jki}\right) = 2\sum_{\phi_{jki} \in Z} \delta \hat{G} - 2\sum_{\phi_{jki} \in Z} u_{jki}$$
-
-But now remember that if we've maximized w.r.t $Z$ that:
-
-$$\sum_{\phi_{jki} \in Z} u_{jki} = 0$$
-
-which means that in order for our partial derivative above to be zero (and therefore our error term be at a minimum) that $\delta \hat{G}=0$. And this means our iteration will have stopped!
-
-\newpage
-
-\begin{procedure}{Fitting a Log Odds Model}
-\begin{enumerate}
-\item Collect decisions $D_j$ and corresponding features $\eta_{jk}$, options $v_{jk}$, and selection $s_j$. 
-\item Make an initial guess $\psi_{jk0} = 0$.
-\item Fit $\hat{G}_i$ on the pairs of $\psi_{jk(i-1)},\eta_{jk}$ using MSE to produce the $\phi_{jki}$.
-\item Generate the $u_{jki} = \alpha_i \partial_{\phi_{jki}} \ln{\mathcal{L}}$ and produce a new set of $\psi_{jki}$.
-\item Repeat 3 and 4 until, varying $\alpha_i$ until convergence.
-
-$$\partial_{\phi_{js_ji}}\ln{\mathcal{L}}=1-P_i'(v_{js_j} | \eta_{js_j})$$
-
-$$\partial_{\phi_{j\not{s_j}i}}\ln{\mathcal{L}}=-P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$ 
-\end{enumerate}
-
-\end{procedure}
-
 \end{document}

### 2024-06-21 report/Report.tex
main

@@
 
-\subsection{Fitting a Log Odds Model}
+\subsection{Fitting an Odds Model}
 
@@
 =
-1-P_i'(v_{js_j} | \eta_{js_j})$$
+1-P_i'(v_{js_j} | \eta_{js_j}) $$
 
@@
 
-$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\left( 1-P_i'(v_{js_j} | \eta_{js_j}) \right)-\sum_{\phi_{j\not{s_j}i}\in Z}P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
+$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)-\sum_{\phi_{j\not{s_j}i}\in Z}P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
 
@@
 
+There is however an issue with this approach. Recall that our gradient is:
+
+$$\partial_Z \ln{\mathcal{L}}=\sum_{\phi_{js_ji} \in Z}\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)-\sum_{\phi_{j\not{s_j}i}\in Z}P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
+
+Let's however look at the second derivative of $\ln{\mathcal{L}}$:
+
+$$\partial_{\phi_{jk}} P_i'(v_{jk} | \eta_{jk})=\partial_{\phi_{jk}} \frac{e^{\phi_{jk}}}{\sum_p e^{\phi_{jp}}}=(1-P_i'(v_{jk} | \eta_{jk}))P_i'(v_{jk} | \eta_{jk})$$
+
+$$\partial^2_Z \ln{\mathcal{L}}=-\sum_{\phi_{js_ji} \in Z}\left(1-P_i'(v_{js_j} | \eta_{js_j}) \right)P_i'(v_{js_j} | \eta_{js_j})-\sum_{\phi_{j\not{s_j}i}\in Z}\left(1-P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}}) \right)P_i'(v_{j\not{s_j}} | \eta_{j\not{s_j}})$$
+
+What's important to note is that this function will be near 0 at $P\approx 1$ or $P\approx 0$ and will have its largest magnitude near $P \approx 0.5$.
+Why is this an issue?
+Well we know that as we get close to our maximum value for $\ln{\mathcal{L}}$ that our steps towards that maximum will get smaller and smaller.
+This is just because our steps are based on a derivative and we are seeking where the derivative is zero.
+Put another way getting to our max gets harder the closer we get to that maximum.
+
+Now normally one deals with this by using the second derivative (the curvature) as a kind of correction.
+As you get closer to your maximum, you can use your curvature to guide how quickly you can move.
+If the curvature is very small in magnitude you can move more quickly because the odds of you overshooting your maximum are smaller.
+If your curvature is very large you'll slow things down because you know only small steps are required to make big changes to the derivative (the thing we are ultimately trying to set to a specific value here).
+I.e.
+you can modulate your step size by the inverse of the curvature.
+
+However, in our case we don't actually know what the curvature is because we don't know what $Z$ is!
+Therefore we have to plan for the most volatile case - the case where $P\approx 0.5$.
+And that means that for any of our $Z$'s where $P$ is approaching 1 or 0 we'll be moving at a snail's pace (convergence will take forever).
+
+Now in case you're wondering whether we could solve the $P\approx 0.5$ case first and then move onto the others remember that anytime we change any of the $\phi$ all the other probabilities change.
+So we have to solve this problem all at once.
+Because of that and because we cannot depend on knowing the $Z$ (different $Z$ can give the same value $\phi$) we're stuffed with having to take extraordinarily long convergence times.
+And this more or less means we've got no shot of using this in practice.
+\newline
+
+Log odds modeling doesn't work in practice for non-parametric models.
+
 \end{document}

### 2024-06-21 report/Report.tex
logarch

@@
 \subsection{Fitting an Odds Model}
+
+Our objective in using machine learning (ML) is to reduce the time spent looking for the particular form of $G(\eta_k)$ by instead allowing that form to be fit given the data at hand.
+Deep Learning (DL) models are especially well suited to this problem as they are both parametric (see Section \ref{convergence issues} for why non-parametric models are not well suited to our problem) and have been shown to be able to represent just about any function given a large enough network \cite{LiquetMokaNazarathy2024DeepLearning}.
+There is also a robust field around probabilistic deep learning that, as we shall see in a moment, we can take advantage of \cite{durr}.
+\newline
+
+Taking a look at traditional probabilistic classifiers we can see that they have two things in common with our intended log odds modeling.
+
+First, probabilistic classifiers use the categorical cross entropy cost function to optimize their weights.
+This cost function happens to be just another name for optimizing for the Log Likelihood of the data \cite{durr}.
+Therefore probabilistic DL shares the same objective function as we do.
+
+Second, the final layer in a probabilistic DL network is the one that produces a probability for each class.
+This layer uses the softmax function as it's activation function which if if $a_i=W_i x+b_i$ is given by - $\frac{e^{a_i}}{\sum_k e^{a_k}}$.
+This means that if we can get our $\psi_k$ to be the $a_i$ that we'll be, in effect, training a log odds model.
+
+We can do this by taking advantage of Keras' Functional API \cite{kerasfunctional} which allows us to split our network into a branches and also share layers between branches.
+Specifically we can follow the following steps:
+
+\begin{enumerate}
+\item Define a series of layers that represent our underlying log odds model (these layers should end in a layer of output size 1 which will represent our $\psi_k$).
+\item Split our training decisions into $N$ choices.
+\item Create a branch in our overall model for each of these $N$ choices.
+\item Converge the branches at a softmax layer where the weights are the identity matrix $I$ and the biases all 0 (this ensures we directly pass through our $\psi_k$ to the softmax function).
+\item Use categorical cross entropy as our fitness function.
+\end{enumerate}
+
+This architecture is illustrated by Fig. \ref{fig:logarch}.
+
+\begin{figure}[h!] 
+  \includegraphics[width=\linewidth]{logarch.png}
+  \caption{Overall Architecture}
+  \label{fig:logarch}
+\end{figure}
+
+The main advantage of this architecture as opposed to a more classic approach to probabilistic DL is the fact that the weights are shared on each of the branches.
+This has the effect of drastically reducing the number of parameters in our network meaning that we can use far less data to train this model then we'd need to train a full bore, fully connected, probabilistic model.
+Each of our columns in Fig.
+\ref{fig:logarch} is in fact the same model and therefore updates coming from each of the columns goes to all of the columns.
+
+This, then, solves one of our issues - reducing the variance in the model by switching to the log odds approach.
+But the other issue we're trying to solve is being able to predict on variable numbers of choices per decision.
+How does that work out here?\newline
+
+Well, in the case of training we know the maximum number of choices in any of our training decisions.
+Therefore we can set $N$ to this maximum and for decisions where we have fewer than $N$ choices we can simply provide some kind of default "do not choose" feature for the "missing" choices.
+
+Later during inference on new data we can simply take our shared layers and use them to predict on each of the choices we're presented with.
+Just another demonstration that while we're using the architecture in Fig.
+\ref{fig:logarch} in order to \textit{train} the log odds model, the actual log odds model is simply the shared layers in a single column behind the softmax layer.
+
+\newpage
+\section{Appendices}
+
+\subsection{Convergence Issue with Non-Parametric Log Odds Models} \label{convergence issues}
 
@@
 
+\newpage
+
+\section{Bibliography}
+
+\bibliographystyle{apalike}
+\bibliography{reference}
+
 \end{document}

### 2024-06-21 mirrorverse/log_odds_model/log_odds_model.py
fixed tests

@@ -67,14 +67,7 @@ def get_proposed_log_odds(dataframe, learning_rate):
         (1 - dataframe["_selected"]) * dataframe["probability"]
     )
 
-    dataframe["curvature"] = dataframe["probability"] * (1 - dataframe["probability"])
-    dataframe["total_curvature"] = dataframe.groupby("log_odds")["curvature"].transform(
-        "mean"
-    )
-
-    dataframe["step"] = (
-        dataframe["partial"] / dataframe["total_curvature"] * learning_rate
-    )
+    dataframe["step"] = dataframe["partial"] * learning_rate
 
     # propose new utility values
     dataframe["proposed"] = dataframe["log_odds"] + dataframe["step"]
@@ -116,8 +109,8 @@ class LogOddsModel:
         diagnostics = []
 
         # setup an initial guess
-        X_train["log_odds"] = 3
-        X_test["log_odds"] = 3
+        X_train["log_odds"] = 0
+        X_test["log_odds"] = 0
 
         get_probability(X_train)
         get_probability(X_test)
@@ -132,59 +125,11 @@ class LogOddsModel:
             }
         )
 
-        CHANGE = None
-
         for i in tqdm(range(1, iterations + 1)):
-            print(get_central_likelihood(X_train))
-            update = X_train[["_selected", "_decision", "proposed"]].rename(
-                {"proposed": "log_odds"}, axis=1
-            )
-            get_probability(update)
-            print(get_central_likelihood(update))
-
-            # TODO: remove me
-            print(X_train.groupby("log_odds")["curvature"].max().describe())
-
-            # df = X_train.groupby(features)["step"].mean().reset_index()
-            # df["iteration"] = i
-            # output_file = "tmp.csv"
-            # if i == 1:
-            #    df.to_csv(output_file, index=False)
-            # else:
-            #    df.to_csv(output_file, mode="a", header=False, index=False)
-            try:
-                self.model.fit(X_train[features], X_train["proposed"])
-            except:
-                break
-
-            old = X_train["log_odds"]
+            self.model.fit(X_train[features], X_train["proposed"])
 
             X_train["log_odds"] = self.model.predict(X_train[features])
             X_test["log_odds"] = self.model.predict(X_test[features])
-
-            # print(np.abs(X_train["log_odds"] - old).describe())
-            # change = (
-            #    np.sqrt(np.sum((X_train["log_odds"] - old) ** 2)) / X_train.shape[0]
[…]

### 2024-06-21 mirrorverse/models/chinook_depth/warehouse.py

@@ -126,9 +126,8 @@ def join_in_context_data(input_files, output_file):
         ]
     )
 
-    data['home_region'] = data['home_region'].fillna('UNKNOWN')
+    data["home_region"] = data["home_region"].fillna("UNKNOWN")
     for option in ["SEAK", "UNKNOWN", "WA/OR", "BC"]:
         data[option.lower()] = (data["home_region"] == option).astype(float)
-    data[['home_region', 'seak', 'unknown', 'wa/or', 'bc']]
 
     data.to_csv(output_file, index=False)

### 2024-06-24 report/Report.tex
wip

@@
 \ref{fig:logarch} in order to \textit{train} the log odds model, the actual log odds model is simply the shared layers in a single column behind the softmax layer.
+
+\newpage
+\section{Simulation}
+
+\subsection{Dealing with Scale}
+
+\subsubsection{Dealing with Many-World}
+
+We're dealing with states in, states out, with associated relative abundances. Therefore a group by at the end keeps the many world constrained instead of having some $D^n$ problem.
+
+\subsubsection{Parallelism}
+
+Because each state determines it's own outcomes we can parallelize things dramatically.
+As the number of states we're working with gets bigger (as we add more detail to the sim) this will become more and more important.
+Furthermore it means we can deal with our simulation as a single step that gets repeated on itself.
+We'll use pyspark to make sure that when the parallelization is needed it is close at hand.
+
+\subsubsection{Exploring Data}
+There's a ridiculous amount of data that comes out of this and only with a DB can we have any hope of exploring it.
+The data can be stored away, what we want can be queried, and then that can be uploaded into standard tools for viewing.
+This keeps the memory requirements of the exploration application to a minimum while still having all the data at one's fingertips.
+
+\subsubsection{Process to Make this Work}
+
+\begin{enumerate}
+\item Determine what will key your state
+\item Build feature functions
+\item Build a decision exploder
+\item Softmax
+\item Uploader (so we can get matrixed information)
+\item Aggregator (common across all sims although sometimes skippable)
+\end{enumerate}
 

### 2024-06-24 proposal/Proposal.tex
edits

@@
 
-The ultimate objective of this research is to provide a new pathway by which collections of high resolution spatio-temporal movement data that are representative of a species can be used to make a direct impact on fisheries management.
+The ultimate goal of this research is to provide a new pathway by which collections of high resolution spatio-temporal movement data that are representative of a species can be used to make a direct impact on fisheries management.
 It is hoped that this will provide an incentive for creating/enlarging such datasets.
@@
 
-The overview of the components of this simulation are outlined in Fig.
-\ref{fig:simulation}.
-Boxes with rounded edges would be inputs specified by the user.
-The hexagons are functions provided by the simulation and the trapezoids are data generated during the main loop.
+The overview of the components of this simulation are outlined in Fig. \ref{fig:simulation}.
 
@@
   \caption{Simulation Outline}
+  \medskip
+	\small
+	Boxes with rounded edges would be inputs specified by the user. The hexagons are functions 		provided by the simulation and the trapezoids are data generated during the main loop. 
   \label{fig:simulation}
@@
 
-With this simulation we will attempt to demonstrate the following:
+Some specific use cases we want to demonstrate with this simulation are:
 
 \begin{enumerate}
-\item \textbf{Connectivity:} by looking at the vector flows of fish throughout the space we will be able to identify which areas are highly connected and which are more or less separated by environment, cohort, or other factors of fish behavior.
-\item \textbf{Vulnerability:} by especially looking at changes in depth occupancy and preference for habitat we will attempt to quantify the degree of vulnerability in different parts of the space. 
-\item \textbf{Cumulative Mortality:} we would like to demonstrate that we can model the effect of fishing the same cohorts through space as opposed to distributing the fishing mortality across more of the stock.
-\item \textbf{Protected Areas/Times:} we will attempt to demonstrate that we can simulate the effects of MPA or other kinds of closures in space and time. 
-\item \textbf{Cohort Bias:} we will show (models willing) that we can understand and model differences in the above by cohort.
+\item \textbf{Understanding Connectivity:} by looking at the vector flows of fish throughout the space we will be able to identify which areas are highly connected and which are more or less separated by environment, cohort, or other factors of fish behavior.
+\item \textbf{Understanding Vulnerability:} by especially looking at changes in depth occupancy and preference for habitat we will attempt to quantify the degree of vulnerability to different kinds of fishing gear as a function of space and time.
+\item \textbf{Cumulative Mortality:} by being able to follow cohorts through space and time we should be able to show how different fishing pressure scenarios affect the cumulative fishing mortality.
+\item \textbf{Protected Areas/Times:} Of those fishing pressure scenarios, one of especial interest is modeling Marine Protected Areas (MPAs) or other kinds of closures in space and time.
 \item \textbf{Uncertainty:} we will show that such a simulation can be run at scale, bootstrapping can be applied, and therefore uncertainty in simulation outcomes quantified. 
+\end{enumerate}
+
+To give a specific example of what this might look like in practice, if we were interested in understanding how to protect one specific genetic cohort by creating fishing closures we would:
+
+\begin{enumerate}
+\item Input a distribution of the cohort at some initial time point as relative abundance per spatial grid cell.
+\item Input a time period over which to run the simulation (perhaps the fishing season in the area under study).
+\item Collect and input any environmental data needed for the movement model indexed by space and time.
+\item Then we might run the simulation without any fishing pressure to understand how the cohort naturally distributes itself and generate hypotheses for closures from this.
+\item We could then input fishing pressure (as a mortality function) by space and time for a few different scenarios where some would include our closures.
+\item Run the simulation for each of these scenarios and compare resulting abundance trends. 
 \end{enumerate}
@@
 
-\subsection{Objective 1}
+\subsection{Objective 1} \label{objective1}
 
@@
 So how can we compare the predictive power of different $\theta$?
-Well as we find $\theta$ that are more predictive the corresponding $F$'s will have greater and greater likelihoods $\mathcal{L}$, given the data.
+Well as we find $\theta$ that are more predictive the corresponding $F$s will have greater and greater likelihoods $\mathcal{L}$, given the data.
 But this presents us with a chicken and the egg style problem - to know the value of $\theta$ we must know $F$, but to know $F$ we must have fit $\hat{F}$ which requires a reasonable amount of data to have been collected.
-I.e.
-we cannot know in advance if the data we are collecting is going to be useful or not.
-Instead anyone doing modeling finds themselves in the cycle illustrated by Fig.
-\ref{fig:model_cycle}.
+I.e., we cannot know in advance if the data we are collecting is going to be useful or not.
+Instead anyone doing modeling finds themselves in the cycle (Fig.
+\ref{fig:model_cycle}).
 
@@
   \caption{Modeling Cycle}
+  \medskip
+	\small
+	The four major steps in building good models. Note that while one must always start at (1) this process is indeed a cycle.
   \label{fig:model_cycle}
@@
 
-The second issue comes from the "curse of dimensionality".
-Basically this is an issue where as the dimensionality of $\theta$ increases, the amount of data required to find real relationships in that data increases exponentially.
+The second issue comes from the "curse of dimensionality." Basically this is an issue where as the dimensionality of $\theta$ increases, the amount of data required to find real relationships in that data increases exponentially.
 This is a particular problem for behavioral modeling because we in all likelihood need to include data on each of the possible decisions before us.
-So if I can choose between 5 different drinks I need to provide features on each of those five drinks.
+So if I can choose between five different drinks I need to provide features on each of those five drinks.
 This leads to very high dimensional spaces very quickly.
@@
 
-We can get around both of these problems by introducing what we will term an "odds model".
-Specifically if we imagine $\theta$ now as a variable length vector composed of vectors $\eta_k$ for each of our possible outcomes $v_k$ then the odds model is given by the function:
+We can get around both of these problems by introducing what we will term an "odds model." Specifically if we imagine $\theta$ now as a variable length vector composed of vectors $\eta_k$ for each of our possible outcomes $v_k$ then the odds model is given by the function:
 
@@
 
-From studies that started with Beverton and Holt's work on recruitment in fisheries, we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to a pattern as illustrated in Fig.
-\ref{fig:recruitment} \cite{waltersmartell}.
+From studies that started with Beverton and Holt's work on recruitment in fisheries, we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to an asymptotic relationship between spawning biomass and recruits \cite{waltersmartell} (Fig.
+\ref{fig:recruitment}).
+What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
+Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \cite{waltersmartell}.
+\newline
 
@@
 \end{figure}
-
-What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
-Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \cite{waltersmartell}.
-\newline
 
@@
 This means that in order to maintain our SSB over time we have to be careful about how we target each age group.
-This gets even more complicated for fish like Pacific salmon or many species of groupers where life history is intimately entwined with spawning.
+This gets even more complicated for fish like Pacific salmon (\textit{Oncorhynchus}) or many species of groupers (\textit{Epinephelinae}) where life history is intimately entwined with spawning.
 The salmon obviously only breed once \cite{NOAA24} and some species of groupers are protogynous hermaphrodites \cite{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex.
@@
 
-The other observation is that many species of fish actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \cite{lorenzen2010} \cite{salmonplan} \cite{prince2010}.
+The other observation is that many species of fishes actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \cite{lorenzen2010} \cite{salmonplan} \cite{prince2010}.
 Pacific salmon are once again a good example in that they home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \cite{erisman2017}.
@@
 What does this mean?
-It means that if we trust our Beverton-Holt style model, instead of having a single SSB to manage we really have several SSB's - one for each sub stock.
+It means that if we trust our Beverton-Holt style model, instead of having a single SSB to manage we really have several SSBs - one for each sub stock.
 Obviously, if we set one global piece of management like an Allowable Biological Catch (ABC) that makes sense for a single SSB but then have biased fishing mortality where one sub stock is getting fished out far more than the others we may drop below the SSB for that one stock without actually exceeding our expected ABC for the whole fishery.
@@
 All of this to say that understanding heterogeneity in vulnerability by things like age class and sub stock is of the utmost importance for managing a fishery and truly ensuring an appropriate amount of SSB remains year over year.
-
 It is also worth noting that this knowledge around changes in selectivity in space and time is not only useful for considerations around SSB but also for other management issues such as bycatch avoidance and growth overfishing.
@@
 \item \textbf{Satellite:} On pop-up the tags then broadcast data (usually summarized) to Argos satellites which is then relayed to a database for retrieval.
-Such tags usually also broadcast some positioning information which allows for direct retrieval and access to the full archived dataset.
+Such tags usually also broadcast some information that can be used to deduce position as well which allows for direct retrieval and access to the full archived dataset.
 \end{enumerate}
@@
 Therefore they should provide a good challenge to attempts at behavioral modeling and simulation.
-It is the author's hope that patterns useful to Chinook management issues may arise from this research but it is not a necessity as the intention here is to develop tools that justify additional data collection for species where spatial management is of interest.
+It is the author's hope that patterns useful to Chinook management issues may arise from this research but it is not a necessity as the intention here is to develop tools that justify additional data collection for species for which spatial management is of interest.
 
@@
 
+\subsection{Summary}
+
+The sum outcome of the work proposed is the creation of a simulation that can be used to explore, propose, and (in simulation) test management options.
+The advantages of having such a simulation are many (as is described in the following situations) and therefore it comes as no surprise that the notion of using a simulation to drive understanding of management options is in no way novel.
+Indeed as is demonstrated in the following sections many different kinds of simulation (or analogs) already exist.
+Assessment paradigms are a form of simulation, movement models based on more traditional tagging methods can be used in simulation, and individual based models of several sorts have lent themselves to simulations used to assess changes to environment or management as well.
+Therefore the question is - what advantages come from the proposed direction?
+
+\begin{enumerate}
+\item \textbf{Isolating the Study of Movement.} In machine learning there is a well known phenomenon where features that are even mildly correlated with one another can compensate for one another in model fitting.
+A simple example would be something like height representing the "feature impact" of sex in a model of human health.
+Therefore it is always dubious to treat the relationship between a feature and a target as an actual indication of the true relationship.
+Within the spatial assessment paradigms in use today many different parameters of life history are fit together - movement, spatial heterogeneity in recruitment or mortality, fishing pressure, and so on.
+Because these are all fit as one model it is possible if not likely that features are "compensating" for one another and obscuring the actual relationships between the underlying processes and the outcomes.
+The simulation proposed herein fits movement on its own which should lead to a greater confidence in its representation of actual fish movement patterns.
+It also makes the debugging and improvement of those models much more clear and straightforward.
+\item \textbf{Generalized Strategy.} Individual based models (IBMs) also solve the problem mentioned above as they directly model movement as well.
+However IBMs are notoriously specific to the problem at hand.
+Furthermore most IBMs represent a specific hypothesis of behavior that has been coded up and as was mentioned in Section \ref{objective1} building models this way takes an enormous amount of time.
+The strategy proposed herein provides a generalized tool set for going from data, to a model, and then from there to a spatial simulation.
+And this means that researchers can contribute to building up a common tool set instead of having to start from ground zero each round.
+\item \textbf{Repeated Usefulness.} There are obviously more straightforward ways to answer specific spatial questions than by using the framework proposed here.
+For example if trying to understand the value of a Marine Protected Area (MPA) one could simply study general abundances changes over the course of a year or apply more traditional tagging methodologies.
+The purpose of this framework would be to provide a centralized datastore, model, and simulation that could be used across many different problems.
+\item \textbf{Interpretation and Communication.} Current standard spatial assessments depend on a movement matrix $X$ that defines how cohorts move between grids in the assessment at each time point.
+Such a matrix is ultimately an abstract mathematical concept that while useful in assessment makes the communication and interpretation of the underlying movement model difficult.
+Questions such as - what is driving the movement pattern seen - require actual parametrization of that movement matrix $X$.
+The strategy we are proposing provides this and therefore makes communication and feedback from stakeholders of the fishery far easier.
+Given engagement is such an important part of management this is paramount.
+\end{enumerate}
+
+Obviously these each of these individual pros are not unique to the strategy proposed herein.
+IBMs, spatial assessment models, and more ad-hoc approaches all share some number of these advantages.
+However the strategy proposed herein presents an opportunity to gain all of these advantages at once which should make it a useful additional to the tools fisheries science already possesses.
+\newpage
+
 \subsection{Spatial Management}
@@
 Fisheries management is as diverse a subject as the species and cultures to which it is applied \cite{waltersmartell} \cite{king}.
-It simultaneously has to deal with creatures from Bluefin Tuna who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \cite{prince2010}.
+It simultaneously has to deal with creatures from Bluefin Tuna (\textit{Thunnus thynnus}) who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \cite{prince2010}.
 There are species that school and species that consume those schools.
-Some live entirely at sea, some in inland rivers and lakes, and some, like Salmon and eels spend time in both.
+Some live entirely at sea, some in inland rivers and lakes, and some, like salmon and eels spend time in both.
 There are species whose lives last the span of a year to others that can live longer than any us human beings.
@@
 \item \textbf{Real Time Management} - This is a particularly interesting kind of management as instead of trying to predict where closures are needed, in this strategy metrics from the fishers themselves help dictate where closures are needed temporarily.
-For example if bycatch avoidance is the goal then catch logs (that are usually reported with some frequency in these cases) are used to determine areas where bycatch abundance has spiked and then the fishery can call for a temporary closure last days to months in order to avoid that aggregation.
+For example if bycatch avoidance is the goal then catch logs (that are usually reported with some frequency in these cases) are used to determine areas where bycatch abundance has spiked and then the fishery can call for a temporary closure lasting days to months in order to avoid that aggregation.
 In general this strategy is comprised of measuring some overall catch statistic and then responding to that statistic in real time as needs be.
@@
 
-Every assessment paradigm is of course a simplification of reality but what's important to know is whether those simplifications are obscuring important structure of the stock and sub-stocks.
-I.e.
-are the simplifications introducing bias or errors that would lead to contradictory results from sound management based on the advice of the assessment \cite{cadrin2020}?
+Every assessment paradigm is of course a simplification of reality but what's important to know is whether those simplifications are obscuring important structure of the stock and sub-stocks, i.e., are the simplifications introducing bias or errors that would lead to contradictory results from sound management based on the advice of the assessment \cite{cadrin2020}?
 
-Given that ultimately we do \textit{not} know the underlying structure (models such as these are built specifically to elucidate such meaningful structure) we find ourselves with a "Catch-22".
-We need spatial components to understand spatial structure, but need to know the spatial structure in order to understand what simplifications are allowable.
+Given that ultimately we do \textit{not} know the underlying structure (models such as these are built specifically to elucidate such meaningful structure) we find ourselves with a "Catch-22." We need spatial components to understand spatial structure, but need to know the spatial structure in order to understand what simplifications are allowable.
 Note that it is not enough to simply design the most complicated spatial model possible given the data will likely not exist to fit that model with any degree of confidence.
@@
 
-\subsubsection{Struggles in Building IBM's}
+\subsubsection{Struggles in Building IBMs}
 
@@
 \begin{enumerate}
-\item \textbf{High Precision as a Requirement} - IBM's typically predict what an animal \textit{will} do as opposed to the likelihoods of what it \textit{could} do.
+\item \textbf{High Precision as a Requirement} - IBMs typically predict what an animal \textit{will} do as opposed to the likelihoods of what it \textit{could} do.
 This creates the requirement for capturing \textit{all} the meaningful drivers of variability in the behavior of the animal.
@@
 And this can cascade into greater and greater requirements for modeling the "internal state" of an animal.
-\item \textbf{Obscurity} - given the inherent complexity of IBM's that arises from the point above it gets increasingly difficult to make IBM's clear to new users or potential developers.
+\item \textbf{Obscurity} - given the inherent complexity of IBMs that arises from the point above it gets increasingly difficult to make IBMs clear to new users or potential developers.
 There can often be many dozens of different parameters, functions, and "internal states" to keep track of without even considering all of the external modeling that has to go in place to provide for all of this.
 And this makes for not just complicated descriptions but equally if not more complicated code.
-\item \textbf{Hand Made} - almost all IBM's have been hand-coded in the sense that every functional relationship is specifically crafted and outlined by the modeler.
-This is obviously a very laborious process but it also means that no two IBM's are the same and that lack of generalization or abstraction makes exchanging lessons learned near difficult to impossible.
+\item \textbf{Hand Made} - almost all IBMs have been hand-coded in the sense that every functional relationship is specifically crafted and outlined by the modeler.
+This is obviously a very laborious process but it also means that no two IBMs are the same and that lack of generalization or abstraction makes exchanging lessons learned near difficult to impossible.
 \item \textbf{No Generalization of Fitting or Diagnostics} - because each IBM is effectively unique the methods for fitting, the data used, and the diagnostics developed are all unique as well.
@@
 
-Juvenile Chinook are born in the rivers where their parents spawned.
+Juvenile Chinook are generally born in the rivers where their parents spawned (although some genetic mixing does occur through straying).
 Sometime in their first or second year of life they transition to the marine environment where they will spend the rest of their life before maturation \cite{oceanchinook}.
@@
 
-Fish are released at weights of around 6-9 grams and a large number of these have a micro-wire with an identifiable code etched into them inserted into their snouts.
+Fish are released at weights of around 6-9g and a large number of these have a micro-wire with an identifiable code etched into them inserted into their snouts.
 These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook sub-populations \cite{shelton2019}.
@@
 These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook form so many different genetic sub-populations \cite{salmonplan} - a fact that further complicates management.
-Indeed several sub populations are actually listed as endangered under the Endangered Species Act \cite{NOAA24}.
+Indeed several sub populations are actually listed as endangered and threatened under the Endangered Species Act \cite{NOAA24}.
 

### 2024-06-24 proposal/Proposal.tex
wip

@@
 
-Chapter 2 will be concerned with taking these models for Chinook in the Gulf of Alaska and building a spatio-temporal movement simulation that can be used to investigate a series of management questions.
-It is worth noting that we will use the simulation to ask questions that may come up in general management settings even if they may not be directly applicable to the issues currently faced by Chinook management in the GOA.
+Chapter 2 will be concerned with taking these models for Chinook salmon in the Gulf of Alaska and building a spatio-temporal movement simulation that can be used to investigate a series of management questions.
+It is worth noting that we will use the simulation to ask questions that may come up in general management settings even if they may not be directly applicable to the issues currently faced by Chinook salmon management in the GOA.
 This is just to demonstrate general applicability.
@@
 
-\subsection{Chinook in the GOA}
+\subsection{Chinook salmon in the GOA}
 
-At this point in time the main interest here is that Chinook present an interesting model species for testing out the efficacy of these models and simulations.
-Chinook are highly motile and undergo clear migration behaviors \cite{tucker2019} \cite{langan2024}.
+At this point in time the main interest here is that Chinook salmon present an interesting model species for testing out the efficacy of these models and simulations.
+Chinook salmon are highly motile and undergo clear migration behaviors \cite{tucker2019} \cite{langan2024}.
 Therefore they should provide a good challenge to attempts at behavioral modeling and simulation.
-It is the author's hope that patterns useful to Chinook management issues may arise from this research but it is not a necessity as the intention here is to develop tools that justify additional data collection for species for which spatial management is of interest.
+It is the author's hope that patterns useful to Chinook salmon management issues may arise from this research but it is not a necessity as the intention here is to develop tools that justify additional data collection for species for which spatial management is of interest.
 
@@
 
-\subsection{Chinook Ecology and Management}
+\subsection{Chinook salmon Ecology and Management}
 
@@
 
-Juvenile Chinook are generally born in the rivers where their parents spawned (although some genetic mixing does occur through straying).
+Juvenile Chinook salmon are generally born in the rivers where their parents spawned (although some genetic mixing does occur through straying).
 Sometime in their first or second year of life they transition to the marine environment where they will spend the rest of their life before maturation \cite{oceanchinook}.
@@
 
-Given this homing behavior, Chinook populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies.
-Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook is highly dependent on the region of origin for the fish in question \cite{shelton2019} \cite{tucker2019}.
+Given this homing behavior, Chinook salmon populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies.
+Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook salmon is highly dependent on the region of origin for the fish in question \cite{shelton2019} \cite{tucker2019}.
 
@@
 
-Studies like these point to Chinook being a highly mobile, spatially heterogeneous species.\newline
+Studies like these point to Chinook salmon being a highly mobile, spatially heterogeneous species.\newline
 
-Broadly distributed through out the Northern pacific, Chinook are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side \cite{oceanchinook}.
+Broadly distributed through out the Northern pacific, Chinook salmon are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side \cite{oceanchinook}.
 
@@
 
-Chinook are well known piscivores with several studies of stomach contents finding primarily fish and squid as their main sources of prey.
+Chinook salmon are well known piscivores with several studies of stomach contents finding primarily fish and squid as their main sources of prey.
 It has been noted that this diet is similar to that of Coho salmon but that it is unlikely that the two species experience much cross-species competition for these resources \cite{oceanchinook}.
 
-Chinook themselves are prey for animal groups such as Salmon sharks, whales, and dolphins but it is still unclear how much mortality in later life stages is due to predation as opposed to other sources.
+Chinook salmon themselves are prey for animal groups such as Salmon sharks, whales, and dolphins but it is still unclear how much mortality in later life stages is due to predation as opposed to other sources.
 Size selective predation on earlier life stages is a well known driver of natural mortality \cite{oceanchinook}.
@@
 
-Chinook have been the subject of hatchery programs since the 1950s with the past decade seeing hatchery releases of 254million individuals on average.
+Chinook salmon have been the subject of hatchery programs since the 1950s with the past decade seeing hatchery releases of 254million individuals on average.
 The vast majority of these individuals (approx.
@@
 Fish are released at weights of around 6-9g and a large number of these have a micro-wire with an identifiable code etched into them inserted into their snouts.
-These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook sub-populations \cite{shelton2019}.
+These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook salmon sub-populations \cite{shelton2019}.
 The etchings provide a unique label for each cohort from each hatchery.
@@
 
-Given the highly mobile nature of Chinook independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Fishery Management Council.
+Given the highly mobile nature of Chinook salmon independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Fishery Management Council.
 Federal management councils (NOAA for the United States) then collaborate with the Pacific Fishery Management Council to set harvest quotas and other management objectives.
@@
 
-These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook form so many different genetic sub-populations \cite{salmonplan} - a fact that further complicates management.
+These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook salmon form so many different genetic sub-populations \cite{salmonplan} - a fact that further complicates management.
 Indeed several sub populations are actually listed as endangered and threatened under the Endangered Species Act \cite{NOAA24}.
 
-This combined with other threatened stocks \cite{NOAA24} means that bycatch avoidance is a serious issue for Chinook management and for other overlapping fisheries such as the Walleye Pollock fishery \cite{ianelli2015}.
+This combined with other threatened stocks \cite{NOAA24} means that bycatch avoidance is a serious issue for Chinook salmon management and for other overlapping fisheries such as the Walleye Pollock fishery \cite{ianelli2015}.
 
@@
 Finally, Chinook salmon are especially prized for their large size which has led to serious questions about size selective mortality from fishing as both recreational and commercial fisheries have incentives for targeting larger fish.
-While it is not yet known whether fisheries are responsible, there is clear evidence that fish size in Chinook populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook's primary predators \cite{kendall2011}\cite{oceanchinook}.
+While it is not yet known whether fisheries are responsible, there is clear evidence that fish size in Chinook salmon populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook salmon's primary predators \cite{kendall2011}\cite{oceanchinook}.
 \newline
 
-All in all then, given their ecology and the corresponding complexities of management, Chinook represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
+All in all then, given their ecology and the corresponding complexities of management, Chinook salmon represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
 

### 2024-06-25 mirrorverse/models/chinook_depth/spark.py
wip

@@ -1,0 +1,332 @@
+import os
+import uuid
+import shutil
+
+import numpy as np
+import pandas as pd
+import tensorflow as tf
+import h3
+from functools import partial
+from scipy.stats import norm
+from pyspark.sql.types import (
+    StructType,
+    StructField,
+    FloatType,
+    IntegerType,
+    TimestampType,
+)
+from pyspark.sql.functions import udf, pandas_udf, lit
+from suntimes import SunTimes
+
+
+def load_from_database(spark, connection, sql, schema):
+    return (
+        spark.read.format("jdbc")
+        .option("url", connection)
+        .option("query", sql)
+        .option("customSchema", schema)
+        .load()
+    )
+
+
+def update_schema(dataframe, new_fields):
+    return StructType(dataframe.schema.fields + new_fields)
+
+
+def apply_to_create_columns(dataframe, func, input_cols, new_fields):
+    args = [dataframe[col] for col in input_cols]
+    if len(new_fields) > 1:
+        func = udf(func, StructType(new_fields))
+        dataframe = dataframe.withColumn("_tmp_struct", func(*args))
+        for field in new_fields:
+            dataframe = dataframe.withColumn(
+                field.name, dataframe["_tmp_struct"][field.name]
+            )
+        dataframe = dataframe.drop("_tmp_struct")
+    else:
+        func = udf(func, new_fields[0].dataType)
+        dataframe = dataframe.withColumn(new_fields[0].name, func(*args))
+    return dataframe
+
+
+def vectorized_create_column(dataframe, pandas_udf, input_cols, field):
+    return dataframe.withColumn(
+        field, pandas_udf(*[dataframe[col] for col in input_cols])
+    )
+
+
+def join(dataframe, other, on, how="inner"):
+    dataframe = dataframe.repartition(on)
+    other = other.repartition(on)
+    return dataframe.join(other, on=on, how=how)
+
+
+def load_training_states(spark, connection):
+    sql = """
+    select 
+
+        d.epoch,
+        t.h3_level_4_key,
+
+        d.depth,
+        "t" || t.tag_key as _individual,
+        ROW_NUMBER() over () as _decision
+
+    from 
+        tag_depths d
+        inner join tag_tracks t 
+            on d.tag_key = t.tag_key
+            and d.date_key = t.date_key
+    where
+        d.depth is not null
+    """
[…]

### 2024-06-26 mirrorverse/models/chinook_depth/spark.py
wip

@@ -1,6 +1,7 @@
 import os
 import uuid
 import shutil
+import json
 
 import numpy as np
 import pandas as pd
@@ -15,7 +16,7 @@ from pyspark.sql.types import (
     IntegerType,
     TimestampType,
 )
-from pyspark.sql.functions import udf, pandas_udf, lit
+from pyspark.sql.functions import udf, pandas_udf, lit, avg, std
 from suntimes import SunTimes
 
 
@@ -225,7 +226,7 @@ def create_common_context(states, elevation, depth_classes):
         ["datetime", "sunrise", "sunset", "daytime"],
         "period_progress",
     )
-    # states = join(states, elevation, on="h3_level_4_key")
+    states = join(states, elevation, on="h3_level_4_key")
     return states
 
 
@@ -238,7 +239,8 @@ def create_choices(states, depth_classes):
 def create_features(states, depth_classes, features):
     for i, _ in enumerate(depth_classes):
         for feature in features:
-            states = states.withColumn(f"{feature}_{i}", states[feature])
+            if feature != "depth_class":
+                states = states.withColumn(f"{feature}_{i}", states[feature])
     return states
 
 
@@ -253,6 +255,47 @@ def train_test_split(states, split):
     train_states = states.filter(states["_individual"].isin(train_individuals))
     test_states = states.filter(states["_individual"].isin(test_individuals))
     return train_states, test_states
+
+
+def explode_func(N, features, iterator):
+    for dataframe in iterator:
+        for i in range(N):
+            sub_dataframe = dataframe[[f"{feature}_{i}" for feature in features]]
+            sub_dataframe = sub_dataframe.rename(
+                {f"{feature}_{i}": feature for feature in features}, axis=1
+            )
+            for feature in features:
+                sub_dataframe[feature] = sub_dataframe[feature].astype(float)
+            yield sub_dataframe
+
+
+def explode(states, N, features):
+    return states.mapInPandas(
+        partial(explode_func, N, features),
+        schema=StructType([StructField(feature, FloatType()) for feature in features]),
+    )
+
+
+def get_normalization_parameters(states, N, features):
+    states = explode(states, N, features)
+    avg_aggs = [avg(feature).alias(feature) for feature in features]
+    std_aggs = [std(feature).alias(feature) for feature in features]
+    return {
+        "avg": states.agg(*avg_aggs).collect()[0].asDict(),
+        "std": states.agg(*std_aggs).collect()[0].asDict(),
+    }
[…]

### 2024-06-26 mirrorverse/models/chinook_depth/keras.py
wip

@@ -1,0 +1,69 @@
+import os
+
+from functools import partial
+import tensorflow as tf
+import numpy as np
+from tensorflow.keras.utils import to_categorical
+from tensorflow.keras.models import Model
+from tensorflow.keras.layers import Dense, Input, concatenate
+
+
+def split_data(N, features, data):
+    inputs = {}
+    for i in range(N):
+        input = tf.stack(
+            [tf.cast(data[f"{feature}_{i}"], tf.float32) for feature in features]
+        )
+        inputs[f"input_{i}"] = input
+    label = to_categorical(data["_selected"], num_classes=N)
+    return inputs, label
+
+
+def load_data(data_dir, N, features, batch_size, shuffle_buffer_size):
+    feature_description = {
+        "_selected": tf.io.FixedLenFeature([], tf.int64),
+    }
+    for i in range(N):
+        for feature in features:
+            feature_description[f"{feature}_{i}"] = tf.io.FixedLenFeature(
+                [], tf.float32
+            )
+
+    def _parse_function(proto):
+        return tf.io.parse_single_example(proto, feature_description)
+
+    tfrecord_files = []
+    for path in os.listdir(data_dir):
+        if path.endswith(".tfrecord"):
+            tfrecord_files.append(os.path.join(data_dir, path))
+
+    data = tf.data.TFRecordDataset(tfrecord_files)
+    data = data.map(_parse_function)
+    data = data.map(partial(split_data, N, features))
+    data = data.shuffle(buffer_size=shuffle_buffer_size)
+    data = data.batch(batch_size=batch_size)
+    data = data.prefetch(buffer_size=tf.data.experimental.AUTOTUNE)
+    return data
+
+
+def build_model(N, features, layers, final_activation="linear"):
+    layers.append(Dense(1, activation=final_activation))
+    inputs = [Input(shape=(len(features),), name=f"input_{i}") for i in range(N)]
+    outcomes = []
+    for input in inputs:
+        last_layer = input
+        for layer in layers:
+            last_layer = layer(last_layer)
+        outcomes.append(last_layer)
+
+    outcomes = concatenate(outcomes)
+
+    output_layer = Dense(N, activation="softmax")
+    output = output_layer(outcomes)
+    output_layer.set_weights([np.eye(N), np.zeros(N)])
+    output_layer.trainable = False
+
+    model = Model(inputs=inputs, outputs=output)
+    model.compile(optimizer="adam", loss="categorical_crossentropy")
+
+    return model, layers

### 2024-06-26 mirrorverse/models/chinook_depth/spark.py

@@ -3,6 +3,7 @@ import uuid
 import shutil
 import json
 
+import click
 import numpy as np
 import pandas as pd
 import tensorflow as tf
@@ -309,7 +310,7 @@ def serialize_example(depth_classes, features, row):
     for i, _ in enumerate(depth_classes):
         current_feature = {
             f"{feature}_{i}": tf.train.Feature(
-                float_list=tf.train.FloatList(value=[row.month])
+                float_list=tf.train.FloatList(value=[row[f"{feature}_{i}"]])
             )
             for feature in features
         }
@@ -346,10 +347,6 @@ def build_training_data(
     split,
     overwrite=False,
 ):
-    # Sanitize Inputs
-    depth_classes = np.array(eval(depth_classes))
-    features = eval(features)
-
     states = load_training_states(spark, connection)
     elevation = load_elevation(spark, connection)
 

### 2024-06-27 mirrorverse/models/chinook_depth/spark.py
working infer

@@ -3,7 +3,6 @@ import uuid
 import shutil
 import json
 
-import click
 import numpy as np
 import pandas as pd
 import tensorflow as tf
@@ -16,6 +15,7 @@ from pyspark.sql.types import (
     FloatType,
     IntegerType,
     TimestampType,
+    LongType,
 )
 from pyspark.sql.functions import udf, pandas_udf, lit, avg, std
 from suntimes import SunTimes
@@ -83,6 +83,11 @@ def load_training_states(spark, connection):
         d.depth is not null
     """
     schema = "epoch INTEGER, h3_level_4_key BIGINT, depth FLOAT, _individual STRING, _decision INTEGER"
+    return load_from_database(spark, connection, sql, schema)
+
+
+def load_infer_states(spark, connection, sql):
+    schema = "epoch INTEGER, h3_level_4_key BIGINT"
     return load_from_database(spark, connection, sql, schema)
 
 
@@ -175,13 +180,7 @@ def get_period_progress(date, sunrise, sunset, daytime):
     return (1 - hours_to_transition / interval).astype(float)
 
 
-def create_common_context(states, elevation, depth_classes):
-    states = apply_to_create_columns(
-        states,
-        partial(get_depth_class, depth_classes),
-        ["depth"],
-        [StructField("_selected", FloatType())],
-    )
+def create_common_context(states, elevation):
     states = apply_to_create_columns(
         states,
         h3_to_geo,
@@ -337,6 +336,61 @@ def save_to_tfrecord(dataframe, features, output_dir, depth_classes, overwrite=F
     rdd.foreachPartition(partial(write_partition, output_dir))
 
 
+def infer_map(depth_classes, features, model, iterator):
+    for dataframe in iterator:
+        inputs = {}
+        for i, _ in enumerate(depth_classes):
+            sub_features = [f"{feature}_{i}" for feature in features]
+            inputs[f"input_{i}"] = dataframe[sub_features]
+        predictions = model.predict(inputs)
+        for i, _ in enumerate(depth_classes):
+            dataframe[f"probability_{i}"] = predictions[:, i]
+        yield dataframe
+
+
+def infer(states, depth_classes, features, model):
+    schema = StructType(
+        states.schema.fields
+        + [
+            StructField(f"probability_{i}", FloatType())
+            for i, _ in enumerate(depth_classes)
+        ]
+    )
+    states = states.mapInPandas(
[…]

### 2024-06-28 proposal/Proposal.tex
small edit

@@
 \begin{enumerate}
-\item \textbf{Isolating the Study of Movement.} In machine learning there is a well known phenomenon where features that are even mildly correlated with one another can compensate for one another in model fitting.
+\item \textbf{Isolating the Study of Movement (\S \ref{spatial_assessments}).} In machine learning there is a well known phenomenon where features that are even mildly correlated with one another can compensate for one another in model fitting.
 A simple example would be something like height representing the "feature impact" of sex in a model of human health.
@@
 It also makes the debugging and improvement of those models much more clear and straightforward.
-\item \textbf{Generalized Strategy.} Individual based models (IBMs) also solve the problem mentioned above as they directly model movement as well.
+\item \textbf{Generalized Strategy (\S \ref{modeling_movement}).} Individual based models (IBMs) also solve the problem mentioned above as they directly model movement as well.
 However IBMs are notoriously specific to the problem at hand.
@@
 The purpose of this framework would be to provide a centralized datastore, model, and simulation that could be used across many different problems.
-\item \textbf{Interpretation and Communication.} Current standard spatial assessments depend on a movement matrix $X$ that defines how cohorts move between grids in the assessment at each time point.
+\item \textbf{Interpretation and Communication (\S \ref{spatial_assessments}, \ref{spatial_management}).} Current standard spatial assessments depend on a movement matrix $X$ that defines how cohorts move between grids in the assessment at each time point.
 Such a matrix is ultimately an abstract mathematical concept that while useful in assessment makes the communication and interpretation of the underlying movement model difficult.
@@
 
-Obviously these each of these individual pros are not unique to the strategy proposed herein.
+Obviously each of these individual pros are not unique to the strategy proposed herein.
 IBMs, spatial assessment models, and more ad-hoc approaches all share some number of these advantages.
-However the strategy proposed herein presents an opportunity to gain all of these advantages at once which should make it a useful additional to the tools fisheries science already possesses.
+However the strategy proposed here presents an opportunity to gain all of these advantages at once which should make it a useful additional to the tools fisheries science already possesses.
 \newpage
 
-\subsection{Spatial Management}
+\subsection{Spatial Management}\label{spatial_management}
 
@@
 
-\subsection{Spatial Assessments}
+\subsection{Spatial Assessments}\label{spatial_assessments}
 
@@
 
-\subsection{Modeling Movement}
+\subsection{Modeling Movement}\label{modeling_movement}
 

### 2024-06-28 mirrorverse/models/chinook_depth/keras.py
linting

@@ -1,3 +1,9 @@
+"""
+Keras Code for Chinook Depth Model
+"""
+
+# pylint: disable=invalid-name, import-error, no-name-in-module
+
 import os
 
 from functools import partial
@@ -8,7 +14,16 @@ from tensorflow.keras.models import Model
 from tensorflow.keras.layers import Dense, Input, concatenate
 
 
+# pylint: disable=redefined-builtin
 def split_data(N, features, data):
+    """
+    Inputs:
+    - N: int, number of choices
+    - features: list of strings, names of features
+    - data: dict, data loaded from tfrecord file
+
+    Splits the data into inputs and labels for the model
+    """
     inputs = {}
     for i in range(N):
         input = tf.stack(
@@ -20,6 +35,16 @@ def split_data(N, features, data):
 
 
 def load_data(data_dir, N, features, batch_size, shuffle_buffer_size):
+    """
+    Inputs:
+    - data_dir: str, path to directory containing tfrecord files
+    - N: int, number of choices
+    - features: list of strings, names of features
+    - batch_size: int, batch size
+    - shuffle_buffer_size: int, size of buffer for shuffling data
+
+    Returns a tf.data.Dataset object containing the data
+    """
     feature_description = {
         "_selected": tf.io.FixedLenFeature([], tf.int64),
     }
@@ -47,6 +72,15 @@ def load_data(data_dir, N, features, batch_size, shuffle_buffer_size):
 
 
 def build_model(N, features, layers, final_activation="linear"):
+    """
+    Inputs:
+    - N: int, number of choices
+    - features: list of strings, names of features
+    - layers: list of keras layers
+    - final_activation: str, activation function for final layer
+
+    Returns a keras model
+    """
     layers.append(Dense(1, activation=final_activation))
     inputs = [Input(shape=(len(features),), name=f"input_{i}") for i in range(N)]
     outcomes = []

### 2024-06-28 mirrorverse/models/chinook_depth/spark.py

@@ -1,13 +1,19 @@
+"""
+Spark Code for Chinook Depth Model
+"""
+
+# pylint: disable=invalid-name, import-error, no-name-in-module, no-member
+
 import os
 import uuid
 import shutil
 import json
+from functools import partial
 
 import numpy as np
 import pandas as pd
 import tensorflow as tf
 import h3
-from functools import partial
 from scipy.stats import norm
 from pyspark.sql.types import (
     StructType,
@@ -17,11 +23,20 @@ from pyspark.sql.types import (
     TimestampType,
     LongType,
 )
+
+# pylint: disable=redefined-outer-name
 from pyspark.sql.functions import udf, pandas_udf, lit, avg, std
 from suntimes import SunTimes
 
 
 def load_from_database(spark, connection, sql, schema):
+    """
+    Inputs:
+    - spark: SparkSession
+    - connection: str, connection string
+    - sql: str, SQL query
+    - schema: str, schema string
+    """
     return (
         spark.read.format("jdbc")
         .option("url", connection)
@@ -32,10 +47,27 @@ def load_from_database(spark, connection, sql, schema):
 
 
 def update_schema(dataframe, new_fields):
+    """
+    Inputs:
+    - dataframe: DataFrame
+    - new_fields: list of StructFields
+
+    Returns a new schema with the new fields added to the original schema.
+    """
     return StructType(dataframe.schema.fields + new_fields)
 
 
 def apply_to_create_columns(dataframe, func, input_cols, new_fields):
+    """
+    Inputs:
+    - dataframe: DataFrame
+    - func: function
+    - input_cols: list of str, input columns
+    - new_fields: list of StructFields
+
+    Returns a new DataFrame with the new fields created by applying
+    the function to the input columns.
+    """
     args = [dataframe[col] for col in input_cols]
     if len(new_fields) > 1:
         func = udf(func, StructType(new_fields))
@@ -52,18 +84,46 @@ def apply_to_create_columns(dataframe, func, input_cols, new_fields):
 
 
 def vectorized_create_column(dataframe, pandas_udf, input_cols, field):
+    """
+    Inputs:
+    - dataframe: DataFrame
+    - pandas_udf: function, pandas UDF
+    - input_cols: list of str, input columns
+    - field: str, new field name
+
+    Returns a new DataFrame with the new field created by
+    applying the pandas UDF to the input columns.
+    """
     return dataframe.withColumn(
[…]

### 2024-06-28 report/Report.tex

--- 2024-06-21 report/Report.tex
@@
 \ref{fig:logarch} in order to \textit{train} the log odds model, the actual log odds model is simply the shared layers in a single column behind the softmax layer.
+
+\newpage
+\section{Simulation}
+
+\subsection{Dealing with Scale}
+
+\subsubsection{Key Issues}
+
+There are a few key issues in trying to build spatio-temporal simulations using log odds models.
+
+\begin{enumerate}
+\item \textbf{Many World Problems.} \textit{Issue: Exponentially growing state space.
+Solution: Grouping abundance by state keys.} \newline If we took the our log odds model outcomes at face value we'd have a simulation that quickly (and exponentially) careened out of control.
+Why is this?
+Well if we started with $S$ distinct states and each of these generated $C$ choices, in the next step we'd have $SC$ states.
+Then in the next step $SC^2$ states.
+And so on until by step $n$ we have $SC^n$ states!
+Obviously this kind of exponential explosion of the state space is totally unmanageable.
+However we can get around this by recognizing that at the end of each step many of the states will overlap.
+For example if we are just predicting which new grid cells the fish are moving into, fish that start in nearby cells will move into the same grid cells as one another.
+Therefore we can group by the new resulting states, accumulate abundances from the various individual decisions, and keep our state space under control.
+\item \textbf{Combinatorics of State.} \textit{Issue: Problem scales by multiplying dimensions.
+Solution: Use cluster compute to take advantage of high parallelism in the problem.} \newline The size of each step in the simulation can get exceptionally large very quickly due to the number of dimensions that can be present in the state and the fact that these dimensions multiply on one another.
+So if I have 500 grid points, 50 distinct genetic cohorts, 10 different age groups, 3 size bins per age group, and 2 sexes that's a state space of $500 \bullet 50 \bullet 10 \bullet 3 \bullet 2 = 1,500,000$ distinct states to keep track of and model.
+And that would represent a relatively coarse description of the problem with no internal states present at all.
+Each additional dimension we add does not just add to this scale, it multiplies the scale.
+Very quickly then we will run into simulations that simply do not fit into a single machine's memory.
+However note that we can deal with each of these states independently which means that the parallelization opportunity is very high.
+Therefore if we take advantage of distributed compute paradigms like Spark we can parallelize the problems across as many machines as we like thereby preventing the bottleneck on single machine memory.
+\item \textbf{Enormity of Results.} \textit{Issue: Adding time and context increases the scale of the data even further.
+Solution: Using DBMS, optimized file storage systems, and a clear dimension vs fact delineation.} \newline In our example above we have 1.5 million entries per timestamp.
+If we were interested in this over the course of a year and wanted to present results every 3 hours that would mean multiplying our 1.5 million states by approximately 3,000 different time points putting us at an easy 4.5 billion data points.
+Add into this the fact that useful exploration requires a lot of meta data (polygons, human readable timestamps, context like day/night, month of the year, country or state, etc.) and the sheer size of the data we would like to present and explore is quite daunting.
+We can do two things to make this data actually manageable.
+First is to take advantage of an actual database management system (DBMS).
+By writing there researchers can then interact with the data through queries, pull only what they need, and visualize just that.
+Then, in order to keep the data in simulation at a minimum we can separate our information into facts and dimensions \cite{datawarehouse}, facts being what is strictly needed for simulation and dimensions being repeated context that can be joined in later.
+\end{enumerate}
+
+In summary then we want to ensure we obey the following design principles:
+
+\begin{enumerate}
+\item Collapse state at the end of each step.
+\item Keep our simulation parallelizable by state and use cluster compute paradigms to allow for multi-machine parallelization.
+\item Read from and write to databases or optimized file storage and retrieval systems.
+\item Keep clear distinctions between facts and dimensions to reduce bloat in simulation while allowing for easy human exploration in database. 
+\end{enumerate}
+
+\subsubsection{Kinds of Information}
+
+As we go through simulation there are a few key kinds of information each of which has specific implications for handling in simulation:
+
+\begin{enumerate}
+\item \textbf{Keys.} These are the information needed to identify a specific state.
+It includes spatial keys, temporal keys, demographic keys, and any internal state keys.
+We will parallelize along and group by these keys.
+\item \textbf{Abundance.} This is a single metric that lets us know how much of an individual state there is. 
+\item \textbf{Joined Context.} This is any information that will be joined to the spatial or temporal keys and cannot be generated from the information already given.
+Examples would be things like bathymetry or temperatures.
+Because this information has to be joined on special consideration must be given for the sake of performance.
+If the data is small enough broadcast joins are best as they don't require reshuffling of the state information.
+Otherwise it is important that the states are partitioned by the join keys for this information so that as little shuffling as possible needs to happen during the joins.
+\item \textbf{Built Context.} This is information that can be built from state keys or joined context.
+Examples would be sunset/sunrise times, season, etc.
+Whenever this kind of computation can be vectorized it should be for the sake of performance.
+\item \textbf{Choice Independent Context.} This is information about the individual state itself that can be generated before choices are determined.
+This should obviously be computed before choices are generated.
+\item \textbf{Choice Dependent Context.} This is any information specific to the choice in question and must be generated after choices have been built. 
+\item \textbf{Model Features.} This is information built from context that has to be normalized before being sent to the model.
+\end{enumerate}
+
+From this we get a specific series of steps to follow:
+
+\begin{enumerate}
+\item Select or receive initial states (from DBMS).
+\item Partition by state keys and join keys.
+\item Join context.
+\item Build choice independent context.
+\item Build choices.
+\item Build choice dependent context.
+\end{enumerate}
+
+At this point it is important to know whether we are building data for training or running inference. In the former case (training):
+
+\begin{enumerate}
+\item Split on individuals input to create train and validation data.
+\item Create and save normalization parameters from training data.
+\item Build and normalize model features.
+\item Capture choice selected.
+\item Serialize and write to TFRecord (an optimized file storage for building deep learning models)
+\end{enumerate}
+
+In the latter case (inference):
+
+\begin{enumerate}
+\item Build and normalize model features.
+\item Apply the model to produce probabilities of selection.
+\item Multiply probabilities by abundance.
+\item Group by new states.
+\item Write to DBMS.
+\end{enumerate}
+
+As a final note it is important to recognize that the choices must be kept together until after model application in order to ensure the same decision stays on the same partition all the way until inference.
+This leads to very broad (but no less massive) datasets.
 

### 2024-07-09 report/Report.tex
report + tuning

@@
 \tableofcontents
+
 \newpage
 
-\section{Log Odds Modeling}
+\section{Modeling Depth Occupancy}
+
+\subsection{Introduction}
+
+The ocean is ultimately a three dimensional world.
+As such understanding such things as vulnerability (the ease of making a catch) and selectivity (the specificity of that catch) require an understanding not just of the latitudinal and longitudinal distribution of fishes but of their depth occupancy as well.
+For example, if a standardized trawl always samples at a specific depth but depending on placement in space and time that depth is not always as occupied, then vulnerability and abundance become intertwined - i.e., lower or higher catch does not directly mean lower or higher abundance.
+
+We can define depth occupancy as the proportions of fish, of a given species or group, within a series of depth bins at a given time and place.
+The question is how to measure or model depth occupancy.
+Clearly, catch per unit of effort (CPUE) already has both abundance and vulnerability folded into it and so while it can give us relative "catchable" abundance, it does not give us the information about true abundance that we would need to start constructing a notion of true "vulnerability".
+Besides, even if we could compute that vulnerability, vulnerability is itself a function of more than just depth occupancy.
+Sizes, temperature, and age are likely determinants of vulnerability as well.
+As such we must go looking for data elsewhere to fit depth occupancy models.
+
+One potential data source is electronic tagging data.
+These tags specifically measure depths for individual fish across the lifetime of the tag and then, either through retrieval or upload, make that data accessible to researchers.
+The issue here of course is that this data comes from individual fish whereas we are interested in overall group depth occupancy.
+However this limitation can be overcome if we simply are careful about our timescale of interest.
+\newline
+
+To see this consider a highly motile fish like Chinook salmon (\textit{Oncorhynchus tshawytscha}).
+Analysis of movements across different depth levels indicates that within a few short hours these fish can move from depths as shallow as a few meters to many hundreds of meters below the surface.
+This in turn means that from a practical perspective where the fish is now is not an absolute determinant of where it will be in three hours from a physical or energetics perspective.
+If the fish wishes to be at a different depth, in three hours it certainly can be.
+
+However, if we instead were trying to look at an interval along the lines of a few minutes then certainly the fish are simply not fast enough to cover the hundreds of meters of depth available to them and therefore where the fish is now \textit{is} a necessary piece of information for knowing where the fish will be later.
+In other words, if we choose an interval of time large enough that the fish's current location in the water column is not necessary for predicting its upcoming position then we end up modeling depth occupancy preference.
+And across many different individuals this preference will be equivalent to proportion.
+\newline
+
+\begin{objective}
+To model depth occupancy proportions using electronic tagging data by modeling depth occupancy preference at a large enough time scale that would allow full depth mixing. 
+\end{objective}
+
+We have, at this point, identified the data that we wish to use and the kind of model we wish to build.
+We have not yet described the kinds of information or features that we will attempt to use in this model.
+This is just as an important consideration because the potential applications of our model is dependent on what information we choose to include.
+For example, if we wanted to use this model to help inform multi-year management policy but the model required detailed information about primary productivity, unless we also have models of primary productivity that be forecast multiple years ahead, our model is useless to us.
+Therefore we must first determine what specific applications we want to apply this model too.
+
+For now we will assume two applications:
+
+\begin{enumerate}
+\item \textbf{Selectivity/Bycatch Avoidance} - increases selectivity is a win-win outcome for any fishing operation. It can increase the likelihood of a catch while also reducing bycatch.
+\item \textbf{Abundance Estimation} - in order to go from catches to abundance across space and time it is important to understand the differing levels of vulnerability inherent in each of the catches.
+\end{enumerate}
+
+This means that we will want to limit ourselves to information that would be present either to a fishing fleet or to those attempting to create an index of abundance. \newline
+
+Finally, it is obvious that such models would need to be built per species and therefore, if the data becomes available, there will be numerous such models to build.
+Therefore it is also our intention to provide a clear, easy to follow guide on how to organize and then execute on these kinds of models.
+It is the hope that, as with standardize machine learning workflows, this will increase the productivity and quality of the models build in the future.
+
+\begin{objective}
+To provide a clear framework for building and organizing these models. 
+\end{objective}
+
+\subsection{Log Odds Modeling}
 
@@
 
-\subsection{Fitting an Odds Model}
+\subsubsection{Fitting an Odds Model}
 
@@
 \ref{fig:logarch} in order to \textit{train} the log odds model, the actual log odds model is simply the shared layers in a single column behind the softmax layer.
-
-\newpage
-\section{Simulation}
 
@@
 This leads to very broad (but no less massive) datasets.
+
+\subsection{The Depth Model}
+
+With the context of how to build a log odds model out of the way we can now turn to a specific depth occupancy model - one for Chinook salmon (Oncorhynchus tshawytscha) in the Gulf of Alaska (GOA).
+\newline
+
+\textbf{The Data:} The data used were pop-up satellite archival tag (PSAT) data collected from 111 Chinook salmon caught and released in the GOA.
+These tags recorded depth, temperature and light levels before releasing from the fish and returning to the surface where they broadcast that data to the Argos satellite system.
+Finally proprietary software was then used to turn the light levels, temperature, and depth data into most likely paths (in longitude and latitude) for the fish in question \cite{PSAT}.
+As such the data we ended up with for each fish was, by epoch, likely position, depth and temperature.
+\newline
+
+\textbf{The Features:} We then used that data along with elevation data from the General Bathymetric Chart of the Oceans to compute the following features:
+
+\begin{enumerate}
+\item Month (integer from 1 - 12)
+\item Daytime (boolean indicating if it was between sunrise and sunset)
+\item Period Progress (float from 0 - 1 representing how far through day or night it currently was)
+\item Elevation (the average depth in meters in the H3 resolution 4 cell the fish was most likely in) 
+\item Depth Class (depth bin the fish was found in)
+\end{enumerate}
+
+The last feature is worth special note.
+The depth data used for all fish was the lower resolution aggregated data uploaded to the Argos satellite system.
+However we also had a few fish for which tags had been retrieved and the actual measurements themselves were accessible.
+In comparing the aggregations to the real time measurements it was found that the aggregation seemed to represent a binned, first measurement in the interval.
+That is to say that each tag seemed to have a set of aggregation bins (those bins were not necessarily the same per tag or over time for the same tag) and would bin the first measurement from each aggregation interval into each bin.
+The result was that aggregated depth measurements looked like the fish was follow a very jagged step function whereas the real time measurements had the kind of continuity you would expect from actual animal behavior.
+By plotting the measured values against the aggregated values it was found that in general one could expect the real measurements to be within $\pm10$\% of the aggregated value.
+Therefore when computing the depth class we modeled depth occupancy at a moment in time as a distribution and sampled the specific depth class used in training from that distribution.
+\newline
+
+\textbf{Training:}
+
+\begin{enumerate}
+\item Features were computed for all time steps for all fish.
+\item Training and Test sets were split by taking 80\% of the fish for training. Fish were binned exclusively into one or the other set. 
+\item All features were normalized to have a mean of 0 and a standard deviation of 1. 
+\item Hyperparameter tuning of batch size, model depth, and layer sizes was used to determine the best model given the performance on the test set. 
+\end{enumerate}
+
+\subsection{Results}
 

### 2024-07-11 report/Report.tex
wip

@@
 \begin{enumerate}
+\item We chose depth classes of 25, 50, 75, 100, 150, 200, 250, 300, 400, and 500 meters.
 \item Features were computed for all time steps for all fish.
@@
 \subsection{Results}
+\begin{figure}[h] 
+\begin{center}
+  \includegraphics[width=85mm]{images/jan_prob_depth.png}
+ \end{center}
+  \caption{Probability at Depth for January}
+  \medskip
+	\small
+	The probability at depth by local hour for an H3 cell at 58N, 146W in the month of January. Note the movement away from the surface during the day.
+  \label{fig:jan_prob}
+\end{figure}
+
+\begin{figure}[h] 
+\begin{center}
+  \includegraphics[width=85mm]{images/jul_prob_depth.png}
+
+\end{center}
+  \caption{Probability at Depth for July}
+  \medskip
+	\small
+	The probability at depth by local hour for an H3 cell at 58N, 146W in the month of July.
+Note that the probability near the surface is higher than in January \ref{fig:jan_prob} and that the shift downwards reflects the longer day during the summer.
+  \label{fig:jul_prob}
+\end{figure}
+
+\begin{figure}[h] 
+\begin{center}
+  \includegraphics[width=\linewidth]{images/jan_diurnal_map.png}
+
+\end{center}
+  \caption{Map of Depth Occupancy (January)}
+  \medskip
+	\small
+	The cumulative probability for depth classes above and below 100m for January. Shows both during the day and at night.
+  \label{fig:jul_depth_map}
+\end{figure}
+
+After hyper parameter tuning we found that the best model had three layers of size 16, 8, and 16 and a batch size of 1,000.
+This model achieved a categorical cross entropy score on the test set of 1.32 and on the training set of 1.4 (suggesting more variability in the train than test set).
+A model with no predictive power would've gotten a cross entropy score of $\ln{0.1}=-2.3$ meaning that the model is a significant improvement over a random guess.
+
+Interesting patterns emerged in the predicted results.
+First there are clear diurnal patterns in the predictions (Figs.
+\ref{fig:jan_prob}, \ref{fig:jul_prob}) with predictions showing a descent away from the surface during the day.
+Furthermore, as expected, the predictions show a tendency toward shallower depths as the predictions move towards the coast with the diurnal pattern becoming less obvious as one does so (Fig.
+\ref{fig:jul_depth_map}).
 

### 2024-07-19 book/Perspectives on Stewardship.tex
what it means to manage

@@
 
+\chapter{What it Means to Manage}
+
+"Taking a close look at what's around us, there is some sort of a harmony. It is the harmony of overwhelming and collective murder." - Werner Herzog on the Amazon rainforest, 1981. \newline
+
+I absolutely love this quote.
+Not because I dislike rainforests or am a sadist but because it is illuminating an essential part of the natural world that we rarely think about.
+If I bring up rainforests, the arctic, or the oceans, to most people that is synonymous with beautiful butterflies enjoying the sunlight filtering down through a magnificent canopy, polar bear cubs playing in the snow, or reefs teeming with life.
+If ever something dark and dismal is presented it is us tearing down trees, melting the bears' home, or bleaching the reefs.
+We are the violence and we must stop it.
+
+However, in truth the natural world is a massive struggle for survival that takes no prisoners.
+Bears eat the seals that look so cute to us.
+Trees try to crowd out the light that would feed the children of their competitors.
+Insects go about parasitizing each other and many other creatures.
+It is a massive race to get before you're gotten.
+The harmony is not one of collective, atruistic cooperation but of struggle.
+
+Yet it is precisely this struggle that creates the very sustainability and stability that we depend upon and seek to emulate.
+Plants take up nutrients and energy which becomes food for hungry grazers who in turn become the prey of other creatures or disease and so it goes until all is returned to the earth where scavengers and eventually microorganisms return those nutrients back to the soil so that the process can start anew.
+The cycle of life and death is the cycle of renewal.
+
+Likewise stability is a result of these interplays, specifically in the form of negative feedback loops.
+Predators, food availability, and disease regulates the abundance of prey while the abundance of disease and prey availability regulates the abundance of predators.
+A gross over simplification to be sure, but it gets the point across - nature regulates itself through feedback.
+\newline
+
+Nature then is not in harmony with itself because of some kind of stasis or inherent permanence but instead because hundreds of thousands of different dynamic elements are trying to gain ground and through their collective competition create the cycles that allow for sustainability and the negative feedback loops that allow for stability.
+Nature is the sustainability engine it is through its dynamic non-linearity.
+\newline
+
+With this in mind I believe the place of management becomes clear.
+But let me illustrate with an example.
+Waste from human processes (better known today as pollution) has become a problem as of late.
+Run off enters our streams, estuaries, and oceans and has been known to create devastating blooms of algae that choke out the local life.
+What has happened here?
+Quite simply our interaction with the feedback loops present sent them cascading to a new "stable" point.
+When the waste showed up in the quantities it did microorganisms, who are known for their exceptional ability to multiply, were able to increase in quantity at rates that far exceeded any of their predators ability to compensate.
+The feedback loop could not keep pace.
+Eventually the mass of microorganisms got so large that they starved the area of much needed oxygen and their would be regulators died off instead.
+
+Consider instead what would have happened if that waste had more slowly leached into the system.
+If the rate was slow enough then these blooms would not have happened.
+It is not that the waste ended up their that is the immediate problem it is the fact that it came in so fast that it completely overwhelmed the feedback loop that was the problem.
+
+We human beings \textit{will} produce waste and we \textit{will} consume resources from the earth.
+There is no such thing as a humanity separate from its biosphere.
+We are as much a part of the wild as anything else.
+The question to us is not whether we will engage with and shift the world by our presence (as compared to our absence) it is rather a question of whether the way we engineer our interactions results in \textit{dynamic} stability or not.
+Do we engineer interactions that mesh well with the feedback cycles out there or do we topple them over and race towards a great unknown?
+\newline
+
+\textit{Good management means engineering interactions with our dynamic, non-linear biosphere that take advantage of the cycles and feedback loops present to achieve dynamic stability and sustainability.
+Change is inevitable, instability is not.} \newline
+
+In order for us to be good managers then, we need to understand the feedback loops and cycles that are present.
+And this means looking at the world around us and ask - "what jobs do you do?" For example much of the issues that have arisen with industrial farming have come from not appreciating all the kinds of value the biosphere has been/was bringing us in the first place \cite{biomimicry}.
+Soil erosion, nitrogen fixing, pest mitigation, weather resistance, all of these are examples of "jobs" we've realized the biosphere has already figured out only after problems have cropped up from our having eliminated the species doing those jobs.
+Soil erosion is preventable if you have plants that can cover the ground and slow the water as it hits the earth as rain.
+Not as much nitrogen fertilizer is needed if nitrogen fixers are planted in fields as well whatever else is being grown there.
+Diversity in crops helps prevent plagues and just generally ensures not all of our eggs have been placed in one basket.
+
+Likewise following the chain of feedback loops and "jobs" performed helps ensure we don't insert ourselves into an ecosystem attempting to achieve one kind of value while accidentally destroying another.
+Catching one kind of fish in great quantities that is prey for another valuable kind of fish can result in amping up one fishery while decimating another.
+Clearing predators in order to bump up hunting can eventually just result in poorer and poorer quality game as the predators are no longer around to put adaption pressure on the populations that are left.
+Removing wetlands in order to get more farmland and simply removing nurseries for valuable fish species.
+All in all it is important to recognize that not only are their many "jobs" being done but there are usually many values being derived from ecosystems simultaneously and unless we recognize these we are bound to erase things we depend upon by accident.
+\newline
+
+\textit{Good management means understanding the full breadth of these systems, both in terms of all of the "jobs" performed to keep the system stable and sustainable, but also in terms of the various "jobs" performed that bring direct human value.
+That which goes unnoticed ends up being that which is eventually lost.
+}\newline
+
+However this requirement for breadth presents us with a rather big problem.
+There are quite literally millions of different species distributed across our planet and that means there are an absolutely massive number of interactions, cycles, and feedback loops.
+Indeed this number gets even more severe if you think about the fact that each of these kinds of interactions play out in many different place at many different times across the globe.
+How can we possibly know all of this?
+Naively, if we were trying to describe each of these interactions and processes in hyper-specific detail we know, for a fact, the task would be impossible - such knowledge would be more or less equivalent to omniscience.
+However we probably do not need such painfully precise detail.
+
+To illustrate, consider your car (or someone else's if you do not have one).
+Cars these days are extremely complicated.
+There are loads of different systems operating at all times when you are driving one.
+Pumps, engine(s), coolant systems, hydraulics, differentials, navigation systems, stereos, safety systems, lights...
+the list goes on and on.
+Yet you do not have to know all of these in all of their detail to drive the car around safely.
+Indeed driving a car is relatively easy.
+Building a car from scratch, however, takes the industry of many, many souls.
+The key is that you know what things you \textit{need} to pay attention to and what things you can leave as a black box.
+You know you need to steer with the steering wheel, use the gas and break pedals, indicate, and in certain more specific cases use the windshield wipers, understand that the oil light means its time to go to the mechanic and so on.
+The operating abstraction of the car means that you needn't know how every last component in your car works to operate it.
+
+In a similar fashion we need an operating abstraction of ecosystems.
+We need to know what kinds of knowledge we need to get our hands on to operate responsibly, but also what information can be left a mystery so that the problem become tractable at all.
+We need to know how to break down stability and sustainability in a system to its basic, abstract parts.
+Thankfully such a field exists!
+Or rather, such fields exist and they are the fields of cybernetics, systems thinking/design, chaos theory and stability theory.
+These four deal with how to understand and abstract non-linear systems and their dynamics.
+It will be (and to some extent already is) in applying and advancing these fields that a theory of ecological management arises.
+\newline
+
+\textit{Good management takes advantage of the fields of cybernetics, systems thinking/design, chaos theory, and stability theory to provide operating abstractions of ecosystems that makes the problem of management tractable.}\newline
+
+In sum then management is about taking the mathematical fields outlined above to produce an operating abstraction of ecological stability and sustainability, doing the research to describe the various "job" and values in the terms outlined by that operating abstraction, and then designing the human engagement in that system to achieve those values with dynamic stability and sustainability.
+Theory to knowledge to practice.
+
+\bibliographystyle{plain}
+\bibliography{reference}
 \end{document}

### 2024-07-19 book/Perspectives on Stewardship.tex

@@
-\documentclass[11pt,a5paper]{book}
+\documentclass[11pt,a5paper]{article}
 \usepackage[utf8]{inputenc}
@@
 
-\chapter{What it Means to Manage}
+\section{What it Means to Manage}
 
@@
 If the rate was slow enough then these blooms would not have happened.
-It is not that the waste ended up their that is the immediate problem it is the fact that it came in so fast that it completely overwhelmed the feedback loop that was the problem.
+It is not that the waste ended up there that is the immediate problem it is the fact that it came in so fast that it completely overwhelmed the feedback loop.
 
@@
 We are as much a part of the wild as anything else.
-The question to us is not whether we will engage with and shift the world by our presence (as compared to our absence) it is rather a question of whether the way we engineer our interactions results in \textit{dynamic} stability or not.
+The question to us is not whether we will engage with and shift the world by our presence (as compared to our absence) but rather whether the way we engineer our interactions results in \textit{dynamic} stability or not.
 Do we engineer interactions that mesh well with the feedback cycles out there or do we topple them over and race towards a great unknown?
@@
 In order for us to be good managers then, we need to understand the feedback loops and cycles that are present.
-And this means looking at the world around us and ask - "what jobs do you do?" For example much of the issues that have arisen with industrial farming have come from not appreciating all the kinds of value the biosphere has been/was bringing us in the first place \cite{biomimicry}.
+And this means looking at the world around us and asking - "what jobs do you do?" For example much of the issues that have arisen with industrial farming have come from not appreciating all the kinds of value the biosphere has been/was bringing us in the first place \cite{biomimicry}.
 Soil erosion, nitrogen fixing, pest mitigation, weather resistance, all of these are examples of "jobs" we've realized the biosphere has already figured out only after problems have cropped up from our having eliminated the species doing those jobs.
@@
 Removing wetlands in order to get more farmland and simply removing nurseries for valuable fish species.
-All in all it is important to recognize that not only are their many "jobs" being done but there are usually many values being derived from ecosystems simultaneously and unless we recognize these we are bound to erase things we depend upon by accident.
+All in all it is important to recognize that not only are there many "jobs" being done but there are usually many values being derived from ecosystems simultaneously and unless we recognize these we are bound to erase things we depend upon by accident.
 \newline
@@
 There are quite literally millions of different species distributed across our planet and that means there are an absolutely massive number of interactions, cycles, and feedback loops.
-Indeed this number gets even more severe if you think about the fact that each of these kinds of interactions play out in many different place at many different times across the globe.
+Indeed this number gets even more severe if you think about the fact that each of these kinds of interactions play out in many different places at many different times across the globe.
 How can we possibly know all of this?
-Naively, if we were trying to describe each of these interactions and processes in hyper-specific detail we know, for a fact, the task would be impossible - such knowledge would be more or less equivalent to omniscience.
+Naively, if we were trying to describe each of these interactions and processes in hyper-specific detail we know, for a fact, that the task would be impossible - such knowledge would be more or less equivalent to omniscience.
 However we probably do not need such painfully precise detail.
@@
 The key is that you know what things you \textit{need} to pay attention to and what things you can leave as a black box.
-You know you need to steer with the steering wheel, use the gas and break pedals, indicate, and in certain more specific cases use the windshield wipers, understand that the oil light means its time to go to the mechanic and so on.
+You know you need to steer with the steering wheel, use the gas and break pedals, indicate, and in certain more specific cases use the windshield wipers, understand that the oil light means its time to go to the mechanic, and so on.
 The operating abstraction of the car means that you needn't know how every last component in your car works to operate it.
@@
 Thankfully such a field exists!
-Or rather, such fields exist and they are the fields of cybernetics, systems thinking/design, chaos theory and stability theory.
+Or rather, such fields exist and they are the fields of cybernetics, systems thinking/design, chaos theory, and stability theory.
 These four deal with how to understand and abstract non-linear systems and their dynamics.
@@
 Theory to knowledge to practice.
+\newline
+
+Our next step then in understanding what it means to manage is to find that operating abstraction that will allow us to avoid the proverbial boiling of the ocean.
 

### 2024-07-20 proposal/Proposal.tex

@@
 \newtheorem{model}{Model}
+
+\usepackage[authoryear]{natbib}
 
@@
 The outcome will be a package in Python for building, diagnosing, and running odds models.
+Note that at this point the purpose is to demonstrate how to build odds models for the purpose of building simulations.
+It will be in the next chapter that specific research outcomes directly relevant to management will manifest themselves.
+The models' sole purpose will be in powering the simulations.
 
@@
 
-Probabilistic machine classification \cite{durr} of the kind we were pointing to in the last section has two pitfalls when it comes to modeling behavior.
+Probabilistic machine classification \citep{durr} of the kind we were pointing to in the last section has two pitfalls when it comes to modeling behavior.
 
@@
 
-From studies that started with Beverton and Holt's work on recruitment in fisheries, we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to an asymptotic relationship between spawning biomass and recruits \cite{waltersmartell} (Fig.
+From studies that started with Beverton and Holt's work on recruitment in fisheries, we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to an asymptotic relationship between spawning biomass and recruits \citep{waltersmartell} (Fig.
 \ref{fig:recruitment}).
 What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
-Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \cite{waltersmartell}.
+Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \citep{waltersmartell}.
 \newline
@@
 
-The implication for fisheries management is pretty clear - sustainable fishing is in large part driven by just maintaining the spawning stock biomass (SSB) above that knee in the curve \cite{waltersmartell} \cite{king}.
+The implication for fisheries management is pretty clear - sustainable fishing is in large part driven by just maintaining the spawning stock biomass (SSB) above that knee in the curve \citep{waltersmartell} \citep{king}.
 Indeed this is well noted as the basis of why fisheries \textit{can} exist in the first place.
@@
 This gets even more complicated for fish like Pacific salmon (\textit{Oncorhynchus}) or many species of groupers (\textit{Epinephelinae}) where life history is intimately entwined with spawning.
-The salmon obviously only breed once \cite{NOAA24} and some species of groupers are protogynous hermaphrodites \cite{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex.
+The salmon obviously only breed once \citep{NOAA24} and some species of groupers are protogynous hermaphrodites \citep{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex.
 Clearly understanding and managing our age class selectivity matters when it comes to maintaining a healthy SSB for years to come.
@@
 
-The other observation is that many species of fishes actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \cite{lorenzen2010} \cite{salmonplan} \cite{prince2010}.
-Pacific salmon are once again a good example in that they home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \cite{erisman2017}.
-Research also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior \cite{armstrong2001}.
+The other observation is that many species of fishes actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \citep{lorenzen2010} \citep{salmonplan} \citep{prince2010}.
+Pacific salmon are once again a good example in that they home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \citep{erisman2017}.
+Research also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior \citep{armstrong2001}.
 
@@
 Obviously, if we set one global piece of management like an Allowable Biological Catch (ABC) that makes sense for a single SSB but then have biased fishing mortality where one sub stock is getting fished out far more than the others we may drop below the SSB for that one stock without actually exceeding our expected ABC for the whole fishery.
-And this can certainly happen as different sub stocks will often exhibit quite different spatial behaviors in movement \cite{tucker2019} \cite{punt2019} \cite{cadrin2020}.
+And this can certainly happen as different sub stocks will often exhibit quite different spatial behaviors in movement \citep{tucker2019} \citep{punt2019} \citep{cadrin2020}.
 \newline
@@
 
-Pop-up Satellite Archival Tags (PSATs) are a great way of collecting behavior data \cite{PSAT}. They are defined by three key properties:
+Pop-up Satellite Archival Tags (PSATs) are a great way of collecting behavior data \citep{PSAT}. They are defined by three key properties:
 
@@
 
-The latter two are then used along with tag release location to predict animal movement (latitude and longitude) using a proprietary statistical model \cite{PSAT}.
+The latter two are then used along with tag release location to predict animal movement (latitude and longitude) using a proprietary statistical model \citep{PSAT}.
 
 We will look at movements on a day by day basis and depths at the transmitted resolution (2-5 minutes).
-For those tags retrieved, a much higher resolution is available (1-3 seconds) however most data is only guaranteed from transmission and not from retrieval \cite{PSAT}.
+For those tags retrieved, a much higher resolution is available (1-3 seconds) however most data is only guaranteed from transmission and not from retrieval \citep{PSAT}.
 \newline
@@
 At this point in time the main interest here is that Chinook salmon present an interesting model species for testing out the efficacy of these models and simulations.
-Chinook salmon are highly motile and undergo clear migration behaviors \cite{tucker2019} \cite{langan2024}.
+Chinook salmon are highly motile and undergo clear migration behaviors \citep{tucker2019} \citep{langan2024}.
 Therefore they should provide a good challenge to attempts at behavioral modeling and simulation.
@@
 
-Fisheries management is as diverse a subject as the species and cultures to which it is applied \cite{waltersmartell} \cite{king}.
-It simultaneously has to deal with creatures from Bluefin Tuna (\textit{Thunnus thynnus}) who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \cite{prince2010}.
+Fisheries management is as diverse a subject as the species and cultures to which it is applied \citep{waltersmartell} \citep{king}.
+It simultaneously has to deal with creatures from Bluefin Tuna (\textit{Thunnus thynnus}) who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \citep{prince2010}.
 There are species that school and species that consume those schools.
@@
 And to capture this vast array of diversity are similarly diverse strategies and gears.
-From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial bands, and everything in between the variety in strategies that exist to catch fish are numerous \cite{waltersmartell} \cite{king}.
+From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial bands, and everything in between the variety in strategies that exist to catch fish are numerous \citep{waltersmartell} \citep{king}.
 
@@
 What requirements of management are there?
-\cite{waltersmartell} \cite{king}
+\citep{waltersmartell} \citep{king}
 
@@
 
-As with any such endeavor that can come in so many forms and captures so many different people, interests, and potential problems and opportunities, the best creative solutions come when everyone is working together to build and understand those solutions \cite{prince2010} \cite{waltersmartell}.
+As with any such endeavor that can come in so many forms and captures so many different people, interests, and potential problems and opportunities, the best creative solutions come when everyone is working together to build and understand those solutions \citep{prince2010} \citep{waltersmartell}.
 And of utmost importance here is the inclusion of those who actually go out and do the fishing.
@@
 
-First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world \cite{prince2010} \cite{lorenzen2010}.
+First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world \citep{prince2010} \citep{lorenzen2010}.
 By having simulations that can simply be shown and laid down on a map that places the spatio-temporal modeling in familiar context, one should be able to make objections or confirmation much easier.
@@
 While we've already noted that the creative solutions in fisheries management are as diverse as the people, species, and objectives present, there are some specific kinds of strategies that are used across the world to do spatial management.
-By and large they are comprised of performing spatial closures, however those closures differ with regards to who they effect and for how long \cite{selig2016} \cite{little2014}.
+By and large they are comprised of performing spatial closures, however those closures differ with regards to who they effect and for how long \citep{selig2016} \citep{little2014}.
 \newline
@@
 
-Unsurprisingly then, for any kind of spatial management tool in use today, spatio-temporal understanding of the stock is key to both definition and efficacy \cite{selig2016} \cite{little2014}.
+Unsurprisingly then, for any kind of spatial management tool in use today, spatio-temporal understanding of the stock is key to both definition and efficacy \citep{selig2016} \citep{little2014}.
 Therefore it seems likely that bringing high resolution spatio-temporal data to the bear on the issue can only help make such techniques more nuanced, precise, and thereby effective.
@@
 Fisheries assessments are one way in which unknown parameters important to management are estimated.
-By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so on arise from the state model \cite{king}.
+By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so on arise from the state model \citep{king}.
 Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit.
-The observation model on the other hand is encoded as a set of distributions dependent on those parameters (and including some parameters of its own) thus allowing for a likelihood to be computed against known data for any given set of parameters \cite{sippel2014}.
+The observation model on the other hand is encoded as a set of distributions dependent on those parameters (and including some parameters of its own) thus allowing for a likelihood to be computed against known data for any given set of parameters \citep{sippel2014}.
 This, in turn, allows for maximum likelihood estimation (MLE) routines to then fit the likely parameters of the assessment.
@@
 Vulnerability might then be modeled to capture a couple different kinds of gear, provide an expected distribution for catches given your underlying abundance by cohort, and then "simulate" a few years.
-Matching this back to real catch data would then allow you to estimate your natural mortality and recruitment by year \cite{king} \cite{ASAP}.
+Matching this back to real catch data would then allow you to estimate your natural mortality and recruitment by year \citep{king} \citep{ASAP}.
 \newline
@@
 
-As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit \cite{sippel2014} \cite{punt2019} \cite{goethel2023}.
-However from studies of fish behavior and the fact that fish obviously don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \cite{lorenzen2010}.
+As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit \citep{sippel2014} \citep{punt2019} \citep{goethel2023}.
+However from studies of fish behavior and the fact that fish obviously don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \citep{lorenzen2010}.
 
-Therefore in recent years there has been work in creating new assessments that are able to capture spatial heterogeneity and movement \cite{punt2019} \cite{cadrin2020}.
+Therefore in recent years there has been work in creating new assessments that are able to capture spatial heterogeneity and movement \citep{punt2019} \citep{cadrin2020}.
 There are effectively two different kinds of heterogeneity that such models have to capture - that in recruitment and that in post recruitment movement and connectivity (or lack thereof).
@@
 
-Different assessment paradigms will have varying degree of resolution in the sub-regions, will estimate certain parameters freely to not at all, leave $X$ free or constrain it on other kinds of assumptions (indeed some assessments don't model movement at all) \cite{punt2019} \cite{goethel2023}.
+Different assessment paradigms will have varying degree of resolution in the sub-regions, will estimate certain parameters freely to not at all, leave $X$ free or constrain it on other kinds of assumptions (indeed some assessments don't model movement at all) \citep{punt2019} \citep{goethel2023}.
 
-Ultimately what is bounding the freedom in parametrizing these assessments is the availability of data to actually fit all of these parameters \cite{goethel2023} - a subject to which we turn next.
+Ultimately what is bounding the freedom in parametrizing these assessments is the availability of data to actually fit all of these parameters \citep{goethel2023} - a subject to which we turn next.
 
@@
 
-Therefore a huge step forward in being able to run these kinds of assessments was the development of "integrated assessments" - assessments that can take advantage of pretty much any kind of data collected from the fishery \cite{sippel2014}.
+Therefore a huge step forward in being able to run these kinds of assessments was the development of "integrated assessments" - assessments that can take advantage of pretty much any kind of data collected from the fishery \citep{sippel2014}.
 The process here falls neatly into our earlier description of how an assessment works - for some new kind of data (say acoustic tag data) one must add a new component to our observation model that describes the distribution that would generate our acoustic tag data.
@@
 
-There are of course issues that arise in terms of how to weight one data type against another when doing MLE \cite{sippel2014} \cite{punt2019} but by and large integrated assessments are a boon because they increase the data available to fit assessments and have led to several tag integrated assessments that include electronic tag data in addition to more standard mark and recapture tag data.
+There are of course issues that arise in terms of how to weight one data type against another when doing MLE \citep{sippel2014} \citep{punt2019} but by and large integrated assessments are a boon because they increase the data available to fit assessments and have led to several tag integrated assessments that include electronic tag data in addition to more standard mark and recapture tag data.
 
@@
 
-With the recency of their development and overall complexity it has been noted that diagnosing goodness of fit and relevance of spatial assessments is an art that is still in development (although several papers exist that give guidance on best practices in choosing and using spatial assessment paradigms \cite{goethel2023} \cite{punt2019}).
+With the recency of their development and overall complexity it has been noted that diagnosing goodness of fit and relevance of spatial assessments is an art that is still in development (although several papers exist that give guidance on best practices in choosing and using spatial assessment paradigms \citep{goethel2023} \citep{punt2019}).
 
-Every assessment paradigm is of course a simplification of reality but what's important to know is whether those simplifications are obscuring important structure of the stock and sub-stocks, i.e., are the simplifications introducing bias or errors that would lead to contradictory results from sound management based on the advice of the assessment \cite{cadrin2020}?
+Every assessment paradigm is of course a simplification of reality but what's important to know is whether those simplifications are obscuring important structure of the stock and sub-stocks, i.e., are the simplifications introducing bias or errors that would lead to contradictory results from sound management based on the advice of the assessment \citep{cadrin2020}?
 
@@
 
-The way out of this hole, as many authors have noted, is to use simulations \cite{cadrin2020} \cite{goethel2023}.
+The way out of this hole, as many authors have noted, is to use simulations \citep{cadrin2020} \citep{goethel2023}.
 Specifically one can use simulations to do hypothesis testing of the form - if my stock had spatial structure of the form X would my specific assessment capture that at all or would it introduce biases and/or errors as a result?
@@
 
-For such things as fisheries assessments this is particularly concerning as the specific ranges of parameters derived from the models are used directly to set such things as allowable biological catch \cite{king}.
+For such things as fisheries assessments this is particularly concerning as the specific ranges of parameters derived from the models are used directly to set such things as allowable biological catch \citep{king}.
 Yet given we are often estimating recruitment, abundance, mortality, and in the case of spatial assessments things like movement and spatial heterogeneity all in the same model the only guarantee we have is that this choice of parameters fits the current data, not that they are parameters that are indicative of reality itself.
@@
 
-It has been noted that one of the most difficult aspects to fit is the movement matrix $X$ described above \cite{punt2019}.
+It has been noted that one of the most difficult aspects to fit is the movement matrix $X$ described above \citep{punt2019}.
 So here too is another application of the model proposed in this document.
@@
 
-There are many decisions that have to be made when designing a spatial assessment \cite{cadrin2020} \cite{goethel2023} and some of these would directly benefit from the kind of view our proposed model would provide:
+There are many decisions that have to be made when designing a spatial assessment \citep{cadrin2020} \citep{goethel2023} and some of these would directly benefit from the kind of view our proposed model would provide:
 
@@
 Modeling movement has traditionally divided into two categories - the Eulerian and Lagrangian approaches.
-In brief, the former deals with movement as a series of flows throughout a space (that may or may not be discretized) while the latter treats each animal independently \cite{waltersmartell}.
+In brief, the former deals with movement as a series of flows throughout a space (that may or may not be discretized) while the latter treats each animal independently \citep{waltersmartell}.
 We will provide a brief introduction to both approaches.
@@
 
-The typical strategy behind the Eulerian approach is to subdivide the region in question into discrete subregions and then model movements between the boundaries of these regions as the result of advection and diffusion where advection is a directed movement and diffusion is not \cite{sibert1999} \cite{lehodey2008}.
+The typical strategy behind the Eulerian approach is to subdivide the region in question into discrete subregions and then model movements between the boundaries of these regions as the result of advection and diffusion where advection is a directed movement and diffusion is not \citep{sibert1999} \citep{lehodey2008}.
 
-These advection and diffusion terms are represented by partial differential equations where each may be parametrized by life history or environmental factors depending on the model in question \cite{lehodey2008}.
+These advection and diffusion terms are represented by partial differential equations where each may be parametrized by life history or environmental factors depending on the model in question \citep{lehodey2008}.
 Data is included in the model as boundary conditions, for example where we were fish released and where were they recovered.
-Finally time too is discretized in order to allow finite difference methods to be used to solve the PDE's in question \cite{sibert1999}.
+Finally time too is discretized in order to allow finite difference methods to be used to solve the PDE's in question \citep{sibert1999}.
 
@@
 The advantage of the Eulerian approach is that it is specifically attempting to catch population level movement patterns and therefore does not need as thorough parametrization as the Langragian approaches.
-It is also therefore less data intense in its requirements and generally stock assessments are only interested in population or sub-population level movement anyways \cite{lehodey2008}.
+It is also therefore less data intense in its requirements and generally stock assessments are only interested in population or sub-population level movement anyways \citep{lehodey2008}.
 
@@
 
-The approaches used to build these kinds of models are extremely variable \cite{grimm} (a point we'll get back to momentarily) but follow the general paradigm of:
+The approaches used to build these kinds of models are extremely variable \citep{grimm} (a point we'll get back to momentarily) but follow the general paradigm of:
 
@@
 All of these steps are presented so vaguely because there really are no general paradigms for how these are done.
-For example, in the last step - tuning the parameters - strategies range from integration of deterministic models against boundary conditions \cite{phillipsa2018}, to "by-eye" tuning \cite{goodwin2006}, to weighted averages across multiple different allowable parameters \cite{burke2016}.
+For example, in the last step - tuning the parameters - strategies range from integration of deterministic models against boundary conditions \citep{phillipsa2018}, to "by-eye" tuning \citep{goodwin2006}, to weighted averages across multiple different allowable parameters \citep{burke2016}.
 \newline
 
-Such approaches however have been shown to be able to model very fine scale movement dynamics and have been especially useful in applications in rivers where structures like dams are being designed and the ecological impacts are wanting to be understood \cite{goodwin2006} \cite{grimm}.
+Such approaches however have been shown to be able to model very fine scale movement dynamics and have been especially useful in applications in rivers where structures like dams are being designed and the ecological impacts are wanting to be understood \citep{goodwin2006} \citep{grimm}.
 These kinds of fine scale dynamics are not possible with Eulerian approaches.
@@
 
-Finally it has been noted that one can derive Eulerian models from Lagrangian ones making the latter perhaps more general \cite{phillipsa2018}. \newline
+Finally it has been noted that one can derive Eulerian models from Lagrangian ones making the latter perhaps more general \citep{phillipsa2018}. \newline
 
@@
 
-There are a series of key issues \cite{grimm} that have arisen in building IBM's of which we will note a few that relate to our proposal:
+There are a series of key issues \citep{grimm} that have arisen in building IBM's of which we will note a few that relate to our proposal:
 
@@
 
-Chinook salmon (\textit{Oncorhynchus tshawytscha}) are the largest species of Pacific salmon \cite{oceanchinook}.
+Chinook salmon (\textit{Oncorhynchus tshawytscha}) are the largest species of Pacific salmon \citep{oceanchinook}.
 Like other Pacific salmon the species is both anadromous and semelparous leading to a very interesting and distinct life history.
@@
 Juvenile Chinook salmon are generally born in the rivers where their parents spawned (although some genetic mixing does occur through straying).
-Sometime in their first or second year of life they transition to the marine environment where they will spend the rest of their life before maturation \cite{oceanchinook}.
+Sometime in their first or second year of life they transition to the marine environment where they will spend the rest of their life before maturation \citep{oceanchinook}.
 This leads to a specific system of enumerating age that includes both years in freshwater and years at sea.
@@
 
-The immature fish then spend time (up to five years) growing in the marine environment before maturing and then migrating back to the river where they were born to breed and subsequently die \cite{oceanchinook}.
-The timing of such homing behavior is clearly season dependent with species further north having to avoid ice build up during the cooler months \cite{oceanchinook}.
-The mechanism behind this homing ability is still a question under active research \cite{bracis2012}.
+The immature fish then spend time (up to five years) growing in the marine environment before maturing and then migrating back to the river where they were born to breed and subsequently die \citep{oceanchinook}.
+The timing of such homing behavior is clearly season dependent with species further north having to avoid ice build up during the cooler months \citep{oceanchinook}.
+The mechanism behind this homing ability is still a question under active research \citep{bracis2012}.
 \newline
@@
 Given this homing behavior, Chinook salmon populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies.
-Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook salmon is highly dependent on the region of origin for the fish in question \cite{shelton2019} \cite{tucker2019}.
+Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook salmon is highly dependent on the region of origin for the fish in question \citep{shelton2019} \citep{tucker2019}.
 
-Beyond this, large scale analysis of fisheries dependent data has also demonstrated some of the likely migration patterns the species exhibits during its rearing phase at sea \cite{langan2024}.
+Beyond this, large scale analysis of fisheries dependent data has also demonstrated some of the likely migration patterns the species exhibits during its rearing phase at sea \citep{langan2024}.
 
@@
 
-Broadly distributed through out the Northern pacific, Chinook salmon are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side \cite{oceanchinook}.
+Broadly distributed through out the Northern pacific, Chinook salmon are known to spawn in rivers spanning from the Sacramento River in central California to the Kotzebue Sound in Alaska on the eastern side of the pacific and primarily in the Kamchatka Peninsula on the western side \citep{oceanchinook}.
 
-Marine residency in the North Eastern pacific is primarily coastal from California up to Western Alaska and then spans into much of the Eastern Bering Sea \cite{oceanchinook} \cite{langan2024}.
+Marine residency in the North Eastern pacific is primarily coastal from California up to Western Alaska and then spans into much of the Eastern Bering Sea \citep{oceanchinook} \citep{langan2024}.
 \newline
@@
 Chinook salmon are well known piscivores with several studies of stomach contents finding primarily fish and squid as their main sources of prey.
-It has been noted that this diet is similar to that of Coho salmon but that it is unlikely that the two species experience much cross-species competition for these resources \cite{oceanchinook}.
+It has been noted that this diet is similar to that of Coho salmon but that it is unlikely that the two species experience much cross-species competition for these resources \citep{oceanchinook}.
 
 Chinook salmon themselves are prey for animal groups such as Salmon sharks, whales, and dolphins but it is still unclear how much mortality in later life stages is due to predation as opposed to other sources.
-Size selective predation on earlier life stages is a well known driver of natural mortality \cite{oceanchinook}.
+Size selective predation on earlier life stages is a well known driver of natural mortality \citep{oceanchinook}.
 
-Other drivers of natural mortality include a variety of known diseases, starvation during periods of low ocean productivity, and potential competition for food with species like whales \cite{oceanchinook}.\newline
+Other drivers of natural mortality include a variety of known diseases, starvation during periods of low ocean productivity, and potential competition for food with species like whales \citep{oceanchinook}.\newline
 
@@
 The vast majority of these individuals (approx.
-200million) have come from hatchery programs in the United States with the remaining significant portion from programs in Canada \cite{oceanchinook}.
+200million) have come from hatchery programs in the United States with the remaining significant portion from programs in Canada \citep{oceanchinook}.
 
 Fish are released at weights of around 6-9g and a large number of these have a micro-wire with an identifiable code etched into them inserted into their snouts.
-These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook salmon sub-populations \cite{shelton2019}.
+These are the tags driving the Coded Wire Tag program in the Northeastern Pacific that has become a major source of demographic and movement data used in managing and understanding Chinook salmon sub-populations \citep{shelton2019}.
 The etchings provide a unique label for each cohort from each hatchery.
@@
 The initial purpose of these hatcheries programs was to supplement losses in juveniles due to targeting of adult fish in river systems, however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact.
-For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments \cite{oceanchinook}.
+For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments \citep{oceanchinook}.
 \newline
 
-There are valuable tribal, commercial, and recreational fisheries of Chinook salmon throughout the Pacific \cite{salmonplan}.
+There are valuable tribal, commercial, and recreational fisheries of Chinook salmon throughout the Pacific \citep{salmonplan}.
 
@@
 Federal management councils (NOAA for the United States) then collaborate with the Pacific Fishery Management Council to set harvest quotas and other management objectives.
-The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals \cite{NOAA22}.
+The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals \citep{NOAA22}.
 
-These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook salmon form so many different genetic sub-populations \cite{salmonplan} - a fact that further complicates management.
-Indeed several sub populations are actually listed as endangered and threatened under the Endangered Species Act \cite{NOAA24}.
+These management goals as well as designations of statuses such as "overfished" have to be stated on the sub-population level given the fact that Chinook salmon form so many different genetic sub-populations \citep{salmonplan} - a fact that further complicates management.
+Indeed several sub populations are actually listed as endangered and threatened under the Endangered Species Act \citep{NOAA24}.
 
-This combined with other threatened stocks \cite{NOAA24} means that bycatch avoidance is a serious issue for Chinook salmon management and for other overlapping fisheries such as the Walleye Pollock fishery \cite{ianelli2015}.
+This combined with other threatened stocks \citep{NOAA24} means that bycatch avoidance is a serious issue for Chinook salmon management and for other overlapping fisheries such as the Walleye Pollock fishery \citep{ianelli2015}.
 
-To manage this diversity the United States uses a variety of tools including but not limited to - differentiated management zones, minimum size limits, bag limits for recreational fisheries, allowed seasons and quotas, and fishing gear restrictions \cite{salmonplan}.
+To manage this diversity the United States uses a variety of tools including but not limited to - differentiated management zones, minimum size limits, bag limits for recreational fisheries, allowed seasons and quotas, and fishing gear restrictions \citep{salmonplan}.
 \newline
@@
 Finally, Chinook salmon are especially prized for their large size which has led to serious questions about size selective mortality from fishing as both recreational and commercial fisheries have incentives for targeting larger fish.
-While it is not yet known whether fisheries are responsible, there is clear evidence that fish size in Chinook salmon populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook salmon's primary predators \cite{kendall2011}\cite{oceanchinook}.
+While it is not yet known whether fisheries are responsible, there is clear evidence that fish size in Chinook salmon populations has been decreasing steadily over the years which has potential impacts on both the fisheries themselves (in terms of optimizing yields) and for Chinook salmon's primary predators \citep{kendall2011}\citep{oceanchinook}.
 \newline
 
-All in all then, given their ecology and the corresponding complexities of management, Chinook salmon represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
+Being an interesting species to model, however, is not enough for our purposes.
+Given that we want to build machine learning models and model driven simulations we also need a species with available and high quality data.
+Chinook salmon in the GOA fit this bill nicely.
+There is quite a lot of electronic tagging data that has been collected for them (CITE), all of which is fisheries independent allowing for closer examination of Chinook salmon behavior, and there is an abundance of other diverse data types including coded wire tag (CWT) data, fishery independent trawl data, and so on.
+This availability of data means that we can focus on the modeling aspects of this problem as opposed to data acquisition.
+\newline
+
+All in all then, given their ecology, data availability, and the corresponding complexities of management, Chinook salmon represent an ideal model species for trying to understand and quantify spatial patterns of movement and behavior.
 

### 2024-07-27 book/Perspectives on Stewardship.tex
draft

@@
 
+\newpage
+
+\section{Stability and Control}
+
+What is a system?
+A system can be a lot of different things but fundamentally it is a set of states and how those states change.
+This allows us to define a pretty simple looking definition of a system.
+Suppose our state can be represented by a vector $x$.
+Following in the tradition of cybernetics\cite{cybernetics} we'll refer to this as our operand.
+Then our system simply transforms this operand into what we'll call a transform $x'$ which is just the new state of the system.
+The system therefore can be defined by $x'=T(x)$.
+Note here we're assuming that we're interested in our system as a set of discreet time steps.
+Also, for now we'll assume that $T$ is single valued which means that for each $x$ there is one and only one transform $x'$.
+Continuing in the language of cybernetics we call $T$ a transducer.
+So any system is simply a transducer.
+
+Let's define a simple example transducer for ourselves. Suppose our $x$ is a vector of the number of fish at ages 0, 1, 2, and so on. That is:
+
+$$x=[n_0, n_1, ..., n_A]$$
+
+where $A$ is the maximum age of the species in question.
+
+Then suppose our transducer is given by:
+
+$$n_i'=T_i(n_{i-1})=n_{i-1}M \forall i > 0$$
+
+and
+
+$$n_0'=R$$
+
+where $R > 0$ and $M < 1$.
+
+This represents a system where each age group experiences the same mortality $M$ and we have constant recruitment of baby fish each year.
+A simple system indeed but one that will give us something more concrete to work with as we go along.
+\newline
+
+Now our first question is - are there stable states within this system?
+What is a stable state?
+Well one situation that would be certainly stable is if there existed some $x$ such that $T(x)=x$ i.e.
+the system doesn't change the state.
+In our case this will be true if:
+
+$$n_i=RM^i$$
+
+This the represents an equilibrium point.
+We can however broaden this definition and ask whether there are sets of states $S=\{x\}$ such that for any $x\in S$ we also have that $T(x)\in S$.
+This would also be a kind of equilibrium in the sense that once the system enters $S$ it will never leave.
+Why broaden the definition like this?
+Well consider a system over three states $a$, $b$, and $c$ s.t.
+$T(a)=b$, $T(b)=a$ and $T(c)=a$.
+There is obviously no equilibrium point in this system as it will just volley back and forth between $a$ and $b$ but the set $S=\{a,b\}$ is still a smaller set than the whole domain of $T$ and clearly is stable within itself.
+In other words this definition allows us to still talk about stability even for oscillatory systems.
+
+A set $S$ is therefore considered \textbf{stable} if for all $x\in S$ it follows that $T(x)\in S$. If your system enters a stable set it will never leave that set. \newline
+
+This however doesn't seem to really be enough to consider a system as a whole "stable".
+For example the transducer defined by $x'=T(x)=2x$ has a stable set $S=\{0\}$ but so long as $x\neq 0$ that system will never go to $S$.
+So what we want is not only a stable set $S$ but another set $U$ s.t.
+our system will tend towards $S$.
+
+One way we could define this would be to ask if there is a set $U$ s.t.
+the under operation of our transducer the system state gets closer to $S$ and the new state is also in $U$.
+Then, by induction we would be able to say that asymptotically states in $U$ would go to $S$.
+To put it more precisely we'd be looking for another stable region $U$ (because we need points in $U$ to lead to other points in $U$ under transformation by our system) such that $d(x, S) > d(T(x), S) \forall x \in U$.
+Were $d(x, S)$ is the distance between $x$ and $S$.
+
+Sometimes however $d$ is such a nasty function that evaluating that definition directly is quite difficult.
+When that is the case we can take advantage of an insight by Aleksandr Lyapunov\cite{stability}.
+Suppose that we have a function $V(x)$ defined on $U$ such that $V(x)=0$ when $x\in S$, $V(x)>0$ when $x\notin S$ and $V(x) > V(T(x)) \forall x\in U - S$.
+If these assumptions are true it also holds that the system tends toward $S$ while in $U$ and given $U$ is itself stable any point in $U$ \textit{will} asymptotically lead toward $S$.
+The $V$ are called Lyapunov Functions.
+
+So in general we can now say that if there exists a stable set $U$ over which a lyapunov function can be defined relative to another stable set $S$ that $S$ is an \textbf{attractor} which is asymptotically stable.
+\newline
+
+So is our example system's equilibrium an attractor? Well if our equilibrium point is $s$, given some $x=[n_0, n_1, ..., n_A]$ we have:
+
+$$x_i-s_i=n_i-RM^i$$
+
+and
+
+$$T_i(x_i) - s_i = n_{i-1}M-RM^i \forall i>0$$
+$$T_0(x_0) - s_0 = R - R = 0$$
+
+Effectively then the difference for each element where $i>0$ is getting multiplied by $M$ and $n_0$ is getting set precisely to $R$.
+Therefore the distance from equilibrium is certainly dropping under $T$ and so our $s$ is indeed an attractor.
+\newline
+
+At this point it seems our system is so simple it offers no interesting behavior.
+Surely if there were no fish to speak of we'd be in trouble!
+We can model this by creating a new piece-wise transducer where if the sum of $x$ is below some value $R$ begins to decrease with the sum of $x$.
+In other words if $\sum_i n_i < N$ then our $T_0$ becomes:
+
+$$T_0=R  \frac{\sum_i n_i}{N}$$
+
+In this case it will be noted we get a new stable set $x=0$.
+Our old stable set is also still there but now there is this "danger zone" as well.
+Indeed one can work out that there are sets $U$ and lyapunov functions for both of our equilibrium points that make both of them attractors.
+
+This is really important to highlight because it shows that a system does not just exhibit one kind of behavior.
+One can in general have multiple attractors and so understanding our system means dividing up its domain into regions of attraction to specific attractors and regions that are either unattracted or just simply diverge.
+\newline
+
+Alright so far we've spoken of stable sets, attractors, and the fact that our system exhibits different behaviors based on the region it starts in.
+But we haven't yet talked about control.
+Control however is just the introduction of some new component to the system.
+I.e.
+rather than just having $T$ there is some second function $C$ which takes the output of $T$ and changes the states itself.
+However $C(T(x))=x'$ is itself just a new transducer.
+So everything we've talked about still applies.
+Where then is the control?
+Well usually $C$ will have "inputs".
+For example suppose that we've decided to apply a fishing mortality pressure $F$ to our earlier super simple system.
+I.e.
+we now have:
+
+$$n_i'=T_i(n_{i-1})=n_{i-1}M \forall i > 0$$
+
+and
+
+$$n_0'=RF$$
+
+Well we can see that as before we have an attractor now defined by:
+
+$$n_i=RM^iF^{i+1}$$
+
+The control here comes in the fact that we can \textit{choose} what $F$ should be.
+If $F$ is set to 0 then we arrive back at the old system, but as we vary $F$ the attractor will move.
+This is how we exert control.
+Note however that the set of attractors is simply a line through our state space.
+That is because we have a single control parameter.
+If we had two distinct parameters our set of equilibria would define a surface, if we had three a volume, and so on.
+I.e.
+the degrees of freedom in our control is limited by the degrees of freedom in our inputs.
+This is a restatement of the notion of Requisite Variety\cite{cybernetics}.
+Practically in our case it means that right now we can control how many fish we catch at equilibrium but not the age distribution of those fish.
+We'd need to add addition, independent control parameters to allow for that.
+\newline
+
+Alright let's summarize what we now know:
+
+\begin{enumerate}
+\item \textbf{Transducers:} Any system (including ones with regulators) can be defined as a transducer.
+\item \textbf{Stability:} A set $S$ is considered to be stable if for all $x\in S$ it is also true that $T(x)\in S$.
+\item \textbf{Attractors:} A set $S$ is an attractor if there exists a stable set $U$ and a lyapunov function of $x\in U$ relative to set $S$. 
+\item \textbf{Modes of Behavior:} In general a transducer $T$ may have multiple stable sets as well as regions that show no attraction.
+Therefore attraction is dependent both on the system and the starting state $x$.
+\item \textbf{Control:} Control comes from the inclusion of input parameters to the system. The result of control is the movement of attractors and their regions of attraction. 
+\item \textbf{Requisite Variety:} The degrees of freedom in control is limited by the degrees of freedom in the input parameters. 
+\item \textbf{Laissez-Faire:} Given the above it should be clear now that control in a deterministic system is not about constantly grappling with that system.
+Instead it is about designing a system with input parameters that give the degrees of freedom and movement in attractors that's desired.
+Then, once those parameters have been set there is nothing left to do but allow the system to follow its course to stability.
+\end{enumerate}
+
+That last point should feel like it stands in stark contrast to reality.
+Certainly life is not so simple that you can just design systems and never worry about them again.
+Yet if $T$ can be trusted it seems like that is the case.
+But what happens if $T$ can't be trusted?
+What happens if it changes under our feet?
+Well that brings us directly to the question of systems with noise.
+
+\newpage
+
 \bibliographystyle{plain}

### 2024-08-08 book/Perspectives on Stewardship.tex
draft

@@
 
+\section{Understanding Noise by Understanding Inputs}
+
+Noise can be considered in a lot of different ways.
+However it seems practical to take a leaf out of the study of feedback control in linear time invariant systems \cite{feedback}.
+At first look these systems seem to be setup quite differently (from a mathematical) perspective than the ones we just went over in the last section.
+Specifically they are described by a series of ordinary differential equations that are concerned with an input function $u$, an output function $y$, and some parameter (usually time) $t$.
+One classic example is the dampened spring:
+
+$$m\partial^2y+b\partial y+ky=u$$
+
+This is no longer a direct definition of our transducer but instead a "constraint" governing what our transducer can look like.
+However given $u$ and the appropriate initial conditions this will give you a single $y(t)$ that describes the spring's path through state space.
+
+As such we can think of $y$ as a specific pathway, determined by $u$, and parametrized by $t$.
+What is $u$ then?
+Well it is a force applied to our spring over time and so we can consider it equivalent to setting the input on the transducer from one moment to the next - $u(t)$ is the output of an "input transducer" over time.
+What's important to notice here however is that $u$ is independent of $y$, i.e.
+$u$ is specified without knowing what $y$ is.\newline
+
+How does this relate to noise?
+Well consider the fact that while a die is represented by some kind of distribution that in reality when the dice are rolled we get a specific set of dice rolls.
+Distributions exist to help us explain phenomena over many repeats of the same "experiment" but they don't really "exist" in real life - we are only ever left with one "pathway" through the state space that those distributions describe.
+So, if we had some gambling system where the "noise" was generated by dice rolls we can imagine that for any run of the experiment there is a specific $u$ that describes the rolls we got.
+This $u$ then feeds into our specific gambling system to then give us our path through the state space $y$.
+If we then want to consider what would happen under another set of dice rolls we'd construct a different $u$ and see what new $y$ we get out of it.
+But in each case our overall system of $u$ and $y$ stays deterministic and follows all the rules we illuminated in the last section.
+
+What we want to study then is how our system $y$ behaves under different input pathways $u$ and determine which $u$'s lead to bad outcomes. \newline
+
+There is however a problem.
+In both this case and in the case of building lyupanov functions for the sake of studying stability there is no silver bullet, no single method (or at least no one has found one yet).
+A great quote from "Feedback Control of Dynamic Systems" is that "the theory only gives the engineer a hunting license to look for a [lyapunov] function" \cite{feedback}.
+So what are we to do?
+Well once again we can take a leaf out of the LTI (linear time invariant) systems book.
+\newline
+
+Their way of dealing with this is to take advantage of the specific properties that make an LTI and LTI - and that's linearity and time invariance.
+This more or less means that if $u_1$ is one input with response $y_1$ and $u_2$ is an input with response $y_2$ that the response of $au_1 + bu_2$ (where $a$ and $b$ are constants) is given by $ay_1 + by_2$.
+In other words if a specific input can be represented as the sum of a series of inputs, its response is equal to the sum of the responses of the individual inputs.
+
+So what about the response to the dirac delta function? Suppose we call that $h(t)$. Then because the dirac delta function happens to be an impulse of size 1 that last only infinitesimally long that:
+
+$$u(t) = \int_{-\infty}^{\infty}u(\tau) \delta(\tau)d\tau$$
+
+and therefore:
+
+$$y(t) = \int_{-\infty}^{\infty}u(\tau) h(t-\tau)d\tau$$
+
+More or less all this is saying is that because our system in linear and time invariant that the response to some arbitrary input is the "sum" of the impulse responses multiplied by the magnitude of the input at that time carried forward (or backward) to the current time.
+
+This form in and of itself isn't so interesting but the magic happens when we consider what happens in our input is of the form $e^{st}$.
+
+ $$y(t) = \int_{-\infty}^{\infty}e^{s\tau} h(t-\tau)d\tau$$
+
+ which is equivalent under a change of variable to:
+
+ $$y(t) = \int_{-\infty}^{\infty}e^{s(t-\tau)} h(\tau)d\tau = \int_{-\infty}^{\infty}e^{-s\tau} h(\tau)d\tau e^{st}=H(s)e^{-st}$$
+
+ where $H(s)$ is the Laplace transform of $h(t)$.
+But notice that we've just shown that for an input $e^{st}$ our system will produce an output that is just some scalar multiple of the input!
+So what does this $H(s)$ look like?
+Well we can solve for it in our original equation:
+
+ $$ms^2H(s)e^{st}+bsH(s)e^{st}+kH(s)e^{st}=e^{st}$$
+
+ $$H(s) = \frac{1}{ms^2+bs+k}$$
+
+This $H(s)$ (also known as the transfer function) tells us everything we need to know about how the systems responds to different frequencies of input $s$.
+For example if $s$ is a root of the above denominator we know we're going to have significant issues.
+And because we can imagine any arbitrary input $u$ as being some combinations of frequencies we've also sorted out how to tell what properties of the input could give us issues.
+Very useful indeed.\newline
+
+Notice though that in order to arrive at this result we had to lean heavily on the system being LTI.
+For some other kind of system none of this would apply.
+And that's just the nature of studying systems - to find their specific properties there has to be some inherent structure (beyond just having a transducer of some kind) that you can leverage to pull out the characteristics of those systems.
+\newline
+
+For us then it means that we need to find specific properties of ecosystem subsystems that will allow us to define them as a specific type(s) of system and then leverage those properties into understanding the characteristics of the inputs that could lead to issues.
+Only by getting specific are we going to be able to make any more progress.
+
+\newpage
+
 \bibliographystyle{plain}

### 2024-09-26 proposal/Proposal.tex
edits

@@
 \maketitle
+\newpage
+
+\section{Summary}
+
+Vulnerability of fish stocks to various human and environmental impacts is tied to the spatio-temporal distribution and movement of those stocks.
+Therefore understanding these patterns is important to improving management practice today.
+Central to understanding these patterns is the creation of behavior models that can be used to test movement hypotheses and predict future and current movement behavior.
+Yet, as of today, there is no common methodology for building and using such models.
+The purpose of this study is to provide such a common methodology that takes advantage of modern computing solutions such as machine learning and the cloud.
+It is hoped that by providing such a strategy that (1) the barrier to producing movement models will be greatly reduced and (2) a common language for discussing, evaluating, and using these models will arise.
+
 \newpage
@@
 
-\textbf{Target:} Depth range. \newline
+\textbf{Target:} Discrete depth ranges. \newline
 
@@
 
-\textbf{Target:} Spatial grid cell selection as movement. Any state feature updates needed to capture internal states of a cohort. \newline
+\textbf{Target:} Spatial grid cell selection as movement. Also if clearly different modes of movement are found (migration vs foraging for example) we'll predict the mode as well. \newline
 
@@
 \item Splitting Data into Train, Validation, Test
-\item Training and Hyperparameter Tuning Using Standard Model Types
+\item Training and Hyperparameter Tuning Using Standard Neural Net Architectures
 \item Diagnoses of Fit and Comparison of Different Model Types
@@
 	\small
-	Boxes with rounded edges would be inputs specified by the user. The hexagons are functions 		provided by the simulation and the trapezoids are data generated during the main loop. 
+	Boxes with rounded edges would be inputs specified by the user.
+The hexagons are functions 		provided by the simulation and the trapezoids are data generated during the main loop.
+It is important to note that this design is anticipatory, not all mortality or environmental data may be available at this time in which case spatially homogeneous inputs would be provided.
   \label{fig:simulation}
@@
 \item \textbf{Understanding Connectivity:} by looking at the vector flows of fish throughout the space we will be able to identify which areas are highly connected and which are more or less separated by environment, cohort, or other factors of fish behavior.
-\item \textbf{Understanding Vulnerability:} by especially looking at changes in depth occupancy and preference for habitat we will attempt to quantify the degree of vulnerability to different kinds of fishing gear as a function of space and time.
-\item \textbf{Cumulative Mortality:} by being able to follow cohorts through space and time we should be able to show how different fishing pressure scenarios affect the cumulative fishing mortality.
+\item \textbf{Understanding Vulnerability:} by especially looking at changes in depth occupancy and preference for habitat we will attempt to quantify the degree of vulnerability to different kinds of activities as a function of space and time.
+Specific vulnerabilities of interest might include: vulnerability to specific fishing fleets, vulnerability to naval training activities, vulnerability to marine mammals.
+\item \textbf{Cumulative Mortality:} by being able to follow cohorts through space and time we should be able to show how different fishing pressure scenarios might affect the cumulative fishing mortality.
+Note that the data for specific fleets may be unavailable but that our interest here is to demonstrate that given data we can make these kinds of predictions.
 \item \textbf{Protected Areas/Times:} Of those fishing pressure scenarios, one of especial interest is modeling Marine Protected Areas (MPAs) or other kinds of closures in space and time.
+Given the motility of Chinook, how effective would marine protected areas be?
 \item \textbf{Uncertainty:} we will show that such a simulation can be run at scale, bootstrapping can be applied, and therefore uncertainty in simulation outcomes quantified. 
@@
 \ref{fig:recruitment}).
-What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits.
-Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \citep{waltersmartell}.
+What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits (there is some variation on this theme depending on the precise stock recruitment model used).
+Given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \citep{waltersmartell}.
 \newline
@@
 Indeed this is well noted as the basis of why fisheries \textit{can} exist in the first place.
-If taking spawners always led to decreases in recruits, we'd quickly fish a stock out of existence.
+If taking spawners always led to significant decreases in recruits, we'd quickly fish a stock out of existence.
 It is only through the compensation we get on the asymptote that sustainable fisheries are possible.
+However, two very simple observations of fish biology will show how actually managing SSB can be made difficult.
 \newline
-
-However, two very simple observations of fish biology will show how actually managing SSB can be made difficult.
 
@@
 This gets even more complicated for fish like Pacific salmon (\textit{Oncorhynchus}) or many species of groupers (\textit{Epinephelinae}) where life history is intimately entwined with spawning.
-The salmon obviously only breed once \citep{NOAA24} and some species of groupers are protogynous hermaphrodites \citep{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex.
+The salmon only breed once \citep{NOAA24} and some species of groupers are protogynous hermaphrodites \citep{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex.
 Clearly understanding and managing our age class selectivity matters when it comes to maintaining a healthy SSB for years to come.
@@
 Pacific salmon are once again a good example in that they home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \citep{erisman2017}.
-Research also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior \citep{armstrong2001}.
+Research also indicates that even fish like Atlantic herring likely form distinct genetic stocks due to differences in homing during breeding and school behavior \citep{armstrong2001}.
 
@@
 
-All of this to say that understanding heterogeneity in vulnerability by things like age class and sub stock is of the utmost importance for managing a fishery and truly ensuring an appropriate amount of SSB remains year over year.
+All of this to say that understanding heterogeneity in selectivity by things like age class and sub stock is of the utmost importance for managing a fishery and truly ensuring an appropriate amount of SSB remains year over year.
 It is also worth noting that this knowledge around changes in selectivity in space and time is not only useful for considerations around SSB but also for other management issues such as bycatch avoidance and growth overfishing.
 
-What kind of tooling do we need in order to address these issues and test out ideas for "selectivity aware" fisheries management? Clearly we need some kind of reliable simulation. \newline
+What kind of tooling do we need in order to address these issues and test out ideas for fisheries management? Clearly we need some kind of reliable simulation. \newline
 
@@
 
-Clearly then this kind of data suites our modeling and simulation purposes well as we get behavior outcomes directly.
+Clearly then this kind of data suits our modeling and simulation purposes well as we get behavior outcomes directly.
 
@@
 
-Fisheries management is as diverse a subject as the species and cultures to which it is applied \citep{waltersmartell} \citep{king}.
+Fisheries management is as diverse a subject as the species and contexts to which it is applied \citep{waltersmartell} \citep{king}.
 It simultaneously has to deal with creatures from Bluefin Tuna (\textit{Thunnus thynnus}) who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \citep{prince2010}.
@@
 And to capture this vast array of diversity are similarly diverse strategies and gears.
-From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial bands, and everything in between the variety in strategies that exist to catch fish are numerous \citep{waltersmartell} \citep{king}.
+From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial fleets, and everything in between the variety in strategies that exist to catch fish are numerous \citep{waltersmartell} \citep{king}.
 
@@
 In this way mapped out views of the simulation work to be a useful device for consensus building among multiple groups of stakeholders.
-Not everyone may be familiar with the nuances of a specific kind of model but everyone in the problem can point to a map and understand what's going on.
+Not everyone may be familiar with the nuances of a specific kind of model but everyone with an interest can point to a map and understand what's going on.
 \newline
@@
 For example, suppose that we have a more standard model that has been hand crafted by a scientist familiar with the species in question.
-They may have included specific reactions to, say, oceanic currents, temperatures, and primary productivity.
+They may have included specific associations with, say, oceanic currents, temperatures, and primary productivity.
 But then if someone looking at the simulation points out that they think the species' behavior is also being driven by day/night patterns the scientist then has to go back to the drawing board, figure out how to add this into their overall model, code it up, tune parameters, and then return - all without any real guarantee that whoever provided the feedback is right.
@@
 
-Finally it is worth noting that the potential in combining this modeling with fisheries dependent information is quite interesting on its own.
+Finally it is worth noting that the potential in combining this modeling with fishery dependent information is quite interesting on its own.
 Any forecasting model is always going to have issues predicting farther and farther into the future but fisheries dependent data could provide a means for making corrections to keep the model in line.
@@
 Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit.
-The observation model on the other hand is encoded as a set of distributions dependent on those parameters (and including some parameters of its own) thus allowing for a likelihood to be computed against known data for any given set of parameters \citep{sippel2014}.
+The observation model on the other hand is encoded as a set of distributions dependent on those parameters (and including some parameters of its own) thus allowing for the likelihood of the parameters given the data to be computed \citep{sippel2014}.
 This, in turn, allows for maximum likelihood estimation (MLE) routines to then fit the likely parameters of the assessment.
@@
 
-In short the pattern is (1) design a sim that has your unknown statistics as parameters, (2) specify how data is drawn from the sim, (3) fit your parameters using MLE.
+In short the pattern is (1) design a simulation that has your unknown statistics as parameters, (2) specify how data is drawn from the simulation, (3) fit your parameters to your data using MLE.
 Then those desired statistics are extracted from the assessment (usually with some notion of uncertainty attached to them) and used to inform management.
@@
 
-As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit \citep{sippel2014} \citep{punt2019} \citep{goethel2023}.
+As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock of interest is bounded by the area under assessment, is well mixed, and constitutes a single unit \citep{sippel2014} \citep{punt2019} \citep{goethel2023}.
 However from studies of fish behavior and the fact that fish obviously don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \citep{lorenzen2010}.
@@
 By having a movement simulation based on real observations of the species in question we would be able to help build a high quality simulation of at least the last of those three.
-Rather than having to assume one form of a movement or another we'd instead be able to see whether the assessment paradigm in question can capture the relevant aspects of structure in the \textit{real} movement of our stocks and sub-stocks.
+Rather than having to assume one form of movement or another we'd instead be able to see whether the assessment paradigm in question can capture the relevant aspects of structure in the \textit{real} movement of our stocks and sub-stocks.
 And this in turn would help in selecting the assessments best suited to the job (as well as illustrate what uncertainties exist if a lack of specific kinds of data requires us to use an over-simplified assessment paradigm).
@@
 But this is obviously non-sensical.
-However because the model is deterministic in nature the modeler cannot depend on stochasticity to save the day - instead they must find whatever is driving the difference between staying put and moving on.
+However, because the model is deterministic in nature the modeler cannot depend on stochasticity to save the day - instead they must find whatever is driving the difference between staying put and moving on.
 And this can cascade into greater and greater requirements for modeling the "internal state" of an animal.
@@
 The timing of such homing behavior is clearly season dependent with species further north having to avoid ice build up during the cooler months \citep{oceanchinook}.
-The mechanism behind this homing ability is still a question under active research \citep{bracis2012}.
+The mechanism behind this natal homing ability is still a question under active research \citep{bracis2012}.
 \newline
 
-Given this homing behavior, Chinook salmon populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies.
+Given this homing behavior, Chinook salmon populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategy.
 Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook salmon is highly dependent on the region of origin for the fish in question \citep{shelton2019} \citep{tucker2019}.
@@
 
-The initial purpose of these hatcheries programs was to supplement losses in juveniles due to targeting of adult fish in river systems, however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact.
+The initial purpose of these hatchery programs was to supplement losses in juveniles due to targeting of adult fish in river systems, however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact.
 For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments \citep{oceanchinook}.
@@
 
-There are valuable tribal, commercial, and recreational fisheries of Chinook salmon throughout the Pacific \citep{salmonplan}.
-
-Given the highly mobile nature of Chinook salmon independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Fishery Management Council.
-Federal management councils (NOAA for the United States) then collaborate with the Pacific Fishery Management Council to set harvest quotas and other management objectives.
+There are valuable tribal, commercial, and recreational fisheries of Chinook salmon throughout the Pacific \citep{salmonplan} yet given the highly mobile nature of Chinook salmon independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Salmon Commission.
+Federal management councils (NOAA for the United States) then collaborate with the Pacific Salmon Commission to set harvest quotas and other management objectives.
 The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals \citep{NOAA22}.
@@
 
-\section{Appendices}
-
-\subsection{Value of Log Likelihood Maximization}
-
-\begin{theorem}
-Suppose we have a set of possible outcomes $V=\lbrace v_k \rbrace$ with probabilities $P_k$. Such that:
-
-$$\sum_k P_k = P$$
-
-Next suppose we concoct a new series of probabilities $U_k = P_k \alpha_k$ s.t.
-
-$$\sum_k U_k = P-\epsilon$$
-
-where $\epsilon \geq 0$. There does not exist a set of $\alpha_k$ s.t.
-
-$$\prod_k \left(\frac{P_k\alpha_k}{P_k}\right)^{P_kN}>1$$
-
-\end{theorem}
-
-\begin{proof}
-
-We proceed by induction. \newline
-
-First note that from the equation above we get:
-
-$$\sum_k P_k \ln \alpha_k > 0$$
-
-and subtracting $\sum_k U_k$ we arrive at:
-
-$$\sum_k P_k \left( \ln \alpha_k - \alpha_k \right) > -P+\epsilon$$
-
-Now suppose for $k$ outcomes we know no such $\alpha_k$ exist as to satisfy the above. Suppose we incorporate a new $k+1$ term s.t:
-
-$$P_{k+1} + \sum_k {P_k} = P_{k+1} + P$$
-
-and:
-
-$$P_{k+1}\alpha_{k+1} + \sum_k {P_k}\alpha_k = P_{k+1} + P - \epsilon = P_{k+1}\alpha_{k+1} + \left(P - P_{k+1}(\alpha_{k+1}-1) \right) - \epsilon$$
-
-Now we need:
-
-$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) + \sum_k P_k \left( \ln \alpha_k - \alpha_k \right) > -P_{k+1}-P+\epsilon$$
-
-However given our inductive assumption we know that at best,
-
-$$\sum_k P_k \left( \ln \alpha_k - \alpha_k \right)=-P$$
-
-therefore at best:
-
-$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P > -P - P_{k+1} + \epsilon$$
-
-Furthermore the easiest case for us is if $\epsilon=0$. If it cannot be satisfied than this certainly doesn't work for $\epsilon>0$. So let's consider:
-
-$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P > -P - P_{k+1}$$
-
-or equivalently:
-
-$$\ln \alpha_{k+1}> \alpha_{k+1}-1$$
-
-Now if $\alpha_{k+1}=1$ we have:
-
-$$\ln 1 = 1-1$$
-
-Further consider the derivatives of each side:
-
-$$\partial_{\alpha}\ln \alpha_{k+1}=\frac{1}{\alpha_{k+1}}$$
-
-$$\partial_{\alpha}(\alpha_{k+1}-1) = 1$$
-
-If $\alpha_{k+1}> 1$ then the log component rises more slowly than the constant component.
-I.e.
-our left side will be larger than the right.
-Likewise if $\alpha_{k+1}<1$ the log component will shrink faster that the constant component which means it will also be less than the constant component.
-Therefore given our assumptions:
-
-$$\ln \alpha_{k+1} \not\ge \alpha_{k+1}-1$$
-
-and so:
-
-$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P \not\ge -P - P_{k+1} + \epsilon$$
-
-So much for the $k+1$th case. What about $k=1$. This is trivial because any $\alpha_1$ that satisfies:
-
-$$\sum_k U_k = P-\epsilon$$
-
-must be less than or equal to $1$ and therefore our product of quotients:
-
-$$\prod_k \left(\frac{P_k\alpha_k}{P_k}\right)^{P_kN}>1$$
-
-must be less than or equal to one.
-
-\end{proof}
-
-With this proof in hand we can now show how in the limit as the number of samples taken $N$ goes to infinity our likelihood is only maximized if $\hat{F} \rightarrow F$. \newline
-
-For some given information $\theta$ and outcomes $V=\lbrace v_k \rbrace$ we have, in gathering our data, observed $v_k$ $N_k$ times. Therefore our overall likelihood is:
-
-$$\mathcal{L} = \prod_k \hat{F}(v_k, \theta)^{N_k}$$
-
-Now suppose that we represent:
-
-$$\hat{F}(v_k, \theta)=F(v_k, \theta)\alpha_k$$
-
-That is the ratio of our likelihoods between $\hat{F}$ and $F$ is given by:
-
-$$\prod_k \left(\frac{F(v_k, \theta)\alpha_k}{F(v_k, \theta)}\right)^{N_k}$$
-
-but we also know that:
-
-$$\sum_k F(v_k, \theta) = \sum_k F(v_k, \theta)\alpha_k = 1$$
-
-and that $\lim_{N\rightarrow \inf}N_k = F(v_k, \theta)N$ so that we have:
-
-$$\prod_k \left(\frac{F(v_k, \theta)\alpha_k}{F(v_k, \theta)}\right)^{F(v_k, \theta)N}$$
-
-which is now in the exact same form as our theorem above.
-And we now know that this ratio is maximized when $\alpha_k\equiv 1$.
-So as $N\rightarrow \inf$ a higher $\mathcal{L}$ means we are closer to approximating $F$.
-
-\newpage
-
 \section{Bibliography}

### 2024-09-26 proposal/Proposal.tex
edits

@@
 
-{\Huge \bf \@title\par}
+{\Large \bf \@title\par}
+\vspace{1cm}
+
+{\Large \bf (Proposal) \par}
 \vspace{2.5cm}
@@
 
-\vspace{2.5cm}
-{\Large Dr. Andrew Seitz}\hspace{2cm}{\Large Dr. Curry Cunningham}\\[2cm]{\Large Michael Courtney, M.S.}\\[2cm]
+\vspace{1cm}
+{\Large Dr. Andrew Seitz}\hspace{2cm}{\Large Dr. Curry Cunningham}\\[2cm]{\Large Michael Courtney, M.S.}\\[1cm]
 College of Fisheries and Ocean Sciences\\
@@
 
-\title{Master's Research Proposal (DRAFT)}
+\title{Demonstration of a Standardized Methodology for Building and Using Individual Movement Models using Chinook Salmon Pop-up Archival Tagging Data from the Gulf of Alaska}
 
@@
 
-\section{Summary}
+\section{Purpose}
 
@@
 
-The ultimate goal of this research is to provide a new pathway by which collections of high resolution spatio-temporal movement data that are representative of a species can be used to make a direct impact on fisheries management.
-It is hoped that this will provide an incentive for creating/enlarging such datasets.
-
 \begin{objective}
-To provide tooling that allows for using machine learning methods to fit models of the form
+Movement modeling is fundamentally about modeling the selection of movement outcomes.
+So, given a set of choices $v_k$ and a set of features $n_k$ describing those choices, we want a standardized method to fit, evaluate, and use models of the form:
 
@@
 
-that maximize the following objective:
+where $\psi_k$ is the "odds for" choice $v_k$. Such a model should maximize the following objective:
 
@@
 $$P'(v_i|\eta_i) = \frac{\psi_i}{\sum_k \psi_k}$$
-
-These models will be known as odds models as they predict the "odds for" each outcome $v_k$ given the information contained in $\eta_k$.
 
@@
 \item \textbf{Protected Areas/Times:} Of those fishing pressure scenarios, one of especial interest is modeling Marine Protected Areas (MPAs) or other kinds of closures in space and time.
-Given the motility of Chinook, how effective would marine protected areas be?
+Given the motility of Chinook, how must effective protected areas/times be designed?
 \item \textbf{Uncertainty:} we will show that such a simulation can be run at scale, bootstrapping can be applied, and therefore uncertainty in simulation outcomes quantified. 

### 2024-09-30 docs/conf.py
cleanup

@@ -1,40 +1,0 @@
-# Configuration file for the Sphinx documentation builder.
-#
-# For the full list of built-in configuration values, see the documentation:
-# https://www.sphinx-doc.org/en/master/usage/configuration.html
-
-# -- Project information -----------------------------------------------------
-# https://www.sphinx-doc.org/en/master/usage/configuration.html#project-information
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(".."))
-
-project = "mirrorverse"
-copyright = "2024, Marcel Gietzmann-Sanders"
-author = "Marcel Gietzmann-Sanders"
-
-# -- General configuration ---------------------------------------------------
-# https://www.sphinx-doc.org/en/master/usage/configuration.html#general-configuration
-
-extensions = [
-    "sphinx.ext.todo",
-    "sphinx.ext.viewcode",
-    "sphinx.ext.autodoc",
-    "sphinx.ext.graphviz",
-    "pyan.sphinx",
-]
-
-templates_path = ["_templates"]
-exclude_patterns = ["_build", "Thumbs.db", ".DS_Store"]
-
-
-# -- Options for HTML output -------------------------------------------------
-# https://www.sphinx-doc.org/en/master/usage/configuration.html#options-for-html-output
-
-html_theme = "sphinx_rtd_theme"
-html_static_path = ["_static"]
-
-add_module_names = False
-graphviz_output_format = "svg"

### 2024-09-30 mirrorverse/blob/states.py

@@ -1,39 +1,0 @@
-"""
-Split Train and Test
-"""
-
-import click
-import pandas as pd
-import numpy as np
-
-
-@click.command()
-@click.option("--data_path", "-d", help="path to data file", required=True)
-@click.option("--output_dir", "-o", help="directory to store outputs", required=True)
-@click.option("--training_size", "-ts", help="training size", type=float)
-def main(data_path, output_dir, training_size):
-    """
-    Main function for splitting data into training and testing sets.
-    """
-
-    print("Reading Data...")
-    data = pd.read_csv(data_path)
-
-    print("Splitting into Train and Test...")
-    training_ptt = set(
-        np.random.choice(
-            data["_identifier"].unique(),
-            round(data["_identifier"].unique().shape[0] * training_size),
-            replace=False,
-        )
-    )
-    testing_ptt = set(data["_identifier"].unique()) - training_ptt
-    print(len(training_ptt), len(testing_ptt))
-
-    training_data = data[data["_identifier"].isin(training_ptt)]
-    testing_data = data[data["_identifier"].isin(testing_ptt)]
-
-    print("Saving...")
-    training_data.to_csv(f"{output_dir}/training_states.csv", index=False)
-    if testing_ptt:
-        testing_data.to_csv(f"{output_dir}/testing_states.csv", index=False)

### 2024-09-30 mirrorverse/blob/tree/blob_root.py

@@ -1,248 +1,0 @@
-"""
-The Blob
-"""
-
-from time import time
-
-import click
-import pandas as pd
-import numpy as np
-from sklearn.model_selection import KFold
-from tqdm import tqdm
-
-from mirrorverse.tree import DecisionTree
-from mirrorverse.chinook.utils import diff_heading
-from mirrorverse.utility import get_proposed_utility
-
-
-class BlobBuilder:
-    """
-    Blob choice builder.
-    """
-
-    STATE = ["type"]
-    CHOICE_STATE = []
-    COLUMNS = ["heading", "type"]
-
-    def __init__(self, enrichment):
-        pass
-
-    def __call__(self, state, choice_state):
-        slices = 24
-
-        choices = pd.DataFrame(
-            {"heading": np.linspace(2 * np.pi / slices, 2 * np.pi, slices)}
-        )
-
-        choices["type"] = state["type"]
-
-        return choices
-
-
-class BlobGround:
-    """
-    Build the Blob Data
-    """
-
-    def __init__(self, favored_heading):
-        self.favored_heading = favored_heading
-        self.choice_builder = BlobBuilder({})
-        self.choices = self.choice_builder({"type": 0}, {})
-        self.choices["diff_heading"] = self.choices["heading"].apply(
-            lambda x: diff_heading(self.favored_heading, x)
-        )
-        self.choices["utility"] = self.choices["diff_heading"].apply(
-            lambda x: 0.5 ** (np.abs(x) / (np.pi / 4))
-        )
-        self.choices["probability"] = (
-            self.choices["utility"] / self.choices["utility"].sum()
-        )
-
-    def new_state(self, state):
-        """
-        Builds a new state for the Blob.
-
-        State should have x, y, type
-        """
-        heading = np.random.choice(
-            self.choices["heading"], p=self.choices["probability"]
-        )
-        x, y = state["x"], state["y"]
-        return {
-            "x": x + np.cos(heading),
-            "y": y + np.sin(heading),
-            "heading": heading,
-            "type": state["type"],
-        }
-
-
-class BlobRoot(DecisionTree):
-    """
-    Blob Model
-    """
-
-    BUILDERS = [BlobBuilder]
-    FEATURE_COLUMNS = ["heading", "type"]
-    OUTCOMES = ["heading"]
-    BRANCHES = {}
-    PARAM_GRID = {"n_estimators": [10, 20], "min_samples_leaf": [25, 50]}
[…]

### 2024-09-30 mirrorverse/chinook/db.py

@@ -1,112 +1,0 @@
-"""
-Queries for Chinook Model
-"""
-
-import click
-import pandas as pd
-
-from mirrorverse.warehouse.utils import get_engine
-
-
-def get_elevation():
-    """
-    Retrieve elevation information from the warehouse
-
-    Returns pd.DataFrame
-    """
-    sql = """
-    select
-        elevation,
-        h3_level_4_key as h3_index
-    from 
-        elevation
-    """
-    return pd.read_sql_query(sql, get_engine())
-
-
-def get_surface_temperature():
-    """
-    Retrieve surface temperature information from the warehouse
-
-    Returns pd.DataFrame
-    """
-    sql = """
-    select
-        s.temperature_c,
-        s.h3_level_4_key as h3_index,
-        d.month
-    from 
-        surface_temperature s
-        inner join dates d
-        on s.date_key = d.date_key
-    """
-    return pd.read_sql_query(sql, get_engine())
-
-
-def get_tag_tracks():
-    """
-    Retrieve tag tracks information from the warehouse
-
-    Returns pd.DataFrame
-    """
-    sql = """
-    select
-        tt.tag_key,
-        tt.latitude,
-        tt.longitude,
-        d.date,
-        d.year,
-        d.month,
-        d.day
-    from 
-        tag_tracks tt 
-        inner join dates d
-        on tt.date_key = d.date_key
-    """
-    return pd.read_sql_query(sql, get_engine())
-
-
-def get_context():
-    """
-    Retrieve context information from the warehouse
-
-    Returns pd.DataFrame
-    """
-    sql = """
-    select
-        t.tag_key,
-        t.fork_length_cm,
-        coalesce(hr.home_region, 'Unknown') as home_region
-    from 
-        tags t
-        left join home_regions hr
-            on hr.tag_key = t.tag_key
-    """
-    return pd.read_sql_query(sql, get_engine())
-
-
-@click.command()
-@click.option("--elevation", "-e", required=True, help="file to save elevation data")
-@click.option(
-    "--surface_temperature",
-    "-s",
-    required=True,
-    help="file to save surface temperature data",
-)
-@click.option("--tag_tracks", "-t", required=True, help="file to save tag tracks data")
-@click.option("--context", "-c", required=True, help="file to save context data")
-def main(elevation, surface_temperature, tag_tracks, context):
-    """
-    Retrieve data from the warehouse and save it to files
-    """
[…]

### 2024-09-30 mirrorverse/chinook/simulate.py

@@ -1,145 +1,0 @@
-"""
-Code for simulating using the Chinook decision tree.
-"""
-
-# pylint: disable=duplicate-code
-
-import os
-import pickle
-from multiprocessing import Pool
-
-import click
-import pandas as pd
-import h3
-
-
-from mirrorverse.chinook.tree.run_or_drift import RunOrDriftBranch
-from mirrorverse.chinook.states import (
-    get_elevation,
-    get_surface_temps,
-)
-
-
-def simulate(args):
-    """
-    Input:
-    - ptt: ptt
-    - h3_index (str): starting h3 index
-    - date (pd.Timestamp): starting date
-    - home_region (str): the home region of the fish
-    - fork_length_cm (float): fork length of the fish
-    - steps (int): number of steps to simulate
-    - decision_tree (DecisionTree): decision tree to use
-
-    Returns a DataFrame with the simulated data.
-    """
-    ptt, h3_index, date, home_region, fork_length_cm, steps, decision_tree = args
-    state = {
-        "drifting": True,
-        "steps_in_state": 1,
-        "h3_index": h3_index,
-        "month": date.month,
-        "mean_heading": 0,
-        "home_region": home_region,
-        "fork_length_cm": fork_length_cm,
-    }
-    row = dict(state)
-    row["ptt"] = ptt
-    row["date"] = date
-    rows = [row]
-    for _ in range(steps):
-        choice_state = {}
-        decision_tree.choose(state, choice_state)
-
-        date = date + pd.Timedelta(days=1)
-        steps_in_state = (
-            state["steps_in_state"] + 1
-            if state["drifting"] == choice_state["drifting"]
-            and (
-                not choice_state["drifting"]
-                and state["mean_heading"] == choice_state["mean_heading"]
-            )
-            else 1
-        )
-        new_state = {
-            "drifting": choice_state["drifting"],
-            "steps_in_state": steps_in_state,
-            "h3_index": choice_state["h3_index"],
-            "month": date.month,
-            "heading": 0 if choice_state["drifting"] else choice_state["heading"],
-            "mean_heading": (
-                0 if choice_state["drifting"] else choice_state["mean_heading"]
-            ),
-            "home_region": home_region,
-            "fork_length_cm": fork_length_cm,
-        }
-
-        row = dict(new_state)
-        row["ptt"] = ptt
-        row["date"] = date
-        rows.append(row)
-
-        state = new_state
-
[…]

### 2024-09-30 mirrorverse/chinook/train.py

@@ -1,55 +1,0 @@
-"""
-Code for training the models for the Chinook decision tree.
-"""
-
-# pylint: disable=duplicate-code
-
-import pickle
-
-import click
-import pandas as pd
-
-from mirrorverse.chinook.tree.drift_movement import train_drift_movement_model
-from mirrorverse.chinook.tree.run_movement import train_run_movement_model
-from mirrorverse.chinook.tree.run_heading import train_run_heading_model
-from mirrorverse.chinook.tree.run_or_drift import train_run_or_drift_model
-from mirrorverse.chinook.tree.run_stay_or_go import train_run_stay_or_go_model
-from mirrorverse.chinook.states import get_elevation, get_surface_temps
-
-
-@click.command()
-@click.option("--node", "-n", help="node to train on", required=True)
-@click.option("--train_data_path", "-tr", help="path to training data", required=True)
-@click.option("--test_data_path", "-te", help="path to testing data", required=True)
-@click.option("--temps_path", "-t", help="path to surface temps file", required=True)
-@click.option("--elevation_path", "-e", help="path to elevation file", required=True)
-@click.option("--model_dir", "-m", help="directory to save models in", required=True)
-def main(node, train_data_path, test_data_path, temps_path, elevation_path, model_dir):
-    """
-    Main function for training the models for the Chinook decision tree.
-    """
-    pd.options.mode.chained_assignment = None
-
-    print("Pulling Enrichment...")
-    enrichment = {
-        "elevation": get_elevation(elevation_path),
-        "surface_temps": get_surface_temps(temps_path),
-        "neighbors": {},
-    }
-
-    print("Loading Data...")
-    training_data = pd.read_csv(train_data_path)
-    testing_data = pd.read_csv(test_data_path)
-
-    model_export = {
-        "DriftMovementLeaf": train_drift_movement_model,
-        "RunMovementLeaf": train_run_movement_model,
-        "RunHeadingBranch": train_run_heading_model,
-        "RunOrDriftBranch": train_run_or_drift_model,
-        "RunStayOrGoBranch": train_run_stay_or_go_model,
-    }[node](training_data, testing_data, enrichment)
-
-    print("Exporting Models...")
-    model_export_path = f"{model_dir}/{node}.pkl"
-    with open(model_export_path, "wb") as fh:
-        pickle.dump(model_export, fh)

### 2024-09-30 mirrorverse/chinook/tree/run_heading.py

@@ -1,260 +1,0 @@
-"""
-Run Heading Model for Chinook salmon
-"""
-
-# pylint: disable=duplicate-code, protected-access
-
-from time import time
-
-import pandas as pd
-import numpy as np
-import h3
-from tqdm import tqdm
-from sklearn.model_selection import KFold
-
-from mirrorverse.tree import DecisionTree
-from mirrorverse.utility import get_proposed_utility
-from mirrorverse.chinook import utils
-from mirrorverse.chinook.tree.run_stay_or_go import RunStayOrGoBranch
-
-
-class RunHeadingChoiceBuilder:
-    """
-    Run Heading choice builder for Chinook salmon.
-    """
-
-    STATE = ["h3_index", "month", "mean_heading", "drifting", "home_region"]
-    CHOICE_STATE = []
-    COLUMNS = [
-        "mean_heading",
-        "elevation",
-        "diff_elevation",
-        "temp",
-        "last_mean_heading",
-        "was_drifting",
-        "month",
-        "unknown",
-        "seak",
-        "wa/or",
-        "bc",
-    ]
-
-    def __init__(self, enrichment):
-        self.surface_temps = enrichment["surface_temps"]
-        self.elevation = enrichment["elevation"]
-
-    def __call__(self, state, choice_state):
-        slices = 24
-
-        choices = pd.DataFrame(
-            {"mean_heading": np.linspace(2 * np.pi / slices, 2 * np.pi, slices)}
-        )
-
-        # we only want immediate neighbors in this case
-        neighbors = pd.DataFrame({"h3_index": list(h3.k_ring(state["h3_index"], 1))})
-        start_lat, start_lon = h3.h3_to_geo(state["h3_index"])
-        neighbors["h3_heading"] = neighbors.apply(
-            lambda row: utils.get_heading(
-                start_lat, start_lon, *h3.h3_to_geo(row["h3_index"])
-            ),
-            axis=1,
-        )
-        choices = choices.merge(neighbors, how="cross")
-        choices["diff_heading"] = choices.apply(
-            lambda row: utils.diff_heading(row["mean_heading"], row["h3_heading"]),
-            axis=1,
-        )
-        choices = choices.sort_values("diff_heading", ascending=True).drop_duplicates(
-            "mean_heading", keep="first"
-        )
-        del choices["h3_heading"]
-        del choices["diff_heading"]
-
-        choices["month"] = state["month"]
-        choices = choices.merge(
[…]

### 2024-09-30 mirrorverse/chinook/tree/run_or_drift.py

@@ -1,210 +1,0 @@
-"""
-RunOrDrift model for Chinook salmon.
-"""
-
-# pylint: disable=duplicate-code, protected-access
-
-from time import time
-
-from tqdm import tqdm
-import pandas as pd
-from sklearn.model_selection import KFold
-
-from mirrorverse.tree import DecisionTree
-from mirrorverse.utility import get_proposed_utility
-from mirrorverse.chinook.tree.run_heading import RunHeadingBranch
-from mirrorverse.chinook.tree.drift_movement import DriftMovementLeaf
-
-
-class RunOrDriftBuilder:
-    """
-    Run or Drift choice builder for Chinook salmon.
-    """
-
-    STATE = ["drifting", "steps_in_state", "month", "home_region"]
-    CHOICE_STATE = []
-    COLUMNS = [
-        "was_drifting",
-        "steps_in_state",
-        "drift",
-        "month",
-        "unknown",
-        "seak",
-        "wa/or",
-        "bc",
-    ]
-
-    def __init__(self, enrichment):
-        pass
-
-    def __call__(self, state, choice_state):
-        choices = pd.DataFrame(
-            [
-                {
-                    "was_drifting": state["drifting"],
-                    "steps_in_state": state["steps_in_state"],
-                    "month": state["month"],
-                    "drift": True,
-                },
-                {
-                    "was_drifting": state["drifting"],
-                    "steps_in_state": state["steps_in_state"],
-                    "month": state["month"],
-                    "drift": False,
-                },
-            ]
-        )
-
-        one_is_true = False
-        for option in ["SEAK", "Unknown", "WA/OR", "BC"]:
-            choices[option.lower()] = state["home_region"] == option
-            one_is_true |= state["home_region"] == option
-        assert one_is_true
-
-        return choices
-
-
-class RunOrDriftBranch(DecisionTree):
-    """
-    Run or Drift model for Chinook salmon.
-    """
-
-    BUILDERS = [RunOrDriftBuilder]
-    FEATURE_COLUMNS = [
-        "was_drifting",
-        "steps_in_state",
-        "drift",
-        "month",
-        "unknown",
-        "seak",
-        "wa/or",
-        "bc",
-    ]
-    OUTCOMES = ["drifting"]
-    BRANCHES = {
-        "run": RunHeadingBranch,
-        "drift": DriftMovementLeaf,
-    }
[…]

### 2024-09-30 mirrorverse/docks/api.py

@@ -1,11 +1,0 @@
-"""
-To keep commands.py from changing everytime we add some new importer.
-"""
-
-from mirrorverse.docks.elevation import pull_elevation_data
-from mirrorverse.docks.surface_temperature import pull_surface_temperature_data
-
-FILE_IMPORTERS = {
-    "elevation": pull_elevation_data,
-    "surface_temperature": pull_surface_temperature_data,
-}

### 2024-09-30 mirrorverse/graph.py

@@ -1,70 +1,0 @@
-"""
-For graphing out decision trees
-"""
-
-import click
-import graphviz
-
-from mirrorverse.chinook.tree.run_or_drift import RunOrDriftBranch
-
-
-def build_graph(decision_tree):
-    """
-    Input:
-    - decision_tree (DecisionTree): the decision tree to graph
-
-    Build a graph of the decision tree
-    """
-    dot = graphviz.Digraph(decision_tree.__name__, format="png")
-
-    def build_node_label(root):
-        states = set()
-        choice_states = set()
-        for builder in root.BUILDERS:
-            states.update(builder.STATE)
-            choice_states.update(builder.CHOICE_STATE)
-
-        features = root.FEATURE_COLUMNS
-        outcomes = root.OUTCOMES
-
-        label = f"""<
-        <FONT>{root.__name__}</FONT>
-        <BR/>
-        <FONT POINT-SIZE="10">States: {sorted(states)}</FONT>
-        <BR/>
-        <FONT POINT-SIZE="10">Choice States: {sorted(choice_states)}</FONT>
-        <BR/>
-        <FONT POINT-SIZE="10">Features: {sorted(features)}</FONT>
-        <BR/>
-        <FONT POINT-SIZE="10"><B>Outcomes:</B> {sorted(outcomes)}</FONT>
-        >"""
-        return label
-
-    def build_nodes(root, dot):
-        label = build_node_label(root)
-        dot.node(root.__name__, label=label, shape="box")
-        for choice, branch in root.BRANCHES.items():
-            build_nodes(branch, dot)
-            if len(root.BRANCHES) > 1:
-                dot.edge(root.__name__, branch.__name__, label=choice)
-            else:
-                dot.edge(root.__name__, branch.__name__)
-
-    build_nodes(decision_tree, dot)
-
-    return dot
-
-
-@click.command()
-@click.option("--decision_tree", "-d", help="decision tree to graph", required=True)
-def main(decision_tree):
-    """
-    Main function for graphing out decision trees
-    """
-    manifest = {
-        "RunOrDriftBranch": RunOrDriftBranch,
-    }
-    assert decision_tree in manifest
-
-    dot = build_graph(manifest[decision_tree])
-    dot.render()

### 2024-09-30 mirrorverse/grouper.py

@@ -1,295 +1,0 @@
-"""
-Grouper - to help us group data by divergence
-or conformance with a utility model.
-"""
-
-import os
-import random
-from multiprocessing import Pool
-
-import click
-import pandas as pd
-import numpy as np
-from tqdm import tqdm
-
-from mirrorverse.log_odds_model.log_odds_model import get_central_likelihood
-
-
-# pylint: disable=missing-class-docstring, missing-function-docstring
-class Individual(list):
-    def __init__(self, *args, **kwargs):
-        super().__init__(*args, **kwargs)
-
-        self.fitness = None
-
-    def invalidate(self):
-        self.fitness = None
-
-    def clone(self):
-        return Individual(self)
-
-
-def create_individual(identifiers, groups):
-    """
-    Inputs:
-    - identifiers (list): list of identifiers
-    - groups (list): list of groups
-
-    Returns a list of which group each identifier belongs to.
-    """
-    return Individual([np.random.choice(groups) for _ in identifiers])
-
-
-def mutate(individual, groups, mutation_gene_rate):
-    """
-    Inputs:
-    - individual (list): the individual to mutate
-    - groups (list): the groups to choose from
-    - mutation_gene_rate (float): the rate of individual gene
-        mutation
-
-    Mutates the individual by changing one of its elements.
-    """
-    individual.invalidate()
-    for i, _ in enumerate(individual):
-        if np.random.random() < mutation_gene_rate:
-            individual[i] = np.random.choice(groups)
-
-
-def crossover(ind1, ind2, crossover_gene_rate):
-    """
-    Inputs:
-    - ind1 (list): the first individual
-    - ind2 (list): the second individual
-    - crossover_gene_rate (float): the rate of crossover
-
-    Crosses over the two individuals by swapping elements.
-    """
-    ind1.invalidate()
-    ind2.invalidate()
-    for i, _ in enumerate(ind1):
-        if np.random.random() < crossover_gene_rate:
-            ind1[i], ind2[i] = ind2[i], ind1[i]
-
-
-def get_central_likelihoods(groups, identifiers, data):
-    """
-    Inputs:
-    - groups (list): the groups
-    - identifiers (list): the identifiers
-    - data (pd.DataFrame): the data to evaluate over
-        should have columns "_identifier", "selected",
-        and "probability"
-
-    Returns the central likelihoods of the groups and the
-    counts of the groups.
-
-    dict, dict
-    """
[…]

### 2024-09-30 mirrorverse/log_odds_model/search.py

@@ -1,85 +1,0 @@
-"""
-Hyperparameter search for the odds model.
-"""
-
-# pylint: disable=invalid-name
-
-import numpy as np
-import pandas as pd
-
-from mirrorverse.log_odds_model.log_odds_model import LogOddsModel
-
-
-def build_randomized_param_sets(param_grids, M, max_attempts):
-    """
-    Inputs:
-    - param_grids: dict, parameter names to grid of values
-    - M: int, number of parameter sets to generate
-    - max_attempts: int, maximum number of attempts to generate a unique parameter set
-
-    Outputs:
-    - list of dicts, parameter sets
-    """
-    param_sets = []
-    attempts = 0
-    while len(param_sets) < M:
-        assert attempts < max_attempts
-
-        param_set = {}
-        for param, grid in param_grids.items():
-            param_set[param] = np.random.choice(grid)
-        if param_set in param_sets:
-            attempts += 1
-        else:
-            attempts = 0
-            param_sets.append(param_set)
-
-    return param_sets
-
-
-def randomized_odds_model_search(
-    train_data,
-    test_data,
-    iterations,
-    model_class,
-    param_grids,
-    M,
-    max_attempts,
-):
-    """
-    Inputs:
-    - train_data: pd.DataFrame, training data
-    - test_data: pd.DataFrame, testing data
-    - iterations: int, number of iterations to run
-    - model_class: class, model class to use
-    - param_grids: dict, parameter names to grid of values
-    - M: int, number of parameter sets to generate
-    - max_attempts: int, maximum number of attempts to generate a unique parameter set
-
-    Outputs:
-    - list of dicts, parameter sets
-    - pd.DataFrame, diagnostics
-    """
-    param_sets = build_randomized_param_sets(param_grids, M, max_attempts)
-    diagnostics_array = []
-    for i, param_set in enumerate(param_sets):
-        print(f"Running parameter set {i + 1}/{M}")
-        to_pass = {k: v for k, v in param_set.items() if k != "learning_rate"}
-        model = LogOddsModel(model_class(**to_pass))
-        model.fit(train_data, test_data, param_set["learning_rate"], iterations)
-        diagnostics = model.diagnostics
-        for param, val in param_set.items():
-            diagnostics[param] = val
-        diagnostics["_param_set"] = i
-        diagnostics_array.append(diagnostics)
-
-    diagnostics = pd.concat(diagnostics_array)
-
[…]

### 2024-09-30 mirrorverse/log_odds_model/tests/test_odds_model.py

@@ -1,150 +1,0 @@
-"""
-Odds Model Tests
-"""
-
-# pylint: disable=missing-function-docstring, invalid-name, missing-class-docstring
-
-import unittest
-
-import pandas as pd
-import numpy as np
-from sklearn.ensemble import RandomForestRegressor
-
-from mirrorverse.log_odds_model.log_odds_model import (
-    get_central_likelihood,
-    get_probability,
-    get_proposed_log_odds,
-    LogOddsModel,
-)
-
-
-def test_get_central_likelihood():
-    X = pd.DataFrame(
-        {
-            "_selected": [False, True, True, False],
-            "probability": [0.1, 0.2, 0.3, 0.4],
-        }
-    )
-    likelihood = get_central_likelihood(X)
-    assert likelihood == np.exp((np.log(0.2) + np.log(0.3)) / 2)
-
-
-def test_get_probability():
-    X = pd.DataFrame(
-        {
-            "_decision": [0, 1, 1, 2],
-            "log_odds": [np.log(1), np.log(2), np.log(3), np.log(4)],
-        }
-    )
-    get_probability(X)
-    print(X["probability"] - [1, 2 / 5, 3 / 5, 1])
-    assert np.abs(X["probability"] - [1.0, 2 / 5, 3 / 5, 1.0]).max() < 10**-10
-    assert np.abs(X["sum_odds"] - [1, 5, 5, 4]).max() < 10**-10
-
-
-def test_get_proposed_log_odds():
-    X = pd.DataFrame(
-        {
-            "_decision": [0, 1, 1, 2],
-            "_selected": [True, False, True, True],
-            "log_odds": [0, 1, 2, 3],
-            "odds": [1, 2, 3, 4],
-            "sum_odds": [1, 5, 5, 4],
-            "probability": [1, 2 / 5, 3 / 5, 1],
-        }
-    )
-    learning_rate = 0.9
-    get_proposed_log_odds(X, learning_rate)
-    assert (
-        abs(X["partial"] - [0, -1 / 5 * 2, 1 / 5 * 5 / 3 * (1 - 3 / 5) * 3, 0]).max()
-        < 10**-10
-    )
-    assert (X["step"] == X["partial"] * learning_rate).all()
-    assert (X["proposed"] == X["log_odds"] + X["step"]).all()
-
-
-class TestOddsModel(unittest.TestCase):
-
-    def setUp(self):
-        self.X_train = pd.concat(
-            [
-                pd.DataFrame(  # 75% of the time 0 is selected
-                    {
[…]

### 2024-09-30 mirrorverse/models/chinook_depth/api.py

@@ -1,87 +1,0 @@
-"""
-API for the Chinook Depth Model
-"""
-
-# pylint: disable=eval-used
-
-import click
-
-from mirrorverse.models.chinook_depth.warehouse import (
-    load_depth_data,
-    add_depth_classes,
-    load_context_data,
-    join_in_context_data,
-)
-from mirrorverse.models.chinook_depth.time import add_time_features
-from mirrorverse.models.chinook_depth.choices import fill_out_choices, downsample_data
-from mirrorverse.models.chinook_depth.model import (
-    split_data,
-    train_model,
-    do_search,
-)
-from mirrorverse.models.chinook_depth.grouping import merge_back
-
-
-@click.command()
-@click.option("--function", required=True, type=str)
-@click.option("--output_file", type=str)
-@click.option("--output_files", type=str)
-@click.option("--input_file", type=str)
-@click.option("--input_files", type=str)
-@click.option("--depth_classes", type=str)
-@click.option("--train_fraction", type=float)
-@click.option("--features", type=str)
-@click.option("--learning_rate", type=float)
-@click.option("--iterations", type=int)
-@click.option("--size_per_individual", type=int)
-@click.option("--num_param_sets", type=int)
-def main(**kwargs):
-    """
-    Just a Router
-    """
-    functions = {
-        "load_depth_data": (load_depth_data, ["output_file"]),
-        "add_depth_classes": (
-            add_depth_classes,
-            ["input_file", "depth_classes", "output_file"],
-        ),
-        "load_context_data": (
-            load_context_data,
-            ["output_file"],
-        ),
-        "join_in_context_data": (
-            join_in_context_data,
-            ["input_files", "output_file"],
-        ),
-        "add_time_features": (
-            add_time_features,
-            ["input_file", "output_file"],
-        ),
-        "fill_out_choices": (
-            fill_out_choices,
-            ["input_file", "output_file", "depth_classes"],
-        ),
-        "split_data": (
-            split_data,
-            ["input_file", "output_file", "train_fraction"],
-        ),
-        "train_model": (
-            train_model,
-            ["input_files", "output_files", "features", "iterations"],
-        ),
-        "downsample_data": (
-            downsample_data,
-            ["input_file", "output_file", "size_per_individual"],
-        ),
-        "do_search": (
-            do_search,
-            ["input_files", "output_files", "features", "iterations", "num_param_sets"],
-        ),
-        "merge_back": (
-            merge_back,
[…]

### 2024-09-30 mirrorverse/models/chinook_depth/choices.py

@@ -1,62 +1,0 @@
-"""
-Creates Full Array of Choices
-"""
-
-# pylint: disable=eval-used
-
-import pandas as pd
-import numpy as np
-
-
-def downsample_data(input_file, output_file, size_per_individual):
-    """
-    Inputs:
-    - input_file: str, path to the input file
-    - output_file: str, path to save the output file
-    - size_per_individual: int, number of samples to take per individual
-
-    Downsamples the data and saves it to a csv file.
-    """
-    data = pd.read_csv(input_file)
-    print("Current Size: ", len(data))
-    num_ids = len(data["tag_key"].unique())
-    print("Expected Size: ", num_ids * size_per_individual)
-    for i, _identifier in enumerate(data["tag_key"].unique()):
-        df = data[data["tag_key"] == _identifier]
-        df = df.sample(size_per_individual, replace=True)
-        if i == 0:
-            df.to_csv(output_file, index=False)
-        else:
-            df.to_csv(output_file, mode="a", header=False, index=False)
-
-
-def fill_out_choices(input_file, output_file, depth_classes):
-    """
-    Inputs:
-    - input_file: str, path to the input file
-    - output_file: str, path to save the output file
-    - depth_classes: str, string of depth classes as an array
-
-    Fills out the choices with all possible depth classes and saves it to a csv file.
-    """
-    depth_classes = np.array(eval(depth_classes))
-    choices = pd.read_csv(input_file)
-    choices = (
-        choices.reset_index(drop=True)
-        .reset_index()
-        .rename(
-            {
-                "index": "_decision",
-                "depth_class": "selected_class",
-                "tag_key": "_identifier",
-            },
-            axis=1,
-        )
-    )
-    all_choices = choices[["_decision"]].merge(
-        pd.DataFrame({"depth_class": depth_classes}), how="cross"
-    )
-    choices = choices.merge(all_choices, how="outer", on="_decision")
-    choices["_selected"] = choices["depth_class"] == choices["selected_class"]
-    del choices["selected_class"]
-    choices.to_csv(output_file, index=False)

### 2024-09-30 mirrorverse/models/chinook_depth/keras.py

@@ -1,103 +1,0 @@
-"""
-Keras Code for Chinook Depth Model
-"""
-
-# pylint: disable=invalid-name, import-error, no-name-in-module
-
-import os
-
-from functools import partial
-import tensorflow as tf
-import numpy as np
-from tensorflow.keras.utils import to_categorical
-from tensorflow.keras.models import Model
-from tensorflow.keras.layers import Dense, Input, concatenate
-
-
-# pylint: disable=redefined-builtin
-def split_data(N, features, data):
-    """
-    Inputs:
-    - N: int, number of choices
-    - features: list of strings, names of features
-    - data: dict, data loaded from tfrecord file
-
-    Splits the data into inputs and labels for the model
-    """
-    inputs = {}
-    for i in range(N):
-        input = tf.stack(
-            [tf.cast(data[f"{feature}_{i}"], tf.float32) for feature in features]
-        )
-        inputs[f"input_{i}"] = input
-    label = to_categorical(data["_selected"], num_classes=N)
-    return inputs, label
-
-
-def load_data(data_dir, N, features, batch_size, shuffle_buffer_size):
-    """
-    Inputs:
-    - data_dir: str, path to directory containing tfrecord files
-    - N: int, number of choices
-    - features: list of strings, names of features
-    - batch_size: int, batch size
-    - shuffle_buffer_size: int, size of buffer for shuffling data
-
-    Returns a tf.data.Dataset object containing the data
-    """
-    feature_description = {
-        "_selected": tf.io.FixedLenFeature([], tf.int64),
-    }
-    for i in range(N):
-        for feature in features:
-            feature_description[f"{feature}_{i}"] = tf.io.FixedLenFeature(
-                [], tf.float32
-            )
-
-    def _parse_function(proto):
-        return tf.io.parse_single_example(proto, feature_description)
-
-    tfrecord_files = []
-    for path in os.listdir(data_dir):
-        if path.endswith(".tfrecord"):
-            tfrecord_files.append(os.path.join(data_dir, path))
-
-    data = tf.data.TFRecordDataset(tfrecord_files)
-    data = data.map(_parse_function)
-    data = data.map(partial(split_data, N, features))
-    data = data.shuffle(buffer_size=shuffle_buffer_size)
-    data = data.batch(batch_size=batch_size)
-    data = data.prefetch(buffer_size=tf.data.experimental.AUTOTUNE)
-    return data
-
-
-def build_model(N, features, layers, final_activation="linear"):
-    """
[…]

### 2024-09-30 mirrorverse/models/chinook_depth/spark.py

@@ -1,701 +1,0 @@
-"""
-Spark Code for Chinook Depth Model
-"""
-
-# pylint: disable=invalid-name, import-error, no-name-in-module, no-member
-
-import os
-import uuid
-import shutil
-import json
-from functools import partial
-
-import numpy as np
-import pandas as pd
-import tensorflow as tf
-import h3
-from scipy.stats import norm
-from pyspark.sql.types import (
-    StructType,
-    StructField,
-    FloatType,
-    IntegerType,
-    TimestampType,
-    LongType,
-)
-
-# pylint: disable=redefined-outer-name
-from pyspark.sql.functions import udf, pandas_udf, lit, avg, std
-from suntimes import SunTimes
-
-
-def load_from_database(spark, connection, sql, schema):
-    """
-    Inputs:
-    - spark: SparkSession
-    - connection: str, connection string
-    - sql: str, SQL query
-    - schema: str, schema string
-    """
-    return (
-        spark.read.format("jdbc")
-        .option("url", connection)
-        .option("query", sql)
-        .option("customSchema", schema)
-        .load()
-    )
-
-
-def update_schema(dataframe, new_fields):
-    """
-    Inputs:
-    - dataframe: DataFrame
-    - new_fields: list of StructFields
-
-    Returns a new schema with the new fields added to the original schema.
-    """
-    return StructType(dataframe.schema.fields + new_fields)
-
-
-def apply_to_create_columns(dataframe, func, input_cols, new_fields):
-    """
-    Inputs:
-    - dataframe: DataFrame
-    - func: function
-    - input_cols: list of str, input columns
-    - new_fields: list of StructFields
-
-    Returns a new DataFrame with the new fields created by applying
-    the function to the input columns.
-    """
-    args = [dataframe[col] for col in input_cols]
-    if len(new_fields) > 1:
-        func = udf(func, StructType(new_fields))
-        dataframe = dataframe.withColumn("_tmp_struct", func(*args))
-        for field in new_fields:
-            dataframe = dataframe.withColumn(
-                field.name, dataframe["_tmp_struct"][field.name]
-            )
-        dataframe = dataframe.drop("_tmp_struct")
-    else:
-        func = udf(func, new_fields[0].dataType)
-        dataframe = dataframe.withColumn(new_fields[0].name, func(*args))
-    return dataframe
-
-
-def vectorized_create_column(dataframe, pandas_udf, input_cols, field):
-    """
-    Inputs:
[…]

### 2024-09-30 mirrorverse/models/chinook_depth/warehouse.py

@@ -1,133 +1,0 @@
-"""
-Loading and Prepping Data for the Chinook Depth Model
-"""
-
-# pylint: disable=eval-used
-
-import pandas as pd
-import numpy as np
-from scipy.stats import norm
-
-from mirrorverse.warehouse.utils import get_engine
-
-
-def load_depth_data(output_file):
-    """
-    Inputs:
-    - output_file: str, path to save the output file
-
-    Loads the depth data from the warehouse and saves it to a csv file.
-    """
-    sql = """
-    select 
-        tag_key,
-        date_key,
-        depth,
-        epoch
-    from 
-        tag_depths
-    """
-    depth = pd.read_sql(sql, get_engine())
-    depth = depth[~np.isnan(depth["depth"])]
-    depth["tag_key"] = "tk" + depth["tag_key"].astype(str)
-    depth.to_csv(output_file, index=False)
-
-
-def select_a_class(depth, depth_classes):
-    """
-    Inputs:
-    - depth: float, the depth of the fish as recorded
-    - depth_classes: np.array, the depth classes to choose from
-
-    Outputs:
-    - int, the selected depth class
-
-    Selects a depth class based on the depth of the fish.
-
-    It turns out that PSAT summary data bins the depth into
-    intervals so the actual depth is not known. However
-    given the recorded depth we can estimate the depth classes
-    it could belong to and the likelihoods of each.
-    """
-    sd = (
-        depth * 0.08 / 1.96
-    )  # ~two standard deviations gives our 95% confidence interval
-    if sd == 0:
-        division = np.zeros(len(depth_classes))
-        division[0] = 1
-    else:
-        # we're going to assume the depth classes are sorted
-        z = (depth_classes - depth) / sd
-        division = norm.cdf(z)
-        division[1:] = division[1:] - division[:-1]
-    # if there aren't quite enough depth classes the
-    # probabilities may not sum to 1, so we'll normalize
-    division = division / division.sum()
-    return np.random.choice(depth_classes, p=division)
-
-
-def add_depth_classes(input_file, depth_classes, output_file):
-    """
-    Inputs:
-    - input_file: str, path to the input file
-    - depth_classes: str, string of depth classes as an array
-    - output_file: str, path to save the output file
-
-    Adds a depth class to the depth data and saves it to a csv file.
-    """
-    depth_data = pd.read_csv(input_file)
-    depth_classes = np.array(eval(depth_classes))
[…]

### 2024-09-30 mirrorverse/tests/test_grouper.py

@@ -1,94 +1,0 @@
-"""
-Grouper Tests
-"""
-
-# pylint: disable=missing-function-docstring
-
-import pandas as pd
-import numpy as np
-
-from mirrorverse.grouper import (
-    Individual,
-    create_individual,
-    mutate,
-    crossover,
-    get_central_likelihoods,
-    evaluate,
-)
-
-
-def test_clone():
-    individual = Individual(np.zeros(10000))
-    individual.fitness = 0.1
-    clone = individual.clone()
-    assert len(clone) == 10000
-    assert clone.fitness is None
-    assert individual.fitness == 0.1
-
-
-def test_create_individual():
-    identifiers = [1, 2, 3, 4, 5]
-    groups = [1, 2, 3]
-    individual = create_individual(identifiers, groups)
-    assert len(individual) == len(identifiers)
-    assert set(individual) - set(groups) == set()
-
-
-def test_mutate():
-    individual = Individual(np.zeros(10000))
-    individual.fitness = 0.1
-    groups = [0, 1]
-    mutation_rate = 0.2
-    mutate(individual, groups, mutation_rate)
-    assert len(individual) == 10000
-    assert set(individual) - set(groups) == set()
-    # note that half of the mutations will be
-    # still zeros
-    assert abs(np.mean(individual) - 0.1) < 0.01
-    assert individual.fitness is None
-
-
-def test_crossover():
-    ind1 = Individual(np.zeros(10000))
-    ind1.fitness = 0.1
-    ind2 = Individual(np.ones(10000))
-    ind2.fitness = 0.2
-    crossover_rate = 0.2
-    crossover(ind1, ind2, crossover_rate)
-    assert len(ind1) == 10000
-    assert len(ind2) == 10000
-    assert set(ind1) - set([0, 1]) == set()
-    assert set(ind2) - set([0, 1]) == set()
-    assert abs(np.mean(ind1) - 0.2) < 0.02
-    assert abs(np.mean(ind2) - 0.8) < 0.02
-    assert ind1.fitness is None
-    assert ind2.fitness is None
-
-
-def test_get_central_likelihoods():
-    groups = [1, 1, 2, 2, 3]
-    identifiers = [1, 2, 3, 3, 4]
-    data = {
-        "_identifier": [1, 2, 3, 3, 4],
-        "_selected": [False, True, False, True, True],
[…]

### 2024-09-30 mirrorverse/tests/test_utility.py

@@ -1,182 +1,0 @@
-"""
-Tests for utility functions.
-"""
-
-# pylint: disable=missing-function-docstring, missing-class-docstring, invalid-name
-
-import pandas as pd
-from pandas.testing import assert_frame_equal
-from sklearn.metrics import mean_absolute_error
-from sklearn.ensemble import RandomForestRegressor
-
-from mirrorverse.utility import get_proposed_utility, train_utility_model
-
-
-def test_get_proposed_utility():
-    dataframe = pd.DataFrame(
-        {
-            "_decision": [0, 0, 1, 2, 2, 3, 3, 4],
-            "selected": [0, 1, 1, 1, 0, 1, 0, 1],
-            "utility": [3, 1, 3, 3, 1, 3, 1, 1],
-        }
-    )
-    results = get_proposed_utility(dataframe)
-    expected = pd.DataFrame(
-        {
-            "_decision": [0, 0, 1, 2, 2, 3, 3, 4],
-            "selected": [0, 1, 1, 1, 0, 1, 0, 1],
-            "utility": [3, 1, 3, 3, 1, 3, 1, 1],
-            "sum_utility": [4, 4, 3, 4, 4, 4, 4, 1],
-            "probability": [0.75, 0.25, 1.0, 0.75, 0.25, 0.75, 0.25, 1.0],
-            "score": [-0.75, 0.75, 0.0, 0.25, -0.25, 0.25, -0.25, 0.0],
-            "proposed": [0.75, 1.75, 3.0, 3.75, 0.75, 3.75, 0.75, 1.0],
-        }
-    )
-    assert_frame_equal(results, expected)
-
-
-class FakeModel:
-    def __init__(self):
-        self.model = None
-
-    def fit(self, X, y):
-        X = X.copy()
-        X["target"] = y
-        self.model = X.groupby("feature").agg({"target": "mean"}).reset_index()
-
-    def predict(self, X):
-        return X.merge(self.model, on="feature", how="left")["target"]
-
-
-def test_train_utility_model_1():
-    dataframe = pd.DataFrame(
-        {
-            "_decision": [0, 0, 1, 2, 2, 3, 3, 4],
-            "selected": [0, 1, 1, 1, 0, 1, 0, 1],
-            "utility": [3, 1, 3, 3, 1, 3, 1, 1],
-            "feature": [
-                "pizza",
-                "sandwich",
-                "pizza",
-                "pizza",
-                "sandwich",
-                "pizza",
[…]

### 2024-09-30 mirrorverse/tree.py

@@ -1,262 +1,0 @@
-"""
-The Base Tree Class
-"""
-
-import os
-import json
-
-import pandas as pd
-import numpy as np
-
-from sklearn.ensemble import RandomForestRegressor
-from sklearn.model_selection import GridSearchCV
-from sklearn.metrics import explained_variance_score
-
-from mirrorverse.utility import train_utility_model, get_proposed_utility
-
-
-# pylint: disable=no-member, invalid-name, attribute-defined-outside-init
-class DecisionTree:
-    """
-    This class is not complete and requires subclassing
-    that implements the following:
-
-    ATTRIBUTES:
-    - BUILDERS: list of choice builders
-    - BRANCHES: dictionary of DecisionTrees that will
-        be called based on the choice made
-    - FEATURE_COLUMNS: list of columns that will be used
-        by the model to predict the utility of choices
-    - OUTCOMES: a list of keys that will be added
-        to the choice_state to track the outcome of the
-        choice made
-    - PARAM_GRID: dictionary of hyperparameters to be
-        searched over in the grid search
-    - CV: KFold object to use in the grid search
-
-    STATIC METHODS:
-    - get_identifier(choice) - given the chosen vector,
-        return a unique identifier for the branch to be
-        called next (indexes into BRANCHES)
-    - _stich_selection(choices, selection) - given the
-        choices built from a state and the selection passed
-        to _build_model_data, return a dataframe with a
-        column "selected" that is 1 for the choice selected
-        and 0 for all other choices
-    - update_branch(self, choice, choice_state) - given the
-        choice made and the choice_state, update the choice_state
-        to reflect the choice made. This choice state will be
-        passed to the next branch.
-    """
-
-    def __init__(self, enrichment):
-        self.builders = [builder(enrichment) for builder in self.BUILDERS]
-        self.branches = {
-            identifier: branch(enrichment)
-            for identifier, branch in self.BRANCHES.items()
-        }
-
-    def get_choices(self, state, choice_state):
-        """
-        Input:
-        - state (dict): dictionary of state variables
-        - choice_state (dict): dictionary of choice state variables
-
-        Builds the full set of choices given the
-        state, choice_state, and builders
-        """
-        choices = []
-        for builder in self.builders:
-            choices += [builder(state, choice_state)]
-        return pd.concat(choices).reset_index(drop=True)
-
-    def choose(self, state, choice_state):
-        """
-        Input:
-        - state (dict): dictionary of state variables
[…]

### 2024-09-30 mirrorverse/utils.py

@@ -1,26 +1,0 @@
-"""
-Mirrorverse utility functions
-"""
-
-import os
-import pickle
-import click
-
-
-@click.command()
-@click.option("--model_dir", "-m", help="Directory to save models to")
-def bundle_models(model_dir):
-    """
-    Input:
-    - model_dir (str): directory to pull models from and
-        save models to
-    """
-    model_export = {}
-    for file_path in os.listdir(model_dir):
-        if file_path.endswith(".pkl"):
-            with open(f"{model_dir}/{file_path}", "rb") as fh:
-                sub_model_export = pickle.load(fh)
-                model_export.update(sub_model_export)
-    model_export_path = f"{model_dir}/models.pkl"
-    with open(model_export_path, "wb") as fh:
-        pickle.dump(model_export, fh)

### 2024-09-30 mirrorverse/warehouse/api.py

@@ -1,39 +1,0 @@
-"""
-To keep commands.py from changing everytime we add some new table.
-"""
-
-from functools import partial
-
-from mirrorverse.warehouse.etls.dimensions.dates import build_dates
-from mirrorverse.warehouse.etls.dimensions.spatial import build_spatial
-from mirrorverse.warehouse.etls.facts.cwt import format_cwt_recoveries_data
-from mirrorverse.warehouse.etls.dimensions.cwt import (
-    build_cwt_locations,
-    build_cwt_tags,
-)
-from mirrorverse.warehouse.etls.dimensions.tags import build_tags
-from mirrorverse.warehouse.etls.facts.tags import (
-    format_tag_tracks,
-    format_home_regions,
-    format_tag_depths,
-)
-from mirrorverse.warehouse.etls.facts.elevation import format_elevation
-from mirrorverse.warehouse.etls.facts.surface_temperature import (
-    format_surface_temperature,
-)
-
-DIMENSION_FORMATTERS = {
-    "dates": build_dates,
-    "cwt_locations": build_cwt_locations,
-    "cwt_tags": build_cwt_tags,
-    "h3_level_4": partial(build_spatial, 4),
-    "tags": build_tags,
-}
-FACT_FORMATTERS = {
-    "cwt_recoveries": format_cwt_recoveries_data,
-    "tag_tracks": format_tag_tracks,
-    "elevation": format_elevation,
-    "surface_temperature": format_surface_temperature,
-    "home_regions": format_home_regions,
-    "tag_depths": format_tag_depths,
-}

### 2024-09-30 mirrorverse/warehouse/commands.py

@@ -1,99 +1,0 @@
-"""
-Click Commands
-"""
-
-import json
-from time import time
-
-import click
-import pandas as pd
-from sqlalchemy.orm import Session
-from eralchemy2 import render_er
-
-from mirrorverse.warehouse.etls.missing_dimensions import get_primary_key
-from mirrorverse.warehouse.models import ModelBase
-from mirrorverse.warehouse.utils import upload_dataframe, get_engine
-from mirrorverse.warehouse.api import FACT_FORMATTERS, DIMENSION_FORMATTERS
-
-
-MODEL_KEY = {model.__tablename__: model for model in ModelBase.__subclasses__()}
-
-
-@click.command()
-@click.option("--table", "-t", help="The table to upload to", required=True)
-@click.option("--file_path", "-f", help="Path to the raw data", required=True)
-@click.option("--output_path", "-o", help="Path to the output data", required=True)
-def upload_facts(table, file_path, output_path):
-    """
-    Format and upload the data.
-    """
-    dataframe = pd.read_csv(file_path)
-    formatted = FACT_FORMATTERS[table](dataframe)
-
-    model = MODEL_KEY[table]
-    session = Session(get_engine())
-    upload_dataframe(session, model, formatted)
-    session.close()
-
-    status = {
-        "status": "success",
-        "timestamp": time(),
-    }
-    # pylint: disable=unspecified-encoding
-    with open(output_path, "w") as fh:
-        json.dump(status, fh)
-
-
-@click.command()
-@click.option("--table", "-t", help="The table to upload to", required=True)
-@click.option(
-    "--missing_dimensions_path",
-    "-m",
-    help="Path to the missing dimensions data",
-    required=True,
-)
-@click.option("--file_path", "-f", help="Path to the raw data", required=False)
-@click.option("--output_path", "-o", help="Path to the output data", required=True)
-def upload_dimensions(table, missing_dimensions_path, file_path, output_path):
-    """
-    Build the missing dimensions for a given fact table.
-    """
-    model = MODEL_KEY[table]
-    primary_key = get_primary_key(model)
-
-    # pylint: disable=unspecified-encoding
-    with open(missing_dimensions_path, "r") as fh:
-        missing_dimensions = json.load(fh)
-
-    missing_keys = missing_dimensions[primary_key]
-    build_func = DIMENSION_FORMATTERS[table]
-
-    if file_path:
-        dataframe = pd.read_csv(file_path)
[…]

### 2024-09-30 mirrorverse/warehouse/data/elevation.py

@@ -1,25 +1,0 @@
-"""
-Fake Elevation Data for Testing
-"""
-
-import pandas as pd
-
-ELEVATION_DATA = pd.DataFrame(
-    [
-        {
-            "latitude": 46.6,
-            "longitude": -133.2,
-            "elevation": 100,
-        },
-        {
-            "latitude": 46.6,
-            "longitude": -133.2,
-            "elevation": 200,
-        },
-        {
-            "latitude": 56.6,
-            "longitude": -133.2,
-            "elevation": 100,
-        },
-    ]
-)

### 2024-09-30 mirrorverse/warehouse/data/surface_temperature.py

@@ -1,36 +1,0 @@
-"""
-Fake Surface Temperature Data for Testing
-"""
-
-from datetime import datetime
-
-import pandas as pd
-
-SURFACE_TEMPERATURE_DATA = pd.DataFrame(
-    [
-        {
-            "latitude": 46.6,
-            "longitude": -133.2,
-            "temperature_c": 5,
-            "date": datetime(2021, 1, 1),
-        },
-        {
-            "latitude": 46.6,
-            "longitude": -133.2,
-            "temperature_c": 10,
-            "date": datetime(2021, 1, 1),
-        },
-        {
-            "latitude": 56.6,
-            "longitude": -133.2,
-            "temperature_c": 5,
-            "date": datetime(2021, 1, 1),
-        },
-        {
-            "latitude": 56.6,
-            "longitude": -133.2,
-            "temperature_c": 10,
-            "date": datetime(2021, 2, 1),
-        },
-    ]
-)

### 2024-09-30 mirrorverse/warehouse/data/tags.py

@@ -1,116 +1,0 @@
-"""
-Fake Tag Data for Testing
-"""
-
-import pandas as pd
-import numpy as np
-
-TAGS_DATA = pd.DataFrame(
-    [
-        {
-            "Ptt": "205415",
-            "tag.model": "MiniPAT",
-            "time.series.resolution.min": 10.0,
-            "Region": "Kodiak",
-            "fork.length.cm": 80.0,
-            "deploy.date.GMT": "2020-10-13 20:17",
-            "deploy.latitude": 55.3,
-            "deploy.longitude": -151.2,
-            "end.date.time.GMT": "2021-04-19 03:00",
-            "hypothetical.data.retrieved": 0.658,
-            "data.type": "Transmitted",
-            "End.Latitude": 45.7,
-            "End.Longitude": -124.6,
-        },
-        {
-            "Ptt": "142100",
-            "tag.model": "Xtag",
-            "time.series.resolution.min": 15.0,
-            "Region": "Dutch Harbor",
-            "fork.length.cm": 75.0,
-            "deploy.date.GMT": "2015-12-02 19:15",
-            "deploy.latitude": 59.3,
-            "deploy.longitude": -166.6,
-            "end.date.time.GMT": "2016-01-22 03:45",
-            "hypothetical.data.retrieved": 0.827,
-            "data.type": "Transmitted",
-            "End.Latitude": 59.5,
-            "End.Longitude": -168.4,
-        },
-        {
-            "Ptt": "239204",
-            "tag.model": "MiniPAT",
-            "time.series.resolution.min": 5.0,
-            "Region": "Craig",
-            "fork.length.cm": 75.0,
-            "deploy.date.GMT": "2022-05-31 16:00",
-            "deploy.latitude": 55.6,
-            "deploy.longitude": -134.3,
-            "end.date.time.GMT": "2022-07-31 01:35",
-            "hypothetical.data.retrieved": 0.836,
-            "data.type": "Transmitted",
-            "End.Latitude": 51.6,
-            "End.Longitude": -130.9,
-        },
-    ]
-)
-
-TAG_TRACKS_DATA = pd.DataFrame(
-    [
-        {
-            "Ptt": "211761",
-            "Date": "2021-05-01",
-            "Most.Likely.Latitude": 46.6,
-            "Most.Likely.Longitude": -133.2,
-        },
-        {
[…]

### 2024-09-30 mirrorverse/warehouse/etls/dimensions/cwt.py

@@ -1,66 +1,0 @@
-"""
-CWT Dimensions ETLs
-"""
-
-import numpy as np
-
-from mirrorverse.warehouse.etls.dimensions.spatial import add_spatial_keys_to_facts
-from mirrorverse.warehouse.models import CWTLocations, CWTTags
-
-
-def build_cwt_locations(missing_keys, raw_data):
-    """
-    Input:
-    - missing_keys (list): A list of missing location keys
-    - raw_data (pd.DataFrame): Raw data from the CWT Website
-
-    Returns a pd.DataFrame
-    """
-
-    dataframe = raw_data.rename(
-        {
-            "location_code": "cwt_location_key",
-            "name": "cwt_location_name",
-            "rmis_latitude": "latitude",
-            "rmis_longitude": "longitude",
-        },
-        axis=1,
-    )
-
-    assert dataframe.shape[0] == dataframe.drop_duplicates("cwt_location_key").shape[0]
-
-    dataframe = dataframe[dataframe["cwt_location_key"].isin(missing_keys)]
-
-    dataframe = dataframe[
-        ~np.isnan(dataframe["latitude"]) & ~np.isnan(dataframe["longitude"])
-    ]
-
-    dataframe = add_spatial_keys_to_facts(dataframe)
-
-    columns = [column.key for column in CWTLocations.__table__.columns]
-    return dataframe[columns]
-
-
-def build_cwt_tags(missing_keys, raw_data):
-    """
-    Input:
-    - missing_keys (list): A list of missing tag keys
-    - raw_data (pd.DataFrame): Raw data from the CWT Website
-
-    Returns a pd.DataFrame
-    """
-
-    dataframe = raw_data.rename(
-        {
-            "tag_code_or_release_id": "cwt_tag_key",
-            "release_location_code": "cwt_release_location_key",
-        },
-        axis=1,
-    )
-
-    assert dataframe.shape[0] == dataframe.drop_duplicates("cwt_tag_key").shape[0]
-
-    dataframe = dataframe[dataframe["cwt_tag_key"].isin(missing_keys)]
-
-    columns = [column.key for column in CWTTags.__table__.columns]
-    return dataframe[columns]

### 2024-09-30 mirrorverse/warehouse/etls/dimensions/tests/test_tags.py

@@ -1,50 +1,0 @@
-"""
-Tests for Tags ETLs
-"""
-
-# pylint: disable=missing-function-docstring, missing-class-docstring, duplicate-code
-
-import pandas as pd
-from pandas.testing import assert_frame_equal
-
-from mirrorverse.warehouse.etls.dimensions.tags import build_tags
-from mirrorverse.warehouse.data.tags import TAGS_DATA
-
-
-def test_build_tags():
-    missing_keys = ["205415", "239204"]
-    results = build_tags(missing_keys, TAGS_DATA)
-    expected = pd.DataFrame(
-        [
-            {
-                "tag_key": "205415",
-                "tag_model": "MiniPAT",
-                "time_resolution_min": 10.0,
-                "fork_length_cm": 80.0,
-                "deploy_date_key": 1602547200,
-                "deploy_latitude": 55.3,
-                "deploy_longitude": -151.2,
-                "deploy_h3_level_4_key": 594997496145510399,
-                "end_date_key": 1618790400,
-                "end_latitude": 45.7,
-                "end_longitude": -124.6,
-                "end_h3_level_4_key": 595195451188183039,
-            },
-            {
-                "tag_key": "239204",
-                "tag_model": "MiniPAT",
-                "time_resolution_min": 5.0,
-                "fork_length_cm": 75.0,
-                "deploy_date_key": 1653955200,
-                "deploy_latitude": 55.6,
-                "deploy_longitude": -134.3,
-                "deploy_h3_level_4_key": 594804342876274687,
-                "end_date_key": 1659225600,
-                "end_latitude": 51.6,
-                "end_longitude": -130.9,
-                "end_h3_level_4_key": 594801946284523519,
-            },
-        ]
-    )
-    assert set(results.columns) == set(expected.columns)
-    assert_frame_equal(results[expected.columns].reset_index(drop=True), expected)

### 2024-09-30 mirrorverse/warehouse/etls/facts/cwt.py

@@ -1,39 +1,0 @@
-"""
-ETLs for CWT Fact Data
-"""
-
-import pandas as pd
-
-from mirrorverse.warehouse.etls.dimensions.dates import add_date_keys_to_facts
-from mirrorverse.warehouse.models.facts import CWTRecoveries
-
-
-def format_cwt_recoveries_data(raw_data):
-    """
-    Input:
-    - raw_data (pd.DataFrame): Raw data from the CWT Website
-    """
-
-    # some basic renaming
-    dataframe = raw_data.rename(
-        {
-            "recovery_location_code": "cwt_recovery_location_key",
-            "species": "species_key",
-            "reporting_agency": "cwt_reporting_agency_key",
-            "number_cwt_estimated": "number_estimated",
-            "tag_code": "cwt_tag_key",
-        },
-        axis=1,
-    )
-
-    # deal with their very weird date column
-    dataframe["valid_date"] = dataframe["recovery_date"].apply(lambda x: x > 10000000)
-    dataframe = dataframe[dataframe["valid_date"]]
-    dataframe["recovery_date"] = pd.to_datetime(
-        dataframe["recovery_date"].astype(str), format="%Y%m%d"
-    )
-    dataframe = add_date_keys_to_facts(dataframe, "recovery_date", "recovery_date_key")
-
-    # filter down to the columns in our table
-    columns = [column.key for column in CWTRecoveries.__table__.columns]
-    return dataframe[columns]

### 2024-09-30 mirrorverse/warehouse/etls/tests/test_missing_dimensions.py

@@ -1,67 +1,0 @@
-"""
-Test Missing Dimensions ETLs
-"""
-
-# pylint: disable=missing-function-docstring, duplicate-code, missing-class-docstring
-
-import unittest
-
-from sqlalchemy.orm import Session
-
-from mirrorverse.warehouse.etls.missing_dimensions import (
-    get_primary_key,
-    get_associated_dimensions,
-    get_missing_dimension_keys,
-)
-from mirrorverse.warehouse.etls.facts.cwt import format_cwt_recoveries_data
-from mirrorverse.warehouse.models import (
-    ModelBase,
-    CWTRecoveries,
-    Dates,
-    CWTLocations,
-    CWTTags,
-)
-from mirrorverse.warehouse.data.cwt import CWT_RETRIEVAL_DATA
-from mirrorverse.warehouse.utils import get_engine, upload_dataframe
-
-
-def test_get_primary_key():
-    assert get_primary_key(Dates) == "date_key"
-
-
-def test_get_associated_dimensions():
-    keys, models = get_associated_dimensions(CWTRecoveries)
-    assert set(keys) == set(
-        [
-            "cwt_tag_key",
-            "recovery_date_key",
-            "cwt_recovery_location_key",
-        ]
-    )
-    assert set(models) == set(
-        [
-            CWTTags,
-            Dates,
-            CWTLocations,
-        ]
-    )
-
-
-class TestMissingDimensions(unittest.TestCase):
-
-    def setUp(self):
-        engine = get_engine(db_url="sqlite://")
-        ModelBase.metadata.create_all(engine)
-        self.session = Session(engine)
-
-    def tearDown(self):
-        self.session.rollback()
-        self.session.close()
-
-    def test_get_missing_dimension_keys(self):
-        formatted = format_cwt_recoveries_data(CWT_RETRIEVAL_DATA)
-        upload_dataframe(self.session, CWTRecoveries, formatted)
-        missing_keys = get_missing_dimension_keys(
-            Dates, CWTRecoveries, "recovery_date_key", self.session
-        )
-        assert missing_keys == {"date_key": [1609459200]}

### 2024-09-30 mirrorverse/warehouse/models/__init__.py

@@ -1,24 +1,0 @@
-"""
-Models!
-"""
-
-# pylint: disable=wrong-import-position, cyclic-import
-
-from sqlalchemy.orm import DeclarativeBase
-
-
-class ModelBase(DeclarativeBase):
-    """
-    The base to which all other models are attached
-    """
-
-
-from .dimensions import Dates, CWTLocations, CWTTags, H3Level4, Tags
-from .facts import (
-    CWTRecoveries,
-    TagTracks,
-    TagDepths,
-    Elevation,
-    SurfaceTemperature,
-    HomeRegions,
-)

### 2024-09-30 mirrorverse/warehouse/models/dimensions.py

@@ -1,91 +1,0 @@
-"""
-Dimension Tables
-"""
-
-from datetime import date
-
-from typing import Optional
-from sqlalchemy import String, ForeignKey
-from sqlalchemy.orm import Mapped, mapped_column
-
-from mirrorverse.warehouse.models import ModelBase
-
-
-class Dates(ModelBase):
-    """
-    Dates Dimension Table
-    """
-
-    __tablename__ = "dates"
-
-    date_key: Mapped[int] = mapped_column(primary_key=True)
-
-    date: Mapped[date]
-    day: Mapped[int]
-    month: Mapped[int]
-    year: Mapped[int]
-
-
-class CWTLocations(ModelBase):
-    """
-    CWT Locations Dimension Table
-    """
-
-    __tablename__ = "cwt_locations"
-
-    cwt_location_key: Mapped[str] = mapped_column(String(19), primary_key=True)
-
-    cwt_location_name: Mapped[str]
-    longitude: Mapped[float]
-    latitude: Mapped[float]
-    h3_level_4_key: Mapped[int] = mapped_column(ForeignKey("h3_level_4.h3_level_4_key"))
-
-
-class CWTTags(ModelBase):
-    """
-    CWT Tags Dimension Table
-    """
-
-    __tablename__ = "cwt_tags"
-
-    cwt_tag_key: Mapped[str] = mapped_column(String(10), primary_key=True)
-
-    cwt_release_location_key: Mapped[str] = mapped_column(
-        ForeignKey("cwt_locations.cwt_location_key")
-    )
-    run: Mapped[Optional[int]]
-
-
-class H3Level4(ModelBase):
-    """
-    H3 Level 4 Dimension Table
-    """
-
-    __tablename__ = "h3_level_4"
-
-    h3_level_4_key: Mapped[int] = mapped_column(primary_key=True)
-
-    geometry: Mapped[str]
-
-
-class Tags(ModelBase):
-    """
-    Tags Dimension Table
-    """
-
-    __tablename__ = "tags"
-
-    tag_key: Mapped[str] = mapped_column(primary_key=True)
-    tag_model: Mapped[str]
-    time_resolution_min: Mapped[int]
-    fork_length_cm: Mapped[float]
-    deploy_latitude: Mapped[float]
-    deploy_longitude: Mapped[float]
-    deploy_h3_level_4_key: Mapped[int] = mapped_column(
[…]

### 2024-09-30 mirrorverse/warehouse/tests/test_tags.py

@@ -1,174 +1,0 @@
-"""
-Tests of Integrated Tag Data
-"""
-
-# pylint: disable=missing-function-docstring, missing-class-docstring, duplicate-code
-
-import unittest
-
-import pandas as pd
-from pandas.testing import assert_frame_equal
-from sqlalchemy import select
-from sqlalchemy.orm import Session
-
-from mirrorverse.warehouse.models import ModelBase
-from mirrorverse.warehouse.utils import get_engine, upload_dataframe
-
-from mirrorverse.warehouse.data.tags import (
-    TAGS_DATA,
-    TAG_TRACKS_DATA,
-    HOME_REGIONS_DATA,
-    TAG_DEPTHS_DATA,
-)
-from mirrorverse.warehouse.models import Tags, TagTracks, HomeRegions, TagDepths
-from mirrorverse.warehouse.etls.dimensions.tags import (
-    build_tags,
-)
-from mirrorverse.warehouse.etls.facts.tags import (
-    format_tag_tracks,
-    format_home_regions,
-    format_tag_depths,
-)
-
-
-class TestTags(unittest.TestCase):
-
-    def setUp(self):
-        engine = get_engine(db_url="sqlite://")
-        ModelBase.metadata.create_all(engine)
-        self.session = Session(engine)
-
-    def tearDown(self):
-        self.session.rollback()
-        self.session.close()
-
-    def test_upload_tags(self):
-        missing_keys = ["205415", "239204"]
-        formatted = build_tags(missing_keys, TAGS_DATA)
-        upload_dataframe(self.session, Tags, formatted)
-        stmt = select(Tags)
-        results = pd.read_sql_query(stmt, self.session.bind)
-        expected = pd.DataFrame(
-            [
-                {
-                    "tag_key": "205415",
-                    "tag_model": "MiniPAT",
-                    "time_resolution_min": 10,
-                    "fork_length_cm": 80.0,
-                    "deploy_date_key": 1602547200,
-                    "deploy_latitude": 55.3,
-                    "deploy_longitude": -151.2,
-                    "deploy_h3_level_4_key": 594997496145510399,
-                    "end_date_key": 1618790400,
-                    "end_latitude": 45.7,
-                    "end_longitude": -124.6,
-                    "end_h3_level_4_key": 595195451188183039,
-                },
-                {
-                    "tag_key": "239204",
-                    "tag_model": "MiniPAT",
-                    "time_resolution_min": 5,
[…]

### 2024-09-30 rebuild_db.sh

@@ -1,44 +1,0 @@
-#!/bin/bash
-
-export DATABASE_URL="sqlite:////workspaces/mirrorverse/mirrorverse.db"
-
-# Drop the database
-rm mirrorverse.db
-
-# build the fact tables
-dvc repro -Rf pipelines/upload/tag_tracks/
-dvc repro -Rf pipelines/upload/home_regions/
-dvc repro -Rf pipelines/upload/surface_temperature/
-dvc repro -Rf pipelines/upload/elevation/
-
-# build the tags table
-echo "missing_dimensions: ../tag_tracks/missing_dimensions.json" > pipelines/upload/tags/params.yaml
-echo "data: data/HMM.Inventory_CSV_Marcel_2.12.2024.csv" >> pipelines/upload/tags/params.yaml
-dvc repro -R pipelines/upload/tags/
-
-echo "missing_dimensions: ../home_regions/missing_dimensions.json" > pipelines/upload/tags/params.yaml
-echo "data: data/HMM.Inventory_CSV_Marcel_2.12.2024.csv" >> pipelines/upload/tags/params.yaml
-dvc repro -R pipelines/upload/tags/
-
-# build the spatial table
-echo "missing_dimensions: ../tag_tracks/missing_dimensions.json" > pipelines/upload/spatial/params.yaml
-dvc repro -R pipelines/upload/spatial/
-
-echo "missing_dimensions: ../tags/missing_dimensions.json" > pipelines/upload/spatial/params.yaml
-dvc repro -R pipelines/upload/spatial/
-
-echo "missing_dimensions: ../surface_temperature/missing_dimensions.json" > pipelines/upload/spatial/params.yaml
-dvc repro -R pipelines/upload/spatial/
-
-echo "missing_dimensions: ../elevation/missing_dimensions.json" > pipelines/upload/spatial/params.yaml
-dvc repro -R pipelines/upload/spatial/
-
-# build the dates table
-echo "missing_dimensions: ../tag_tracks/missing_dimensions.json" > pipelines/upload/dates/params.yaml
-dvc repro -R pipelines/upload/dates/
-
-echo "missing_dimensions: ../tags/missing_dimensions.json" > pipelines/upload/dates/params.yaml
-dvc repro -R pipelines/upload/dates/
-
-echo "missing_dimensions: ../surface_temperature/missing_dimensions.json" > pipelines/upload/dates/params.yaml
-dvc repro -R pipelines/upload/dates/

### 2024-09-30 models/depth_model/glue/depth_features.py
depth model

@@ -1,0 +1,50 @@
+import sys
+from awsglue.transforms import *
+from awsglue.utils import getResolvedOptions
+from pyspark.context import SparkContext
+from awsglue.context import GlueContext
+from awsglue.job import Job
+import add_depth_class
+from awsglue import DynamicFrame
+
+def sparkSqlQuery(glueContext, query, mapping, transformation_ctx) -> DynamicFrame:
+    for alias, frame in mapping.items():
+        frame.toDF().createOrReplaceTempView(alias)
+    result = spark.sql(query)
+    return DynamicFrame.fromDF(result, glueContext, transformation_ctx)
+args = getResolvedOptions(sys.argv, ['JOB_NAME'])
+sc = SparkContext()
+glueContext = GlueContext(sc)
+spark = glueContext.spark_session
+job = Job(glueContext)
+job.init(args['JOB_NAME'], args)
+
+# Script generated for node AWS Glue Data Catalog
+AWSGlueDataCatalog_node1725837638799 = glueContext.create_dynamic_frame.from_catalog(database="haven", table_name="mgietzmann_tag_depths", transformation_ctx="AWSGlueDataCatalog_node1725837638799")
+
+# Script generated for node SQL Query
+SqlQuery0 = '''
+select 
+    tag_key,
+    epoch - epoch % 300 as epoch,
+    avg(depth) as depth,
+    avg(temperature) as temperature
+from 
+    myDataSource 
+where 
+    depth is not null 
+    and temperature is not null
+group by 
+    1, 2
+'''
+SQLQuery_node1725838239173 = sparkSqlQuery(glueContext, query = SqlQuery0, mapping = {"myDataSource":AWSGlueDataCatalog_node1725837638799}, transformation_ctx = "SQLQuery_node1725838239173")
+
+# Script generated for node Add Depth Class
+AddDepthClass_node1725837653823 = SQLQuery_node1725838239173.add_depth_class(depth_classes="25,50,75,100,150,200,250,300,400,500", depth_column="depth")
+
+# Script generated for node Amazon S3
+AmazonS3_node1725837948039 = glueContext.getSink(path="s3://haven-database/tag_depth_features_mk1/", connection_type="s3", updateBehavior="LOG", partitionKeys=["tag_key"], enableUpdateCatalog=True, transformation_ctx="AmazonS3_node1725837948039")
[…]

### 2024-09-30 models/depth_model/hypercube/read_hypercube.py

@@ -1,0 +1,33 @@
+sql = """
+with base as (
+    select 
+        _decision,
+        _choice,
+        depth_class * std_depth_class + avg_depth_class as depth_class,
+        period_progress * std_period_progress + avg_period_progress as period_progress,
+        elevation * std_elevation + avg_elevation as elevation,
+        month * std_month + avg_month as month,
+        daytime
+    from 
+        depth_model_hypercube_mk3
+), probs as (
+    select 
+        _decision,
+        _choice,
+        probability
+    from
+        depth_model_mk3_hypercube_inference
+)
+select 
+    base.*,
+    probs.probability as probability
+from
+    base
+    inner join probs 
+        on base._decision = probs._decision 
+        and base._choice = probs._choice
+
+"""
+hypercube = db.read_data(sql)
+print(hypercube.shape)
+hypercube.head()

### 2024-09-30 models/depth_model/results/build_grid.py
grid

@@ -1,0 +1,9 @@
+import plotly.express as px
+
+df = hypercube[hypercube['period_progress'] == 0.5]
+df['month'] = df['month'].astype(int)
+fig = px.line(
+    df, x='elevation', y='probability', color='depth_class', facet_row='daytime', facet_col='month',
+    width=3000, height=500, log_x=True
+)
+fig.show()

### 2024-09-30 docs/proposal/Proposal.tex

--- 2024-09-26 proposal/Proposal.tex
@@
 
-\title{Demonstration of a Standardized Methodology for Building and Using Individual Movement Models using Chinook Salmon Pop-up Archival Tagging Data from the Gulf of Alaska}
+\title{Demonstrating a Standardized IBM Methodology using Chinook Salmon PSAT Data from the Gulf of Alaska}
 
@@
 \begin{objective}
-Movement modeling is fundamentally about modeling the selection of movement outcomes.
+\textbf{Standardized Decision Models} Movement (and more generally decision) modeling is fundamentally about modeling the selection of outcomes.
 So, given a set of choices $v_k$ and a set of features $n_k$ describing those choices, we want a standardized method to fit, evaluate, and use models of the form:
@@
 \begin{objective}
-To provide tooling that allows for running spatio-temporal simulations of management strategies using movement models as defined above. Such simulations should allow for:
+\textbf{Standardized Decision Simulations} Given an individual in a specific situation, the decision models described above produce a set of probabilities for each potential next outcome $v_k$.
+However there is nothing preventing us from taking each of those outcomes, considering them each a new situation, and applying the models again.
+Carried out repeatedly we will end up with a probabilistic diffusion of the original individual across its decision space.
+Furthermore, if we can do this for one individual we can do it for an entire distribution of individuals.
+In sum, using this diffusion modeling approach we can simulate the movement through a decision space of an entire population.
 
-\begin{enumerate}
-\item Flexible initial assumptions that capture heterogeneity in the stock.
-\item The ability to impose spatio-temporal mortality (both fishing and natural).
-\item The ability to capture heterogeneity in fishing vulnerability.
-\end{enumerate}
+Our second objective then is to build such a diffusion model.
+However, we would like to do it in such a way that life history, mortality, and environmental models can be provided as input as well.
+This diffusion model will then allow us to explore the spatio-temporal behavior of a stock under both nominal conditions as well as under various exploratory management strategies.
 
-Likewise these simulations need to be quick enough to allow for capturing uncertainty in initial conditions through bootstrapping.\newline
-
-These simulations will be known as diffusion models as they model the probabilistic behavior of cohorts as opposed to individual fish or schools. 
 \end{objective}
@@
 
-\textbf{Target:} Discrete depth ranges. \newline
+\textbf{Choices:} Discrete depth ranges. \newline
 
@@
 
-\textbf{Target:} Spatial grid cell selection as movement. Also if clearly different modes of movement are found (migration vs foraging for example) we'll predict the mode as well. \newline
+\textbf{Choices:} Spatial grid cell selection as movement. Also if clearly different modes of movement are found (migration vs foraging for example) we'll predict the mode as well. \newline
 
@@
 
-In building these models our primary purpose is to demonstrate the tooling exists to prudently build odds models.
+In building these models our primary purpose is to demonstrate the tooling exists to prudently build movement models.
 We can frame this goal in terms of a "standard" machine learning pipeline and work-flow.
@@
 \item Training and Hyperparameter Tuning Using Standard Neural Net Architectures
-\item Diagnoses of Fit and Comparison of Different Model Types
+\item Diagnoses of Fit and Comparison of Different Models
 \item Identifying Sample Groups that are More or Less Fit
+\item Exploring Discovered Patterns
 \end{enumerate}
@@
 
-The outcome will be a package in Python for building, diagnosing, and running odds models.
-Note that at this point the purpose is to demonstrate how to build odds models for the purpose of building simulations.
+The outcome will be a package in Python for building, diagnosing, and running movement models.
+Note that at this point the purpose is to demonstrate how to build movement models for the purpose of building simulations.
 It will be in the next chapter that specific research outcomes directly relevant to management will manifest themselves.
-The models' sole purpose will be in powering the simulations.
 
@@
 
-Chapter 2 will be concerned with taking these models for Chinook salmon in the Gulf of Alaska and building a spatio-temporal movement simulation that can be used to investigate a series of management questions.
-It is worth noting that we will use the simulation to ask questions that may come up in general management settings even if they may not be directly applicable to the issues currently faced by Chinook salmon management in the GOA.
-This is just to demonstrate general applicability.
+Chapter 2 will be concerned with taking these movement models for Chinook salmon in the Gulf of Alaska and building a diffusion model.
+We will then demonstrate its capabilities by exploring a series of management questions.
+It is worth noting that these questions have been chosen based on their facility in demonstrating the aspects of the simulation as opposed to their priority to current Chinook salmon management in the GOA.
 \newline
 
-The overview of the components of this simulation are outlined in Fig. \ref{fig:simulation}.
+The overview of the components of this diffusion model are outlined in Fig. \ref{fig:simulation}.
 
@@
 	Boxes with rounded edges would be inputs specified by the user.
-The hexagons are functions 		provided by the simulation and the trapezoids are data generated during the main loop.
-It is important to note that this design is anticipatory, not all mortality or environmental data may be available at this time in which case spatially homogeneous inputs would be provided.
+The hexagons are functions 		provided by the diffusion model and the trapezoids are data generated during the main loop.
+It is important to note that in this design we are planning for the most data rich case.
+However, depending on the application, mortality or environmental data may in fact have quite limited availability.
   \label{fig:simulation}
@@
 
-Some specific use cases we want to demonstrate with this simulation are:
+The specific questions we would like to answer with this simulation are:
 
@@
 To give a specific example of what this might look like in practice, if we were interested in understanding how to protect one specific genetic cohort by creating fishing closures we would:
-
-\begin{enumerate}
-\item Input a distribution of the cohort at some initial time point as relative abundance per spatial grid cell.
-\item Input a time period over which to run the simulation (perhaps the fishing season in the area under study).
-\item Collect and input any environmental data needed for the movement model indexed by space and time.
-\item Then we might run the simulation without any fishing pressure to understand how the cohort naturally distributes itself and generate hypotheses for closures from this.
-\item We could then input fishing pressure (as a mortality function) by space and time for a few different scenarios where some would include our closures.
-\item Run the simulation for each of these scenarios and compare resulting abundance trends. 
-\end{enumerate}
 

### 2024-12-17 models/movement_model/move_it_move_it/model/layers.py
2024-12-16

@@ -4,11 +4,15 @@ LAYERS = {
     "Dropout1": lambda: Dropout(0.1),
     "Dropout2": lambda: Dropout(0.2),
     "Dropout3": lambda: Dropout(0.3),
+    "Dropout4": lambda: Dropout(0.4),
+    "Dropout5": lambda: Dropout(0.5),
+    "Dropout6": lambda: Dropout(0.6),
     "D2": lambda: Dense(2, activation='relu'),
     "D4": lambda: Dense(4, activation='relu'),
     "D8": lambda: Dense(8, activation='relu'),
     "D16": lambda: Dense(16, activation='relu'),
     "D24": lambda: Dense(24, activation='relu'),
     "D32": lambda: Dense(32, activation='relu'),
+    "D48": lambda: Dense(48, activation='relu'),
     "D64": lambda: Dense(64, activation='relu'),
 }

### 2024-12-17 models/movement_model/move_it_move_it/model/sample_models.py

@@ -32,30 +32,25 @@ def build_randomized_param_sets(param_grids, M, max_attempts):
 @click.option('-n', '--num_sets', required=True, type=int)
 def main(num_sets):
     grids = {
-        "batch_size": [500, 600, 700],
-        "epochs": [15, 20, 25],
-        "layer1": [8, 16, 16],
-        "layer2": [8, 8, 16, 16, 16],
-        "layer3": [4, 8, 16, 0, 0, 0],
+        "batch_size": [500],
+        "random_seed": list(range(5)),
+        "epochs": [50],
+        "dropout": [2, 3],
+        "num_layers": [2, 3, 4],
+        "layer_size": [16, 24]
     }
-    grids = {
-        "batch_size": [500, 600, 700],
-        "epochs": [125],
-        "dropout": [0, 1, 2, 3],
-        "layer_size": [8, 16, 24, 32]
-    }
-    param_sets = build_randomized_param_sets(grids, num_sets, 10)
+    param_sets = build_randomized_param_sets(grids, num_sets, 100)
     for param_set in param_sets:
         dropout = param_set["dropout"]
         layer_size = param_set["layer_size"]
         if dropout == 0:
             param_set["layers"] = [
-                f"D{layer_size}", f"D{layer_size}", f"D{layer_size}"
-            ]
+                f"D{layer_size}",
+            ] * param_set["num_layers"]
         else:
             param_set["layers"] = [
-                f"D{layer_size}", f"Dropout{dropout}", f"D{layer_size}", f"Dropout{dropout}", f"D{layer_size}", f"Dropout{dropout}"
-            ]
+                f"D{layer_size}", f"Dropout{dropout}",
+            ] * param_set["num_layers"]
 
     print(param_sets)
     with open("param_sets.json", "w") as f:

### 2024-12-18 docs/papers/log_odds/Paper.tex
wip

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[super]{nth}
\usepackage{amsthm}
\usepackage{bm}
\newtheorem{theorem}{Theorem}
\newtheorem{objective}{Objective}
\newtheorem{model}{Model}
\usepackage{xcolor}
\definecolor{light-gray}{gray}{0.95}
\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}
\usepackage{listings}
\usepackage{placeins}

\usepackage[authoryear]{natbib}

\makeatletter
\renewcommand{\maketitle}{
\begin{center}

\pagestyle{empty}

{\LARGE \bf \@title\par}
\vspace{1cm}

{\Large Marcel Gietzmann-Sanders, Michael Courtney, Andrew Seitz, Curry Cunningham}\\[1cm]

University of Alaska Fairbanks

\end{center}
}\makeatother

\title{Technique for Dimensionality Reduction in Decision Modeling}

\date{2024}
\setcounter{tocdepth}{2}
\begin{document}
\maketitle

\section*{Introduction}

\subsection*{Value of Probabilistic Modeling}

\section*{Theory}

Standard probabilistic deep learning networks are typically framed as a classification problem, using categorical cross-entropy as the loss function (\cite{durr}). Each output neuron represents a potential choice, with the model predicting the probability of each choice being correct based on this loss formulation. For these choices, we provide the network with features encapsulating the relevant information. Training is then comprised of providing a series such decisions.

However, this formulation introduces a critical challenge: if there are $N$ features per choice and $M$ potential choices, the overall dimensionality of the input space becomes $N \cdot M$. Adding even a single feature increases the dimensionality by $M$ not just 1.

This growth poses a significant challenge due to the "curse of dimensionality", where the amount of data required to effectively train models grows exponentially with the dimensionality of the input space (CITE).

\subsection*{Log-Odds Modeling}

To address this issue, we propose an alternative framing. Instead of predicting the probabilities directly, we predict the log-odds $\phi_m$ for each choice and calculate the probability $p_m$ using the softmax function:

$$p_m = \frac{e^{\phi_m}}{\sum_{m=1}^{M}e^{\phi_m}}$$

This approach reduces the feature space dimensionality to $N$ and effectively increases the number of training examples by a factor of $M$.

We can implement this log-odds model using standard probabilistic deep learning techniques by replicating the log-odds computation across all $M$ choices. The outputs are fed into a softmax layer with $M$ units, where the layer's weights are set to the identity matrix and biases are set to zero. Using categorical cross-entropy as the loss function ensures compatibility with standard probabilistic deep learning while enabling us to train the log-odds weights and significantly reduce the problem's dimensionality.

\subsection*{Contrast Sampling}

As $M$ grows large, a practical issue arises: for each training example, most instances of the internal log-odds model would ideally report very low log-odds, resulting in low probabilities. Ideally, only one choice should produce $p_m=1$. This is analogous to a class imbalance problem, where the model becomes prone to predicting the most common class (CITE).

To address this, we balance the training data. Instead of presenting the model with full decisions containing all $M$ choices, we create training pairs, or contrasts, where each pair consists of one selected choice and one unselected choice. This approach is valid because the log-odds model focuses on the relative likelihood of choices, making the number of choices considered at any one time irrelevant.

The primary risk in using contrasts is introducing bias by disproportionately sampling certain combinations of choices. To mitigate this, we randomly sample pairs from each decision, ensure an equal number of contrasts per decision, and an equal number of decisions per individual. This preserves the balance across the training data and avoids skewing the model's predictions.

\section*{Application}

We consider the tracks of 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) recorded using

\subsection*{Data}

\subsection*{Model}

\subsection*{Training}

\section*{Results}

\section*{Discussion}

\bibliographystyle{apalike}
\bibliography{reference}

\end{document}

### 2024-12-20 docs/papers/log_odds/Paper.tex
wip

@@
 
-We consider the tracks of 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) recorded using
-
 \subsection*{Data}
 
-\subsection*{Model}
+We consider a series of tracks from 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) caught and monitored between 2013 and 2022 (CITE).
+These tracks were obtained from pop-up satellite archival tags which collect temperature, light level, and depth information at specified (sub day) intervals.
+This data is then passed through a proprietary algorithm from Wildlife Computers to determine likely longitude and latitude during each day of of monitoring (\cite{PSAT}).
+\newline
+
+Environmental data was derived from the Global Ocean Biogeochemistry Hindcast dataset (10.48670/moi-00019) and the Global Ocean Physics Reanalysis (10.48670/moi-00021) from the E.U.
+Copernicus Marine Service Information.
+Net primary production (mg/m3/day) and mixed layer thickness (m) were aggregated per Uber h3 resolution 4 cell in the Northern Pacific.
+
+\subsection*{Formulation}
+
+The resolution 4 Uber h3 cell containing each salmon location was identified and then, assuming a maximum travel distance of 100km (centeroid to centeroid) all adjacent cells within the 100km were identified as choices (including the currently occupied cell).
+In general this represented $\sim 19$ choices per decision with the intention being to predict the probability of moving to any particular cell.
+Training data was derived by identifying the actual cell moved to.
+
+\subsection*{Features}
+
+Movement heading in radians and distance to the centroid of the choice cell were computed and then mixed layer thickness and net primary production were joined to the choices on cell and day.
+
+Distance was normalized to a range of 0-1 by division by 100, while mixed layer thickness and net primary production were both log-scaled and then centered at zero.
+
+\subsection*{Contrast Sampling}
+
+After inspecting the distribution of number of choices per salmon and number of choices per decision, we decided on random sampling (with replacement) 200 decisions per individual and 19 choices per decision.
+
+Over a test/validation/training split of 20, 20, 71 this resulted in 345,800 contrasts of which 269,800 were used in training and the rest in validation (no contrasts were necessary for the test set).
+
+Note that only 14,200 training examples would've been available to a traditional probabilistic approach representing a large increase in the number of available training examples.
 
 \subsection*{Training}
+
+Three sets models were trained, one including only distance, one with both distance and movement heading, and one with all four features.
+Note that while the feature dimensions of these models are 1, 2, and 4 respectively, given the maximum number of choices per decision seen was 33 they dimensionality of a standard probabilistic model would've been 33, 66, and 132 representing a large reduction in the dimensionality of our feature spaces.
+
+Architectures/hyperparameters for the log-odds component of the model were parametrized in the following ways:
+
+\begin{center}
+\begin{tabular}{| c | c |} 
+\hline 
+Component & Options \\
+\hline
+Layers & 2, 3, 4 \\ 
+Units per Layer & 16, 24, 32 \\
+Dropout \% per Layer & 20, 30 \\ 
+Batch Size & 500 \\
+\hline
+\end{tabular}
+\end{center}
+
+With 10 models trained for each combination. Models were trained in Keras using an Adam optimizer with default settings.
+
+Models for each set of features we selected on the basis of the loss over the validation set of contrasts.
 

### 2024-12-20 models/movement_model/move_it_move_it/local_training/layers.py

@@ -1,0 +1,18 @@
+from tensorflow.keras.layers import Dense, Dropout
+
+LAYERS = {
+    "Dropout1": lambda: Dropout(0.1),
+    "Dropout2": lambda: Dropout(0.2),
+    "Dropout3": lambda: Dropout(0.3),
+    "Dropout4": lambda: Dropout(0.4),
+    "Dropout5": lambda: Dropout(0.5),
+    "Dropout6": lambda: Dropout(0.6),
+    "D2": lambda: Dense(2, activation='relu'),
+    "D4": lambda: Dense(4, activation='relu'),
+    "D8": lambda: Dense(8, activation='relu'),
+    "D16": lambda: Dense(16, activation='relu'),
+    "D24": lambda: Dense(24, activation='relu'),
+    "D32": lambda: Dense(32, activation='relu'),
+    "D48": lambda: Dense(48, activation='relu'),
+    "D64": lambda: Dense(64, activation='relu'),
+}

### 2024-12-20 models/movement_model/move_it_move_it/model/grid_search.py

@@ -1,0 +1,57 @@
+import click
+import json
+import numpy as np
+from copy import copy
+
+def build_gridded_param_sets(param_grids):
+    """
+    Inputs:
+    - param_grids: dict, parameter names to grid of values
+    - M: int, number of parameter sets to generate
+    - max_attempts: int, maximum number of attempts to generate a unique parameter set
+
+    Outputs:
+    - list of dicts, parameter sets
+    """
+    param_sets = [{}]
+    for param, grid in param_grids.items():
+        new_param_sets = []
+        for el in grid:
+            for set in param_sets:
+                new_set = copy(set)
+                new_set[param] = el
+                new_param_sets.append(new_set)
+        param_sets = new_param_sets
+
+    return param_sets
+
+@click.command()
+def main():
+    grids = {
+        "batch_size": [500],
+        "random_seed": list(range(10)),
+        "epochs": [100],
+        "dropout": [2, 3],
+        "num_layers": [2, 3, 4],
+        "layer_size": [16, 24, 32]
+    }
+    param_sets = build_gridded_param_sets(grids)
+    for param_set in param_sets:
+        dropout = param_set["dropout"]
+        layer_size = param_set["layer_size"]
+        if dropout == 0:
+            param_set["layers"] = [
+                f"D{layer_size}",
+            ] * param_set["num_layers"]
+        else:
+            param_set["layers"] = [
+                f"D{layer_size}", f"Dropout{dropout}",
+            ] * param_set["num_layers"]
+
+    print(param_sets)
+    print(len(param_sets))
+    with open("param_sets.json", "w") as f:
+        json.dump(param_sets, f, indent=4, sort_keys=True)
+
+if __name__ == '__main__':
+    main()

### 2024-12-20 models/movement_model/move_it_move_it/model/grid_search.py
finished for the week

@@ -28,12 +28,13 @@ def build_gridded_param_sets(param_grids):
 @click.command()
 def main():
     grids = {
-        "batch_size": [500],
-        "random_seed": list(range(10)),
+        "batch_size": [10000],
+        "random_seed": list(range(5)),
         "epochs": [100],
-        "dropout": [2, 3],
-        "num_layers": [2, 3, 4],
-        "layer_size": [16, 24, 32]
+        "dropout": [0],
+        "num_layers": [3, 4],
+        "layer_size": [24, 32],
+        "learning_rate": [0.0005]
     }
     param_sets = build_gridded_param_sets(grids)
     for param_set in param_sets:
@@ -47,6 +48,8 @@ def main():
             param_set["layers"] = [
                 f"D{layer_size}", f"Dropout{dropout}",
             ] * param_set["num_layers"]
+        param_set["optimizer"] = "Adam"
+        param_set["optimizer_kwargs"] = {"learning_rate": param_set["learning_rate"]}
 
     print(param_sets)
     print(len(param_sets))

### 2024-12-23 docs/papers/log_odds/Paper.tex
results

@@
 
-Over a test/validation/training split of 20, 20, 71 this resulted in 345,800 contrasts of which 269,800 were used in training and the rest in validation (no contrasts were necessary for the test set).
+Over a test/validation/training split of 40, 71 this resulted in 345,800 contrasts of which 269,800 were used in training and the rest in validation.
 
@@
 \hline
-Layers & 2, 3, 4 \\ 
-Units per Layer & 16, 24, 32 \\
-Dropout \% per Layer & 20, 30 \\ 
-Batch Size & 500 \\
+Layers & 3, 4 \\ 
+Units per Layer & 24, 32 \\
+Batch Size & 10000 \\
+Learning Rate & 0.0005 \\
 \hline
@@
 
-With 10 models trained for each combination. Models were trained in Keras using an Adam optimizer with default settings.
+With 5 models trained for each combination. Models were trained in Keras using an Adam optimizer for 100 epochs.
 
@@
 
+In training the metric of interest is the contrasts normalized log probability (C-NLP) - the average log probability per contrast.
+For an equivalent metric over all of the decisions we also computed the average log probability per decision for each individual and then computed an average over those across individuals (in order to not favor individuals with many decisions).
+This is the D-NLP reported in the table below.
+
+\begin{center}
+\begin{tabular}{| c | c | c | c | c |}
+\hline 
+Model & Train C-NLP & Val C-NLP & Train D-NLP & Val D-NLP \\
+\hline
+No Model & -0.693 & -0.693 & -2.944 & -2.944 \\
+Model 1 & -0.172 & -0.154 & -1.336 & -1.223 \\
+Model 2 & -0.156 & -0.150 & -1.281 & -1.200 \\
+Model 3 & -0.147 & -0.146 & -1.248 & -1.180 \\
+\hline
+\end{tabular}
+\end{center}
+
+"No Model" assumes all decisions are equally likely, Model 1 is the distance only model, Model 2 adds the movement heading, and Model 3 adds the net productivity and mixed layer thickness features.
+\newline
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[height=70mm]{figures/ll_increase.png}
+  \caption{Log likelihood increase (unnormalized) per individual of each model change.}
+  \label{fig:ll_increase}
+\end{figure}
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[height=70mm]{figures/ll_increase_no_movement.png}
+  \caption{Log likelihood increase (unnormalized) per individual of each model change in cases where the selection involved no movement.}
+  \label{fig:ll_increase_no_movement}
+\end{figure}
+
+\FloatBarrier
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[height=70mm]{figures/ll_increase_movement.png}
+  \caption{Log likelihood increase (unnormalized) per individual of each model change in cases where the selection involved movement.}
+  \label{fig:ll_increase_movement}
+\end{figure}
+
+Figure \ref{fig:ll_increase} shows the unnormalized changes in log likelihood for each individual colored by the region in which they were initially tagged as we go from one model to the next.
+Figures \ref{fig:ll_increase_no_movement} and \ref{fig:ll_increase_movement} break this down for decisions that resulted in no movement and movement respectively.
+Only individuals in the validation set were considered.
+
+The appendix contains a series of mapped examples (figures \ref{fig:chignik_map}-\ref{fig:yakutat_map}) from each of these regions (besides EBS) that shows how the log likelihood per decision (when moving) changed in moving from model 1 to model 2 and then from model 2 to model 3.
+As in the above, only individuals in the validation set were considered.\newline
+
+Finally as much of the behavior seems to be modulated by movement distances (or lack of movement) a summary table of empirical likelihood of movement of a specific distance is given (taken over all individuals).
+
+\begin{center}
+\begin{tabular}{| c | c |}
+\hline
+Distance Bin (km) & Likelihood of Selection \\
+\hline
+No Movement & 65.6\% \\
+Up to 50km & 32.5\% \\
+50km to 100km & 1.9\% \\
+\hline
+
+\end{tabular}
+\end{center}
+
 \section*{Discussion}
@@
 \bibliography{reference}
+\newpage
+\section*{Appendix: Mapped Examples}
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[width=140mm]{figures/chignik_map.png}
+  \caption{Log likehood changes per decision - Chignik}
+  \label{fig:chignik_map}
+\end{figure}
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[width=140mm]{figures/kodiak_map.png}
+  \caption{Log likehood changes per decision - Kodiak}
+  \label{fig:kodiak_map}
+\end{figure}
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[width=140mm]{figures/nanwalek_map.png}
+  \caption{Log likehood changes per decision - Nanwalek}
+  \label{fig:nanwalek_map}
+\end{figure}
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[width=140mm]{figures/sitka_map.png}
+  \caption{Log likehood changes per decision - Sitka}
+  \label{fig:sitka_map}
+\end{figure}
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[width=140mm]{figures/unalaska_map.png}
+  \caption{Log likehood changes per decision - Unalaska}
+  \label{fig:unalaska_map}
+\end{figure}
+
+\begin{figure}[h!] 
+	\centering
+  \includegraphics[width=140mm]{figures/yakutat_map.png}
+  \caption{Log likehood changes per decision - Yakutat}
+  \label{fig:yakutat_map}
+\end{figure}
 

### 2024-12-24 docs/papers/log_odds/Paper.tex
wip

@@
 
-\title{Technique for Dimensionality Reduction in Decision Modeling}
+\title{A Framework for Studying Behavior Using Deep Learning}
 
@@
 \subsection*{Value of Probabilistic Modeling}
+
+\section*{Framing}
+
+This methodology centers on a two-stage cycle of discovery: (1) testing hypotheses about covariates that are believed to predict behaviors and (2) using the results to generate new hypotheses.
+
+In the first stage, we model behaviors as decisions made from a set of choices, each associated with hypothesized covariates.
+Then, using machine learning, we build a predictive model to evaluate the likelihood of these selections based on the covariates.
+
+In the second stage, we evaluate how the model shifts the likelihood of specific individuals or decision groups relative to baseline models.
+Improved performance reveals where the covariates more effectively predict behavior, while underperformance indicates missing covariates.
+This focus on areas of "greatest possible effect" provides case studies for generating new hypotheses, enabling iterative refinement of the process.
+
+The keystone of this entire process becomes having a clear, quick means of producing machine learning models from our framing of decisions and choices and our covariates, an issue which we turn to next.
 
@@
 
+\subsection*{Alternative Approaches to Reducing Dimensionality}
+There are various strategies for reducing the dimensionality of one's data.
+One straightforward method is to limit the choices available.
+For instance if making decisions among a set of directions, one could reduce the precision of the angles allowed.
+While this approach reduces the model's specificity, it significantly simplifies the dimensionality of the problem, which is often a key challenge in probabilistic deep learning.
+
+The methodology proposed, however, achieves dimensionality reduction without sacrificing specificity.
+That said, there are scenarios where some choices are unlikely to be relevant.
+In such cases, dimensions corresponding to these irrelevant choices can simply be discarded, further simplifying the model.
+\newline
+
+Another effective strategy to enhance the data set is leveraging the order invariance of choices in traditional probabilistic problem framing.
+Specifically, the order in which choices are presented to the model should not matter.
+For instance, whether a particular choice appears in the first or the thirteenth position should have no impact on the model's operation.
+This property allows for data augmentation by reordering choices.
+
+In essence, for each training example, $M!$ (factorial of the number of choices) augmented examples can be created.
+While this exponentially increases the size of the data set as the feature space grows, it provides a mechanism to generate additional training examples from the same data.
+
+The issue with this approach is that as your augmented data size grows to match the needs of the greater dimensionality so too does the time required for training.
+So while the problem remains theoretically possible, the exponential uptick in time complexity poses a significant practical issue.
+The proposed approach does not suffer from this issue and remains linear in time with the amount of data provided.
 \section*{Application}
@@
 \begin{center}
-\begin{tabular}{| c | c |}
+\begin{tabular}{| c | c | c |}
 \hline
-Distance Bin (km) & Likelihood of Selection \\
+Distance Bin (km) & Likelihood of Selection & \# Training Decisions \\
 \hline
-No Movement & 65.6\% \\
-Up to 50km & 32.5\% \\
-50km to 100km & 1.9\% \\
+No Movement & 65.6\% & 3163 \\
+Up to 50km & 32.5\% & 1567 \\
+50km to 100km & 1.9\% & 92 \\
 \hline
@@
 \section*{Discussion}
+
+\subsection*{Model One: Distance as a Key Feature}
+Starting with Model One, we observe a significant improvement in log-likelihood across all individuals in the validation set.
+This outcome is intuitive, given that the most probable action for these salmonids is to remain in place.
+This behavior highlights the importance of distance as a feature, as it dramatically reduces the likelihood of alternative actions.
+Beyond staying in place, most movement is limited to a range of 0 to 50 kilometers, with only a small percentage extending further.
+
+This demonstrates that a baseline model treating all locations as equally likely is a gross oversimplification of salmon behavior.
+Introducing distance as a feature alone vastly improves the model’s performance, as it aligns more closely with observed patterns.
+
+\subsection*{Model Two: Movement Heading}
+Model Two adds movement heading as a feature, offering further improvement, though not as dramatic as Model One's addition of distance.
+Movement heading enables the model to learn general directional trends in salmon migration, which are particularly influenced by the geography and location of the fish.
+
+Examining changes in log-likelihood between Models One and Two, we see a general pattern: southeasterly movements gain likelihood, while northern, northwesterly, and northeasterly movements often see decreases.
+This aligns with the data, as the farthest-moving fish predominantly migrate southeast.
+However, the inclusion of movement heading introduces variability—while most fish experience an increase in likelihood, some see decreases due to the model imposing a general directional preference that may not align with all individual behaviors.
+
+\subsection*{Model Three: Environmental Features}
+To refine the model further, we incorporate environmental features such as net primary productivity and mixed layer thickness in Model Three.
+These features provide information about productivity in the fish's habitat.
+Their inclusion results in another improvement in log-likelihood, particularly for fish previously associated with less-likely movements (e.g., westward or northward).
+This suggests a correlation between environmental productivity and movement behavior.
+
+However, as with movement heading, the addition of these features benefits some individuals while diminishing the likelihoods for others.
+This variability reflects the nuanced interactions between salmonid behavior and environmental conditions, reinforcing the importance of continuing to explore exemplar cases where the model performs poorly.
+
+\subsection*{Iterative Refinement Through Exemplars}
+At each stage, we identify individuals with the lowest log-likelihood improvements compared to their baselines.
+These exemplars provide valuable insights into potential missing covariates or unmodeled behaviors.
+By iteratively incorporating new features and examining their effects, we uncover diverse behavioral patterns and refine the model's predictive capability.
+
+Throughout this process, the value of contrast as a guide is evident.
+Improvements in normalized log-likelihood consistently align with better decision predictions, demonstrating that the model is effectively trained to assess the odds of observed behaviors.
+This contrast-driven approach allows for incremental learning and refinement, ensuring the model captures key patterns while maintaining flexibility to incorporate new information.
+
+This methodology proves powerful in understanding salmonid movement by enabling rapid integration of new features, establishment of revised baselines, and identification of exemplar behaviors for further analysis.
+By iteratively refining the model, we continue to discover and represent the complex ecological and behavioral dynamics underlying salmonid movement.
+
+\subsection*{Computational Performance and Efficiency}
+The computational performance of these log-odds models proved to be efficient and straightforward.
+All models were trained in the cloud using a two-core, 4GB machine, with training times ranging from 15 to 30 minutes over 100 epochs.
+The variation in training time depended on the complexity of the model and the number of features included.
+
+This efficiency scales well in a cloud environment, as parallelizing model training across multiple machines is both feasible and straightforward.
+By distributing the workload, all models for various subsets of the data can be trained simultaneously, maintaining the same 15 to 30-minute window for training and arriving at a new baseline.
+
+Within the proposed workflow—where a baseline model is built based on hypothesized important features, and poorly performing individuals are identified as exemplars for further investigation—the computational time spent on training is a minor component.
+The iterative process of identifying features, gathering new data, and refining the model dominates the time and energy expenditure.
+
+As such, the short training time ensures that computational efficiency is not a bottleneck, allowing for rapid iteration and refinement.
+
+Given the combination of short training times, scalability in cloud environments, and ease of integration into the iterative process, these log-odds models are highly efficient tools for data exploration and feature discovery.
+The author concludes that their use is well-suited to workflows requiring repeated refinement and baseline adjustments.
 

### 2024-12-25 docs/papers/log_odds/Paper.tex
wip

--- 2024-12-23 docs/papers/log_odds/Paper.tex
@@
 
-\title{Technique for Dimensionality Reduction in Decision Modeling}
+\title{A Framework for Studying Animal Behavior Using Deep Learning}
 
@@
 
+\begin{center}
+This study explores using deep learning to build Individual-Based Models (IBMs).
+Using Chinook salmon (\textit{Oncorhynchus tshawytscha}) PSAT data as a case study, we introduce a methodology that leverages deep learning to transform the construction of IBMs into a systematic tool for investigating behaviors and their covariates.
+By employing a "log-odds model" to reduce dimensionality, this approach positions modeling as a central driver of discovery, enabling efficient hypothesis testing and iterative refinement.
+\end{center}
+
 \section*{Introduction}
 
-\subsection*{Value of Probabilistic Modeling}
+Individual-Based Models (IBMs) have emerged as a critical tool for studying ecological systems, particularly for capturing the nuanced behaviors and interactions of individual organisms within their environments (CITE).
+With them, investigators can propose hypotheses on how fish behave and then test those hypotheses against actual individual behaviors as observed in the wild (CITE).
+
+However, current methods for building IBMs are largely bespoke (CITE), tailored to specific systems or questions, and often require extensive manual development and calibration.
+This lack of standardization poses significant challenges to their broader adoption as a foundational methodology for studying behavior (CITE).
+Each model requires unique assumptions, data integrations, and computational workflows, making it difficult to generalize findings across studies or species.
+Moreover, the high dimensionality and computational demands of these models—exacerbated by the need to simulate individual decisions across time and space—can limit their scalability and applicability in data-limited scenarios (CITE).
+
+Machine learning offers a way in which to at least partially address the bespoke nature of traditional Individual-Based Models (IBMs).
+By framing individual behaviors as probabilistic decisions among consecutive sets of choices each described by a series of features machine learning can be invoked to largely automate the process of building the decision "engine" behind IBM's.
+Machine learning excels at identifying patterns in high-dimensional data (CITE), enabling models to dynamically incorporate a wide range of environmental and behavioral covariates without requiring custom coding for every new hypothesis.
+Furthermore, by providing a consistent framing, the same systems and tools can be used across a variety of different problems.
+
+Beyond offering a common framework, reducing the time required to build IBMs allows investigators to focus on using the modeling process to quickly test hypotheses and identify areas of uncertainty early in their studies.
+This efficiency enhances the utility of IBMs, making them an even more powerful tool for advancing the understanding of complex behaviors.
+
+In this study we demonstrate how integrating machine learning with Individual-Based Models (IBMs) provides a standardized, efficient approach for testing hypotheses and refining models, enabling rapid exploration of behavior and uncertainty.
+
+\section*{Framing}
+
+This methodology centers on a two-stage cycle of discovery: (1) testing hypotheses about covariates that are believed to predict behaviors and (2) using the results to generate new hypotheses.
+
+In the first stage, we model behaviors as decisions made from a set of choices, each associated with hypothesized covariates.
+Then, using machine learning, we build a predictive model to evaluate the likelihood of these selections based on the covariates.
+
+In the second stage, we evaluate how the model shifts the likelihood of specific individuals or decision groups relative to baseline models.
+Improved performance reveals where the covariates more effectively predict behavior, while underperformance indicates missing covariates.
+This focus on areas of "greatest possible effect" provides case studies for generating new hypotheses, enabling iterative refinement of the process.
+
+The keystone of this entire process becomes having a clear, quick means of producing machine learning models from our framing of decisions and choices and our covariates, an issue which we turn to next.
 
@@
 
+\subsection*{Compute}
+
+Models were trained using AWS Batch using Fargate instances of 2 vcpu's and 4 GB of memory.
+By taking advantage of batch models could be all trained in parallel allowing for short (15-30 minute) turn around times.
+
 \section*{Results}
@@
 \section*{Discussion}
+
+\subsection*{Model One: Distance as a Key Feature}
+Given the observations that most decisions involve an individual staying put (within its h3 cell) and rarely do fish travel to an h3 cell further than 50km away, it was decided that a more reasonable baseline model would be one that included distance traveled as a feature (model 1).
+
+As expected model 1 resulted in a significant improvement in log-likelihood across all individuals in the validation set as compared to a model that gives even odds to each available choice.
+This model therefore provides the baseline for our subsequent models.
+
+\subsection*{Model Two: Movement Heading}
+Beyond this underlying distance feature the interest is in predicting, when fish move, where they will move.
+Of particular interest was the individuals in the sitka and yakutat groups as these groups had several individuals exhibiting what appeared to be very directed and extensive movement.
+
+Upon studying these individuals it was noted that changes in mixed layer thickness seemed to incite movement and the individuals tending to stay near higher primary productivity.
+Therefore a model with net primary productivity and mixed layer thickness was of interest.
+
+However before moving to this model we wanted a baseline model that included movement heading as a feature so that we could determine what aspects of the movement were predicted simply by common headings as opposed to movement specifically related to productivity.
+Model 2 is this distance plus movement heading model.
+
+Looking at figure \ref{fig:ll_increase} we can see that for several individuals in the yakutat group adding movement heading increased the log likelihood meaningfully as compared to model 1 (our baseline).
+Indeed figure \ref{fig:ll_increase_movement} indicates that most of this increase comes from decisions where the individuals moved.
+Looking at figure \ref{fig:yakutat_map} it is clear that the move from model 1 to model 2 is creating a model that's learning a heading south east is more likely than alternative directions with the movements of fish in north or westerly directions actually reducing in log likelihood as compared to the baseline.
+
+This southeasterly pattern is maintained throughout figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} and indeed looking at \ref{fig:ll_increase_movement} there are several groups where the majority of individuals suffer a decrease in log likelihood, in all likelihood due to this common direction not being particularly representative of them.
+
+The sitka group however does see an improvement as well showing that both of these majority south eastern movement groups (sitka and yakutat) can be at least partially explained by a common heading down the coast.
+
+\subsection*{Model Three: Environmental Features}
+With this naive heading baseline in place we can now investigate the extent to which adding in net primary productivity and mixed layer thickness allows our model to distinguish when different headings are taken by different fish, or in fact reinforce the headings already "selected".
+
+Looking at the step from model 2 to model 3 in figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} we see now that several of the movements that saw a drop in log likelihood as compared to the model 1 baseline are actually seeing a rise in log-likelihood under this new model indicating that the new features are allowing us to swing the preferred movement heading conditioned on the environment.
+
+In addition, especially in the sitka group, we see reinforcement/refinement of the decisions as there is an increase in log likelihood up and beyond what was received from model 2.
+That is to say that while there is a general tendency in the south easterly direction, the specific cells chosen are in some way correlated to productivity.
+
+Noteable as well is the fact that the move from model 2 to model 3 had a significant positive impact on the predictability of the decisions from the nanwalek group, much more so than in the step from model 1 to model 2.
+The kodiak group however remains largely unaffected.
+
+\subsection*{Further Questions}
+
+Looking at the cases in figure \ref{fig:ll_increase} in which we see significant drops in log likelihood or cases where there is no meaningful change we can see that while there is predictive power in these features chosen thus far - especially for the yakutat, sitka, and nanwalek groups - there are still plenty of individuals that can act as exemplars for searching for new behavioral drivers.
+
+For example in figure \ref{fig:chignik_map} even with the addition of the productivity features in model 3 the westerly movements are still poorly predicted and would warrant investigation.
+
+Likewise the unalaska group is just poorly explained in general beyond the baseline given by model 1 and would represent another case study for further investigation.
+Perhaps by training a model with the same set of features as model 3 but only over this group we would see a significant improvement in the performance of the model which may lead us to look for demographic or geophysical differences that could explain how two different groups of fish are "using" the same covariates so differently.
+
+Clearly though, the models we have allow us to identify and focus upon individuals that so far resist our attempts to explain them.
+
+\subsection*{Conclusion}
+
+This investigation demonstrates the power and flexibility of an iterative modeling approach for understanding behavior.
+By building baseline models and progressively refining them with additional covariates, we identified both predictive patterns and individuals resistant to explanation, which serve as focal points for further hypothesis generation.
+The efficiency of this process, enabled by quick training times and parallelization, ensures that refinement is limited only by the investigator's ability to propose and test new hypotheses.
+Crucially, at every stage, the models not only advance understanding but remain actionable for predictions, underscoring their dual utility as tools for both discovery and application.
 

### 2024-12-25 docs/papers/log_odds/Paper.tex
wip

@@
 
-Individual-Based Models (IBMs) have emerged as a critical tool for studying ecological systems, particularly for capturing the nuanced behaviors and interactions of individual organisms within their environments (CITE).
-With them, investigators can propose hypotheses on how fish behave and then test those hypotheses against actual individual behaviors as observed in the wild (CITE).
+Individual-Based Models (IBMs) have emerged as a critical tool for studying ecological systems, particularly for capturing the nuanced behaviors and interactions of individual organisms within their environments \citep{grimm}.
+With them, investigators can propose hypotheses on how fish behave and then test those hypotheses against actual individual behaviors as observed in the wild.
 
-However, current methods for building IBMs are largely bespoke (CITE), tailored to specific systems or questions, and often require extensive manual development and calibration.
-This lack of standardization poses significant challenges to their broader adoption as a foundational methodology for studying behavior (CITE).
+However, current methods for building IBMs are largely bespoke, tailored to specific systems or questions, and often require extensive manual development and calibration \citep{grimm}.
+This lack of standardization poses significant challenges to their broader adoption as a foundational methodology for studying behavior \citep{grimm}.
 Each model requires unique assumptions, data integrations, and computational workflows, making it difficult to generalize findings across studies or species.
-Moreover, the high dimensionality and computational demands of these models—exacerbated by the need to simulate individual decisions across time and space—can limit their scalability and applicability in data-limited scenarios (CITE).
 
@@
 By framing individual behaviors as probabilistic decisions among consecutive sets of choices each described by a series of features machine learning can be invoked to largely automate the process of building the decision "engine" behind IBM's.
-Machine learning excels at identifying patterns in high-dimensional data (CITE), enabling models to dynamically incorporate a wide range of environmental and behavioral covariates without requiring custom coding for every new hypothesis.
+Machine learning excels at identifying patterns in high-dimensional data, enabling models to dynamically incorporate a wide range of environmental and behavioral covariates without requiring custom coding for every new hypothesis.
 Furthermore, by providing a consistent framing, the same systems and tools can be used across a variety of different problems.
@@
 
-Standard probabilistic deep learning networks are typically framed as a classification problem, using categorical cross-entropy as the loss function (\cite{durr}).
+Standard probabilistic deep learning networks are typically framed as a classification problem, using categorical cross-entropy as the loss function \citep{durr}.
 Each output neuron represents a potential choice, with the model predicting the probability of each choice being correct based on this loss formulation.
@@
 
-This growth poses a significant challenge due to the "curse of dimensionality", where the amount of data required to effectively train models grows exponentially with the dimensionality of the input space (CITE).
+This growth poses a significant challenge due to the "curse of dimensionality", where the amount of data required to effectively train models grows exponentially with the dimensionality of the input space \citep{curse}.
 
@@
 Ideally, only one choice should produce $p_m=1$.
-This is analogous to a class imbalance problem, where the model becomes prone to predicting the most common class (CITE).
+This is analogous to a class imbalance problem, where the model becomes prone to predicting the most common class.
 
@@
 
-We consider a series of tracks from 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) caught and monitored between 2013 and 2022 (CITE).
+We consider a series of tracks from 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) caught and monitored between 2013 and 2022 \citep{tags1} \citep{tags2}.
 These tracks were obtained from pop-up satellite archival tags which collect temperature, light level, and depth information at specified (sub day) intervals.
-This data is then passed through a proprietary algorithm from Wildlife Computers to determine likely longitude and latitude during each day of of monitoring (\cite{PSAT}).
+This data is then passed through a proprietary algorithm from Wildlife Computers to determine likely longitude and latitude during each day of of monitoring \citep{PSAT}.
 \newline
@@
 
+\newpage
+
+\section*{Code}
+
+The tooling used to build these models as well as the means to deploy them using Amazon Web Services has been packaged at: \newline https://github.com/networkearth/mimic
+
 \bibliographystyle{apalike}

### 2024-12-26 docs/papers/log_odds/Paper.tex
wip

@@
 
-Machine learning offers a way in which to at least partially address the bespoke nature of traditional Individual-Based Models (IBMs).
-By framing individual behaviors as probabilistic decisions among consecutive sets of choices each described by a series of features machine learning can be invoked to largely automate the process of building the decision "engine" behind IBM's.
+Machine learning offers a way to at least partially address the bespoke nature of traditional Individual-Based Models (IBMs).
+By framing individual behaviors as probabilistic decisions among consecutive sets of choices each described by a series of features, machine learning can be invoked to largely automate the process of building the decision "engine" behind IBM's.
 Machine learning excels at identifying patterns in high-dimensional data, enabling models to dynamically incorporate a wide range of environmental and behavioral covariates without requiring custom coding for every new hypothesis.
@@
 
-This methodology centers on a two-stage cycle of discovery: (1) testing hypotheses about covariates that are believed to predict behaviors and (2) using the results to generate new hypotheses.
+The methodology centers on a two-stage cycle of discovery: (1) testing hypotheses about covariates that are believed to predict behaviors and (2) using the results to generate new hypotheses.
 
 In the first stage, we model behaviors as decisions made from a set of choices, each associated with hypothesized covariates.
-Then, using machine learning, we build a predictive model to evaluate the likelihood of these selections based on the covariates.
+Then, using deep learning, we build a predictive model to evaluate the likelihood of these selections based on the covariates.
 
@@
 
-The keystone of this entire process becomes having a clear, quick means of producing machine learning models from our framing of decisions and choices and our covariates, an issue which we turn to next.
+The keystone of this entire process is having a clear, quick means of producing deep learning models from our framing of decisions, choices, and covariates, an issue which we turn to next.
 
@@
 
-This growth poses a significant challenge due to the "curse of dimensionality", where the amount of data required to effectively train models grows exponentially with the dimensionality of the input space \citep{curse}.
+This growth poses a significant challenge due to the "curse of dimensionality", where the amount of data required to effectively train models can grow exponentially with the dimensionality of the input space \citep{curse}.
 
@@
 
-We can implement this log-odds model using standard probabilistic deep learning techniques by replicating the log-odds computation across all $M$ choices.
+We can implement this log-odds model using standard probabilistic deep learning techniques by replicating the "log-odds model" weights across all $M$ choices.
 The outputs are fed into a softmax layer with $M$ units, where the layer's weights are set to the identity matrix and biases are set to zero.
@@
 \subsection*{Alternative Approaches to Reducing Dimensionality}
-There are various strategies for reducing the dimensionality of one's data.
+There are other strategies for reducing the dimensionality of one's data.
 One straightforward method is to limit the choices available.
 For instance if making decisions among a set of directions, one could reduce the precision of the angles allowed.
-While this approach reduces the model's specificity, it significantly simplifies the dimensionality of the problem, which is often a key challenge in probabilistic deep learning.
+While this approach reduces the model's specificity, it significantly simplifies the dimensionality of the problem.
 
-The methodology proposed, however, achieves dimensionality reduction without sacrificing specificity.
+The methodology proposed here, however, achieves dimensionality reduction without sacrificing specificity.
 That said, there are scenarios where some choices are unlikely to be relevant.
 In such cases, dimensions corresponding to these irrelevant choices can simply be discarded, further simplifying the model.
-\newline
 
-Another effective strategy to enhance the data set is leveraging the order invariance of choices in traditional probabilistic problem framing.
+Another effective strategy to enhance the data set is leveraging the order invariance of choices in the traditional probabilistic problem framing.
 Specifically, the order in which choices are presented to the model should not matter.
@@
 In essence, for each training example, $M!$ (factorial of the number of choices) augmented examples can be created.
-While this exponentially increases the size of the data set as the feature space grows, it provides a mechanism to generate additional training examples from the same data.
 
-The issue with this approach is that as your augmented data size grows to match the needs of the greater dimensionality so too does the time required for training.
-So while the problem remains theoretically possible, the exponential uptick in time complexity poses a significant practical issue.
+The issue with this approach is that as your augmented data size grows to match the needs of the greater dimensionality, so too does the time required for training.
+So while the problem remains theoretically possible, the potential exponential uptick in time complexity poses a significant practical issue.
 The proposed approach does not suffer from this issue and remains linear in time with the amount of data provided.
@@
 
-We consider a series of tracks from 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) caught and monitored between 2013 and 2022 \citep{tags1} \citep{tags2}.
+To demonstrate this methodology in practice, we consider a series of tracks from 111 Chinook salmon (\textit{Oncorhynchus tshawytscha}) caught and monitored between 2013 and 2022 \citep{tags1} \citep{tags2}.
 These tracks were obtained from pop-up satellite archival tags which collect temperature, light level, and depth information at specified (sub day) intervals.
@@
 
-Over a test/validation/training split of 40, 71 this resulted in 345,800 contrasts of which 269,800 were used in training and the rest in validation.
+Over a validation/training split of 40, 71 this resulted in 421,800 contrasts of which 269,800 were used in training and the rest in validation.
 
@@
 Three sets models were trained, one including only distance, one with both distance and movement heading, and one with all four features.
-Note that while the feature dimensions of these models are 1, 2, and 4 respectively, given the maximum number of choices per decision seen was 33 they dimensionality of a standard probabilistic model would've been 33, 66, and 132 representing a large reduction in the dimensionality of our feature spaces.
+Note that while the feature dimensions of these models are 1, 2, and 4 respectively, given the maximum number of choices per decision seen was 33, the dimensionality of a standard probabilistic model would've been 33, 66, and 132 representing a large reduction in the dimensionality of our feature spaces.
 
@@
 Models were trained using AWS Batch using Fargate instances of 2 vcpu's and 4 GB of memory.
-By taking advantage of batch models could be all trained in parallel allowing for short (15-30 minute) turn around times.
+By taking advantage of AWS Batch,  models could be all trained in parallel allowing for short (15-30 minute) turn around times.
 
@@
 
-In training the metric of interest is the contrasts normalized log probability (C-NLP) - the average log probability per contrast.
+In training the metric of interest is the contrasts' normalized log probability (C-NLP) - the average log probability per contrast.
 For an equivalent metric over all of the decisions we also computed the average log probability per decision for each individual and then computed an average over those across individuals (in order to not favor individuals with many decisions).
@@
 \subsection*{Model One: Distance as a Key Feature}
-Given the observations that most decisions involve an individual staying put (within its h3 cell) and rarely do fish travel to an h3 cell further than 50km away, it was decided that a more reasonable baseline model would be one that included distance traveled as a feature (model 1).
+Given the observations that most decisions involve an individual staying put (within its h3 cell) and rarely do fish travel to an h3 cell further than 50km away, it was decided that a more reasonable baseline model than even odds would be one that included distance traveled as a feature (model 1).
 
@@
 \subsection*{Model Two: Movement Heading}
-Beyond this underlying distance feature the interest is in predicting, when fish move, where they will move.
+Beyond this underlying distance feature, the interest is in predicting, when fish move, where they will move.
 Of particular interest was the individuals in the sitka and yakutat groups as these groups had several individuals exhibiting what appeared to be very directed and extensive movement.
 
-Upon studying these individuals it was noted that changes in mixed layer thickness seemed to incite movement and the individuals tending to stay near higher primary productivity.
+Upon studying these individuals it was noted that changes in mixed layer thickness seemed to incite movement and the individuals tended to stay (anecdotally) near higher primary productivity.
 Therefore a model with net primary productivity and mixed layer thickness was of interest.
 
-However before moving to this model we wanted a baseline model that included movement heading as a feature so that we could determine what aspects of the movement were predicted simply by common headings as opposed to movement specifically related to productivity.
+However, before moving to this model we wanted a baseline model that included movement heading as a feature so that we could determine what aspects of the movement were predicted simply by common headings as opposed to movement specifically related to productivity.
 Model 2 is this distance plus movement heading model.
@@
 
-This southeasterly pattern is maintained throughout figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} and indeed looking at \ref{fig:ll_increase_movement} there are several groups where the majority of individuals suffer a decrease in log likelihood, in all likelihood due to this common direction not being particularly representative of them.
+This southeasterly pattern is maintained throughout figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} and indeed looking at \ref{fig:ll_increase_movement} there are several groups where the majority of individuals suffer a decrease in log likelihood, in all likelihood due to this common direction not being representative of them.
 
@@
 
-Looking at the step from model 2 to model 3 in figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} we see now that several of the movements that saw a drop in log likelihood as compared to the model 1 baseline are actually seeing a rise in log-likelihood under this new model indicating that the new features are allowing us to swing the preferred movement heading conditioned on the environment.
+Looking at the step from model 2 to model 3 in figures \ref{fig:chignik_map}-\ref{fig:yakutat_map} we see now that several of the movements that saw a drop in log likelihood as compared to the model 1 baseline are actually seeing a rise in log-likelihood under this new model indicating that the new features are allowing us to swing the preferred movement heading in a way that is conditioned on the environment.
 
@@
 
-Looking at the cases in figure \ref{fig:ll_increase} in which we see significant drops in log likelihood or cases where there is no meaningful change we can see that while there is predictive power in these features chosen thus far - especially for the yakutat, sitka, and nanwalek groups - there are still plenty of individuals that can act as exemplars for searching for new behavioral drivers.
+Looking at the cases in figure \ref{fig:ll_increase} in which we see significant drops in log likelihood or cases where there is no meaningful change we can see that while there is predictive power in the features chosen thus far - especially for the yakutat, sitka, and nanwalek groups - there are still plenty of individuals that can act as exemplars for searching for new behavioral drivers.
 
-For example in figure \ref{fig:chignik_map} even with the addition of the productivity features in model 3 the westerly movements are still poorly predicted and would warrant investigation.
+For example in figure \ref{fig:chignik_map}, even with the addition of the productivity features in model 3 the westerly movements are still poorly predicted and would warrant investigation.
 
-Likewise the unalaska group is just poorly explained in general beyond the baseline given by model 1 and would represent another case study for further investigation.
+Likewise the unalaska group is poorly explained in general and would represent another case study for further investigation.
 Perhaps by training a model with the same set of features as model 3 but only over this group we would see a significant improvement in the performance of the model which may lead us to look for demographic or geophysical differences that could explain how two different groups of fish are "using" the same covariates so differently.
 
-Clearly though, the models we have allow us to identify and focus upon individuals that so far resist our attempts to explain them.
+All in all though, the models we have allow us to identify and focus upon individuals that so far resist our attempts to explain them.
 
@@
 
-This investigation demonstrates the power and flexibility of an iterative modeling approach for understanding behavior.
-By building baseline models and progressively refining them with additional covariates, we identified both predictive patterns and individuals resistant to explanation, which serve as focal points for further hypothesis generation.
+This investigation demonstrates the power and flexibility of an iterative modeling approach for modeling behavior.
+By building baseline models and progressively refining them with additional covariates, we identify both predictive patterns and individuals resistant to explanation, which serve as focal points for further hypothesis generation.
 The efficiency of this process, enabled by quick training times and parallelization, ensures that refinement is limited only by the investigator's ability to propose and test new hypotheses.
